### 2021-08-25 Eye-Fitting-Stright-Lines-in-the-Modern-Era.Rmd
Initial commit

---
title: Eye Fitting Straight Lines in the Modern Era

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Emily A. Robinson 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of Statistics, University of Nebraska - Lincoln

- name: Susan VanderPlas 2
  affiliation: Department of Statistics, University of Nebraska - Lincoln

- name: Reka Howard 3
  affiliation: Department of Statistics, University of Nebraska - Lincoln

keywords:
- Graphics
- Regression
- Perception

abstract: |
  Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called ‘You Draw It’. Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. In my research, I created ‘You Draw It’ interactive graphics using the r2d3 package in R statistical software and validated this as a tool and method for testing graphics. I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981) in order to establish you draw it as a tool and method for testing graphics. Participats were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse. A total of 39 participants completed trendlines for 256 unique interactive plots. A linear mixed model and a generalized additive mixed model were used to determine the relationship between the participants drawn points and the best fit lines. Results from my study were consistent with those found in the previous study. When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line. This trend was most prominent when shown data simulated with larger variances. The reproducibility of these results serve as evidence of the reliability of the you draw it method. Future work is necessarry to implement the ‘You Draw It’ tool as a method of testing graphics. [200 word limit]

bibliography: bibliography.bib
output: rticles::asa_article

header-includes:
   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
   - \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
   - \usepackage[capitalise]{cleveref}
   - \newcommand\pcref[1]{(\cref{#1})}
   - \usepackage{algorithm,algpseudocode,booktabs}
---

```{r setup, include = F}
knitr::opts_chunk$set(
  echo = F, eval = T, messages = F, warnings = F,
  fig.width = 6, fig.height = 4,  fig.align = 'center',
  out.width = "\\linewidth", dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45),
  fig.pos = "tbp",
  out.extra = ""
)
```

```{r load-packages, include = F}
library(readr)
library(tidyverse)
library(knitr)
library(gridExtra)
library(patchwork)
library(cowplot)
library(ggforce)
```

# Introduction

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney_subjective_1951; @mosteller_eye_1981]. 
Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. 
In @finney_subjective_1951, it was of interest to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring. 
Guesses are made at the best estimates of certain parameters and these guesses are then used as the basis of a computation which yields a new set of approximation to the parameter estimates; this same procedure is the performed on the new parameter estimates and the computing cycle is repeated until convergence, as determined by the statistician, is reached. 
The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with dose-response relationships. 
One measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods. 
\cref{fig:subjective-judgement} shows a pair of parallel probit responses in a biological assay.
The x-axis is the $\log_{1.5}$ dose level for four dose levels (for example, doses 4, 6, 9, and 13 correspond correspond to equally spaced values on a logarithmic scale, labeled  0, 1, 2, and 3) and the y-axis is the corresponding probit response as calculated in @finney1948table; circles correspond to the test preparation method while the crosses correspond to the standard preparation method.
For these sort of assays, the does-response relationship follows a linear regression of the probit response on the logarithm of the dose levels; the two preparation methods can be constrained to be parallel [@jerne1949validity], limiting the relative potency to one consistent value. 
In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay \pcref{fig:subjective-judgement}.
<!-- \rh{I think more explanation is needed for Figure 3.1 and the corresponding experiment.} -->
<!-- https://www.quantics.co.uk/blog/what-is-relative-potency/ -->
The author then computed one iterative calculation of the relative potency based on starting values as indicated by the pair of lines provided by each participant and compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
Results indicated that one cycle of iterations for calculating the relative potency was sufficient based on the starting values provided by eye from the participants.

```{r subjective-judgement, fig.cap = "Subjective Judgement in Statistical Analysis (1951) Parallel Probits", out.width="50%"}
knitr::include_graphics("images/subjective-judgement-plot.png")
```

Thirty years later, Mosteller et al. (1981), sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
The authors recruited 153 graduate students and post doctoral researchers in Introductory Biostatistics. 
Participants were asked to fit lines by eye to four sets of points \pcref{fig:mosteller-eyefitting-plot} using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
A latin square design [@giesbrecht2004planning] with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
<!-- \rh{Maybe a reference would be nice for the latin square desing. THere might be people not being familiar with experimental design.} -->
It was found that order of presentation had no effect and that participants tended to fit the slope of the first principal component (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
<!-- \pref{fig:ols-vs-pca-example} -->
<!-- \rh{maybe some explanation would be nice for this.} -->

```{r mosteller-eyefitting-plot, fig.cap = "Eye Fitting Straight Lines (1981) Data Sets", out.width="70%"}
knitr::include_graphics("images/eyefitting-straight-lines-plots.png")
```

In 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
\cref{fig:nyt-caraccidents} [@katz_2017] is one such example in which readers are asked to draw the line for the missing years providing what they estimate to be the number of Americans who have died every year from car accidents, since 1990.
After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.

```{r nyt-caraccidents, fig.cap = "New York Times You Draw It Feature", out.width="75%"}
knitr::include_graphics("images/nyt-caraccidents-frame4.png")
```

Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the Pew Research Center create and customize graphics with Data Driven Documents (D3).
In June 2020, the New York Times released a front page displaying figures that represent each of the 100,000 lives lost from the COVID-19 pandemic until this point in time [@NYTrememberinglives]; this visualization was meant to bring about a visceral reaction and resonate with readers. 
During 2021 March Madness, FiveThirtyEight created a roster-shuffling machine which allowed readers to build their own NBA contender through interactivity [@ryanabest_2021].
Data Driven Documents (D3) is an open-source JavaScript based graphing framework created by Mike Bostock during his time working on graphics at the New York Times.
For readers familiar with R, it is notable to consider D3 in JavaScript equivalent to the ggplot2 package in R [@ggplot2].

The goal of this paper is to establish you draw it as a tool for measuring predictions of trends fitted by eye and a method for testing graphics, the second then applies you draw it to test an individual's ability to make predictions for exponentially increasing data.
In order to validate you draw it as a method for testing graphics, the first sub-study, referred to as Eye Fitting Straight Lines in the Modern Era, replicated the experiment and results found in Mosteller et al. (1981).

# Study Development

## Interactive Plot Development

## Data Generation

All data processing was conducted in R before being passed to the D3.js source code. 
A total of $N = 30$ points $(x_i, y_i), i = 1,...N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
Data were simulated based on linear model with additive errors: 
\begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align} 
The parameters $\beta_0$ and $\beta_1$ are selected to replicate Mosteller et al. (1981) with $e_i$ generated by rejection sampling in order to guarantee the points shown align with that of the fitted line. 
An ordinary least squares regression is then fit to the simulated points in order to obtain the best fit line and fitted values in 0.25 increments across the domain, $(x_k, \hat y_{k,OLS}), k = 1, ..., 4 x_{max} +1$.
The data simulation function then outputs a list of point data and line data both indicating the parameter identification, x-value, and corresponding simulated or fitted y value.
The data simulation procedure is described in \cref{alg:eyefitting-algorithm}.

\begin{algorithm}
  \caption{Eye Fitting Straight Lines in the Modern Era Data Simulation}\label{alg:eyefitting-algorithm}
  \begin{algorithmic}[1]
    \Statex \textbullet~\textbf{Input Parameters:} $y_{\bar{x}}$ for calculating the y-intercept, $\beta_0$; slope $\beta_1$; standard deviation from line $\sigma$; sample size of points $N = 30$; domain $x_{min}$ and $x_{max}$; fitted value increment $x_{by} = 0.25$.
    \Statex \textbullet~\textbf{Output Parameters:} List of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
    \State Randomly select and jitter $N = 30$ x-values along the domain, $x_{i=1:N}\in [x_{min}, x_{max}]$.
    \State Determine the y-intercept, $\beta_0$, at x = 0 from the provided slope ($\beta_1$) and y-value at the mean of x ($y_{\bar{x}}$) using point-slope equation of a line.
    \State Generate "good" errors, $e_{i = 1:N}$ based on $N(0,\sigma)$ by setting a constraint requiring the mean of the first $\frac{1}{3}\text{N}$ errors $< |2\sigma|.$
    \State Simulate point data based on $y_i = \beta_0 + \beta_1 x_i + e_i$
    \State Obtain ordinary least squares regression coefficients, $\hat\beta_0$ and $\hat\beta_1$, for the simulated point data using the lm function in the stats package in base R.
    \State Obtain fitted values every 0.25 increment across the domain from the ordinary least squares regression $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$.
    \State Output data list of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
  \end{algorithmic}
\end{algorithm}

```{r eyefitting-parameters}
data.frame(Parm = c("F", "N", "S", "V"),
           y_xbar = c(3.9, 4.11, 3.88, 3.89),
           slope = c(0.66, -0.70, 0.66, 1.98),
           sigma = c(1.98, 2.5, 1.3, 1.5)
           ) %>%
  mutate(Parm = factor(Parm, levels = c("S", "F", "V", "N"))) %>%
  arrange(Parm) %>%
  knitr::kable("latex", 
               digits = 2, 
               escape = F, 
               booktabs = T, 
               linesep = "", 
               align = "c", 
               label = "eyefitting-parameters",
               col.names = c("Parameter Choice", "$y_{\\bar{x}}$", "$\\beta_1$", "$\\sigma$"),
        caption = "Eye Fitting Straight Lines in the Modern Era simulation model parameters")
```

Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller_eye_1981 \pcref{tab:eyefitting-parameters}. 
Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
Parameter choice F produces a trend with a positive slope and a large variance while N has a negative slope and a large variance. 
In comparison, S shows a trend with a positive slope with a small variance and V yields a steep positive slope with a small variance over the domain of 4 to 16. 
\cref{fig:eyefitting-simplot} illustrates an example of simulated data for all four parameter choices intended to reflect the trends seen in \cref{fig:mosteller-eyefitting-plot}.
Aesthetic design choices were made consistent across each of the interactive you draw it plots; the aspect ratio, defining the $x$ to $y$ axis ratio was set to one and the y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range. 
<!-- \rh{I am not sure if readers wll understand "buffer of 50\% on the lower limit", "linear scale option", "start draw point set to 50\%.} -->

```{r eyefitting-simplot, fig.height = 6, fig.width = 6, fig.cap = "Eye Fitting Straight Lines in the Modern Era Simulated Data Example", out.width="75%"}

eyefitting_example_sim <- read.csv("data/youdrawit/youdrawit-eyefitting-simdata-example.csv")
eyefitting_example_simplot <- eyefitting_example_sim %>%
  filter(data == "point_data") %>%
  filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
  mutate(dataset = factor(dataset, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = dataset) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 1) +
  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), ncol = 2) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "none",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        )

eyefitting_example_simplot
```

## Study Design

Participants completed the experiment using a RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
During May 2021, participants were recruited through Twitter, Reddit, and direct email.
A total of 39 individuals completed 256 unique you draw it task plots; all completed you draw it task plots were included in the analysis.

# Results

In addition to the participant drawn points, $(x_k, y_{k,drawn})$, and the ordinary least squares (OLS) regression fitted values, $(x_k, \hat y_{k,OLS})$, a regression equation with a slope based on the first principal component (PCA) was used to calculate fitted values, $(x_k, \hat y_{k,PCA})$.
For each set of simulated data and parameter choice, the PCA regression equation was determined by using the princomp function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
<!-- \rh{Please explain x-rotation and y-rotation of PC1. Maybe the first paragraph of section 3.3.2 could be extended.} -->
Fitted values, $\hat y_{k,PCA}$ are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.

```{r ols-vs-pca-example, fig.height = 6, fig.width = 8, fig.cap="OLS vs PCA Regression Lines", message=FALSE, warning=FALSE, out.width="100%"}
library(ggplot2)
library(magrittr)
library(plyr)

set.seed(2)
corrCoef = 0.5 # sample from a multivariate normal, 10 datapoints
dat = MASS::mvrnorm(10,c(0,0),Sigma = matrix(c(1,corrCoef,2,corrCoef),2,2))
dat[,1] = dat[,1] - mean(dat[,1]) # it makes life easier for the princomp
dat[,2] = dat[,2] - mean(dat[,2])

dat = data.frame(x1 = dat[,1],x2 = dat[,2])

# Calculate the first principle component
# see http://stats.stackexchange.com/questions/13152/how-to-perform-orthogonal-regression-total-least-squares-via-pca
v = dat%>%prcomp%$%rotation
x1x2cor = bCor = v[2,1]/v[1,1]

x1tox2 = coef(lm(x1~x2,dat))
x2tox1 = coef(lm(x2~x1,dat))
slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
                       type=c("Principal Component", "Ordinary Least Squares"))

# We want this to draw the neat orthogonal lines.
pointOnLine = function(inp){
  # y = a*x + c (c=0)
  # yOrth = -(1/a)*x + d
  # yOrth = b*x + d
  x0 = inp[1] 
  y0 = inp[2] 
  a = x1x2cor
  b = -(1/a)
  c = 0
  d = y0 - b*x0
  x = (d-c)/(a-b)
  y = -(1/a)*x+d
  return(c(x,y))
}

points = apply(dat,1,FUN=pointOnLine)

segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "Principal Component"),
                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="Ordinary Least Squares"))

dat %>%
ggplot(aes(x1,x2))+
  geom_point()+
  geom_abline(data=slopeData,aes(slope = slope,intercept=0,color=type, linetype=type), size = 1.2)+
  geom_segment(data=segmeData,aes(x=x,y=y,xend=xend,yend=yend,color=type, linetype=type))+
  facet_grid(.~type)+
  coord_equal()+
  scale_x_continuous("x") +
  scale_y_continuous("y") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "none",
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_blank(),
        # legend.text  = element_text(size = 10),
        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
        # strip.background = element_rect(size = 0.8),
        legend.key.size = unit(1, "line")
        ) +
  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
```

<!-- To calculate the first principal component fit: https://benediktehinger.de/blog/science/scatterplots-regression-lines-and-the-first-principal-component/ -->

<!-- \rh{you have xijk and xk -> a little confusing.} -->
<!-- \rh{you mention "domain of "x". what do you mean by that?} -->
For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$. 
Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
\cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.

```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Eye Fitting Straight Lines in the Modern Era Example", out.width="65%"}
trial.feedback <- read.csv("data/youdrawit/youdrawit-eyefitting-example-feedback.csv") %>%
    mutate(`Parameter Choice` = "F")
trial.sim <- read.csv("data/youdrawit/youdrawit-eyefitting-example-simulated.csv") %>%
    mutate(`Parameter Choice` = "F")

trial.feedback %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
  geom_point(data = trial.sim, aes(y = y)) +
  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both)) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
        # strip.background = element_rect(size = 0.8),
        legend.key.size = unit(1, "line")
        ) +
  scale_x_continuous(limits = c(0,20)) +
  scale_color_manual("", values = c("black", "steelblue", "orange")) +
  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
```

Using the lmer function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS and PCA residuals, constraining the fit to a linear trend. 
Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
The LMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation}
\noindent where

+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $\gamma_0$ is the overall intercept
+ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
+ $\gamma_1$ is the overall slope for $x$
+ $\gamma_{2i}$ is the effect of the parameter choice on the slope
+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $j^{th}$ participant's characteristics
+ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.

Eliminating the linear trend constraint, the bam function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS and PCA residuals to allow for estimation of smoothing splines.
Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
The GAMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation}
\noindent where

+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $\alpha_i$ is the intercept for the parameter choice $i$
+ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
+ $s_{j}$ is the random smoothing spline for each participant.

\cref{fig:eyefitting-lmer-residualplots} and \cref{fig:eyefitting-gamm-residualplots} show the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively. 
Examining the plots, the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
In particular, this trend is more prominent in parameter choices with large variances (F and N).
These results are consistent to those found in @mosteller_eye_1981 indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.

```{r eyefitting-plots}
eyefitting_model_data <- read.csv("data/youdrawit/youdrawit-eyefitting-model-data.csv") %>%
  dplyr::rename(`Parameter Choice` = parm_id)
```

```{r eyefitting-lmer-residualplots, fig.height = 8, fig.width = 8, out.width = "90%", fig.cap = "Eye Fitting Straight Lines in the Modern Era LMM results"}
eyefitting.preds.lmer <- read.csv("data/youdrawit/youdrawit-eyefitting-lmerpred-data.csv")
# Plot Predictions
eyefitting.lmer.plot <- eyefitting.preds.lmer %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.ols, color = "OLS")) +
  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.pca, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))

eyefitting.lmer.plot
```

```{r eyefitting-gamm-residualplots, fig.height = 8, fig.width = 8, out.width = "90%", fig.cap = "Eye Fitting Straight Lines in the Modern Era GAMM results"}
eyefitting.grid.gamm <- read.csv("data/youdrawit/youdrawit-eyefitting-gammpred-data.csv")
eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
  geom_line(aes(y = ols.pred, color = "OLS")) +
  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
  geom_line(aes(y = pca.pred, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
eyefitting.gamm.plot
```

In addition to fitting trend lines over the residuals between the participant drawn values and fitted values, an OLS sum of squares and PCA sums of squares measure was calculated for each you draw it plot.
Sums of squares between the two fits were compared using the lmer function in the lme4 package in R [@lme4] to run a linear mixed model (LMM) with a log transformation.
Parameter choice (S, F, V, N), fit (OLS, PCA), and the interaction between parameter choice and fit were treated as fixed effects with a random participant effect. 
Define $SS_{ijk}$ as the sums of squares for parameter choice $i = 1,2,3,4$, fit $j=1,2$, and participant $k = 1,...,N_{participant}$.
The LMM equation is given by: 
<!-- \rh{in equation 3.4, have epsilon ijk, and on page 56 too.} -->
\begin{equation}
\log\left(SS_{ijk}\right) = \alpha_i + \beta_j + \alpha\beta_{ij} + p_{j} + \epsilon_{ijk}
\end{equation}

+ $\alpha_i$ denotes the effect of the $i^{th}$ parameter choice
+ $\beta_j$ denotes the effect of the $j^{th}$ fit
+ $\alpha\beta_{ij}$ denotes the interaction between the $i^{th}$ parameter choice and $j^{th}$ fit
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $k^{th}$ participant's characteristics
+ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.

The estimated odds ratios between the OLS fit and PCA fit for each parameter choice are displayed in \cref{fig:eyefitting-ss-oddsratio}.
While there is no significant effect of fit for any parameter choices, there is indication of the trend found previously in the residual regression models indicating that participants tend to fit the slope of the first principal component over the slope of the ordinary least squares regression line, particularly for trends with large variation (parameter choices N and F).

```{r eyefitting-ss-oddsratio, fig.cap = "Eye Fitting Straight Lines in the Modern Era Sum of Squares Results", out.width="75%"}
ss.slicediffs <- read.csv("data/youdrawit/youdrawit-ssSlicediffs-lmer.csv")
ss.slicediffs %>%
  ggplot(aes(x = ratio, y = parm_id)) +
  geom_point() +
  geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), width = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 0.5,
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12)
        ) +
  scale_x_continuous("Sum of Squares Odds Ratio \n (OLS vs PCA)", limits = c(0,2.5), breaks = seq(0,2.5,0.5)) +
  scale_y_discrete("Data Set")
```

# Discussion and Conclusion

The intent of this paper was to establish you draw it as a tool for testing graphics then use this tool to determine the cognitive implications of displaying data on the log scale.
Eye Fitting Straight Lines in the Modern Era replicated the results found in Mosteller et al. (1981). 
When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
This trend was most prominent when shown data simulated with larger variances. 
The reproducibility of these results serve as evidence of the reliability of the you draw it method.

# Future Work

### 2021-08-25 Eye-Fitting-Stright-Lines-in-the-Modern-Era.tex
Initial commit

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

% Pandoc citation processing

\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
\usepackage[capitalise]{cleveref}
\newcommand\pcref[1]{(\cref{#1})}
\usepackage{algorithm,algpseudocode,booktabs}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Eye Fitting Straight Lines in the Modern Era}

  \author{
        Emily A. Robinson 1 \thanks{The authors gratefully acknowledge
\ldots{}} \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Susan VanderPlas 2 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Reka Howard 3 \\
    Department of Statistics, University of Nebraska - Lincoln\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Eye Fitting Straight Lines in the Modern Era}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Fitting lines by eye through a set of points has been explored since the
20th century. Common methods of fitting trends by eye involve
maneuvering a string, black thread, or ruler until the fit is suitable,
then drawing the line through the set of points. In 2015, the New York
Times introduced an interactive feature, called `You Draw It'. Readers
are asked to input their own assumptions about various metrics and
compare how these assumptions relate to reality. The New York Times team
utilizes Data Driven Documents (D3) that allows readers to predict these
metrics through the use of drawing a line on their computer screen with
their computer mouse. In my research, I created `You Draw It'
interactive graphics using the r2d3 package in R statistical software
and validated this as a tool and method for testing graphics. I
recruited participants via crowdsourcing websites and replicated the
study found in Eye Fitting Straight Lines (Mosteller et al., 1981) in
order to establish you draw it as a tool and method for testing
graphics. Participats were directed to an RShiny application link and
shown points following a linear trend and asked to draw a line through
the data points using their computer mouse. A total of 39 participants
completed trendlines for 256 unique interactive plots. A linear mixed
model and a generalized additive mixed model were used to determine the
relationship between the participants drawn points and the best fit
lines. Results from my study were consistent with those found in the
previous study. When shown points following a linear trend, participants
tended to fit the slope of the first principal component over the slope
of the least squares regression line. This trend was most prominent when
shown data simulated with larger variances. The reproducibility of these
results serve as evidence of the reliability of the you draw it method.
Future work is necessarry to implement the `You Draw It' tool as a
method of testing graphics. {[}200 word limit{]}
\end{abstract}

\noindent%
{\it Keywords:} Graphics, Regression, Perception
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{finney_subjective_1951, mosteller_eye_1981}. Common methods of
fitting trends by eye involve maneuvering a string, black thread, or
ruler until the fit is suitable, then drawing the line through the set
of points. In \citet{finney_subjective_1951}, it was of interest to
determine the effect of stopping iterative maximum likelihood
calculations after one iteration. Many techniques in statistical
analysis are performed with the aid of iterative calculations such as
Newton's method or Fisher's scoring. Guesses are made at the best
estimates of certain parameters and these guesses are then used as the
basis of a computation which yields a new set of approximation to the
parameter estimates; this same procedure is the performed on the new
parameter estimates and the computing cycle is repeated until
convergence, as determined by the statistician, is reached. The author
was interested in whether one iteration of calculations was sufficient
in the estimation of parameters connected with dose-response
relationships. One measure of interest is the relative potency between a
test preparation of doses and standard preparation of does; relative
potency is calculated as the ratio of two equally effective doses
between the two preparation methods. \cref{fig:subjective-judgement}
shows a pair of parallel probit responses in a biological assay. The
x-axis is the \(\log_{1.5}\) dose level for four dose levels (for
example, doses 4, 6, 9, and 13 correspond correspond to equally spaced
values on a logarithmic scale, labeled 0, 1, 2, and 3) and the y-axis is
the corresponding probit response as calculated in
\citet{finney1948table}; circles correspond to the test preparation
method while the crosses correspond to the standard preparation method.
For these sort of assays, the does-response relationship follows a
linear regression of the probit response on the logarithm of the dose
levels; the two preparation methods can be constrained to be parallel
\citep{jerne1949validity}, limiting the relative potency to one
consistent value. In this study, twenty-one scientists were recruited
via postal mail and asked to ``rule two lines'' in order to judge by eye
the positions for a pair of parallel probit regression lines in a
biological assay (\cref{fig:subjective-judgement}). The author then
computed one iterative calculation of the relative potency based on
starting values as indicated by the pair of lines provided by each
participant and compared these relative potency estimates to that which
was estimated by the full probit technique (reaching convergence through
multiple iterations). Results indicated that one cycle of iterations for
calculating the relative potency was sufficient based on the starting
values provided by eye from the participants.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.5\linewidth,]{images/subjective-judgement-plot}

}

\caption{Subjective Judgement in Statistical Analysis (1951) Parallel Probits}\label{fig:subjective-judgement}
\end{figure}

Thirty years later, Mosteller et al.~(1981), sought to understand the
properties of least squares and other computed lines by establishing one
systematic method of fitting lines by eye. The authors recruited 153
graduate students and post doctoral researchers in Introductory
Biostatistics. Participants were asked to fit lines by eye to four sets
of points (\cref{fig:mosteller-eyefitting-plot}) using an 8.5 x 11 inch
transparency with a straight line etched completely across the middle. A
latin square design \citep{giesbrecht2004planning} with packets of the
set of points stapled together in four different sequences was used to
determine if there is an effect of order of presentation. It was found
that order of presentation had no effect and that participants tended to
fit the slope of the first principal component (error minimized
orthogonally, both horizontal and vertical, to the regression line) over
the slope of the least squares regression line (error minimized
vertically to the regression line).

\begin{figure}[tbp]

{\centering \includegraphics[width=0.7\linewidth,]{images/eyefitting-straight-lines-plots}

}

\caption{Eye Fitting Straight Lines (1981) Data Sets}\label{fig:mosteller-eyefitting-plot}
\end{figure}

In 2015, the New York Times introduced an interactive feature, called
You Draw It
\citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
Readers are asked to input their own assumptions about various metrics
and compare how these assumptions relate to reality. The New York Times
team utilizes Data Driven Documents (D3) that allows readers to predict
these metrics through the use of drawing a line on their computer screen
with their computer mouse. \cref{fig:nyt-caraccidents} \citep{katz_2017}
is one such example in which readers are asked to draw the line for the
missing years providing what they estimate to be the number of Americans
who have died every year from car accidents, since 1990. After the
reader has completed drawing the line, the actual observed values are
revealed and the reader may check their estimated knowledge against the
actual reported data.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.75\linewidth,]{images/nyt-caraccidents-frame4}

}

\caption{New York Times You Draw It Feature}\label{fig:nyt-caraccidents}
\end{figure}

Major news and research organizations such as the New York Times,
FiveThirtyEight, Washington Post, and the Pew Research Center create and
customize graphics with Data Driven Documents (D3). In June 2020, the
New York Times released a front page displaying figures that represent
each of the 100,000 lives lost from the COVID-19 pandemic until this
point in time \citep{NYTrememberinglives}; this visualization was meant
to bring about a visceral reaction and resonate with readers. During
2021 March Madness, FiveThirtyEight created a roster-shuffling machine
which allowed readers to build their own NBA contender through
interactivity \citep{ryanabest_2021}. Data Driven Documents (D3) is an
open-source JavaScript based graphing framework created by Mike Bostock
during his time working on graphics at the New York Times. For readers
familiar with R, it is notable to consider D3 in JavaScript equivalent
to the ggplot2 package in R \citep{ggplot2}.

The goal of this paper is to establish you draw it as a tool for
measuring predictions of trends fitted by eye and a method for testing
graphics, the second then applies you draw it to test an individual's
ability to make predictions for exponentially increasing data. In order
to validate you draw it as a method for testing graphics, the first
sub-study, referred to as Eye Fitting Straight Lines in the Modern Era,
replicated the experiment and results found in Mosteller et al.~(1981).

\hypertarget{study-development}{%
\section{Study Development}\label{study-development}}

\hypertarget{interactive-plot-development}{%
\subsection{Interactive Plot
Development}\label{interactive-plot-development}}

\hypertarget{data-generation}{%
\subsection{Data Generation}\label{data-generation}}

All data processing was conducted in R before being passed to the D3.js
source code. A total of \(N = 30\) points \((x_i, y_i), i = 1,...N\)
were generated for \(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\)
have a linear relationship. Data were simulated based on linear model
with additive errors: \begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align} The parameters \(\beta_0\) and \(\beta_1\) are selected to
replicate Mosteller et al.~(1981) with \(e_i\) generated by rejection
sampling in order to guarantee the points shown align with that of the
fitted line. An ordinary least squares regression is then fit to the
simulated points in order to obtain the best fit line and fitted values
in 0.25 increments across the domain,
\((x_k, \hat y_{k,OLS}), k = 1, ..., 4 x_{max} +1\). The data simulation
function then outputs a list of point data and line data both indicating
the parameter identification, x-value, and corresponding simulated or
fitted y value. The data simulation procedure is described in
\cref{alg:eyefitting-algorithm}.

\begin{algorithm}
  \caption{Eye Fitting Straight Lines in the Modern Era Data Simulation}\label{alg:eyefitting-algorithm}
  \begin{algorithmic}[1]
    \Statex \textbullet~\textbf{Input Parameters:} $y_{\bar{x}}$ for calculating the y-intercept, $\beta_0$; slope $\beta_1$; standard deviation from line $\sigma$; sample size of points $N = 30$; domain $x_{min}$ and $x_{max}$; fitted value increment $x_{by} = 0.25$.
    \Statex \textbullet~\textbf{Output Parameters:} List of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
    \State Randomly select and jitter $N = 30$ x-values along the domain, $x_{i=1:N}\in [x_{min}, x_{max}]$.
    \State Determine the y-intercept, $\beta_0$, at x = 0 from the provided slope ($\beta_1$) and y-value at the mean of x ($y_{\bar{x}}$) using point-slope equation of a line.
    \State Generate "good" errors, $e_{i = 1:N}$ based on $N(0,\sigma)$ by setting a constraint requiring the mean of the first $\frac{1}{3}\text{N}$ errors $< |2\sigma|.$
    \State Simulate point data based on $y_i = \beta_0 + \beta_1 x_i + e_i$
    \State Obtain ordinary least squares regression coefficients, $\hat\beta_0$ and $\hat\beta_1$, for the simulated point data using the lm function in the stats package in base R.
    \State Obtain fitted values every 0.25 increment across the domain from the ordinary least squares regression $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$.
    \State Output data list of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
  \end{algorithmic}
\end{algorithm}

\begin{table}

\caption{\label{tab:eyefitting-parameters}Eye Fitting Straight Lines in the Modern Era simulation model parameters}
\centering
\begin{tabular}[t]{cccc}
\toprule
Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$\\
\midrule
S & 3.88 & 0.66 & 1.30\\
F & 3.90 & 0.66 & 1.98\\
V & 3.89 & 1.98 & 1.50\\
N & 4.11 & -0.70 & 2.50\\
\bottomrule
\end{tabular}
\end{table}

Simulated model equation parameters were selected to reflect the four
data sets (F, N, S, and V) used in \citet{mosteller_eye_1981}
(\cref{tab:eyefitting-parameters}). Parameter choices F, N, and S
simulated data across a domain of 0 to 20. Parameter choice F produces a
trend with a positive slope and a large variance while N has a negative
slope and a large variance. In comparison, S shows a trend with a
positive slope with a small variance and V yields a steep positive slope
with a small variance over the domain of 4 to 16.
\cref{fig:eyefitting-simplot} illustrates an example of simulated data
for all four parameter choices intended to reflect the trends seen in
\cref{fig:mosteller-eyefitting-plot}. Aesthetic design choices were made
consistent across each of the interactive you draw it plots; the aspect
ratio, defining the \(x\) to \(y\) axis ratio was set to one and the
y-axis range extended 10\% beyond (above and below) the range of the
simulated data points to allow for users to draw outside the simulated
data set range.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.75\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}

}

\caption{Eye Fitting Straight Lines in the Modern Era Simulated Data Example}\label{fig:eyefitting-simplot}
\end{figure}

\hypertarget{study-design}{%
\subsection{Study Design}\label{study-design}}

Participants completed the experiment using a RShiny application found
\href{https://shiny.srvanderplas.com/you-draw-it/}{here}. During May
2021, participants were recruited through Twitter, Reddit, and direct
email. A total of 39 individuals completed 256 unique you draw it task
plots; all completed you draw it task plots were included in the
analysis.

\hypertarget{results}{%
\section{Results}\label{results}}

In addition to the participant drawn points, \((x_k, y_{k,drawn})\), and
the ordinary least squares (OLS) regression fitted values,
\((x_k, \hat y_{k,OLS})\), a regression equation with a slope based on
the first principal component (PCA) was used to calculate fitted values,
\((x_k, \hat y_{k,PCA})\). For each set of simulated data and parameter
choice, the PCA regression equation was determined by using the princomp
function in the stats package in base R to obtain the rotation of the
coordinate axes from the first principal component (direction which
captures the most variance). The estimated slope, \(\hat\beta_{1,PCA}\),
is determined by the ratio of the axis rotation in y and axis rotation
in x of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
Fitted values, \(\hat y_{k,PCA}\) are then obtained every 0.25 increment
across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS
regression equation which minimizes the vertical distance of points from
the line and a regression equation with a slope calculated by the first
principal component which minimizes the smallest distance of points from
the line.

\begin{figure}[tbp]

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}

}

\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
\end{figure}

For each participant, the final data set used for analysis contains
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
mixed model and a generalized additive mixed model, comparisons of
vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}

}

\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
\end{figure}

Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
model (LMM) is fit separately to the OLS and PCA residuals, constraining
the fit to a linear trend. Parameter choice, \(x\), and the interaction
between \(x\) and parameter choice were treated as fixed effects with a
random participant effect accounting for variation due to participant.
The LMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

Eliminating the linear trend constraint, the bam function in the mgcv
package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
generalized additive mixed model (GAMM) separately to the OLS and PCA
residuals to allow for estimation of smoothing splines. Parameter choice
was treated as a fixed effect with no estimated intercept and a separate
smoothing spline for \(x\) was estimated for each parameter choice. A
random participant effect accounting for variation due to participant
and a random spline for each participant accounted for variation in
spline for each participant. The GAMM equation for each fit (OLS and
PCA) residuals is given by: \begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\alpha_i\) is the intercept for the parameter choice \(i\)
\item
  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
  participant variation
\item
  \(s_{j}\) is the random smoothing spline for each participant.
\end{itemize}

\cref{fig:eyefitting-lmer-residualplots} and
\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
residuals (vertical deviation of participant drawn points from both the
OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
Examining the plots, the estimated trends of PCA residuals (orange)
appear to align closer to the \(y=0\) horizontal (dashed) line than the
OLS residuals (blue). In particular, this trend is more prominent in
parameter choices with large variances (F and N). These results are
consistent to those found in \citet{mosteller_eye_1981} indicating
participants fit a trend line closer to the estimated regression line
with the slope of the first principal component than the estimated OLS
regression line.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.9\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-lmer-residualplots-1}

}

\caption{Eye Fitting Straight Lines in the Modern Era LMM results}\label{fig:eyefitting-lmer-residualplots}
\end{figure}

\begin{figure}[tbp]

{\centering \includegraphics[width=0.9\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-gamm-residualplots-1}

}

\caption{Eye Fitting Straight Lines in the Modern Era GAMM results}\label{fig:eyefitting-gamm-residualplots}
\end{figure}

In addition to fitting trend lines over the residuals between the
participant drawn values and fitted values, an OLS sum of squares and
PCA sums of squares measure was calculated for each you draw it plot.
Sums of squares between the two fits were compared using the lmer
function in the lme4 package in R \citep{lme4} to run a linear mixed
model (LMM) with a log transformation. Parameter choice (S, F, V, N),
fit (OLS, PCA), and the interaction between parameter choice and fit
were treated as fixed effects with a random participant effect. Define
\(SS_{ijk}\) as the sums of squares for parameter choice
\(i = 1,2,3,4\), fit \(j=1,2\), and participant
\(k = 1,...,N_{participant}\). The LMM equation is given by:
\begin{equation}
\log\left(SS_{ijk}\right) = \alpha_i + \beta_j + \alpha\beta_{ij} + p_{j} + \epsilon_{ijk}
\end{equation}

\begin{itemize}
\tightlist
\item
  \(\alpha_i\) denotes the effect of the \(i^{th}\) parameter choice
\item
  \(\beta_j\) denotes the effect of the \(j^{th}\) fit
\item
  \(\alpha\beta_{ij}\) denotes the interaction between the \(i^{th}\)
  parameter choice and \(j^{th}\) fit
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(k^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

The estimated odds ratios between the OLS fit and PCA fit for each
parameter choice are displayed in \cref{fig:eyefitting-ss-oddsratio}.
While there is no significant effect of fit for any parameter choices,
there is indication of the trend found previously in the residual
regression models indicating that participants tend to fit the slope of
the first principal component over the slope of the ordinary least
squares regression line, particularly for trends with large variation
(parameter choices N and F).

\begin{figure}[tbp]

{\centering \includegraphics[width=0.75\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-ss-oddsratio-1}

}

\caption{Eye Fitting Straight Lines in the Modern Era Sum of Squares Results}\label{fig:eyefitting-ss-oddsratio}
\end{figure}

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and Conclusion}\label{discussion-and-conclusion}}

The intent of this paper was to establish you draw it as a tool for
testing graphics then use this tool to determine the cognitive
implications of displaying data on the log scale. Eye Fitting Straight
Lines in the Modern Era replicated the results found in Mosteller et
al.~(1981). When shown points following a linear trend, participants
tended to fit the slope of the first principal component over the slope
of the least squares regression line. This trend was most prominent when
shown data simulated with larger variances. The reproducibility of these
results serve as evidence of the reliability of the you draw it method.

\bibliographystyle{agsm}
\bibliography{bibliography.bib}

\end{document}

### 2021-09-07 Eye-Fitting-Stright-Lines-in-the-Modern-Era.tex
updated bib. began bullet points.

@@
   \author{
-        Emily A. Robinson 1 \thanks{The authors gratefully acknowledge
-\ldots{}} \\
+        Emily A. Robinson 1 \\
     Department of Statistics, University of Nebraska - Lincoln\\
@@
 utilizes Data Driven Documents (D3) that allows readers to predict these
-metrics through the use of drawing a line on their computer screen with
-their computer mouse. In my research, I created `You Draw It'
-interactive graphics using the r2d3 package in R statistical software
-and validated this as a tool and method for testing graphics. I
-recruited participants via crowdsourcing websites and replicated the
-study found in Eye Fitting Straight Lines (Mosteller et al., 1981) in
-order to establish you draw it as a tool and method for testing
-graphics. Participats were directed to an RShiny application link and
-shown points following a linear trend and asked to draw a line through
-the data points using their computer mouse. A total of 39 participants
-completed trendlines for 256 unique interactive plots. A linear mixed
-model and a generalized additive mixed model were used to determine the
-relationship between the participants drawn points and the best fit
-lines. Results from my study were consistent with those found in the
-previous study. When shown points following a linear trend, participants
-tended to fit the slope of the first principal component over the slope
-of the least squares regression line. This trend was most prominent when
-shown data simulated with larger variances. The reproducibility of these
-results serve as evidence of the reliability of the you draw it method.
-Future work is necessarry to implement the `You Draw It' tool as a
-method of testing graphics. {[}200 word limit{]}
+metrics by drawing a line on their computer screen with their computer
+mouse. In my research, I established `You Draw It' as a method for
+graphical testing by adapting the New York Times feature. I recruited
+participants via crowdsourcing websites and replicated the study found
+in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants
+were directed to an RShiny application link and shown points following a
+linear trend and asked to draw a line through the data points using
+their computer mouse; task plots were generated using the r2d3 package
+in R statistical software. Results from my study were consistent with
+those found in the previous study; when shown points following a linear
+trend, participants tended to fit the slope of the first principal
+component over the slope of the least-squares regression line. This
+trend was most prominent when shown data simulated with larger
+variances. The reproducibility of these results serves as evidence of
+the reliability of the you draw it method. Future work is necessary to
+implement the `You Draw It' tool as a method of testing graphics. {[}200
+word limit{]}
 \end{abstract}
@@
 \noindent%
-{\it Keywords:} Graphics, Regression, Perception
+{\it Keywords:} Graphics, Regression, Graph
+Perception, Scatterplot, Cognitive Bias
 \vfill
@@
 
+Advanced technology and computing power have promoted data visualization
+as a central tool in modern data science. \citet{unwin2020data} defines
+data visualization as the art of drawing graphical charts in order to
+display data. Graphics are useful for data cleaning, exploring data
+structure, and have been an essential component in communicating
+information for the last 200 years \citep{lewandowsky1989perception}.
+Although statistical graphics have become widely used and valued in
+science, business, and in many other aspects of life, as creators of
+graphics, we are too accepting of them as default without asking
+critical questions about the graphics we create or view (Unwin, 2020).
+
+\hypertarget{graph-perception}{%
+\subsection{Graph Perception}\label{graph-perception}}
+
+\hypertarget{testing-statistical-graphics}{%
+\subsection{Testing Statistical
+Graphics}\label{testing-statistical-graphics}}
+
+One way in which we evaluate the effectiveness of charts is through the
+use of graphical tests. We could ask participants to identify
+differences in graphs, read information off of a chart accurately, use
+data to make correct real-world decisions, or predict the next few
+observations. All of these types of tests require different levels of
+use and manipulation of the information being presented in the chart.
+
+\begin{itemize}
+\tightlist
+\item
+  Psychophysics (speed and accuracy)
+\item
+  Lineups
+\end{itemize}
+
+\hypertarget{trend-judgement}{%
+\subsection{Trend Judgement}\label{trend-judgement}}
+
+\begin{itemize}
+\tightlist
+\item
+  \citet{finney1951subjective}
+\item
+  \citet{mosteller_eye_1981}
+\item
+  \citet{ciccione2021can}
+\end{itemize}
+
 Initial studies in the 20th century explored the use of fitting lines by
 eye through a set of points
-\citep{finney_subjective_1951, mosteller_eye_1981}. Common methods of
+\citep{finney1951subjective, mosteller_eye_1981}. Common methods of
 fitting trends by eye involve maneuvering a string, black thread, or
 ruler until the fit is suitable, then drawing the line through the set
-of points. In \citet{finney_subjective_1951}, it was of interest to
+of points. In \citet{finney1951subjective}, it was of interest to
 determine the effect of stopping iterative maximum likelihood
@@
 the positions for a pair of parallel probit regression lines in a
-biological assay (\cref{fig:subjective-judgement}). The author then
-computed one iterative calculation of the relative potency based on
-starting values as indicated by the pair of lines provided by each
-participant and compared these relative potency estimates to that which
-was estimated by the full probit technique (reaching convergence through
-multiple iterations). Results indicated that one cycle of iterations for
-calculating the relative potency was sufficient based on the starting
-values provided by eye from the participants.
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.5\linewidth,]{images/subjective-judgement-plot}
-
-}
-
-\caption{Subjective Judgement in Statistical Analysis (1951) Parallel Probits}\label{fig:subjective-judgement}
-\end{figure}
+biological assay. The author then computed one iterative calculation of
+the relative potency based on starting values as indicated by the pair
+of lines provided by each participant and compared these relative
+potency estimates to that which was estimated by the full probit
+technique (reaching convergence through multiple iterations). Results
+indicated that one cycle of iterations for calculating the relative
+potency was sufficient based on the starting values provided by eye from
+the participants.
 
@@
 graduate students and post doctoral researchers in Introductory
-Biostatistics. Participants were asked to fit lines by eye to four sets
-of points (\cref{fig:mosteller-eyefitting-plot}) using an 8.5 x 11 inch
-transparency with a straight line etched completely across the middle. A
-latin square design \citep{giesbrecht2004planning} with packets of the
-set of points stapled together in four different sequences was used to
-determine if there is an effect of order of presentation. It was found
-that order of presentation had no effect and that participants tended to
-fit the slope of the first principal component (error minimized
-orthogonally, both horizontal and vertical, to the regression line) over
-the slope of the least squares regression line (error minimized
-vertically to the regression line).
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.7\linewidth,]{images/eyefitting-straight-lines-plots}
-
-}
-
-\caption{Eye Fitting Straight Lines (1981) Data Sets}\label{fig:mosteller-eyefitting-plot}
-\end{figure}
+Biostatistics. Participants were asked to fit lines by eye to four
+scatterplots using an 8.5 x 11 inch transparency with a straight line
+etched completely across the middle. A latin square design
+\citep{giesbrecht2004planning} with packets of the set of points stapled
+together in four different sequences was used to determine if there is
+an effect of order of presentation. It was found that order of
+presentation had no effect and that participants tended to fit the slope
+of the first principal component (error minimized orthogonally, both
+horizontal and vertical, to the regression line) over the slope of the
+least squares regression line (error minimized vertically to the
+regression line).
 
@@
 these metrics through the use of drawing a line on their computer screen
-with their computer mouse. \cref{fig:nyt-caraccidents} \citep{katz_2017}
-is one such example in which readers are asked to draw the line for the
-missing years providing what they estimate to be the number of Americans
-who have died every year from car accidents, since 1990. After the
-reader has completed drawing the line, the actual observed values are
-revealed and the reader may check their estimated knowledge against the
-actual reported data.
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.75\linewidth,]{images/nyt-caraccidents-frame4}
-
-}
-
-\caption{New York Times You Draw It Feature}\label{fig:nyt-caraccidents}
-\end{figure}
+with their computer mouse. \citep{katz_2017} is one such example in
+which readers are asked to draw the line for the missing years providing
+what they estimate to be the number of Americans who have died every
+year from car accidents, since 1990. After the reader has completed
+drawing the line, the actual observed values are revealed and the reader
+may check their estimated knowledge against the actual reported data.
 
@@
 
+\hypertarget{research-objective}{%
+\subsection{Research objective}\label{research-objective}}
+
 The goal of this paper is to establish you draw it as a tool for
 measuring predictions of trends fitted by eye and a method for testing
-graphics, the second then applies you draw it to test an individual's
-ability to make predictions for exponentially increasing data. In order
-to validate you draw it as a method for testing graphics, the first
-sub-study, referred to as Eye Fitting Straight Lines in the Modern Era,
-replicated the experiment and results found in Mosteller et al.~(1981).
+graphics. In order to validate you draw it as a method for testing
+graphics, the first sub-study, referred to as Eye Fitting Straight Lines
+in the Modern Era, replicated the experiment and results found in
+Mosteller et al.~(1981).
 
@@
 
-In addition to fitting trend lines over the residuals between the
-participant drawn values and fitted values, an OLS sum of squares and
-PCA sums of squares measure was calculated for each you draw it plot.
-Sums of squares between the two fits were compared using the lmer
-function in the lme4 package in R \citep{lme4} to run a linear mixed
-model (LMM) with a log transformation. Parameter choice (S, F, V, N),
-fit (OLS, PCA), and the interaction between parameter choice and fit
-were treated as fixed effects with a random participant effect. Define
-\(SS_{ijk}\) as the sums of squares for parameter choice
-\(i = 1,2,3,4\), fit \(j=1,2\), and participant
-\(k = 1,...,N_{participant}\). The LMM equation is given by:
-\begin{equation}
-\log\left(SS_{ijk}\right) = \alpha_i + \beta_j + \alpha\beta_{ij} + p_{j} + \epsilon_{ijk}
-\end{equation}
-
-\begin{itemize}
-\tightlist
-\item
-  \(\alpha_i\) denotes the effect of the \(i^{th}\) parameter choice
-\item
-  \(\beta_j\) denotes the effect of the \(j^{th}\) fit
-\item
-  \(\alpha\beta_{ij}\) denotes the interaction between the \(i^{th}\)
-  parameter choice and \(j^{th}\) fit
-\item
-  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
-  the \(k^{th}\) participant's characteristics
-\item
-  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
-\end{itemize}
-
-The estimated odds ratios between the OLS fit and PCA fit for each
-parameter choice are displayed in \cref{fig:eyefitting-ss-oddsratio}.
-While there is no significant effect of fit for any parameter choices,
-there is indication of the trend found previously in the residual
-regression models indicating that participants tend to fit the slope of
-the first principal component over the slope of the ordinary least
-squares regression line, particularly for trends with large variation
-(parameter choices N and F).
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-ss-oddsratio-1}
-
-}
-
-\caption{Eye Fitting Straight Lines in the Modern Era Sum of Squares Results}\label{fig:eyefitting-ss-oddsratio}
-\end{figure}
-
 \hypertarget{discussion-and-conclusion}{%
@@
 
-The intent of this paper was to establish you draw it as a tool for
-testing graphics then use this tool to determine the cognitive
-implications of displaying data on the log scale. Eye Fitting Straight
-Lines in the Modern Era replicated the results found in Mosteller et
-al.~(1981). When shown points following a linear trend, participants
-tended to fit the slope of the first principal component over the slope
-of the least squares regression line. This trend was most prominent when
-shown data simulated with larger variances. The reproducibility of these
-results serve as evidence of the reliability of the you draw it method.
+The intent of this paper was to establish you draw it as a tool. Eye
+Fitting Straight Lines in the Modern Era replicated the results found in
+Mosteller et al.~(1981). When shown points following a linear trend,
+participants tended to fit the slope of the first principal component
+over the slope of the least squares regression line. This trend was most
+prominent when shown data simulated with larger variances. The
+reproducibility of these results serve as evidence of the reliability of
+the you draw it method.
+
+\hypertarget{future-work}{%
+\section{Future Work}\label{future-work}}
+
+Use the method. \emph{Should probably elaborate here..}
 

### 2021-09-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
Spelling error in file name :)

--- 2021-08-25 Eye-Fitting-Stright-Lines-in-the-Modern-Era.Rmd
@@
 - name: Emily A. Robinson 1
-  thanks: The authors gratefully acknowledge ...
+  #thanks: The authors gratefully acknowledge ...
   affiliation: Department of Statistics, University of Nebraska - Lincoln
@@
 - Regression
-- Perception
+- Graph Perception
+- Scatterplot
+- Cognitive Bias
 
@@
 Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-In 2015, the New York Times introduced an interactive feature, called ‘You Draw It’.
+In 2015, the New York Times introduced an interactive feature, called 'You Draw It'.
 Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
-In my research, I created ‘You Draw It’ interactive graphics using the r2d3 package in R statistical software and validated this as a tool and method for testing graphics.
-I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981) in order to establish you draw it as a tool and method for testing graphics.
-Participats were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse.
-A total of 39 participants completed trendlines for 256 unique interactive plots.
-A linear mixed model and a generalized additive mixed model were used to determine the relationship between the participants drawn points and the best fit lines.
-Results from my study were consistent with those found in the previous study.
-When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
+The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics by drawing a line on their computer screen with their computer mouse.
+In my research, I established ‘You Draw It’ as a method for graphical testing by adapting the New York Times feature.
+I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981).
+Participants were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse; task plots were generated using the r2d3 package in R statistical software.
+Results from my study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
 This trend was most prominent when shown data simulated with larger variances.
-The reproducibility of these results serve as evidence of the reliability of the you draw it method.
-Future work is necessarry to implement the ‘You Draw It’ tool as a method of testing graphics.
+The reproducibility of these results serves as evidence of the reliability of the you draw it method.
+Future work is necessary to implement the 'You Draw It' tool as a method of testing graphics.
 [200 word limit]
@@
 # Introduction
+Advanced technology and computing power have promoted data visualization as a central tool in modern data science.
+@unwin2020data defines data visualization as the art of drawing graphical charts in order to display data.
+Graphics are useful for data cleaning, exploring data structure, and have been an essential component in communicating information for the last 200 years [@lewandowsky1989perception].
+Although statistical graphics have become widely used and valued in science, business, and in many other aspects of life, as creators of graphics, we are too accepting of them as default without asking critical questions about the graphics we create or view (Unwin, 2020).
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney_subjective_1951; @mosteller_eye_1981]. 
+## Graph Perception
+
+## Testing Statistical Graphics
+
+One way in which we evaluate the effectiveness of charts is through the use of graphical tests.
+We could ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
+All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
+
++ Psychophysics (speed and accuracy)
++ Lineups
+
+## Trend Judgement
+
++ @finney1951subjective: "rule two lines"
++ @mosteller1981eye: transparancy sheet with a line drawn across middle.
++ @ciccione2021can: used a trackpad to adjust the tilt of a line on screen.
+
++ Ensemble Perception: the visual system can compute averages of various features in parallel across the items in a set
++ Strength of Association: when participants are asked to judge the strength of an association presented in a scatterplot, their judgement is more accurate for higher levels of the Pearson correlation coefficient r, although still affected by the variance of the dataset.
+
+Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. 
 Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. 
-In @finney_subjective_1951, it was of interest to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
+In @finney1951subjective, it was of interest to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
 Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring. 
@@
 For these sort of assays, the does-response relationship follows a linear regression of the probit response on the logarithm of the dose levels; the two preparation methods can be constrained to be parallel [@jerne1949validity], limiting the relative potency to one consistent value.
-In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay \pcref{fig:subjective-judgement}.
-<!-- \rh{I think more explanation is needed for Figure 3.1 and the corresponding experiment.} -->
+In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay.
 <!-- https://www.quantics.co.uk/blog/what-is-relative-potency/ -->
@@
 
-```{r subjective-judgement, fig.cap = "Subjective Judgement in Statistical Analysis (1951) Parallel Probits", out.width="50%"}
-knitr::include_graphics("images/subjective-judgement-plot.png")
-```
-
-Thirty years later, Mosteller et al. (1981), sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
+Thirty years later, @mosteller1981eye, sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
 The authors recruited 153 graduate students and post doctoral researchers in Introductory Biostatistics. 
-Participants were asked to fit lines by eye to four sets of points \pcref{fig:mosteller-eyefitting-plot} using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
+Participants were asked to fit lines by eye to four scatterplots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
 A latin square design [@giesbrecht2004planning] with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
-<!-- \rh{Maybe a reference would be nice for the latin square desing. THere might be people not being familiar with experimental design.} -->
 It was found that order of presentation had no effect and that participants tended to fit the slope of the first principal component (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
-<!-- \pref{fig:ols-vs-pca-example} -->
-<!-- \rh{maybe some explanation would be nice for this.} -->
-
-```{r mosteller-eyefitting-plot, fig.cap = "Eye Fitting Straight Lines (1981) Data Sets", out.width="70%"}
-knitr::include_graphics("images/eyefitting-straight-lines-plots.png")
-```
 
@@
 The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
-\cref{fig:nyt-caraccidents} [@katz_2017] is one such example in which readers are asked to draw the line for the missing years providing what they estimate to be the number of Americans who have died every year from car accidents, since 1990.
+[@katz_2017] is one such example in which readers are asked to draw the line for the missing years providing what they estimate to be the number of Americans who have died every year from car accidents, since 1990.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
 
-```{r nyt-caraccidents, fig.cap = "New York Times You Draw It Feature", out.width="75%"}
+```{r nyt-caraccidents, include = F, fig.cap = "New York Times You Draw It Feature", out.width="75%"}
 knitr::include_graphics("images/nyt-caraccidents-frame4.png")
@@
 
-The goal of this paper is to establish you draw it as a tool for measuring predictions of trends fitted by eye and a method for testing graphics, the second then applies you draw it to test an individual's ability to make predictions for exponentially increasing data.
-In order to validate you draw it as a method for testing graphics, the first sub-study, referred to as Eye Fitting Straight Lines in the Modern Era, replicated the experiment and results found in Mosteller et al.
-(1981).
+## Research objective
 
-# Study Development
+The goal of this paper is to establish you draw it as a tool for measuring predictions of trends fitted by eye and a method for testing graphics.
+In order to validate you draw it as a method for testing graphics, the first sub-study, referred to as Eye Fitting Straight Lines in the Modern Era, replicated the experiment and results found in @mosteller1981eye.
 
-## Interactive Plot Development
+# Methods
 
-## Data Generation
+## Data Simulation
 
@@
 \end{align} 
-The parameters $\beta_0$ and $\beta_1$ are selected to replicate Mosteller et al.
-(1981) with $e_i$ generated by rejection sampling in order to guarantee the points shown align with that of the fitted line.
+The parameters $\beta_0$ and $\beta_1$ are selected to replicate @mosteller1981eye with $e_i$ generated by rejection sampling in order to guarantee the points shown align with that of the fitted line.
 An ordinary least squares regression is then fit to the simulated points in order to obtain the best fit line and fitted values in 0.25 increments across the domain, $(x_k, \hat y_{k,OLS}), k = 1, ..., 4 x_{max} +1$.
@@
 
-Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller_eye_1981 \pcref{tab:eyefitting-parameters}. 
+## Parameter selection
+
+Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}. 
 Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
@@
 Aesthetic design choices were made consistent across each of the interactive you draw it plots; the aspect ratio, defining the $x$ to $y$ axis ratio was set to one and the y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range.
-<!-- \rh{I am not sure if readers wll understand "buffer of 50\% on the lower limit", "linear scale option", "start draw point set to 50\%.} -->
 
@@
 
-## Study Design
+## 'You Draw It' Task Plots
 
-Participants completed the experiment using a RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
-During May 2021, participants were recruited through Twitter, Reddit, and direct email.
-A total of 39 individuals completed 256 unique you draw it task plots; all completed you draw it task plots were included in the analysis.
+## Procedure
+
++ 2 practice trials.
+
+## Participants
+
++ Recruted through Twitter, Reddit, and direct email (May 2021)
++ 39 individuals completed 256 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis.
++ All participants had normal or corrected to normal vision.
++ IRB approval and informed consent.
++ Not paid.
++ Took approximately 15 minutes to complete.
++ Participants completed the experiment using a RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
 
 # Results
+
++ Because of the randomness in the stimulus generation process, on any given trial the actual slope of the linear regression line could differ from its prescribed value.
 
@@
 
-<!-- \rh{you have xijk and xk -> a little confusing.} -->
-<!-- \rh{you mention "domain of "x". what do you mean by that?} -->
 For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
@@
 In particular, this trend is more prominent in parameter choices with large variances (F and N).
-These results are consistent to those found in @mosteller_eye_1981 indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.
+These results are consistent to those found in @mosteller1981eye indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.
 
@@
 
-In addition to fitting trend lines over the residuals between the participant drawn values and fitted values, an OLS sum of squares and PCA sums of squares measure was calculated for each you draw it plot.
-Sums of squares between the two fits were compared using the lmer function in the lme4 package in R [@lme4] to run a linear mixed model (LMM) with a log transformation.
-Parameter choice (S, F, V, N), fit (OLS, PCA), and the interaction between parameter choice and fit were treated as fixed effects with a random participant effect. 
-Define $SS_{ijk}$ as the sums of squares for parameter choice $i = 1,2,3,4$, fit $j=1,2$, and participant $k = 1,...,N_{participant}$.
-The LMM equation is given by: 
-<!-- \rh{in equation 3.4, have epsilon ijk, and on page 56 too.} -->
-\begin{equation}
-\log\left(SS_{ijk}\right) = \alpha_i + \beta_j + \alpha\beta_{ij} + p_{j} + \epsilon_{ijk}
-\end{equation}
-
-+ $\alpha_i$ denotes the effect of the $i^{th}$ parameter choice
-+ $\beta_j$ denotes the effect of the $j^{th}$ fit
-+ $\alpha\beta_{ij}$ denotes the interaction between the $i^{th}$ parameter choice and $j^{th}$ fit
-+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $k^{th}$ participant's characteristics
-+ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.
-
-The estimated odds ratios between the OLS fit and PCA fit for each parameter choice are displayed in \cref{fig:eyefitting-ss-oddsratio}.
-While there is no significant effect of fit for any parameter choices, there is indication of the trend found previously in the residual regression models indicating that participants tend to fit the slope of the first principal component over the slope of the ordinary least squares regression line, particularly for trends with large variation (parameter choices N and F).
-
-```{r eyefitting-ss-oddsratio, fig.cap = "Eye Fitting Straight Lines in the Modern Era Sum of Squares Results", out.width="75%"}
-ss.slicediffs <- read.csv("data/youdrawit/youdrawit-ssSlicediffs-lmer.csv")
-ss.slicediffs %>%
-  ggplot(aes(x = ratio, y = parm_id)) +
-  geom_point() +
-  geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), width = 0.5) +
-  geom_vline(xintercept = 1, linetype = "dashed") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 0.5,
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12)
-        ) +
-  scale_x_continuous("Sum of Squares Odds Ratio \n (OLS vs PCA)", limits = c(0,2.5), breaks = seq(0,2.5,0.5)) +
-  scale_y_discrete("Data Set")
-```
-
 # Discussion and Conclusion
 
-The intent of this paper was to establish you draw it as a tool for testing graphics then use this tool to determine the cognitive implications of displaying data on the log scale.
-Eye Fitting Straight Lines in the Modern Era replicated the results found in Mosteller et al. (1981). 
+The intent of this paper was to establish you draw it as a tool. 
+Eye Fitting Straight Lines in the Modern Era replicated the results found in @mosteller1981eye.
 When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
@@
 # Future Work
+
++ Use the method.
++ Write R package.

### 2021-09-09 Eye-Fitting-Stright-Lines-in-the-Modern-Era.tex
rough rough rough draft results for first part...

@@
 
+\usepackage[dvipsnames]{xcolor} % colors
 \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
@@
 
-Advanced technology and computing power have promoted data visualization
-as a central tool in modern data science. \citet{unwin2020data} defines
-data visualization as the art of drawing graphical charts in order to
-display data. Graphics are useful for data cleaning, exploring data
-structure, and have been an essential component in communicating
-information for the last 200 years \citep{lewandowsky1989perception}.
-Although statistical graphics have become widely used and valued in
-science, business, and in many other aspects of life, as creators of
-graphics, we are too accepting of them as default without asking
-critical questions about the graphics we create or view (Unwin, 2020).
+\begin{itemize}
+\tightlist
+\item
+  What are graphs? Why do we care?
+\end{itemize}
 
@@
 \subsection{Graph Perception}\label{graph-perception}}
+
+\begin{itemize}
+\tightlist
+\item
+  Cleveland and McGill
+\end{itemize}
 
@@
 
-One way in which we evaluate the effectiveness of charts is through the
-use of graphical tests. We could ask participants to identify
-differences in graphs, read information off of a chart accurately, use
-data to make correct real-world decisions, or predict the next few
-observations. All of these types of tests require different levels of
-use and manipulation of the information being presented in the chart.
+Graphical tests are useful for studying the perception of statistical
+graphs. Studies might ask participants to identify differences in
+graphs, read information off of a chart accurately, use data to make
+correct real-world decisions, or predict the next few observations. All
+of these types of tests require different levels of use and manipulation
+of the information being presented in the chart. Early researchers
+studied graphs from a psychological perspective. These studies generally
+tested participants ability to detect a stimulus (or a difference
+between two stimuli) ({\textcolor{blue}{CITATIONS}}). Here we focus on
+the how graphical testing has developed in statistics.
 
-\begin{itemize}
-\tightlist
-\item
-  Psychophysics (speed and accuracy)
-\item
-  Lineups
-\end{itemize}
+A major development in statistical graphics research is Wilkinson's
+Grammar of Graphics \citep{wilkinson2013grammar}. The grammar of
+graphics serves as the fundamental framework for data visualization with
+the notion that graphics are built from the ground up by specifying
+exactly how to create a particular graph from a given data set. Visual
+representations are constructed through the use of ``tidy data'' which
+is characterized as a data set in which each variable is in its own
+column, each observation is in its own row, and each value is in its own
+cell \citep{wickham2016r}. Graphics are viewed as a mapping from
+variables in a data set (or statistics computed from the data) to visual
+attributes such as the axes, colors, shapes, or facets on the canvas in
+which the chart is displayed. Software, such as Hadley Wickham's ggplot2
+\citep{wickham2011ggplot2}, aims to implement the framework of creating
+charts and graphics as the grammar of graphics recommends.
 
-\hypertarget{trend-judgement}{%
-\subsection{Trend Judgement}\label{trend-judgement}}
+One useful tool for testing statistical graphics is the concept of a
+lineup. \citet{buja2009statistical} introduced the lineup protocol in
+which data plots are depicted and interpreted as statistics. Supported
+by the grammar of graphics, a data plot can be characterized as a
+statistic, defined as, ``a functional mapping of a variable or set of
+variables'' \citep{vanderplas2020testing}. This allows the data plot to
+be tested similar to other statistics, by comparing the actual data plot
+to a set of plots with the absence of any data structure we can test the
+likelihood of any perceived structure being significant. The
+construction of data plots as statistics allow for easy experimentation,
+granting researchers the ability to compare the effectiveness of and
+understand the perception of different types of charts
+({\textcolor{blue}{CITATIONS}}). The lineup protocol is one such example
+of the development of tools designed for statistical graphical testing.
+The advancement of graphing software provides the tools necessary to
+develop new methods of testing graphics.
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{finney1951subjective}
-\item
-  \citet{mosteller_eye_1981}
-\item
-  \citet{ciccione2021can}
-\end{itemize}
+\hypertarget{fitting-trends-by-eye}{%
+\subsection{Fitting Trends by Eye}\label{fitting-trends-by-eye}}
 
@@
 eye through a set of points
-\citep{finney1951subjective, mosteller_eye_1981}. Common methods of
-fitting trends by eye involve maneuvering a string, black thread, or
+\citep{finney1951subjective, mosteller1981eye}. Common methods of
+fitting trends by eye involved maneuvering a string, black thread, or
 ruler until the fit is suitable, then drawing the line through the set
-of points. In \citet{finney1951subjective}, it was of interest to
-determine the effect of stopping iterative maximum likelihood
-calculations after one iteration. Many techniques in statistical
-analysis are performed with the aid of iterative calculations such as
-Newton's method or Fisher's scoring. Guesses are made at the best
-estimates of certain parameters and these guesses are then used as the
-basis of a computation which yields a new set of approximation to the
-parameter estimates; this same procedure is the performed on the new
-parameter estimates and the computing cycle is repeated until
-convergence, as determined by the statistician, is reached. The author
-was interested in whether one iteration of calculations was sufficient
-in the estimation of parameters connected with dose-response
-relationships. One measure of interest is the relative potency between a
-test preparation of doses and standard preparation of does; relative
-potency is calculated as the ratio of two equally effective doses
-between the two preparation methods. \cref{fig:subjective-judgement}
-shows a pair of parallel probit responses in a biological assay. The
-x-axis is the \(\log_{1.5}\) dose level for four dose levels (for
-example, doses 4, 6, 9, and 13 correspond correspond to equally spaced
-values on a logarithmic scale, labeled 0, 1, 2, and 3) and the y-axis is
-the corresponding probit response as calculated in
-\citet{finney1948table}; circles correspond to the test preparation
-method while the crosses correspond to the standard preparation method.
-For these sort of assays, the does-response relationship follows a
-linear regression of the probit response on the logarithm of the dose
-levels; the two preparation methods can be constrained to be parallel
-\citep{jerne1949validity}, limiting the relative potency to one
-consistent value. In this study, twenty-one scientists were recruited
-via postal mail and asked to ``rule two lines'' in order to judge by eye
-the positions for a pair of parallel probit regression lines in a
-biological assay. The author then computed one iterative calculation of
-the relative potency based on starting values as indicated by the pair
-of lines provided by each participant and compared these relative
-potency estimates to that which was estimated by the full probit
-technique (reaching convergence through multiple iterations). Results
-indicated that one cycle of iterations for calculating the relative
-potency was sufficient based on the starting values provided by eye from
-the participants.
+of points. Recently, \citet{ciccione2021can} conducted a comprehensive
+set of studies investigating human ability to detect trends in graphical
+representations from a psychophysical approach.
 
-Thirty years later, Mosteller et al.~(1981), sought to understand the
-properties of least squares and other computed lines by establishing one
-systematic method of fitting lines by eye. The authors recruited 153
-graduate students and post doctoral researchers in Introductory
-Biostatistics. Participants were asked to fit lines by eye to four
-scatterplots using an 8.5 x 11 inch transparency with a straight line
-etched completely across the middle. A latin square design
-\citep{giesbrecht2004planning} with packets of the set of points stapled
-together in four different sequences was used to determine if there is
-an effect of order of presentation. It was found that order of
-presentation had no effect and that participants tended to fit the slope
-of the first principal component (error minimized orthogonally, both
-horizontal and vertical, to the regression line) over the slope of the
-least squares regression line (error minimized vertically to the
-regression line).
+In \citet{finney1951subjective}, it was of interest to determine the
+effect of stopping iterative maximum likelihood calculations after one
+iteration. Many techniques in statistical analysis are performed with
+the aid of iterative calculations such as Newton's method or Fisher's
+scoring. The author was interested in whether one iteration of
+calculations was sufficient in the estimation of parameters connected
+with dose-response relationships. One measure of interest is the
+relative potency between a test preparation of doses and standard
+preparation of does; relative potency is calculated as the ratio of two
+equally effective doses between the two preparation methods. In this
+study, twenty-one scientists were recruited via postal mail and asked to
+``rule two lines'' in order to judge by eye the positions for a pair of
+parallel probit regression lines in a biological assay. The author then
+computed one iterative calculation of the relative potency based on
+starting values as indicated by the pair of lines provided by each
+participant and compared these relative potency estimates to that which
+was estimated by the full probit technique (reaching convergence through
+multiple iterations). Results indicated that one cycle of iterations for
+calculating the relative potency was sufficient based on the starting
+values provided by eye from the participants.
+
+\citet{mosteller1981eye}, sought to understand the properties of least
+squares and other computed lines by establishing one systematic method
+of fitting lines by eye. Participants were asked to fit lines by eye to
+four scatter-plots using an 8.5 x 11 inch transparency with a straight
+line etched completely across the middle. A latin square design with
+packets of the set of points stapled together in four different
+sequences was used to determine if there is an effect of order of
+presentation. It was found that order of presentation had no effect and
+that participants tended to fit the slope of the principal axis (error
+minimized orthogonally, both horizontal and vertical, to the regression
+line) over the slope of the least squares regression line (error
+minimized vertically to the regression line). These results support
+previous research on ``ensemble perception'' indicating the visual
+system can compute averages of various features in parallel across the
+items in a set ({\textcolor{blue}{CITATIONS}}).
+
+In \citet{ciccione2021can}, participants were asked to judge trends,
+estimate slopes, and conduct extrapolation. To estimate slopes,
+participants were asked to report the slope of the best-fitting
+regression line using a trackpad to adjust the tilt of a line on screen.
+Results indicated the slopes participants reported were always in excess
+of the ideal slopes, both in the positive and in the negative direction,
+and those biases increase with noise and with number of points. This
+supports the results found in \citet{mosteller1981eye} and suggest that
+participants might use Deming regression when fitting a line to a noisy
+scatterplot.
 
@@
 these metrics through the use of drawing a line on their computer screen
-with their computer mouse. \citep{katz_2017} is one such example in
-which readers are asked to draw the line for the missing years providing
-what they estimate to be the number of Americans who have died every
-year from car accidents, since 1990. After the reader has completed
-drawing the line, the actual observed values are revealed and the reader
-may check their estimated knowledge against the actual reported data.
+with their computer mouse. After the reader has completed drawing the
+line, the actual observed values are revealed and the reader may check
+their estimated knowledge against the actual reported data.
 
-Major news and research organizations such as the New York Times,
-FiveThirtyEight, Washington Post, and the Pew Research Center create and
-customize graphics with Data Driven Documents (D3). In June 2020, the
-New York Times released a front page displaying figures that represent
-each of the 100,000 lives lost from the COVID-19 pandemic until this
-point in time \citep{NYTrememberinglives}; this visualization was meant
-to bring about a visceral reaction and resonate with readers. During
-2021 March Madness, FiveThirtyEight created a roster-shuffling machine
-which allowed readers to build their own NBA contender through
-interactivity \citep{ryanabest_2021}. Data Driven Documents (D3) is an
-open-source JavaScript based graphing framework created by Mike Bostock
-during his time working on graphics at the New York Times. For readers
-familiar with R, it is notable to consider D3 in JavaScript equivalent
-to the ggplot2 package in R \citep{ggplot2}.
+\hypertarget{research-objectives}{%
+\subsection{Research objectives}\label{research-objectives}}
 
-\hypertarget{research-objective}{%
-\subsection{Research objective}\label{research-objective}}
+In this paper, we establish `You Draw It', adapted from the New York
+Times feature, as tool for graphical testing. The `You Draw It' method
+is validated by replicating the study conducted by
+\citet{mosteller1981eye}. The results of our study provide support for
+``ensemble perception''.
 
-The goal of this paper is to establish you draw it as a tool for
-measuring predictions of trends fitted by eye and a method for testing
-graphics. In order to validate you draw it as a method for testing
-graphics, the first sub-study, referred to as Eye Fitting Straight Lines
-in the Modern Era, replicated the experiment and results found in
-Mosteller et al.~(1981).
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-\hypertarget{study-development}{%
-\section{Study Development}\label{study-development}}
+\hypertarget{participants}{%
+\subsection{Participants}\label{participants}}
 
-\hypertarget{interactive-plot-development}{%
-\subsection{Interactive Plot
-Development}\label{interactive-plot-development}}
+Participants were recruited through through Twitter, Reddit, and direct
+email in May 2021. A total of 39 individuals completed 256 unique `You
+Draw It' task plots; all completed you draw it task plots were included
+in the analysis. All participants had normal or corrected to normal
+vision and signed an informed consent form. The experimental tasks took
+approximately 15 minutes to complete. Participants completed the
+experiment on their own computers in an environment of their choosing.
+The experiment was conducted and distributed through an RShiny
+application found
+\href{https://shiny.srvanderplas.com/you-draw-it/}{here}.
+
+\hypertarget{you-draw-it-task}{%
+\subsection{`You Draw It' Task}\label{you-draw-it-task}}
+
+Data Driven Documents (D3), a JavaScript-based graphing framework that
+facilitates user interaction, is used to create the `You Draw It' task
+plots. Integrating this into RShiny using the r2d3 package, participants
+are asked to draw a line using their computer mouse through trend shown
+on their screen. Details of the development of the `You Draw It' task
+plots will be addressed in future work. In the study, participants are
+shown an interactive scatter-plot \cref{fig:ydi-stimuli} along with the
+prompt, ``Use your mouse to fill in the trend in the yellow box
+region.'' The yellow-box region moves along as the participant draws
+their trend-line until the yellow region disappears, indicating the
+participant has filled in the entire domain.
+
+\begin{figure}[tbp]
+
+{\centering \includegraphics[width=1\linewidth,]{images/ydi-stimuli}
+
+}
+
+\caption{'You Draw It' Stimuli}\label{fig:ydi-stimuli}
+\end{figure}
 
@@
 
-All data processing was conducted in R before being passed to the D3.js
-source code. A total of \(N = 30\) points \((x_i, y_i), i = 1,...N\)
-were generated for \(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\)
-have a linear relationship. Data were simulated based on linear model
-with additive errors: \begin{align}
+All data processing was conducted in R statistical software. A total of
+\(N = 30\) points \((x_i, y_i), i = 1,...N\) were generated for
+\(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
+relationship. Data were simulated based on linear model with additive
+errors: \begin{align}
 y_i & = \beta_0 + \beta_1 x_i + e_i \\
@@
 \end{align} The parameters \(\beta_0\) and \(\beta_1\) are selected to
-replicate Mosteller et al.~(1981) with \(e_i\) generated by rejection
+replicate \citet{mosteller1981eye} with \(e_i\) generated by rejection
 sampling in order to guarantee the points shown align with that of the
-fitted line. An ordinary least squares regression is then fit to the
-simulated points in order to obtain the best fit line and fitted values
-in 0.25 increments across the domain,
-\((x_k, \hat y_{k,OLS}), k = 1, ..., 4 x_{max} +1\). The data simulation
-function then outputs a list of point data and line data both indicating
-the parameter identification, x-value, and corresponding simulated or
-fitted y value. The data simulation procedure is described in
-\cref{alg:eyefitting-algorithm}.
+fitted line.
 
-\begin{algorithm}
-  \caption{Eye Fitting Straight Lines in the Modern Era Data Simulation}\label{alg:eyefitting-algorithm}
-  \begin{algorithmic}[1]
-    \Statex \textbullet~\textbf{Input Parameters:} $y_{\bar{x}}$ for calculating the y-intercept, $\beta_0$; slope $\beta_1$; standard deviation from line $\sigma$; sample size of points $N = 30$; domain $x_{min}$ and $x_{max}$; fitted value increment $x_{by} = 0.25$.
-    \Statex \textbullet~\textbf{Output Parameters:} List of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
-    \State Randomly select and jitter $N = 30$ x-values along the domain, $x_{i=1:N}\in [x_{min}, x_{max}]$.
-    \State Determine the y-intercept, $\beta_0$, at x = 0 from the provided slope ($\beta_1$) and y-value at the mean of x ($y_{\bar{x}}$) using point-slope equation of a line.
-    \State Generate "good" errors, $e_{i = 1:N}$ based on $N(0,\sigma)$ by setting a constraint requiring the mean of the first $\frac{1}{3}\text{N}$ errors $< |2\sigma|.$
-    \State Simulate point data based on $y_i = \beta_0 + \beta_1 x_i + e_i$
-    \State Obtain ordinary least squares regression coefficients, $\hat\beta_0$ and $\hat\beta_1$, for the simulated point data using the lm function in the stats package in base R.
-    \State Obtain fitted values every 0.25 increment across the domain from the ordinary least squares regression $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$.
-    \State Output data list of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
-  \end{algorithmic}
-\end{algorithm}
+Simulated model equation parameters were selected to reflect the four
+data sets (F, N, S, and V) used in \citet{mosteller1981eye}
+(\cref{tab:eyefitting-parameters}). Parameter choices F, N, and S
+simulated data across a domain of 0 to 20. Parameter choice F produces a
+trend with a positive slope and a large variance while N has a negative
+slope and a large variance. In comparison, S shows a trend with a
+positive slope with a small variance and V yields a steep positive slope
+with a small variance over the domain of 4 to 16.
+\cref{fig:eyefitting-simplot} illustrates an example of simulated data
+for all four parameter choices intended to reflect the trends in
+\citet{mosteller1981eye}. Aesthetic design choices were made consistent
+across each of the interactive you draw it plots. The y-axis range
+extended 10\% beyond (above and below) the range of the simulated data
+points to allow for users to draw outside the simulated data set range.
 
@@
 
-Simulated model equation parameters were selected to reflect the four
-data sets (F, N, S, and V) used in \citet{mosteller_eye_1981}
-(\cref{tab:eyefitting-parameters}). Parameter choices F, N, and S
-simulated data across a domain of 0 to 20. Parameter choice F produces a
-trend with a positive slope and a large variance while N has a negative
-slope and a large variance. In comparison, S shows a trend with a
-positive slope with a small variance and V yields a steep positive slope
-with a small variance over the domain of 4 to 16.
-\cref{fig:eyefitting-simplot} illustrates an example of simulated data
-for all four parameter choices intended to reflect the trends seen in
-\cref{fig:mosteller-eyefitting-plot}. Aesthetic design choices were made
-consistent across each of the interactive you draw it plots; the aspect
-ratio, defining the \(x\) to \(y\) axis ratio was set to one and the
-y-axis range extended 10\% beyond (above and below) the range of the
-simulated data points to allow for users to draw outside the simulated
-data set range.
-
 \begin{figure}[tbp]
 
-{\centering \includegraphics[width=0.75\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}
+{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}
 
@@
 
-Participants completed the experiment using a RShiny application found
-\href{https://shiny.srvanderplas.com/you-draw-it/}{here}. During May
-2021, participants were recruited through Twitter, Reddit, and direct
-email. A total of 39 individuals completed 256 unique you draw it task
-plots; all completed you draw it task plots were included in the
-analysis.
+This experiment was conducted as part of a larger study; for simplicity,
+we focus on the study design and methods related to the current study.
+Each scatter-plot was the graphical representation of a data set that
+was generated randomly, independently for each participant at the start
+of the experiment. Participants in the study are shown two `You Draw It'
+practice plots in order to train participants in the skills associated
+with executing the task followed by four task plots associated with the
+current study. The order of the task plots was randomly assigned for
+each individual in a completely randomized design.
 
@@
 \section{Results}\label{results}}
+
+\hypertarget{fitted-values}{%
+\subsection{Fitted Values}\label{fitted-values}}
 
@@
 
-\begin{figure}[tbp]
+\begin{itemize}
+\tightlist
+\item
+  Because of the randomness in the stimulus generation process, on any
+  given trial the actual slope of the linear regression line could
+  differ from its prescribed value.
+\end{itemize}
 
-{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}
+\hypertarget{feedback-data}{%
+\subsection{Feedback Data}\label{feedback-data}}
 
-}
+\hypertarget{linear-trend-constraint-performance}{%
+\subsection{Linear Trend Constraint
+Performance}\label{linear-trend-constraint-performance}}
 
-\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
-\end{figure}
+\hypertarget{smoothing-spline-trend-performance}{%
+\subsection{Smoothing Spline Trend
+Performance}\label{smoothing-spline-trend-performance}}
 
-For each participant, the final data set used for analysis contains
-\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
-for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
-\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
-mixed model and a generalized additive mixed model, comparisons of
-vertical residuals in relation to the OLS fitted values
-(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
-values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
-across the domain. \cref{fig:eyefitting-example-plot} displays an
-example of all three fitted trend lines for parameter choice F.
+\hypertarget{discussion-and-conclusion}{%
+\section{Discussion and Conclusion}\label{discussion-and-conclusion}}
 
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}
-
-}
-
-\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
-\end{figure}
-
-Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
-model (LMM) is fit separately to the OLS and PCA residuals, constraining
-the fit to a linear trend. Parameter choice, \(x\), and the interaction
-between \(x\) and parameter choice were treated as fixed effects with a
-random participant effect accounting for variation due to participant.
-The LMM equation for each fit (OLS and PCA) residuals is given by:
-\begin{equation}
-y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
-\end{equation} \noindent where
+\hypertarget{future-work}{%
+\section{Future Work}\label{future-work}}
 
@@
 \item
-  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
-  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
+  Use the method.
 \item
-  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
-  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  x-value corresponding to either the OLS or PCA fit
-\item
-  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
-  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
-  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
-  fit
-\item
-  \(\gamma_0\) is the overall intercept
-\item
-  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
-  V, N) on the intercept
-\item
-  \(\gamma_1\) is the overall slope for \(x\)
-\item
-  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
-\item
-  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
-  \(j^{th}\) participant, and \(k^{th}\) increment
-\item
-  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
-  the \(j^{th}\) participant's characteristics
-\item
-  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
+  Write R package.
 \end{itemize}
-
-Eliminating the linear trend constraint, the bam function in the mgcv
-package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
-generalized additive mixed model (GAMM) separately to the OLS and PCA
-residuals to allow for estimation of smoothing splines. Parameter choice
-was treated as a fixed effect with no estimated intercept and a separate
-smoothing spline for \(x\) was estimated for each parameter choice. A
-random participant effect accounting for variation due to participant
-and a random spline for each participant accounted for variation in
-spline for each participant. The GAMM equation for each fit (OLS and
-PCA) residuals is given by: \begin{equation}
-y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
-\end{equation} \noindent where
-
-\begin{itemize}
-\tightlist
-\item
-  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
-  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
-\item
-  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
-  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  x-value corresponding to either the OLS or PCA fit
-\item
-  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
-  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
-  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
-  fit
-\item
-  \(\alpha_i\) is the intercept for the parameter choice \(i\)
-\item
-  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
-\item
-  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
-  \(j^{th}\) participant, and \(k^{th}\) increment
-\item
-  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
-  participant variation
-\item
-  \(s_{j}\) is the random smoothing spline for each participant.
-\end{itemize}
-
-\cref{fig:eyefitting-lmer-residualplots} and
-\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
-residuals (vertical deviation of participant drawn points from both the
-OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
-Examining the plots, the estimated trends of PCA residuals (orange)
-appear to align closer to the \(y=0\) horizontal (dashed) line than the
-OLS residuals (blue). In particular, this trend is more prominent in
-parameter choices with large variances (F and N). These results are
-consistent to those found in \citet{mosteller_eye_1981} indicating
-participants fit a trend line closer to the estimated regression line
-with the slope of the first principal component than the estimated OLS
-regression line.
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.9\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-lmer-residualplots-1}
-
-}
-
-\caption{Eye Fitting Straight Lines in the Modern Era LMM results}\label{fig:eyefitting-lmer-residualplots}
-\end{figure}
-
-\begin{figure}[tbp]
-
-{\centering \includegraphics[width=0.9\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-gamm-residualplots-1}
-
-}
-
-\caption{Eye Fitting Straight Lines in the Modern Era GAMM results}\label{fig:eyefitting-gamm-residualplots}
-\end{figure}
-
-\hypertarget{discussion-and-conclusion}{%
-\section{Discussion and Conclusion}\label{discussion-and-conclusion}}
-
-The intent of this paper was to establish you draw it as a tool. Eye
-Fitting Straight Lines in the Modern Era replicated the results found in
-Mosteller et al.~(1981). When shown points following a linear trend,
-participants tended to fit the slope of the first principal component
-over the slope of the least squares regression line. This trend was most
-prominent when shown data simulated with larger variances. The
-reproducibility of these results serve as evidence of the reliability of
-the you draw it method.
-
-\hypertarget{future-work}{%
-\section{Future Work}\label{future-work}}
-
-Use the method. \emph{Should probably elaborate here..}
 

### 2021-09-09 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd

@@
 header-includes:
+   - \usepackage[dvipsnames]{xcolor} % colors
    - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
@@
 # Introduction
-Advanced technology and computing power have promoted data visualization as a central tool in modern data science.
-@unwin2020data defines data visualization as the art of drawing graphical charts in order to display data.
-Graphics are useful for data cleaning, exploring data structure, and have been an essential component in communicating information for the last 200 years [@lewandowsky1989perception].
-Although statistical graphics have become widely used and valued in science, business, and in many other aspects of life, as creators of graphics, we are too accepting of them as default without asking critical questions about the graphics we create or view (Unwin, 2020).
+
++ What are graphs? Why do we care?
 
@@
 
++ Cleveland and McGill
+
 ## Testing Statistical Graphics
 
-One way in which we evaluate the effectiveness of charts is through the use of graphical tests.
-We could ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
-All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
+Graphical tests are useful for studying the perception of statistical graphs. 
+Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
+All of these types of tests require different levels of use and manipulation of the information being presented in the chart. Early researchers studied graphs from a psychological perspective. 
+These studies generally tested participants ability to detect a stimulus (or a difference between two stimuli) (\ear{CITATIONS}).
+Here we focus on the how graphical testing has developed in statistics.
 
-+ Psychophysics (speed and accuracy)
-+ Lineups
+A major development in statistical graphics research is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
+The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
+Visual representations are constructed through the use of “tidy data” which is characterized as a data set in which each variable is in its own column, each observation is in its own row, and each value is in its own cell [@wickham2016r].
+Graphics are viewed as a mapping from variables in a data set (or statistics computed from the data) to visual attributes such as the axes, colors, shapes, or facets on the canvas in which the chart is displayed.
+Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the grammar of graphics recommends.
 
-## Trend Judgement
+One useful tool for testing statistical graphics is the concept of a lineup. 
+@buja2009statistical introduced the lineup protocol in which data plots are depicted and interpreted as statistics. 
+Supported by the grammar of graphics, a data plot can be characterized as a statistic, defined as, “a functional mapping of a variable or set of variables” [@vanderplas2020testing]. 
+This allows the data plot to be tested similar to other statistics, by comparing the actual data plot
+to a set of plots with the absence of any data structure we can test the likelihood of any perceived structure being significant. 
+The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts (\ear{CITATIONS}).
+The lineup protocol is one such example of the development of tools designed for statistical graphical testing.
+The advancement of graphing software provides the tools necessary to develop new methods of testing graphics.
 
-+ @finney1951subjective: "rule two lines"
-+ @mosteller1981eye: transparancy sheet with a line drawn across middle.
-+ @ciccione2021can: used a trackpad to adjust the tilt of a line on screen.
-
-+ Ensemble Perception: the visual system can compute averages of various features in parallel across the items in a set
-+ Strength of Association: when participants are asked to judge the strength of an association presented in a scatterplot, their judgement is more accurate for higher levels of the Pearson correlation coefficient r, although still affected by the variance of the dataset.
+## Fitting Trends by Eye
 
 Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. 
-Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. 
+Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+
 In @finney1951subjective, it was of interest to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
 Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
-Guesses are made at the best estimates of certain parameters and these guesses are then used as the basis of a computation which yields a new set of approximation to the parameter estimates; this same procedure is the performed on the new parameter estimates and the computing cycle is repeated until convergence, as determined by the statistician, is reached.
 The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with dose-response relationships. 
 One measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods.
-\cref{fig:subjective-judgement} shows a pair of parallel probit responses in a biological assay.
-The x-axis is the $\log_{1.5}$ dose level for four dose levels (for example, doses 4, 6, 9, and 13 correspond correspond to equally spaced values on a logarithmic scale, labeled  0, 1, 2, and 3) and the y-axis is the corresponding probit response as calculated in @finney1948table; circles correspond to the test preparation method while the crosses correspond to the standard preparation method.
-For these sort of assays, the does-response relationship follows a linear regression of the probit response on the logarithm of the dose levels; the two preparation methods can be constrained to be parallel [@jerne1949validity], limiting the relative potency to one consistent value.
 In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay.
-<!-- https://www.quantics.co.uk/blog/what-is-relative-potency/ -->
 The author then computed one iterative calculation of the relative potency based on starting values as indicated by the pair of lines provided by each participant and compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
@@
 
-Thirty years later, @mosteller1981eye, sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
-The authors recruited 153 graduate students and post doctoral researchers in Introductory Biostatistics. 
-Participants were asked to fit lines by eye to four scatterplots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
-A latin square design [@giesbrecht2004planning] with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
-It was found that order of presentation had no effect and that participants tended to fit the slope of the first principal component (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
+@mosteller1981eye, sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
+Participants were asked to fit lines by eye to four scatter-plots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
+A latin square design with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
+It was found that order of presentation had no effect and that participants tended to fit the slope of the principal axis (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
+These results support previous research on "ensemble perception" indicating the visual system can compute averages of various features in parallel across the items in a set (\ear{CITATIONS}).
+
+In @ciccione2021can, participants were asked to judge trends, estimate slopes, and conduct extrapolation.
+To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a trackpad to adjust the tilt of a line on screen.
+Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatterplot.
 
@@
 The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
-[@katz_2017] is one such example in which readers are asked to draw the line for the missing years providing what they estimate to be the number of Americans who have died every year from car accidents, since 1990.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
 
-```{r nyt-caraccidents, include = F, fig.cap = "New York Times You Draw It Feature", out.width="75%"}
-knitr::include_graphics("images/nyt-caraccidents-frame4.png")
-```
+## Research objectives
 
-Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the Pew Research Center create and customize graphics with Data Driven Documents (D3).
-In June 2020, the New York Times released a front page displaying figures that represent each of the 100,000 lives lost from the COVID-19 pandemic until this point in time [@NYTrememberinglives]; this visualization was meant to bring about a visceral reaction and resonate with readers.
-During 2021 March Madness, FiveThirtyEight created a roster-shuffling machine which allowed readers to build their own NBA contender through interactivity [@ryanabest_2021].
-Data Driven Documents (D3) is an open-source JavaScript based graphing framework created by Mike Bostock during his time working on graphics at the New York Times.
-For readers familiar with R, it is notable to consider D3 in JavaScript equivalent to the ggplot2 package in R [@ggplot2].
-
-## Research objective
-
-The goal of this paper is to establish you draw it as a tool for measuring predictions of trends fitted by eye and a method for testing graphics.
-In order to validate you draw it as a method for testing graphics, the first sub-study, referred to as Eye Fitting Straight Lines in the Modern Era, replicated the experiment and results found in @mosteller1981eye.
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as tool for graphical testing.
+The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye.
+The results of our study provide support for "ensemble perception".
 
@@
 
-## Data Simulation
+## Participants
 
-All data processing was conducted in R before being passed to the D3.js source code. 
+Participants were recruited through through Twitter, Reddit, and direct email in May 2021. 
+A total of 39 individuals completed 256 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
+All participants had normal or corrected to normal vision and signed an informed consent form.
+The experimental tasks took approximately 15 minutes to complete.
+Participants completed the experiment on their own computers in an environment of their choosing.
+The experiment was conducted and distributed through an RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
+
+## 'You Draw It' Task
+
+Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' task plots. 
+Integrating this into RShiny using the r2d3 package, participants are asked to draw a line using their computer mouse through trend shown on their screen. 
+Details of the development of the 'You Draw It' task plots will be addressed in future work. 
+In the study, participants are shown an interactive scatter-plot \cref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
+The yellow-box region moves along as the participant draws their trend-line until the yellow region disappears, indicating the participant has filled in the entire domain.
+
+```{r ydi-stimuli, include = T, fig.cap = "'You Draw It' Stimuli", out.width="100%"}
+knitr::include_graphics("images/ydi-stimuli.png")
+```
+
+## Data Generation
+
+All data processing was conducted in R statistical software. 
 A total of $N = 30$ points $(x_i, y_i), i = 1,...N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
@@
 The parameters $\beta_0$ and $\beta_1$ are selected to replicate @mosteller1981eye with $e_i$ generated by rejection sampling in order to guarantee the points shown align with that of the fitted line.
-An ordinary least squares regression is then fit to the simulated points in order to obtain the best fit line and fitted values in 0.25 increments across the domain, $(x_k, \hat y_{k,OLS}), k = 1, ..., 4 x_{max} +1$.
-The data simulation function then outputs a list of point data and line data both indicating the parameter identification, x-value, and corresponding simulated or fitted y value.
-The data simulation procedure is described in \cref{alg:eyefitting-algorithm}.
 
-\begin{algorithm}
-  \caption{Eye Fitting Straight Lines in the Modern Era Data Simulation}\label{alg:eyefitting-algorithm}
-  \begin{algorithmic}[1]
-    \Statex \textbullet~\textbf{Input Parameters:} $y_{\bar{x}}$ for calculating the y-intercept, $\beta_0$; slope $\beta_1$; standard deviation from line $\sigma$; sample size of points $N = 30$; domain $x_{min}$ and $x_{max}$; fitted value increment $x_{by} = 0.25$.
-    \Statex \textbullet~\textbf{Output Parameters:} List of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
-    \State Randomly select and jitter $N = 30$ x-values along the domain, $x_{i=1:N}\in [x_{min}, x_{max}]$.
-    \State Determine the y-intercept, $\beta_0$, at x = 0 from the provided slope ($\beta_1$) and y-value at the mean of x ($y_{\bar{x}}$) using point-slope equation of a line.
-    \State Generate "good" errors, $e_{i = 1:N}$ based on $N(0,\sigma)$ by setting a constraint requiring the mean of the first $\frac{1}{3}\text{N}$ errors $< |2\sigma|.$
-    \State Simulate point data based on $y_i = \beta_0 + \beta_1 x_i + e_i$
-    \State Obtain ordinary least squares regression coefficients, $\hat\beta_0$ and $\hat\beta_1$, for the simulated point data using the lm function in the stats package in base R.
-    \State Obtain fitted values every 0.25 increment across the domain from the ordinary least squares regression $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$.
-    \State Output data list of point data and line data each indicating the parameter identification, x value, and corresponding simulated or fitted y value.
-  \end{algorithmic}
-\end{algorithm}
+Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}. 
+Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
+Parameter choice F produces a trend with a positive slope and a large variance while N has a negative slope and a large variance. 
+In comparison, S shows a trend with a positive slope with a small variance and V yields a steep positive slope with a small variance over the domain of 4 to 16. 
+\cref{fig:eyefitting-simplot} illustrates an example of simulated data for all four parameter choices intended to reflect the trends in @mosteller1981eye.
+Aesthetic design choices were made consistent across each of the interactive you draw it plots. 
+The y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range.
 
@@
 
-## Parameter selection
+```{r eyefitting-simplot, fig.height = 3, fig.width = 9, fig.cap = "Eye Fitting Straight Lines in the Modern Era Simulated Data Example", out.width="100%"}
 
-Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}. 
-Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
-Parameter choice F produces a trend with a positive slope and a large variance while N has a negative slope and a large variance. 
-In comparison, S shows a trend with a positive slope with a small variance and V yields a steep positive slope with a small variance over the domain of 4 to 16. 
-\cref{fig:eyefitting-simplot} illustrates an example of simulated data for all four parameter choices intended to reflect the trends seen in \cref{fig:mosteller-eyefitting-plot}.
-Aesthetic design choices were made consistent across each of the interactive you draw it plots; the aspect ratio, defining the $x$ to $y$ axis ratio was set to one and the y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range.
+eyefitting_example_sim <- read.csv("data/youdrawit-eyefitting-simdata-example.csv")
 
-```{r eyefitting-simplot, fig.height = 6, fig.width = 6, fig.cap = "Eye Fitting Straight Lines in the Modern Era Simulated Data Example", out.width="75%"}
-
-eyefitting_example_sim <- read.csv("data/youdrawit/youdrawit-eyefitting-simdata-example.csv")
 eyefitting_example_simplot <- eyefitting_example_sim %>%
@@
   geom_point(size = 1) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), ncol = 2) +
+  facet_wrap(~`Parameter Choice`, ncol = 4) +
   theme_bw(base_size = 14) +
@@
   legend.key.size = unit(1, "line")
-        )
-
+) +
+  scale_y_continuous(breaks = seq(-10, 20, 5))
 eyefitting_example_simplot
@@
 
-## 'You Draw It' Task Plots
+## Study Design
 
-## Procedure
-
-+ 2 practice trials.
-
-## Participants
-
-+ Recruted through Twitter, Reddit, and direct email (May 2021)
-+ 39 individuals completed 256 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis.
-+ All participants had normal or corrected to normal vision.
-+ IRB approval and informed consent.
-+ Not paid.
-+ Took approximately 15 minutes to complete.
-+ Participants completed the experiment using a RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
+This experiment was conducted as part of a larger study; for simplicity, we focus on the study design and methods related to the current study. 
+Each scatter-plot was the graphical representation of a data set that was generated randomly, independently for each participant at the start of the experiment.
+Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task followed by four task plots associated with the current study.
+The order of the task plots was randomly assigned for each individual in a completely randomized design.
 
@@
 
-+ Because of the randomness in the stimulus generation process, on any given trial the actual slope of the linear regression line could differ from its prescribed value.
+## Fitted Values
 
@@
           The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
-<!-- \rh{Please explain x-rotation and y-rotation of PC1. Maybe the first paragraph of section 3.3.2 could be extended.} -->
             Fitted values, $\hat y_{k,PCA}$ are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
@@
 
-```{r ols-vs-pca-example, fig.height = 6, fig.width = 8, fig.cap="OLS vs PCA Regression Lines", message=FALSE, warning=FALSE, out.width="100%"}
-library(ggplot2)
-library(magrittr)
-library(plyr)
++ Because of the randomness in the stimulus generation process, on any given trial the actual slope of the linear regression line could differ from its prescribed value.
 
-set.seed(2)
-corrCoef = 0.5 # sample from a multivariate normal, 10 datapoints
-dat = MASS::mvrnorm(10,c(0,0),Sigma = matrix(c(1,corrCoef,2,corrCoef),2,2))
-dat[,1] = dat[,1] - mean(dat[,1]) # it makes life easier for the princomp
-dat[,2] = dat[,2] - mean(dat[,2])
+## Feedback Data
 
-dat = data.frame(x1 = dat[,1],x2 = dat[,2])
+## Linear Trend Constraint Performance
 
-# Calculate the first principle component
-# see http://stats.stackexchange.com/questions/13152/how-to-perform-orthogonal-regression-total-least-squares-via-pca
-v = dat%>%prcomp%$%rotation
-x1x2cor = bCor = v[2,1]/v[1,1]
-
-x1tox2 = coef(lm(x1~x2,dat))
-x2tox1 = coef(lm(x2~x1,dat))
-slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
-                       type=c("Principal Component", "Ordinary Least Squares"))
-
-# We want this to draw the neat orthogonal lines.
-pointOnLine = function(inp){
-  # y = a*x + c (c=0)
-  # yOrth = -(1/a)*x + d
-  # yOrth = b*x + d
-  x0 = inp[1] 
-  y0 = inp[2] 
-  a = x1x2cor
-  b = -(1/a)
-  c = 0
-  d = y0 - b*x0
-  x = (d-c)/(a-b)
-  y = -(1/a)*x+d
-  return(c(x,y))
-}
-
-points = apply(dat,1,FUN=pointOnLine)
-
-segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "Principal Component"),
-                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="Ordinary Least Squares"))
-
-dat %>%
-ggplot(aes(x1,x2))+
-  geom_point()+
-  geom_abline(data=slopeData,aes(slope = slope,intercept=0,color=type, linetype=type), size = 1.2)+
-  geom_segment(data=segmeData,aes(x=x,y=y,xend=xend,yend=yend,color=type, linetype=type))+
-  facet_grid(.~type)+
-  coord_equal()+
-  scale_x_continuous("x") +
-  scale_y_continuous("y") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "none",
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_blank(),
-        # legend.text  = element_text(size = 10),
-        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
-        # strip.background = element_rect(size = 0.8),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
-```
-
-<!-- To calculate the first principal component fit: https://benediktehinger.de/blog/science/scatterplots-regression-lines-and-the-first-principal-component/ -->
-
-For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
-Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
-\cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.
-
-```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Eye Fitting Straight Lines in the Modern Era Example", out.width="65%"}
-trial.feedback <- read.csv("data/youdrawit/youdrawit-eyefitting-example-feedback.csv") %>%
-    mutate(`Parameter Choice` = "F")
-trial.sim <- read.csv("data/youdrawit/youdrawit-eyefitting-example-simulated.csv") %>%
-    mutate(`Parameter Choice` = "F")
-
-trial.feedback %>%
-  ggplot(aes(x = x)) +
-  geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
-  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
-  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
-  geom_point(data = trial.sim, aes(y = y)) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both)) +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "bottom",
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
-        # strip.background = element_rect(size = 0.8),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_x_continuous(limits = c(0,20)) +
-  scale_color_manual("", values = c("black", "steelblue", "orange")) +
-  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
-```
-
-Using the lmer function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS and PCA residuals, constraining the fit to a linear trend. 
-Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
-The LMM equation for each fit (OLS and PCA) residuals is given by:
-\begin{equation}
-y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
-\end{equation}
-\noindent where
-
-+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
-+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $\gamma_0$ is the overall intercept
-+ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
-+ $\gamma_1$ is the overall slope for $x$
-+ $\gamma_{2i}$ is the effect of the parameter choice on the slope
-+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
-+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $j^{th}$ participant's characteristics
-+ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.
-
-Eliminating the linear trend constraint, the bam function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS and PCA residuals to allow for estimation of smoothing splines.
-Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
-A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
-The GAMM equation for each fit (OLS and PCA) residuals is given by:
-\begin{equation}
-y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
-\end{equation}
-\noindent where
-
-+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
-+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $\alpha_i$ is the intercept for the parameter choice $i$
-+ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
-+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
-+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
-+ $s_{j}$ is the random smoothing spline for each participant.
-
-\cref{fig:eyefitting-lmer-residualplots} and \cref{fig:eyefitting-gamm-residualplots} show the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
-Examining the plots, the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
-In particular, this trend is more prominent in parameter choices with large variances (F and N).
-These results are consistent to those found in @mosteller1981eye indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.
-
-```{r eyefitting-plots}
-eyefitting_model_data <- read.csv("data/youdrawit/youdrawit-eyefitting-model-data.csv") %>%
-  dplyr::rename(`Parameter Choice` = parm_id)
-```
-
-```{r eyefitting-lmer-residualplots, fig.height = 8, fig.width = 8, out.width = "90%", fig.cap = "Eye Fitting Straight Lines in the Modern Era LMM results"}
-eyefitting.preds.lmer <- read.csv("data/youdrawit/youdrawit-eyefitting-lmerpred-data.csv")
-# Plot Predictions
-eyefitting.lmer.plot <- eyefitting.preds.lmer %>%
-  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
-  dplyr::rename(`Parameter Choice` = parm_id) %>%
-  ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
-  geom_line(aes(y = emmean.ols, color = "OLS")) +
-  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
-  geom_line(aes(y = emmean.pca, color = "PCA")) +
-  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "right",
-        plot.title   = element_text(size = 12, hjust = 0),
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
-        # strip.background = element_rect(size = 0.5),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_y_continuous("Residual") +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))
-
-eyefitting.lmer.plot
-```
-
-```{r eyefitting-gamm-residualplots, fig.height = 8, fig.width = 8, out.width = "90%", fig.cap = "Eye Fitting Straight Lines in the Modern Era GAMM results"}
-eyefitting.grid.gamm <- read.csv("data/youdrawit/youdrawit-eyefitting-gammpred-data.csv")
-eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
-  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
-  dplyr::rename(`Parameter Choice` = parm_id) %>%
-  ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
-  geom_line(aes(y = ols.pred, color = "OLS")) +
-  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
-  geom_line(aes(y = pca.pred, color = "PCA")) +
-  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "right",
-        plot.title   = element_text(size = 12, hjust = 0),
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
-        # strip.background = element_rect(size = 0.5),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_y_continuous("Residual") +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
-eyefitting.gamm.plot
-```
+## Smoothing Spline Trend Performance
 
 # Discussion and Conclusion
-
-The intent of this paper was to establish you draw it as a tool. 
-Eye Fitting Straight Lines in the Modern Era replicated the results found in @mosteller1981eye.
-When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
-This trend was most prominent when shown data simulated with larger variances.
-The reproducibility of these results serve as evidence of the reliability of the you draw it method.
 

### 2021-09-09 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
rough rough* draft. I need to read through it for readability yet. Also why are intro's so hard??

@@
 
-## Fitted Values
+## Fitted Regression Lines
 
-In addition to the participant drawn points, $(x_k, y_{k,drawn})$, and the ordinary least squares (OLS) regression fitted values, $(x_k, \hat y_{k,OLS})$, a regression equation with a slope based on the first principal component (PCA) was used to calculate fitted values, $(x_k, \hat y_{k,PCA})$.
-          For each set of simulated data and parameter choice, the PCA regression equation was determined by using the princomp function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
+We compare the participant drawn line to two regression lines determined by ordinary least squares regression and regression based on the principal axis (i.e.
+Deming Regression).
+\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
+
+Due to the randomness in the data generation process, the actual slope of the linear regression line fit through the simulated points could differ from the predetermined slope.
+Therefore, we fit an ordinary least squares (OLS) regression to each scatter-plot to obtain fitted values in 0.25 increments across the domain, $\hat y_{k,OLS}$ for $k = 1, ..., 4 x_{max} +1$.
+
+The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
 The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
-            Fitted values, $\hat y_{k,PCA}$ are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
-          \cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.
+Fitted values, $\hat y_{k,PCA}$, are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
 
-+ Because of the randomness in the stimulus generation process, on any given trial the actual slope of the linear regression line could differ from its prescribed value.
+```{r ols-vs-pca-example, fig.height = 3, fig.width = 6, fig.cap="OLS vs PCA Regression Lines", message=FALSE, warning=FALSE, out.width="100%"}
+library(ggplot2)
+library(magrittr)
+library(plyr)
 
-## Feedback Data
+set.seed(2)
+corrCoef = 0.5 # sample from a multivariate normal, 10 datapoints
+dat = MASS::mvrnorm(10,c(0,0),Sigma = matrix(c(1,corrCoef,2,corrCoef),2,2))
+dat[,1] = dat[,1] - mean(dat[,1]) # it makes life easier for the princomp
+dat[,2] = dat[,2] - mean(dat[,2])
 
-## Linear Trend Constraint Performance
+dat = data.frame(x1 = dat[,1],x2 = dat[,2])
 
-## Smoothing Spline Trend Performance
+# Calculate the first principle component
+# see http://stats.stackexchange.com/questions/13152/how-to-perform-orthogonal-regression-total-least-squares-via-pca
+v = dat%>%prcomp%$%rotation
+x1x2cor = bCor = v[2,1]/v[1,1]
+
+x1tox2 = coef(lm(x1~x2,dat))
+x2tox1 = coef(lm(x2~x1,dat))
+slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
+                       type=c("PC", "OLS"))
+
+# We want this to draw the neat orthogonal lines.
+pointOnLine = function(inp){
+  # y = a*x + c (c=0)
+  # yOrth = -(1/a)*x + d
+  # yOrth = b*x + d
+  x0 = inp[1] 
+  y0 = inp[2] 
+  a = x1x2cor
+  b = -(1/a)
+  c = 0
+  d = y0 - b*x0
+  x = (d-c)/(a-b)
+  y = -(1/a)*x+d
+  return(c(x,y))
+}
+
+points = apply(dat,1,FUN=pointOnLine)
+
+segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PC"),
+                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="OLS"))
+
+dat %>%
+ggplot(aes(x1,x2))+
+  geom_point()+
+  geom_abline(data=slopeData,aes(slope = slope,intercept=0,color=type, linetype=type), size = 1.2)+
+  geom_segment(data=segmeData,aes(x=x,y=y,xend=xend,yend=yend,color=type, linetype=type))+
+  facet_grid(.~type)+
+  coord_equal()+
+  scale_x_continuous("x") +
+  scale_y_continuous("y") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "none",
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_blank(),
+        # legend.text  = element_text(size = 10),
+        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+        # strip.background = element_rect(size = 0.8),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
+```
+
+## Data Availability
+
+```{r eyefitting-plots}
+eyefitting_model_data <- read.csv("data/youdrawit-eyefitting-model-data.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+```
+
+For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
+Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
+\cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.
+
+```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Eye Fitting Straight Lines in the Modern Era Example", out.width="65%"}
+trial.feedback <- read.csv("data/youdrawit-eyefitting-example-feedback.csv") %>%
+    mutate(`Parameter Choice` = "F")
+trial.sim <- read.csv("data/youdrawit-eyefitting-example-simulated.csv") %>%
+    mutate(`Parameter Choice` = "F")
+
+trial.feedback %>%
+  ggplot(aes(x = x)) +
+  geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
+  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
+  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
+  geom_point(data = trial.sim, aes(y = y)) +
+  facet_wrap(~`Parameter Choice`) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "bottom",
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+        # strip.background = element_rect(size = 0.8),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_x_continuous(limits = c(0,20)) +
+  scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
+  scale_color_manual("", values = c("black", "steelblue", "orange")) +
+  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
+```
+
+## Linear Trend Constraint
+
+Using the lmer function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS and PCA residuals, constraining the fit to a linear trend. 
+Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
+The LMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
+y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
+\end{equation}
+\noindent where
+
++ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
++ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $\gamma_0$ is the overall intercept
++ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
++ $\gamma_1$ is the overall slope for $x$
++ $\gamma_{2i}$ is the effect of the parameter choice on the slope
++ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $j^{th}$ participant's characteristics
++ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.
+
+```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era LMM results"}
+eyefitting.preds.lmer <- read.csv("data/youdrawit-eyefitting-lmerpred-data.csv")
+# Plot Predictions
+eyefitting.lmer.plot <- eyefitting.preds.lmer %>%
+  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  dplyr::rename(`Parameter Choice` = parm_id) %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
+  geom_line(aes(y = emmean.ols, color = "OLS")) +
+  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
+  geom_line(aes(y = emmean.pca, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "right",
+        plot.title   = element_text(size = 12, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_y_continuous("Residual") +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))
+
+eyefitting.lmer.plot
+```
+
+## Smoothing Spline Trend
+
+Eliminating the linear trend constraint, the bam function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS and PCA residuals to allow for estimation of smoothing splines.
+Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
+A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
+The GAMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
+y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+\end{equation}
+\noindent where
+
++ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
++ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $\alpha_i$ is the intercept for the parameter choice $i$
++ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
++ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
++ $s_{j}$ is the random smoothing spline for each participant.
+
+\cref{fig:eyefitting-lmer-residualplots} and \cref{fig:eyefitting-gamm-residualplots} show the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
+Examining the plots, the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
+In particular, this trend is more prominent in parameter choices with large variances (F and N).
+These results are consistent to those found in @mosteller1981eye indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.
+
+```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era GAMM results"}
+eyefitting.grid.gamm <- read.csv("data/youdrawit-eyefitting-gammpred-data.csv")
+eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
+  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  dplyr::rename(`Parameter Choice` = parm_id) %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
+  geom_line(aes(y = ols.pred, color = "OLS")) +
+  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
+  geom_line(aes(y = pca.pred, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "right",
+        plot.title   = element_text(size = 12, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_y_continuous("Residual") +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
+eyefitting.gamm.plot
+```
 
@@
 
+The intent of research was to establish you draw it as a tool for testing graphics then use this tool to determine the cognitive implications of displaying data on the log scale.
+Eye Fitting Straight Lines in the Modern Era replicated the results found in @mosteller1981eye. 
+When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
+This trend was most prominent when shown data simulated with larger variances. 
+The reproducibility of these results serve as evidence of the reliability of the you draw it method.
+
 # Future Work
 
-+ Use the method.
-+ Write R package.
+This study provided a basis for the use of 'You Draw It' as a tool for testing graphics.
+Further investigation is necessary to implement the method in non-linear settings and with real data.
+This tool could also be used to evaluate human ability to extrapolate data from trends. 
+In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.

### 2021-09-09 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
merged. thanks for catching spelling error! Whoops :)
rough rough* draft. I need to read through it for readability yet. Also why are intro's so hard??

--- 2021-09-09 Eye-Fitting-Stright-Lines-in-the-Modern-Era.tex
@@
 
-{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}
+{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}
 
@@
 
-\hypertarget{fitted-values}{%
-\subsection{Fitted Values}\label{fitted-values}}
+\hypertarget{fitted-regression-lines}{%
+\subsection{Fitted Regression Lines}\label{fitted-regression-lines}}
 
-In addition to the participant drawn points, \((x_k, y_{k,drawn})\), and
-the ordinary least squares (OLS) regression fitted values,
-\((x_k, \hat y_{k,OLS})\), a regression equation with a slope based on
-the first principal component (PCA) was used to calculate fitted values,
-\((x_k, \hat y_{k,PCA})\). For each set of simulated data and parameter
-choice, the PCA regression equation was determined by using the princomp
-function in the stats package in base R to obtain the rotation of the
-coordinate axes from the first principal component (direction which
-captures the most variance). The estimated slope, \(\hat\beta_{1,PCA}\),
-is determined by the ratio of the axis rotation in y and axis rotation
-in x of the first principal component with the y-intercept,
-\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
-using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
-Fitted values, \(\hat y_{k,PCA}\) are then obtained every 0.25 increment
-across the domain from the PCA regression equation,
+We compare the participant drawn line to two regression lines determined
+by ordinary least squares regression and regression based on the
+principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
+illustrates the difference between an OLS regression equation which
+minimizes the vertical distance of points from the line and a regression
+equation based on the principal axis which minimizes the Euclidean
+distance of points (orthogonal) from the line.
+
+Due to the randomness in the data generation process, the actual slope
+of the linear regression line fit through the simulated points could
+differ from the predetermined slope. Therefore, we fit an ordinary least
+squares (OLS) regression to each scatter-plot to obtain fitted values in
+0.25 increments across the domain, \(\hat y_{k,OLS}\) for
+\(k = 1, ..., 4 x_{max} +1\).
+
+The regression equation based on the principal axis was determined by
+using the \texttt{princomp} function in the stats package in base R to
+obtain the rotation of the coordinate axes from the first principal
+component (direction which captures the most variance). The estimated
+slope, \(\hat\beta_{1,PCA}\), is determined by the ratio of the axis
+rotation in y and axis rotation in x of the first principal component
+with the y-intercept, \(\hat\beta_{0,PCA}\) calculated by the
+point-slope equation of a line using the mean of of the simulated
+points, \((\bar x_i, \bar y_i)\). Fitted values, \(\hat y_{k,PCA}\), are
+then obtained every 0.25 increment across the domain from the PCA
+regression equation,
 \(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
-\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS
-regression equation which minimizes the vertical distance of points from
-the line and a regression equation with a slope calculated by the first
-principal component which minimizes the smallest distance of points from
-the line.
+
+\begin{figure}[tbp]
+
+{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}
+
+}
+
+\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
+\end{figure}
+
+\hypertarget{data-availability}{%
+\subsection{Data Availability}\label{data-availability}}
+
+For each participant, the final data set used for analysis contains
+\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
+for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
+\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
+mixed model and a generalized additive mixed model, comparisons of
+vertical residuals in relation to the OLS fitted values
+(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
+values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
+across the domain. \cref{fig:eyefitting-example-plot} displays an
+example of all three fitted trend lines for parameter choice F.
+
+\begin{figure}[tbp]
+
+{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}
+
+}
+
+\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
+\end{figure}
+
+\hypertarget{linear-trend-constraint}{%
+\subsection{Linear Trend Constraint}\label{linear-trend-constraint}}
+
+Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
+model (LMM) is fit separately to the OLS and PCA residuals, constraining
+the fit to a linear trend. Parameter choice, \(x\), and the interaction
+between \(x\) and parameter choice were treated as fixed effects with a
+random participant effect accounting for variation due to participant.
+The LMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
+y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
+\end{equation} \noindent where
 
@@
 \item
-  Because of the randomness in the stimulus generation process, on any
-  given trial the actual slope of the linear regression line could
-  differ from its prescribed value.
+  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
+  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
+\item
+  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
+  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  x-value corresponding to either the OLS or PCA fit
+\item
+  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
+  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
+  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
+  fit
+\item
+  \(\gamma_0\) is the overall intercept
+\item
+  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
+  V, N) on the intercept
+\item
+  \(\gamma_1\) is the overall slope for \(x\)
+\item
+  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
+\item
+  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
+  \(j^{th}\) participant, and \(k^{th}\) increment
+\item
+  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
+  the \(j^{th}\) participant's characteristics
+\item
+  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
 \end{itemize}
 
-\hypertarget{feedback-data}{%
-\subsection{Feedback Data}\label{feedback-data}}
+\begin{figure}[tbp]
 
-\hypertarget{linear-trend-constraint-performance}{%
-\subsection{Linear Trend Constraint
-Performance}\label{linear-trend-constraint-performance}}
+{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-lmer-residualplots-1}
 
-\hypertarget{smoothing-spline-trend-performance}{%
-\subsection{Smoothing Spline Trend
-Performance}\label{smoothing-spline-trend-performance}}
+}
+
+\caption{Eye Fitting Straight Lines in the Modern Era LMM results}\label{fig:eyefitting-lmer-residualplots}
+\end{figure}
+
+\hypertarget{smoothing-spline-trend}{%
+\subsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}
+
+Eliminating the linear trend constraint, the bam function in the mgcv
+package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
+generalized additive mixed model (GAMM) separately to the OLS and PCA
+residuals to allow for estimation of smoothing splines. Parameter choice
+was treated as a fixed effect with no estimated intercept and a separate
+smoothing spline for \(x\) was estimated for each parameter choice. A
+random participant effect accounting for variation due to participant
+and a random spline for each participant accounted for variation in
+spline for each participant. The GAMM equation for each fit (OLS and
+PCA) residuals is given by: \begin{equation}
+y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+\end{equation} \noindent where
+
+\begin{itemize}
+\tightlist
+\item
+  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
+  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
+\item
+  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
+  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  x-value corresponding to either the OLS or PCA fit
+\item
+  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
+  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
+  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
+  fit
+\item
+  \(\alpha_i\) is the intercept for the parameter choice \(i\)
+\item
+  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
+\item
+  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
+  \(j^{th}\) participant, and \(k^{th}\) increment
+\item
+  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
+  participant variation
+\item
+  \(s_{j}\) is the random smoothing spline for each participant.
+\end{itemize}
+
+\cref{fig:eyefitting-lmer-residualplots} and
+\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
+residuals (vertical deviation of participant drawn points from both the
+OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
+Examining the plots, the estimated trends of PCA residuals (orange)
+appear to align closer to the \(y=0\) horizontal (dashed) line than the
+OLS residuals (blue). In particular, this trend is more prominent in
+parameter choices with large variances (F and N). These results are
+consistent to those found in \citet{mosteller1981eye} indicating
+participants fit a trend line closer to the estimated regression line
+with the slope of the first principal component than the estimated OLS
+regression line.
+
+\begin{figure}[tbp]
+
+{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-gamm-residualplots-1}
+
+}
+
+\caption{Eye Fitting Straight Lines in the Modern Era GAMM results}\label{fig:eyefitting-gamm-residualplots}
+\end{figure}
 
@@
 
+The intent of research was to establish you draw it as a tool for
+testing graphics then use this tool to determine the cognitive
+implications of displaying data on the log scale. Eye Fitting Straight
+Lines in the Modern Era replicated the results found in
+\citet{mosteller1981eye}. When shown points following a linear trend,
+participants tended to fit the slope of the first principal component
+over the slope of the least squares regression line. This trend was most
+prominent when shown data simulated with larger variances. The
+reproducibility of these results serve as evidence of the reliability of
+the you draw it method.
+
 \hypertarget{future-work}{%
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Use the method.
-\item
-  Write R package.
-\end{itemize}
+This study provided a basis for the use of `You Draw It' as a tool for
+testing graphics. Further investigation is necessary to implement the
+method in non-linear settings and with real data. This tool could also
+be used to evaluate human ability to extrapolate data from trends. In
+the future, an R package designed for easy implementation of `You Draw
+It' task plots will help make this tool accessible to other researchers.
 

### 2021-09-10 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
rough draft,. still need to add intro though...

@@
 
-+ Cleveland and McGill
++ @cleveland1984graphical
++ @cleveland1985graphical
 
@@
 Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
-All of these types of tests require different levels of use and manipulation of the information being presented in the chart. Early researchers studied graphs from a psychological perspective. 
-These studies generally tested participants ability to detect a stimulus (or a difference between two stimuli) (\ear{CITATIONS}).
+All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
+Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception].
+These studies generally tested participants ability to detect a stimulus or a difference between two stimuli.
 Here we focus on the how graphical testing has developed in statistics.
@@
 to a set of plots with the absence of any data structure we can test the likelihood of any perceived structure being significant. 
-The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts (\ear{CITATIONS}).
+The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
 The lineup protocol is one such example of the development of tools designed for statistical graphical testing.
@@
 It was found that order of presentation had no effect and that participants tended to fit the slope of the principal axis (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
-These results support previous research on "ensemble perception" indicating the visual system can compute averages of various features in parallel across the items in a set (\ear{CITATIONS}).
+These results support previous research on "ensemble perception" indicating the visual system can compute averages of various features in parallel across the items in a set [@chong2003representation; @chong2005statistical; @van2011rapid].
 
@@
 
-In this paper, we establish 'You Draw It', adapted from the New York Times feature, as tool for graphical testing.
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing. 
 The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
-The results of our study provide support for "ensemble perception".
+Based on previous research surrounding "ensemble perception," we hypothesize that regression lines fit by human perception resemble regression lines based on the principal axis rather than an ordinary least squares regression line.
 
@@
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' task plots. 
-Integrating this into RShiny using the r2d3 package, participants are asked to draw a line using their computer mouse through trend shown on their screen. 
+Integrating this into RShiny using the `r2d3` package, participants are asked to draw a trend-line using their computer mouse through a scatter-plot shown on their screen. 
+In the study, participants are shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
+The yellow box region moves along as the participant draws their trend-line until the yellow region disappears, indicating the participant has filled in the entire domain.
 Details of the development of the 'You Draw It' task plots will be addressed in future work.
-In the study, participants are shown an interactive scatter-plot \cref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
-The yellow-box region moves along as the participant draws their trend-line until the yellow region disappears, indicating the participant has filled in the entire domain.
 
-```{r ydi-stimuli, include = T, fig.cap = "'You Draw It' Stimuli", out.width="100%"}
+```{r ydi-stimuli, include = T, fig.cap = "'You Draw It' task plot as shown to particpants during the study.
+The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The last frame (right) illustrates the participants finished trend-line before submission.", out.width="100%"}
 knitr::include_graphics("images/ydi-stimuli.png")
@@
 A total of $N = 30$ points $(x_i, y_i), i = 1,...N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
-Data were simulated based on linear model with additive errors: 
+Data were simulated based on a linear model with additive errors: 
 \begin{align}
@@
 \end{align}
-The parameters $\beta_0$ and $\beta_1$ are selected to replicate @mosteller1981eye with $e_i$ generated by rejection sampling in order to guarantee the points shown align with that of the fitted line.
 
-Simulated model equation parameters were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}. 
+Model equation parameters, $\beta_0$ and $\beta_1$, were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}. 
 Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
@@
 \cref{fig:eyefitting-simplot} illustrates an example of simulated data for all four parameter choices intended to reflect the trends in @mosteller1981eye.
-Aesthetic design choices were made consistent across each of the interactive you draw it plots. 
+Aesthetic design choices were made consistent across each of the interactive 'You Draw It' task plots. 
 The y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range.
@@
                          col.names = c("Parameter Choice", "$y_{\\bar{x}}$", "$\\beta_1$", "$\\sigma$"),
-                         caption = "Eye Fitting Straight Lines in the Modern Era simulation model parameters")
+                         caption = "Designated model equation parameters for simulated data.")
 ```
 
-```{r eyefitting-simplot, fig.height = 3, fig.width = 9, fig.cap = "Eye Fitting Straight Lines in the Modern Era Simulated Data Example", out.width="100%"}
+```{r eyefitting-simplot, fig.height = 3, fig.width = 9, fig.cap = "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.", out.width="100%"}
 
@@
 Each scatter-plot was the graphical representation of a data set that was generated randomly, independently for each participant at the start of the experiment.
-Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task followed by four task plots associated with the current study.
+Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task. 
+During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
+Instructions guide participants to start at the edge of the yellow box, to make sure the yellow region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
+Practice plots are then followed by four 'You Draw It' task plots associated with the current study.
 The order of the task plots was randomly assigned for each individual in a completely randomized design.
@@
 Deming Regression).
-\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
+\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
 
 Due to the randomness in the data generation process, the actual slope of the linear regression line fit through the simulated points could differ from the predetermined slope.
-Therefore, we fit an ordinary least squares (OLS) regression to each scatter-plot to obtain fitted values in 0.25 increments across the domain, $\hat y_{k,OLS}$ for $k = 1, ..., 4 x_{max} +1$.
-
+Therefore, we fit an ordinary least squares (OLS) regression to each scatter-plot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
+Fitted values, $\hat y_{k,OLS}$, are then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
 The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
@@
 
-```{r ols-vs-pca-example, fig.height = 3, fig.width = 6, fig.cap="OLS vs PCA Regression Lines", message=FALSE, warning=FALSE, out.width="100%"}
+```{r ols-vs-pca-example, fig.height = 3, fig.width = 6, fig.cap=" Comparison between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.", message=FALSE, warning=FALSE, out.width="80%"}
 library(ggplot2)
@@
 
-## Data Availability
+## Residual Trends
 
@@
 \cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.
+Data used in the analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/youdrawit-eyefitting-model-data.csv).
 
-```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Eye Fitting Straight Lines in the Modern Era Example", out.width="65%"}
+```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one 'You Draw It' task plot.
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).", out.width="50%"}
 trial.feedback <- read.csv("data/youdrawit-eyefitting-example-feedback.csv") %>%
@@
   ggplot(aes(x = x)) +
+  geom_point(data = trial.sim, aes(y = y), alpha = 0.7) +
   geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
@@
   geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
-  geom_point(data = trial.sim, aes(y = y)) +
   facet_wrap(~`Parameter Choice`) +
@@
 
-## Linear Trend Constraint
+### Linear Trend Constraint
 
-Using the lmer function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS and PCA residuals, constraining the fit to a linear trend. 
+Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS residuals and PCA residuals, constraining the fit to a linear trend. 
 Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
-The LMM equation for each fit (OLS and PCA) residuals is given by:
+The LMM equation for each fit (OLS and PCA) is given by:
 \begin{equation}
@@
 + $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.
+
+Constraining the residual trend to a linear fit, \cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PCA regression line.
+Estimated residual trend lines are overlaid on the observed individual participant residuals. 
+Results indicate the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
+In particular, this trend is more prominent in parameter choices with large variances (F and N).
+These results are consistent to those found in @mosteller1981eye indicating participants fit a trend-line closer to the estimated regression line with the slope of based on the first principal axis than the estimated OLS regression line thus, providing support for "ensemble perception".
 
@@
 
-## Smoothing Spline Trend
+### Smoothing Spline Trend
 
-Eliminating the linear trend constraint, the bam function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS and PCA residuals to allow for estimation of smoothing splines.
+Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PCA residuals to allow for estimation of smoothing splines.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
@@
 
-\cref{fig:eyefitting-lmer-residualplots} and \cref{fig:eyefitting-gamm-residualplots} show the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
-Examining the plots, the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
-In particular, this trend is more prominent in parameter choices with large variances (F and N).
-These results are consistent to those found in @mosteller1981eye indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.
+Allowing for flexibility in the residual trend, \cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PCA regression line.
+Estimated residual trends are overlaid on the observed individual participant residuals. 
+The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that for scatter-plots with more noise (F and N), estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue).
+By fitting smoothing splines, we can determine whether participants naturally fit a straight trend-line to the set of points or whether they deviate throughout the domain.
+In particular, in scatter-plots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and correcting for their fit toward the end of their trend-line.
+In scatter-plots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PCA regression line.
+As participants continued their trend-line, they crossed through the OLS regression line indicating they estimated the slope in the extreme direction.
+These results provide further insight into the curvature humans perceive in a set of points.
 
@@
 
-The intent of research was to establish you draw it as a tool for testing graphics then use this tool to determine the cognitive implications of displaying data on the log scale.
-Eye Fitting Straight Lines in the Modern Era replicated the results found in @mosteller1981eye. 
-When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
+The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics
+The study conducted by @mosteller1981eye was replicated using the 'You Draw It' method providing support for the validity of the tool.
+Our results indicated that when shown points following a linear trend, participants tended to fit a line closer to a regression based on the principal axis over the ordinary least squares regression line.
 This trend was most prominent when shown data simulated with larger variances. 
-The reproducibility of these results serve as evidence of the reliability of the you draw it method.
+These results indicate that humans perform "ensemble perception" in a statistical graphic setting as participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously.
+In addition, we allowed participants to draw trend lines that deviated from a straight line providing insight into the curvature the human eye perceives in a set of points.
 
@@
 
-This study provided a basis for the use of 'You Draw It' as a tool for testing graphics.
+This study provided a basis for the use of 'You Draw It' as a tool for testing graphics as well as provided support for "ensemble perception" in statistical graphics.
 Further investigation is necessary to implement the method in non-linear settings and with real data.

### 2021-09-10 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
rough draft,. still need to add intro though...

@@
 \item
-  Cleveland and McGill
+  \citet{cleveland1984graphical}
+\item
+  \citet{cleveland1985graphical}
 \end{itemize}
@@
 of the information being presented in the chart. Early researchers
-studied graphs from a psychological perspective. These studies generally
-tested participants ability to detect a stimulus (or a difference
-between two stimuli) ({\textcolor{blue}{CITATIONS}}). Here we focus on
-the how graphical testing has developed in statistics.
+studied graphs from a psychological perspective
+\citep{spence1990visual, lewandowsky1989perception}. These studies
+generally tested participants ability to detect a stimulus or a
+difference between two stimuli. Here we focus on the how graphical
+testing has developed in statistics.
 
@@
 understand the perception of different types of charts
-({\textcolor{blue}{CITATIONS}}). The lineup protocol is one such example
-of the development of tools designed for statistical graphical testing.
-The advancement of graphing software provides the tools necessary to
-develop new methods of testing graphics.
+\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
+The lineup protocol is one such example of the development of tools
+designed for statistical graphical testing. The advancement of graphing
+software provides the tools necessary to develop new methods of testing
+graphics.
 
@@
 system can compute averages of various features in parallel across the
-items in a set ({\textcolor{blue}{CITATIONS}}).
+items in a set
+\citep{chong2003representation, chong2005statistical, van2011rapid}.
 
@@
 In this paper, we establish `You Draw It', adapted from the New York
-Times feature, as tool for graphical testing. The `You Draw It' method
+Times feature, as a tool for graphical testing. The `You Draw It' method
 is validated by replicating the study conducted by
-\citet{mosteller1981eye}. The results of our study provide support for
-``ensemble perception''.
+\citet{mosteller1981eye}. Based on previous research surrounding
+``ensemble perception,'' we hypothesize that regression lines fit by
+human perception resemble regression lines based on the principal axis
+rather than an ordinary least squares regression line.
 
@@
 facilitates user interaction, is used to create the `You Draw It' task
-plots. Integrating this into RShiny using the r2d3 package, participants
-are asked to draw a line using their computer mouse through trend shown
-on their screen. Details of the development of the `You Draw It' task
-plots will be addressed in future work. In the study, participants are
-shown an interactive scatter-plot \cref{fig:ydi-stimuli} along with the
-prompt, ``Use your mouse to fill in the trend in the yellow box
-region.'' The yellow-box region moves along as the participant draws
+plots. Integrating this into RShiny using the \texttt{r2d3} package,
+participants are asked to draw a trend-line using their computer mouse
+through a scatter-plot shown on their screen. In the study, participants
+are shown an interactive scatter-plot (\cref{fig:ydi-stimuli}) along
+with the prompt, ``Use your mouse to fill in the trend in the yellow box
+region.'' The yellow box region moves along as the participant draws
 their trend-line until the yellow region disappears, indicating the
-participant has filled in the entire domain.
+participant has filled in the entire domain. Details of the development
+of the `You Draw It' task plots will be addressed in future work.
 
@@
 
-\caption{'You Draw It' Stimuli}\label{fig:ydi-stimuli}
+\caption{'You Draw It' task plot as shown to particpants during the study.
+The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The last frame (right) illustrates the participants finished trend-line before submission.}\label{fig:ydi-stimuli}
 \end{figure}
@@
 \(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
-relationship. Data were simulated based on linear model with additive
+relationship. Data were simulated based on a linear model with additive
 errors: \begin{align}
@@
 \text{with } e_i & \sim N(0, \sigma^2). \nonumber
-\end{align} The parameters \(\beta_0\) and \(\beta_1\) are selected to
-replicate \citet{mosteller1981eye} with \(e_i\) generated by rejection
-sampling in order to guarantee the points shown align with that of the
-fitted line.
+\end{align}
 
-Simulated model equation parameters were selected to reflect the four
-data sets (F, N, S, and V) used in \citet{mosteller1981eye}
-(\cref{tab:eyefitting-parameters}). Parameter choices F, N, and S
-simulated data across a domain of 0 to 20. Parameter choice F produces a
-trend with a positive slope and a large variance while N has a negative
-slope and a large variance. In comparison, S shows a trend with a
-positive slope with a small variance and V yields a steep positive slope
-with a small variance over the domain of 4 to 16.
+Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
+reflect the four data sets (F, N, S, and V) used in
+\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). Parameter
+choices F, N, and S simulated data across a domain of 0 to 20. Parameter
+choice F produces a trend with a positive slope and a large variance
+while N has a negative slope and a large variance. In comparison, S
+shows a trend with a positive slope with a small variance and V yields a
+steep positive slope with a small variance over the domain of 4 to 16.
 \cref{fig:eyefitting-simplot} illustrates an example of simulated data
@@
 \citet{mosteller1981eye}. Aesthetic design choices were made consistent
-across each of the interactive you draw it plots. The y-axis range
-extended 10\% beyond (above and below) the range of the simulated data
-points to allow for users to draw outside the simulated data set range.
+across each of the interactive `You Draw It' task plots. The y-axis
+range extended 10\% beyond (above and below) the range of the simulated
+data points to allow for users to draw outside the simulated data set
+range.
 
@@
 
-\caption{\label{tab:eyefitting-parameters}Eye Fitting Straight Lines in the Modern Era simulation model parameters}
+\caption{\label{tab:eyefitting-parameters}Designated model equation parameters for simulated data.}
 \centering
@@
 
-\caption{Eye Fitting Straight Lines in the Modern Era Simulated Data Example}\label{fig:eyefitting-simplot}
+\caption{Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
 \end{figure}
@@
 practice plots in order to train participants in the skills associated
-with executing the task followed by four task plots associated with the
-current study. The order of the task plots was randomly assigned for
-each individual in a completely randomized design.
+with executing the task. During the practice session, participants are
+provided with instruction prompts accompanied by a .gif and a practice
+plot. Instructions guide participants to start at the edge of the yellow
+box, to make sure the yellow region is moving along with their mouse as
+they draw, and that they can draw over their already drawn line.
+Practice plots are then followed by four `You Draw It' task plots
+associated with the current study. The order of the task plots was
+randomly assigned for each individual in a completely randomized design.
 
@@
 principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
-illustrates the difference between an OLS regression equation which
+illustrates the difference between an OLS regression line which
 minimizes the vertical distance of points from the line and a regression
-equation based on the principal axis which minimizes the Euclidean
-distance of points (orthogonal) from the line.
+line based on the principal axis which minimizes the Euclidean distance
+of points (orthogonal) from the line.
 
@@
 differ from the predetermined slope. Therefore, we fit an ordinary least
-squares (OLS) regression to each scatter-plot to obtain fitted values in
-0.25 increments across the domain, \(\hat y_{k,OLS}\) for
-\(k = 1, ..., 4 x_{max} +1\).
-
-The regression equation based on the principal axis was determined by
-using the \texttt{princomp} function in the stats package in base R to
-obtain the rotation of the coordinate axes from the first principal
-component (direction which captures the most variance). The estimated
-slope, \(\hat\beta_{1,PCA}\), is determined by the ratio of the axis
-rotation in y and axis rotation in x of the first principal component
-with the y-intercept, \(\hat\beta_{0,PCA}\) calculated by the
-point-slope equation of a line using the mean of of the simulated
-points, \((\bar x_i, \bar y_i)\). Fitted values, \(\hat y_{k,PCA}\), are
-then obtained every 0.25 increment across the domain from the PCA
-regression equation,
+squares (OLS) regression to each scatter-plot to obtain estimated
+parameters \(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted
+values, \(\hat y_{k,OLS}\), are then obtained every 0.25 increments
+across the domain from the OLS regression equation,
+\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
+\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
+principal axis was determined by using the \texttt{princomp} function in
+the stats package in base R to obtain the rotation of the coordinate
+axes from the first principal component (direction which captures the
+most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
+determined by the ratio of the axis rotation in y and axis rotation in x
+of the first principal component with the y-intercept,
+\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
+using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
+Fitted values, \(\hat y_{k,PCA}\), are then obtained every 0.25
+increment across the domain from the PCA regression equation,
 \(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
@@
 
-{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}
+{\centering \includegraphics[width=0.8\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}
 
@@
 
-\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
+\caption{ Comparison between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.}\label{fig:ols-vs-pca-example}
 \end{figure}
 
-\hypertarget{data-availability}{%
-\subsection{Data Availability}\label{data-availability}}
+\hypertarget{residual-trends}{%
+\subsection{Residual Trends}\label{residual-trends}}
 
@@
 across the domain. \cref{fig:eyefitting-example-plot} displays an
-example of all three fitted trend lines for parameter choice F.
+example of all three fitted trend lines for parameter choice F. Data
+used in the analyses are available to be downloaded from GitHub
+\href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/youdrawit-eyefitting-model-data.csv}{here}.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}
+{\centering \includegraphics[width=0.5\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}
 
@@
 
-\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
+\caption{Illustrates the data associated with and collected for one 'You Draw It' task plot.
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
 \end{figure}
@@
 \hypertarget{linear-trend-constraint}{%
-\subsection{Linear Trend Constraint}\label{linear-trend-constraint}}
+\subsubsection{Linear Trend Constraint}\label{linear-trend-constraint}}
 
-Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
-model (LMM) is fit separately to the OLS and PCA residuals, constraining
-the fit to a linear trend. Parameter choice, \(x\), and the interaction
-between \(x\) and parameter choice were treated as fixed effects with a
-random participant effect accounting for variation due to participant.
-The LMM equation for each fit (OLS and PCA) residuals is given by:
-\begin{equation}
+Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
+linear mixed model (LMM) is fit separately to the OLS residuals and PCA
+residuals, constraining the fit to a linear trend. Parameter choice,
+\(x\), and the interaction between \(x\) and parameter choice were
+treated as fixed effects with a random participant effect accounting for
+variation due to participant. The LMM equation for each fit (OLS and
+PCA) is given by: \begin{equation}
 y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
@@
 
+Constraining the residual trend to a linear fit,
+\cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line
+of the residuals between the participant drawn points and fitted values
+for both the OLS regression line and PCA regression line. Estimated
+residual trend lines are overlaid on the observed individual participant
+residuals. Results indicate the estimated trends of PCA residuals
+(orange) appear to align closer to the \(y=0\) horizontal (dashed) line
+than the OLS residuals (blue). In particular, this trend is more
+prominent in parameter choices with large variances (F and N). These
+results are consistent to those found in \citet{mosteller1981eye}
+indicating participants fit a trend-line closer to the estimated
+regression line with the slope of based on the first principal axis than
+the estimated OLS regression line thus, providing support for ``ensemble
+perception''.
+
 \begin{figure}[tbp]
@@
 \hypertarget{smoothing-spline-trend}{%
-\subsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}
+\subsubsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}
 
-Eliminating the linear trend constraint, the bam function in the mgcv
-package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
-generalized additive mixed model (GAMM) separately to the OLS and PCA
-residuals to allow for estimation of smoothing splines. Parameter choice
-was treated as a fixed effect with no estimated intercept and a separate
-smoothing spline for \(x\) was estimated for each parameter choice. A
-random participant effect accounting for variation due to participant
-and a random spline for each participant accounted for variation in
-spline for each participant. The GAMM equation for each fit (OLS and
-PCA) residuals is given by: \begin{equation}
+Eliminating the linear trend constraint, the \texttt{bam} function in
+the mgcv package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to
+fit a generalized additive mixed model (GAMM) separately to the OLS
+residuals and PCA residuals to allow for estimation of smoothing
+splines. Parameter choice was treated as a fixed effect with no
+estimated intercept and a separate smoothing spline for \(x\) was
+estimated for each parameter choice. A random participant effect
+accounting for variation due to participant and a random spline for each
+participant accounted for variation in spline for each participant. The
+GAMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
 y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
@@
 
-\cref{fig:eyefitting-lmer-residualplots} and
-\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
-residuals (vertical deviation of participant drawn points from both the
-OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
-Examining the plots, the estimated trends of PCA residuals (orange)
-appear to align closer to the \(y=0\) horizontal (dashed) line than the
-OLS residuals (blue). In particular, this trend is more prominent in
-parameter choices with large variances (F and N). These results are
-consistent to those found in \citet{mosteller1981eye} indicating
-participants fit a trend line closer to the estimated regression line
-with the slope of the first principal component than the estimated OLS
-regression line.
+Allowing for flexibility in the residual trend,
+\cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line
+of the residuals between the participant drawn points and fitted values
+for both the OLS regression line and PCA regression line. Estimated
+residual trends are overlaid on the observed individual participant
+residuals. The results of the GAMM align with those shown in
+\cref{fig:eyefitting-lmer-residualplots} providing support that for
+scatter-plots with more noise (F and N), estimated trends of PCA
+residuals (orange) appear to align closer to the \(y=0\) horizontal
+(dashed) line than the OLS residuals (blue). By fitting smoothing
+splines, we can determine whether participants naturally fit a straight
+trend-line to the set of points or whether they deviate throughout the
+domain. In particular, in scatter-plots with smaller variance (S and V),
+we can see that participants began at approximately the correct starting
+point then deviated away from the fitted regression lines and correcting
+for their fit toward the end of their trend-line. In scatter-plots with
+larger variance (F and N), participants estimated their starting value
+in the extreme direction of the OLS regression line based on the
+increasing or decreasing trend but more accurately represented the
+starting value of the PCA regression line. As participants continued
+their trend-line, they crossed through the OLS regression line
+indicating they estimated the slope in the extreme direction. These
+results provide further insight into the curvature humans perceive in a
+set of points.
 
@@
 
-The intent of research was to establish you draw it as a tool for
-testing graphics then use this tool to determine the cognitive
-implications of displaying data on the log scale. Eye Fitting Straight
-Lines in the Modern Era replicated the results found in
-\citet{mosteller1981eye}. When shown points following a linear trend,
-participants tended to fit the slope of the first principal component
-over the slope of the least squares regression line. This trend was most
-prominent when shown data simulated with larger variances. The
-reproducibility of these results serve as evidence of the reliability of
-the you draw it method.
+The intent of research was to adapt `You Draw It' from the New York
+Times feature as a tool and method for testing graphics The study
+conducted by \citet{mosteller1981eye} was replicated using the `You Draw
+It' method providing support for the validity of the tool. Our results
+indicated that when shown points following a linear trend, participants
+tended to fit a line closer to a regression based on the principal axis
+over the ordinary least squares regression line. This trend was most
+prominent when shown data simulated with larger variances. These results
+indicate that humans perform ``ensemble perception'' in a statistical
+graphic setting as participants minimized the distance from the their
+regression line over both the \(x\) and \(y\) axis simultaneously. In
+addition, we allowed participants to draw trend lines that deviated from
+a straight line providing insight into the curvature the human eye
+perceives in a set of points.
 
@@
 This study provided a basis for the use of `You Draw It' as a tool for
-testing graphics. Further investigation is necessary to implement the
-method in non-linear settings and with real data. This tool could also
-be used to evaluate human ability to extrapolate data from trends. In
-the future, an R package designed for easy implementation of `You Draw
-It' task plots will help make this tool accessible to other researchers.
+testing graphics as well as provided support for ``ensemble perception''
+in statistical graphics. Further investigation is necessary to implement
+the method in non-linear settings and with real data. This tool could
+also be used to evaluate human ability to extrapolate data from trends.
+In the future, an R package designed for easy implementation of `You
+Draw It' task plots will help make this tool accessible to other
+researchers.
 

### 2021-09-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
first draft of manuscript. probably needs a lot of improvements :grimacing: <- also very curious to know if you can put emjoi's into commits!

@@
 
-+ What are graphs? Why do we care?
+Data visualization has become a central tool in modern data science.
+@unwin2020data defines data visualization as the art of graphic displays to show data. 
+Graphics illustrate the relationship between two sets of information or variables. 
+Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.
+When creating a graphic, we must consider the design choices most effective for conveying the intended result.
+In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures. 
+Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
+Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
 
@@
 
-+ @cleveland1984graphical
-+ @cleveland1985graphical
+With the widespread use of graphics, work has been done to understand the effect of charts and graphs on human perception and higher cognitive processing such as learning and analysis [@green2009personal].
+@carpenter1998model established the process of graph comprehension in which viewers interact with charts by first perceptually observing the visual features and later translating to cognitive processing of the information depicted by those features.
+Much of the research on the perception of visual features in charts has been conducted in psychophysics and test for accuracy and quantitative comparisons when understanding a plot. 
+Cleveland and McGill conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons [@cleveland1984graphical].
+For example, it is more effective to display information on an $x$ or $y$ axis rather than using color in order to reduce the visual effort necessary to make numerical comparisons.
+Other early studies evaluated human perception of lines. 
+For instance, @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
+The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
 
@@
 Participants were recruited through through Twitter, Reddit, and direct email in May 2021. 
-A total of 39 individuals completed 256 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
+A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
 All participants had normal or corrected to normal vision and signed an informed consent form.
@@
 
-```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era LMM results"}
+```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
+Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
 eyefitting.preds.lmer <- read.csv("data/youdrawit-eyefitting-lmerpred-data.csv")
@@
 
-```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era GAMM results"}
+```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
+Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
 eyefitting.grid.gamm <- read.csv("data/youdrawit-eyefitting-gammpred-data.csv")

### 2021-09-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
first draft of manuscript. probably needs a lot of improvements :grimacing: <- also very curious to know if you can put emjoi's into commits!

@@
 
-\begin{itemize}
-\tightlist
-\item
-  What are graphs? Why do we care?
-\end{itemize}
+Data visualization has become a central tool in modern data science.
+\citet{unwin2020data} defines data visualization as the art of graphic
+displays to show data. Graphics illustrate the relationship between two
+sets of information or variables. Graphical displays are useful for data
+cleaning, exploratory data analysis, identifying trends, evaluating
+model output, and presenting results. When creating a graphic, we must
+consider the design choices most effective for conveying the intended
+result. In order to evaluate the effectiveness of a graphic, we must
+understand cognitive perception and biases related to different types of
+data structures. Efforts in the field of graphics have developed
+graphical testing tools and methods such as the lineup protocol
+\citep{buja2009statistical} to provide a framework for inferential
+testing. Through experimentation, methods such as the lineup protocol
+allow researchers to conduct studies geared at understanding human
+ability to conduct tasks related to the perception of statistical charts
+such as differentiation, prediction, estimation, and extrapolation.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  \citet{cleveland1984graphical}
-\item
-  \citet{cleveland1985graphical}
-\end{itemize}
+With the widespread use of graphics, work has been done to understand
+the effect of charts and graphs on human perception and higher cognitive
+processing such as learning and analysis \citep{green2009personal}.
+\citet{carpenter1998model} established the process of graph
+comprehension in which viewers interact with charts by first
+perceptually observing the visual features and later translating to
+cognitive processing of the information depicted by those features. Much
+of the research on the perception of visual features in charts has been
+conducted in psychophysics and test for accuracy and quantitative
+comparisons when understanding a plot. Cleveland and McGill conducted a
+series of cognitive tasks designed to establish a hierarchy of visual
+components for making comparisons \citep{cleveland1984graphical}. For
+example, it is more effective to display information on an \(x\) or
+\(y\) axis rather than using color in order to reduce the visual effort
+necessary to make numerical comparisons. Other early studies evaluated
+human perception of lines. For instance, \citet{cleveland1985graphical}
+found that assessing the position of points along an axis is easier than
+determining the slope of a line. The results of these cognitive tasks
+provide some consistent guidance for chart design; however, other
+methods of visual testing can further evaluate design choices and help
+us understand cognitive biases related to the evaluation of statistical
+charts.
 
@@
 Participants were recruited through through Twitter, Reddit, and direct
-email in May 2021. A total of 39 individuals completed 256 unique `You
+email in May 2021. A total of 35 individuals completed 119 unique `You
 Draw It' task plots; all completed you draw it task plots were included
@@
 
-\caption{Eye Fitting Straight Lines in the Modern Era LMM results}\label{fig:eyefitting-lmer-residualplots}
+\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
+Estimated residual trends with 95\% confidence bands are overlaid on the observed individual participant residuals.}\label{fig:eyefitting-lmer-residualplots}
 \end{figure}
@@
 
-\caption{Eye Fitting Straight Lines in the Modern Era GAMM results}\label{fig:eyefitting-gamm-residualplots}
+\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
+Estimated residual trends with 95\% confidence bands are overlaid on the observed individual participant residuals.}\label{fig:eyefitting-gamm-residualplots}
 \end{figure}

### 2021-09-19 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
Some edits to intro/abstract

@@
 abstract: |
-  Fitting lines by eye through a set of points has been explored since the 20th century.
+  \svp{How do statistical regression results compare to intuitive, visually fitted results?} Fitting lines by eye through a set of points has been explored since the 20th century.
 Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It'.
-Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics by drawing a line on their computer screen with their computer mouse.
-In my research, I established ‘You Draw It’ as a method for graphical testing by adapting the New York Times feature.
-I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981).
-Participants were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse; task plots were generated using the r2d3 package in R statistical software.
-Results from my study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
+In 2015, the New York Times introduced an interactive feature, called 'You Draw It'\svp{, where r}eaders are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
+\svp{In this paper, we validate} ‘You Draw It’ as a method for graphical testing, \svp{comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.} Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
 This trend was most prominent when shown data simulated with larger variances.
-The reproducibility of these results serves as evidence of the reliability of the you draw it method.
-Future work is necessary to implement the 'You Draw It' tool as a method of testing graphics.
-[200 word limit]
+\svp{This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.}
 
@@
 
-Data visualization has become a central tool in modern data science.
-@unwin2020data defines data visualization as the art of graphic displays to show data. 
+Data visualization has become a central tool in modern data science.\svp{XXX this is a bit of a lame intro... we might want to spice this up later, but not critical now}
+@unwin2020data defines data visualization as the art of \svp{using} graphic displays to show data.
+\svp{XXX how does this relate - need to make it flow a bit better.
+Add additional clauses or connect this to the next thought somehow.}
 Graphics illustrate the relationship between two sets of information or variables. 
-Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.
+Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.\svp{Right now this reads like bullet points rather than a narrative.}
+
 When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures. 
+\svp{For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering \citep{vanderplas2017clusters}.}
+These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.}
+<!-- In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures.  -->
 Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.

### 2021-09-21 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
knitted Susan's recent changes

@@
 \begin{abstract}
+\svp{How do statistical regression results compare to intuitive, visually fitted results?}
 Fitting lines by eye through a set of points has been explored since the
@@
 then drawing the line through the set of points. In 2015, the New York
-Times introduced an interactive feature, called `You Draw It'. Readers
-are asked to input their own assumptions about various metrics and
-compare how these assumptions relate to reality. The New York Times team
-utilizes Data Driven Documents (D3) that allows readers to predict these
-metrics by drawing a line on their computer screen with their computer
-mouse. In my research, I established `You Draw It' as a method for
-graphical testing by adapting the New York Times feature. I recruited
-participants via crowdsourcing websites and replicated the study found
-in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants
-were directed to an RShiny application link and shown points following a
-linear trend and asked to draw a line through the data points using
-their computer mouse; task plots were generated using the r2d3 package
-in R statistical software. Results from my study were consistent with
-those found in the previous study; when shown points following a linear
-trend, participants tended to fit the slope of the first principal
-component over the slope of the least-squares regression line. This
-trend was most prominent when shown data simulated with larger
-variances. The reproducibility of these results serves as evidence of
-the reliability of the you draw it method. Future work is necessary to
-implement the `You Draw It' tool as a method of testing graphics. {[}200
-word limit{]}
+Times introduced an interactive feature, called `You Draw
+It'\svp{, where r}eaders are asked to input their own assumptions about
+various metrics and compare how these assumptions relate to reality.
+\svp{In this paper, we validate} `You Draw It' as a method for graphical
+testing,
+\svp{comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.}
+Results were consistent with those found in the previous study; when
+shown points following a linear trend, participants tended to fit the
+slope of the first principal component over the slope of the
+least-squares regression line. This trend was most prominent when shown
+data simulated with larger variances.
+\svp{This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.}
 \end{abstract}
@@
 
-Data visualization has become a central tool in modern data science.
-\citet{unwin2020data} defines data visualization as the art of graphic
-displays to show data. Graphics illustrate the relationship between two
-sets of information or variables. Graphical displays are useful for data
-cleaning, exploratory data analysis, identifying trends, evaluating
-model output, and presenting results. When creating a graphic, we must
-consider the design choices most effective for conveying the intended
-result. In order to evaluate the effectiveness of a graphic, we must
-understand cognitive perception and biases related to different types of
-data structures. Efforts in the field of graphics have developed
-graphical testing tools and methods such as the lineup protocol
-\citep{buja2009statistical} to provide a framework for inferential
-testing. Through experimentation, methods such as the lineup protocol
-allow researchers to conduct studies geared at understanding human
-ability to conduct tasks related to the perception of statistical charts
-such as differentiation, prediction, estimation, and extrapolation.
+Data visualization has become a central tool in modern data
+science.{\textcolor{RedOrange}{XXX this is a bit of a lame intro... we might want to spice this up later, but not critical now}}
+\citet{unwin2020data} defines data visualization as the art of
+{\textcolor{RedOrange}{using}} graphic displays to show data.
+{\textcolor{RedOrange}{XXX how does this relate - need to make it flow a bit better. Add additional clauses or connect this to the next thought somehow.}}
+Graphics illustrate the relationship between two sets of information or
+variables. Graphical displays are useful for data cleaning, exploratory
+data analysis, identifying trends, evaluating model output, and
+presenting
+results.{\textcolor{RedOrange}{Right now this reads like bullet points rather than a narrative.}}
+
+When creating a graphic, we must consider the design choices most
+effective for conveying the intended result.
+{\textcolor{RedOrange}{For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering \citep{vanderplas2017clusters}.}}
+These design choices require that we understand the perception and
+visual biases that come into play when creating graphics, and as
+graphics are evaluated visually, we must use human testing to ground our
+understanding in empiricism.\} Efforts in the field of graphics have
+developed graphical testing tools and methods such as the lineup
+protocol \citep{buja2009statistical} to provide a framework for
+inferential testing. Through experimentation, methods such as the lineup
+protocol allow researchers to conduct studies geared at understanding
+human ability to conduct tasks related to the perception of statistical
+charts such as differentiation, prediction, estimation, and
+extrapolation.
 

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on submission

--- 2021-09-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
@@
 - name: Emily A. Robinson 1
-  #thanks: The authors gratefully acknowledge ...
+  # thanks: The authors gratefully acknowledge ...
   affiliation: Department of Statistics, University of Nebraska - Lincoln
@@
 abstract: |
+  How do statistical regression results compare to intuitive, visually fitted results?
 Fitting lines by eye through a set of points has been explored since the 20th century.
 Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It'.
-Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics by drawing a line on their computer screen with their computer mouse.
-In my research, I established ‘You Draw It’ as a method for graphical testing by adapting the New York Times feature.
-I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981).
-Participants were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse; task plots were generated using the r2d3 package in R statistical software.
-Results from my study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
+In 2015, the New York Times introduced an interactive feature, called 'You Draw It', where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
+In this paper, we validate ‘You Draw It’ as a method for graphical testing, comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.
+Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
 This trend was most prominent when shown data simulated with larger variances.
-The reproducibility of these results serves as evidence of the reliability of the you draw it method.
-Future work is necessary to implement the 'You Draw It' tool as a method of testing graphics.
-[200 word limit]
+This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
 
@@
 # Introduction
+We all use statistical graphics, but how do we know that the graphics we use are communicating properly?
+When creating a graphic, we must consider the design choices most effective for conveying the intended result.
+For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
+These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
 
-Data visualization has become a central tool in modern data science.
-@unwin2020data defines data visualization as the art of graphic displays to show data. 
-Graphics illustrate the relationship between two sets of information or variables. 
-Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.
-When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures. 
-Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
-Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
-
-## Graph Perception
-
-With the widespread use of graphics, work has been done to understand the effect of charts and graphs on human perception and higher cognitive processing such as learning and analysis [@green2009personal].
-@carpenter1998model established the process of graph comprehension in which viewers interact with charts by first perceptually observing the visual features and later translating to cognitive processing of the information depicted by those features.
 Much of the research on the perception of visual features in charts has been conducted in psychophysics and test for accuracy and quantitative comparisons when understanding a plot. 
@@
 These studies generally tested participants ability to detect a stimulus or a difference between two stimuli.
-Here we focus on the how graphical testing has developed in statistics.
+In statistical graphics research, @vanderplas2015signs uses methods of psychophysics to estimate the effect of sign illusions by adjusting the sine illusion up and down.
 
@@
 
+Efforts in the field of statistical graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
+Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
 One useful tool for testing statistical graphics is the concept of a lineup.
@@
 The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
-The lineup protocol is one such example of the development of tools designed for statistical graphical testing.
-The advancement of graphing software provides the tools necessary to develop new methods of testing graphics.
+The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
 
@@
 The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
-Based on previous research surrounding "ensemble perception," we hypothesize that regression lines fit by human perception resemble regression lines based on the principal axis rather than an ordinary least squares regression line.
+Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than a ordinary least squares regression.
+In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.
 
@@
 The experimental tasks took approximately 15 minutes to complete.
+While convenience sampling took place, this is primarily a perceptual task and previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
+Data collected in this study serves the purpose of a pilot study intended to lay the foundation for future data collection and provides promising results for understanding the perception of regression.
 Participants completed the experiment on their own computers in an environment of their choosing.
@@
 
-The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics
-The study conducted by @mosteller1981eye was replicated using the 'You Draw It' method providing support for the validity of the tool.
-Our results indicated that when shown points following a linear trend, participants tended to fit a line closer to a regression based on the principal axis over the ordinary least squares regression line.
-This trend was most prominent when shown data simulated with larger variances. 
-These results indicate that humans perform "ensemble perception" in a statistical graphic setting as participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously.
-In addition, we allowed participants to draw trend lines that deviated from a straight line providing insight into the curvature the human eye perceives in a set of points.
+The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
+We provided support for the validity of the 'You Draw It' method by replicating the study found in @mosteller1981eye.
+Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
+Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principle component regression as opposed to ordinary least squares regression.
+Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principle component regression.
+Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
+These results provide support that humans perform "ensemble perception" in a statistical graphic setting.
+We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.
 
@@
 
-This study provided a basis for the use of 'You Draw It' as a tool for testing graphics as well as provided support for "ensemble perception" in statistical graphics.
-Further investigation is necessary to implement the method in non-linear settings and with real data.
+This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
+Further investigation is necessary to implement this method in non-linear settings and with real data.
 This tool could also be used to evaluate human ability to extrapolate data from trends. 
 In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.
+
+\begin{center}
+{\large\bf SUPPLEMENTARY MATERIAL}
+\end{center}
+
+\begin{description}
+
+\item[Participant Data:] De-identified participant data can be found at ...
+
+\item[Data Analysis Code:]
+
+\end{description}

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
working on submission

--- 2021-09-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
@@
 \usepackage{hyperref}
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-% Pandoc citation processing
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
   \author{
-        Emily A. Robinson 1 \\
+        Emily A. Robinson 1 \thanks{The authors gratefully acknowledge
+\ldots{}} \\
     Department of Statistics, University of Nebraska - Lincoln\\
@@
 \begin{abstract}
-Fitting lines by eye through a set of points has been explored since the
-20th century. Common methods of fitting trends by eye involve
-maneuvering a string, black thread, or ruler until the fit is suitable,
-then drawing the line through the set of points. In 2015, the New York
-Times introduced an interactive feature, called `You Draw It'. Readers
-are asked to input their own assumptions about various metrics and
-compare how these assumptions relate to reality. The New York Times team
-utilizes Data Driven Documents (D3) that allows readers to predict these
-metrics by drawing a line on their computer screen with their computer
-mouse. In my research, I established `You Draw It' as a method for
-graphical testing by adapting the New York Times feature. I recruited
-participants via crowdsourcing websites and replicated the study found
-in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants
-were directed to an RShiny application link and shown points following a
-linear trend and asked to draw a line through the data points using
-their computer mouse; task plots were generated using the r2d3 package
-in R statistical software. Results from my study were consistent with
-those found in the previous study; when shown points following a linear
-trend, participants tended to fit the slope of the first principal
-component over the slope of the least-squares regression line. This
-trend was most prominent when shown data simulated with larger
-variances. The reproducibility of these results serves as evidence of
-the reliability of the you draw it method. Future work is necessary to
-implement the `You Draw It' tool as a method of testing graphics. {[}200
-word limit{]}
+How do statistical regression results compare to intuitive, visually
+fitted results? Fitting lines by eye through a set of points has been
+explored since the 20th century. Common methods of fitting trends by eye
+involve maneuvering a string, black thread, or ruler until the fit is
+suitable, then drawing the line through the set of points. In 2015, the
+New York Times introduced an interactive feature, called `You Draw It',
+where readers are asked to input their own assumptions about various
+metrics and compare how these assumptions relate to reality. In this
+paper, we validate `You Draw It' as a method for graphical testing,
+comparing results to the less technological method utilized in
+\citet{mosteller1981eye} and extending that study with formal
+statistical analysis methods. Results were consistent with those found
+in the previous study; when shown points following a linear trend,
+participants tended to fit the slope of the first principal component
+over the slope of the least-squares regression line. This trend was most
+prominent when shown data simulated with larger variances. This study
+reinforces the differences between intuitive visual model fitting and
+statistical model fitting, providing information about human perception
+as it relates to the use of statistical graphics.
 \end{abstract}
@@
 
+--\textgreater{}
+
 \hypertarget{introduction}{%
@@
 
-Data visualization has become a central tool in modern data science.
-\citet{unwin2020data} defines data visualization as the art of graphic
-displays to show data. Graphics illustrate the relationship between two
-sets of information or variables. Graphical displays are useful for data
-cleaning, exploratory data analysis, identifying trends, evaluating
-model output, and presenting results. When creating a graphic, we must
+We all use statistical graphics, but how do we know that the graphics we
+use are communicating properly? When creating a graphic, we must
 consider the design choices most effective for conveying the intended
-result. In order to evaluate the effectiveness of a graphic, we must
-understand cognitive perception and biases related to different types of
-data structures. Efforts in the field of graphics have developed
-graphical testing tools and methods such as the lineup protocol
-\citep{buja2009statistical} to provide a framework for inferential
-testing. Through experimentation, methods such as the lineup protocol
-allow researchers to conduct studies geared at understanding human
-ability to conduct tasks related to the perception of statistical charts
-such as differentiation, prediction, estimation, and extrapolation.
+result. For instance, we must decide to highlight the relationship
+between two variables in a scatterplot by including a trend line, or
+adding color to highlight clustering \citep{vanderplas2017clusters}.
+These design choices require that we understand the perception and
+visual biases that come into play when creating graphics, and as
+graphics are evaluated visually, we must use human testing to ground our
+understanding in empiricism.
 
-\hypertarget{graph-perception}{%
-\subsection{Graph Perception}\label{graph-perception}}
-
-With the widespread use of graphics, work has been done to understand
-the effect of charts and graphs on human perception and higher cognitive
-processing such as learning and analysis \citep{green2009personal}.
-\citet{carpenter1998model} established the process of graph
-comprehension in which viewers interact with charts by first
-perceptually observing the visual features and later translating to
-cognitive processing of the information depicted by those features. Much
-of the research on the perception of visual features in charts has been
-conducted in psychophysics and test for accuracy and quantitative
+Much of the research on the perception of visual features in charts has
+been conducted in psychophysics and test for accuracy and quantitative
 comparisons when understanding a plot. Cleveland and McGill conducted a
@@
 generally tested participants ability to detect a stimulus or a
-difference between two stimuli. Here we focus on the how graphical
-testing has developed in statistics.
+difference between two stimuli. In statistical graphics research,
+\citet{vanderplas2015signs} uses methods of psychophysics to estimate
+the effect of sign illusions by adjusting the sine illusion up and down.
 
@@
 
+Efforts in the field of statistical graphics have developed graphical
+testing tools and methods such as the lineup protocol
+\citep{buja2009statistical} to provide a framework for inferential
+testing. Through experimentation, methods such as the lineup protocol
+allow researchers to conduct studies geared at understanding human
+ability to conduct tasks related to the perception of statistical charts
+such as differentiation, prediction, estimation, and extrapolation
+\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
 One useful tool for testing statistical graphics is the concept of a
@@
 \citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
-The lineup protocol is one such example of the development of tools
-designed for statistical graphical testing. The advancement of graphing
-software provides the tools necessary to develop new methods of testing
-graphics.
+The advancement of graphing software provides the tools necessary to
+develop new methods of testing statistical graphics.
 
@@
 \citet{mosteller1981eye}. Based on previous research surrounding
-``ensemble perception,'' we hypothesize that regression lines fit by
-human perception resemble regression lines based on the principal axis
-rather than an ordinary least squares regression line.
+``ensemble perception,'' we hypothesize that visual regression tends to
+mimic principle component regression rather than a ordinary least
+squares regression. In order to assess this hypothesis, we introduce a
+method for statistically modeling the participant drawn lines using
+generalized additive mixed models.
 
@@
 vision and signed an informed consent form. The experimental tasks took
-approximately 15 minutes to complete. Participants completed the
-experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through an RShiny
-application found
+approximately 15 minutes to complete. While convenience sampling took
+place, this is primarily a perceptual task and previous results have
+found few differences between expert and non-expert participants in this
+context \citep{vanderplas2015spatial}. Data collected in this study
+serves the purpose of a pilot study intended to lay the foundation for
+future data collection and provides promising results for understanding
+the perception of regression. Participants completed the experiment on
+their own computers in an environment of their choosing. The experiment
+was conducted and distributed through an RShiny application found
 \href{https://shiny.srvanderplas.com/you-draw-it/}{here}.
@@
 The intent of research was to adapt `You Draw It' from the New York
-Times feature as a tool and method for testing graphics The study
-conducted by \citet{mosteller1981eye} was replicated using the `You Draw
-It' method providing support for the validity of the tool. Our results
-indicated that when shown points following a linear trend, participants
-tended to fit a line closer to a regression based on the principal axis
-over the ordinary least squares regression line. This trend was most
-prominent when shown data simulated with larger variances. These results
-indicate that humans perform ``ensemble perception'' in a statistical
-graphic setting as participants minimized the distance from the their
-regression line over both the \(x\) and \(y\) axis simultaneously. In
-addition, we allowed participants to draw trend lines that deviated from
-a straight line providing insight into the curvature the human eye
-perceives in a set of points.
+Times feature as a tool and method for testing graphics and introduce a
+method for statistically modeling the participant drawn lines. We
+provided support for the validity of the `You Draw It' method by
+replicating the study found in \citet{mosteller1981eye}. Using
+generalized additive mixed models, we assessed the deviation of the
+participant drawn lines from the statistically fitted regression lines.
+Our results found that when shown points following a linear trend,
+participants visually fit a regression line that mimics the first
+principle component regression as opposed to ordinary least squares
+regression. Data simulated with a larger variance provided strong
+support for a participants tendency to visually fit the first principle
+component regression. Our results indicate that participants minimized
+the distance from the their regression line over both the \(x\) and
+\(y\) axis simultaneously. These results provide support that humans
+perform ``ensemble perception'' in a statistical graphic setting. We
+allowed participants to draw trend lines that deviated from a straight
+line and gained an insight into the curvature the human eye perceives in
+a set of points.
 
@@
 This study provided a basis for the use of `You Draw It' as a tool for
-testing graphics as well as provided support for ``ensemble perception''
-in statistical graphics. Further investigation is necessary to implement
-the method in non-linear settings and with real data. This tool could
-also be used to evaluate human ability to extrapolate data from trends.
-In the future, an R package designed for easy implementation of `You
-Draw It' task plots will help make this tool accessible to other
-researchers.
+testing statistical graphics and introduced a method for statistically
+modeling participant drawn lines using generalized additive mixed
+models. Further investigation is necessary to implement this method in
+non-linear settings and with real data. This tool could also be used to
+evaluate human ability to extrapolate data from trends. In the future,
+an R package designed for easy implementation of `You Draw It' task
+plots will help make this tool accessible to other researchers.
+
+\begin{center}
+{\large\bf SUPPLEMENTARY MATERIAL}
+\end{center}
+
+\begin{description}
+
+\item[Title:] Brief description. (file type)
+
+\item[R-package for  MYNEW routine:] R-package <U+FFFD>MYNEW<U+FFFD> containing code to perform the diagnostic methods described in the article.
+The package also contains all datasets used as examples in the article.
+(GNU zipped tar file)
+
+\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)
+
+\end{description}
 

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on edits.

@@
 keywords:
-- Graphics
-- Regression
+- Cognitive Bias
 - Graph Perception
-- Scatterplot
-- Cognitive Bias
+- Graphical Testing
+- Linear Regression
+- Statistical Graphics
 
@@
 library(ggforce)
+library(here)
 ```
@@
 
-eyefitting_example_sim <- read.csv("data/youdrawit-eyefitting-simdata-example.csv")
+eyefitting_example_sim <- read.csv(here("data/youdrawit-eyefitting-simdata-example.csv"))
 

### 2022-01-24 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
working on edits.

@@
 \noindent%
-{\it Keywords:} Graphics, Regression, Graph
-Perception, Scatterplot, Cognitive Bias
+{\it Keywords:} Cognitive Bias, Graph Perception, Graphical
+Testing, Linear Regression, Statistical Graphics
 \vfill
@@
 \spacingset{1.45} % DON'T change the spacing!
-
---\textgreater{}
 
@@
 
-\item[Title:] Brief description. (file type)
+\item[Participant Data:] De-identified participant data can be found at ...
 
-\item[R-package for  MYNEW routine:] R-package <U+FFFD>MYNEW<U+FFFD> containing code to perform the diagnostic methods described in the article.
-The package also contains all datasets used as examples in the article.
-(GNU zipped tar file)
-
-\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)
+\item[Data Analysis Code:]
 

### 2022-01-26 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
added analysis code files to this repository.

@@
 
-eyefitting_example_sim <- read.csv(here("data/youdrawit-eyefitting-simdata-example.csv"))
+eyefitting_example_sim <- read.csv(here("results/youdrawit-eyefitting-simdata-example.csv"))
 
@@
 \cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.
-Data used in the analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/youdrawit-eyefitting-model-data.csv).
 
@@
 Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).", out.width="50%"}
-trial.feedback <- read.csv("data/youdrawit-eyefitting-example-feedback.csv") %>%
+trial.feedback <- read.csv("results/youdrawit-eyefitting-example-feedback.csv") %>%
     mutate(`Parameter Choice` = "F")
-trial.sim <- read.csv("data/youdrawit-eyefitting-example-simulated.csv") %>%
+trial.sim <- read.csv("results/youdrawit-eyefitting-example-simulated.csv") %>%
     mutate(`Parameter Choice` = "F")
@@
 Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
-eyefitting.preds.lmer <- read.csv("data/youdrawit-eyefitting-lmerpred-data.csv")
+eyefitting.preds.lmer <- read.csv("results/youdrawit-eyefitting-lmerpred-data.csv")
 # Plot Predictions
@@
 Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
-eyefitting.grid.gamm <- read.csv("data/youdrawit-eyefitting-gammpred-data.csv")
+eyefitting.grid.gamm <- read.csv("results/youdrawit-eyefitting-gammpred-data.csv")
 eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
@@
 
-\begin{description}
-
-\item[Participant Data:] De-identified participant data can be found at ...
-
-\item[Data Analysis Code:]
-
-\end{description}
++ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
++ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
++ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://shiny.srvanderplas.com/you-draw-it/)

### 2022-01-26 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
added analysis code files to this repository.

@@
 across the domain. \cref{fig:eyefitting-example-plot} displays an
-example of all three fitted trend lines for parameter choice F. Data
-used in the analyses are available to be downloaded from GitHub
-\href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/youdrawit-eyefitting-model-data.csv}{here}.
+example of all three fitted trend lines for parameter choice F.
 
@@
 
-\begin{description}
-
-\item[Participant Data:] De-identified participant data can be found at ...
-
-\item[Data Analysis Code:]
-
-\end{description}
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Participant Data:} De-identified participant data collected in
+  the study and used for analyses are available to be downloaded from
+  GitHub
+  \href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}{here}.
+\item
+  \textbf{Data Analysis Code:} The code used to replicate the analysis
+  in this paper can be found
+  \href{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}{here}.
+\item
+  \textbf{Study Applet:} The shiny app used to conduct the study can be
+  accessed \href{https://shiny.srvanderplas.com/you-draw-it/}{here}
+\end{itemize}
 

### 2022-01-26 analysis/you-draw-it-eyefitting-analysis.Rmd
added analysis code files to this repository.

---
title: "Eye Fitting Straight Lines in the Modern Era"
subtitle: "Data Analysis"
author: "Emily Robinson"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
library(knitr)
```

<br>

**Rmd file containing code can be accessed [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/analysis).**

## SETUP

+ Load libraries
+ Functions for computing confidence intervals

```{r setup2}
library(tidyverse)
library(kableExtra)
library(readr)
library(here)
library(mgcv)
library(lme4)
library(lmerTest)
library(emmeans)
source(here("analysis/gamm-predict-function.R"))
```

## PARTICIPANT DATA

De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).

```{r data}
eyefitting_model_data  <- read_csv(here("data/youdrawit-eyefitting-model-data.csv"))

factorCols <- c("participantID", "nick_name", "study_starttime", "age", "gender", "academic_study",
                "recruitment", "plotID", "parm_id")
eyefitting_model_data[,factorCols] <- lapply(eyefitting_model_data[,factorCols], factor)
```

```{r data-table, echo = F}
eyefitting_model_data[1:10,] %>%
  kbl() %>%
  kable_paper() %>%
  scroll_box(width = "100%", height = "200px")

eyefitting_model_data <- eyefitting_model_data %>%
  dplyr::mutate(`Parameter Choice` = parm_id)
```

## RAW DATA PLOTS

### OLS with loess smoother

```{r yloess-ols-plot}
eyefitting_model_data %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yloess, group = plotID), alpha = 0.5, color = "steelblue") +
  geom_line(alpha = 0.2, aes(y = yols, group = participantID)) +
  # geom_line(alpha = 0.2, aes(y = ypca, group = participantID), linetype = "dashed") +
  facet_wrap(~ parm_id) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_x_continuous(limits = c(0, 20)) +
  ggtitle("OLS")
```

### PCA with loess smoother

```{r yloess-pca-plot}
eyefitting_model_data %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yloess, group = plotID), alpha = 0.5, color = "steelblue") +
  # geom_line(alpha = 0.2, aes(y = yols, group = participantID)) +
  geom_line(alpha = 0.2, aes(y = ypca, group = participantID)) +
  facet_wrap(~ parm_id) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_x_continuous(limits = c(0, 20)) +
  ggtitle("PCA")
```

### Residuals with loess smoother

+ residualols = ydrawn - yols, denoted $e_{ols}$
+ residualpca = ydrawn - ypca, denoted $e_{pca}$

```{r residual-plot}
eyefitting_model_data %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = residualols.loess, group = plotID, color = "OLS"), alpha = 0.3) +
  geom_line(aes(y = residualpca.loess, group = plotID, color = "PCA"), alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~parm_id) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  scale_x_continuous(limits = c(0, 20)) +
  scale_color_manual("Estimate", values = c("steelblue", "orange"))
```

## LINEAR CONSTRAINT (Linear Mixed Model)

**Treatments:**

+ parm_ID (S, F, V, N)
+ x (0, 20)

**Response:** raw residuals

+ residualols = ydrawn - yols, denoted $e_{ols}$
+ residualpca = ydrawn - ypca, denoted $e_{pca}$

**Experimental Design:**

+ Each participant saw each of the 4 plots (new simulated data for each)

**ANOVA Table:**

```{r, anova-lmer, echo = F}
sv <- c("participantID", "parm_id", "x", "x:parm_id", "residual")
df <- c("(50 - 1) = 49", "(4 - 1) = 3", "1", "(4 - 1) = 3", "143 - by subtraction")
data.frame("SV" = sv, "DF" = df) %>% kable()
```

**LMM Model:**

$$y_{drawn} - y_{ols} = e_{ij,ols} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ij} + \gamma_{2i} x_{ij}\right] + p_{j} + \epsilon_{ij}$$

+ $\gamma_0$ is the overall intercept
+ $\alpha_i$ is the effect of the parameter combination on the intercept (i.e. how much the intercept adjusts for each parameter combo)
+ $\gamma_1$ is the overall slope for x
+ $\gamma_{2i}$ is the effect of the parameter combination on the slope (i.e. think unequal slopes)
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the participant error due to participant variation
+ $\epsilon_{ij} \sim N(0, \sigma^2)$ is the residual error.

```{r lmer, fig.height = 9}
# OLS
eyefitting.ols.lmer <- lmer(residualols ~ -1 + parm_id + x:parm_id + (1|participantID),
                       data = eyefitting_model_data)
anova(eyefitting.ols.lmer)

# PCA
eyefitting.pca.lmer <- lmer(residualpca ~ -1 + parm_id + x:parm_id + (1|participantID),
                            data = eyefitting_model_data)
anova(eyefitting.pca.lmer)

# Obtain Predictions
eyefitting.ols.grid.lmer  <- ref_grid(eyefitting.ols.lmer, at = list(x = seq(1,20,0.5)))
eyefitting.ols.preds.lmer <- emmeans(eyefitting.ols.grid.lmer, ~ parm_id:x, cov.reduce = FALSE) %>% 
  as_tibble()

eyefitting.pca.grid.lmer  <- ref_grid(eyefitting.pca.lmer, at = list(x = seq(1,20,0.5)))
eyefitting.pca.preds.lmer <- emmeans(eyefitting.pca.grid.lmer, ~ parm_id:x, cov.reduce = FALSE) %>% 
  as_tibble()

eyefitting.preds.lmer <- eyefitting.ols.preds.lmer %>%
  full_join(eyefitting.pca.preds.lmer, by = c("x", "parm_id"), suffix = c(".ols", ".pca"))

# write.csv(eyefitting.preds.lmer, "results/youdrawit-eyefitting-lmerpred-data.csv", row.names = F, na = "")

# Plot Predictions
eyefitting.lmer.plot <- eyefitting.preds.lmer %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.ols, color = "OLS")) +
  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.pca, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, ncol = 2) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))

eyefitting.lmer.plot
```

## SMOOTHING SPLINES (Generalized Additive Mixed Model)

**Treatments:**

+ parm_ID (S, F, V, N)
+ x (0, 20)

**Response:** raw residuals

+ residualols = ydrawn - yols, denoted $e_{ols}$
+ residualpca = ydrawn - ypca, denoted $e_{pca}$

**Experimental Design:**

+ Each participant saw each of the 4 plots (new simulated data for each)

**GAMM Model: fit = OLS, PCA**

$$y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})$$

+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $\alpha_i$ is the intercept for the parameter choice $i$
+ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
+ $s_{j}$ is the random smoothing spline for each participant.

```{r gamm, fig.height = 9}
# OLS
eyefitting.ols.gamm <- bam(residualols ~ parm_id + s(x, by = parm_id) + 
                             s(participantID, bs = "re") +
                             s(x,participantID, bs = "re"),
            method = "REML",
            data = eyefitting_model_data)

# evaluate gamm
plot(eyefitting.ols.gamm, pages = 1, all.terms=TRUE)

# PCA
eyefitting.pca.gamm <- bam(residualpca ~ -1 + parm_id + s(x) +
                             s(x, by = parm_id) +
                             s(participantID, bs = "re") +
                             s(x,participantID, bs = "re"),
                           method = "REML",
                           data = eyefitting_model_data)
# evaluate gamm
plot(eyefitting.pca.gamm, pages = 1,all.terms=TRUE)

# Obtain Predictions
eyefitting.grid.gamm <- expand_grid(parm_id = c("S", "V", "F", "N"),
                                    x = seq(0,20, 0.5),
                                    participantID = eyefitting_model_data$participantID[1])

# OLS
eyefitting.ols.preds <- predict_gamm(eyefitting.ols.gamm, newdata = eyefitting.grid.gamm, se = T, re_form = NA)
eyefitting.grid.gamm$ols.pred <- eyefitting.ols.preds$prediction
eyefitting.grid.gamm$ols.lower <- eyefitting.ols.preds$prediction - (1.96 * eyefitting.ols.preds$se)
eyefitting.grid.gamm$ols.upper <- eyefitting.ols.preds$prediction + (1.96 * eyefitting.ols.preds$se)

# PCA
eyefitting.pca.preds <- predict_gamm(eyefitting.pca.gamm, newdata = eyefitting.grid.gamm, se = T, re_form = NA)
eyefitting.grid.gamm$pca.pred <- eyefitting.pca.preds$prediction
eyefitting.grid.gamm$pca.lower <- eyefitting.pca.preds$prediction - (1.96 * eyefitting.pca.preds$se)
eyefitting.grid.gamm$pca.upper <- eyefitting.pca.preds$prediction + (1.96 * eyefitting.pca.preds$se)

# write.csv(eyefitting.grid.gamm, "results/youdrawit-eyefitting-gammpred-data.csv", row.names = F, na = "")

# Plot Predictions
eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
  geom_line(aes(y = ols.pred, color = "OLS")) +
  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
  geom_line(aes(y = pca.pred, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, ncol = 2) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
eyefitting.gamm.plot
```

### 2022-01-26 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
edits to get supplementary material links to work.
proof reading and making edits.

@@
 
-Efforts in the field of statistical graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
+One useful tool for testing statistical graphics is the concept of a lineup.
+@buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
 Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
-One useful tool for testing statistical graphics is the concept of a lineup.
-@buja2009statistical introduced the lineup protocol in which data plots are depicted and interpreted as statistics.
-Supported by the grammar of graphics, a data plot can be characterized as a statistic, defined as, “a functional mapping of a variable or set of variables” [@vanderplas2020testing].
-This allows the data plot to be tested similar to other statistics, by comparing the actual data plot
-to a set of plots with the absence of any data structure we can test the likelihood of any perceived structure being significant.
-The construction of data plots as statistics allow for easy experimentation, granting researchers the ability to compare the effectiveness of and understand the perception of different types of charts [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
 The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
@@
 + **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://shiny.srvanderplas.com/you-draw-it/)
++ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://shiny.srvanderplas.com/you-draw-it/).

### 2022-01-26 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
edits to get supplementary material links to work.
proof reading and making edits.

@@
 
-Efforts in the field of statistical graphics have developed graphical
-testing tools and methods such as the lineup protocol
-\citep{buja2009statistical} to provide a framework for inferential
-testing. Through experimentation, methods such as the lineup protocol
-allow researchers to conduct studies geared at understanding human
-ability to conduct tasks related to the perception of statistical charts
-such as differentiation, prediction, estimation, and extrapolation
-\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
 One useful tool for testing statistical graphics is the concept of a
-lineup. \citet{buja2009statistical} introduced the lineup protocol in
-which data plots are depicted and interpreted as statistics. Supported
-by the grammar of graphics, a data plot can be characterized as a
-statistic, defined as, ``a functional mapping of a variable or set of
-variables'' \citep{vanderplas2020testing}. This allows the data plot to
-be tested similar to other statistics, by comparing the actual data plot
-to a set of plots with the absence of any data structure we can test the
-likelihood of any perceived structure being significant. The
-construction of data plots as statistics allow for easy experimentation,
-granting researchers the ability to compare the effectiveness of and
-understand the perception of different types of charts
+lineup. \citet{buja2009statistical} introduced the lineup protocol to
+provide a framework for inferential testing. Through experimentation,
+methods such as the lineup protocol allow researchers to conduct studies
+geared at understanding human ability to conduct tasks related to the
+perception of statistical charts such as differentiation, prediction,
+estimation, and extrapolation
 \citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
@@
   \textbf{Study Applet:} The shiny app used to conduct the study can be
-  accessed \href{https://shiny.srvanderplas.com/you-draw-it/}{here}
+  accessed \href{https://shiny.srvanderplas.com/you-draw-it/}{here}.
 \end{itemize}

### 2022-01-27 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
proof read of manuscript.

@@
 When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
+For instance, we must decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
 These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
@@
 In @ciccione2021can, participants were asked to judge trends, estimate slopes, and conduct extrapolation.
-To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a trackpad to adjust the tilt of a line on screen.
+To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a track-pad to adjust the tilt of a line on screen.
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatterplot.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot.
 
@@
 The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
-Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than a ordinary least squares regression.
+Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
 In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.
@@
 
-Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' task plots. 
-Integrating this into RShiny using the `r2d3` package, participants are asked to draw a trend-line using their computer mouse through a scatter-plot shown on their screen. 
 In the study, participants are shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
-The yellow box region moves along as the participant draws their trend-line until the yellow region disappears, indicating the participant has filled in the entire domain.
+The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line.
+After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
+Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
+In order to allow for user interaction and data collection, we integrate the D3 visual into RShiny using the `r2d3` package.
 Details of the development of the 'You Draw It' task plots will be addressed in future work.
@@
 Model equation parameters, $\beta_0$ and $\beta_1$, were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
+We obtain $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
 Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
@@
 Aesthetic design choices were made consistent across each of the interactive 'You Draw It' task plots. 
-The y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range.
+The y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range and avoid anchoring their lines to the corners of the plot.
 
@@
 This experiment was conducted as part of a larger study; for simplicity, we focus on the study design and methods related to the current study. 
-Each scatter-plot was the graphical representation of a data set that was generated randomly, independently for each participant at the start of the experiment.
+Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
 Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task. 
@@
 
-We compare the participant drawn line to two regression lines determined by ordinary least squares regression and regression based on the principal axis (i.e.
-Deming Regression).
+We compare the participant drawn line to two regression lines determined by ordinary least squares (OLS) regression and regression based on the principal axis (PCA).
 \cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
@@
 Due to the randomness in the data generation process, the actual slope of the linear regression line fit through the simulated points could differ from the predetermined slope.
-Therefore, we fit an ordinary least squares (OLS) regression to each scatter-plot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
+Therefore, we fit an OLS regression to each scatter-plot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
 Fitted values, $\hat y_{k,OLS}$, are then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
 The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
-The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
+The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
 Fitted values, $\hat y_{k,PCA}$, are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
@@
 
-```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one 'You Draw It' task plot.
+```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one `You Draw It' task plot.
 Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).", out.width="50%"}
@@
 
-+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
-+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $y_{ijk,drawn}$ is the drawn $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value
++ $\hat y_{ijk,fit}$ is the fitted $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted $y$ values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
 + $\gamma_0$ is the overall intercept
@@
 + $\gamma_{2i}$ is the effect of the parameter choice on the slope
-+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $x_{ijk}$ is the $x$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
 + $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $j^{th}$ participant's characteristics
@@
 In particular, this trend is more prominent in parameter choices with large variances (F and N).
-These results are consistent to those found in @mosteller1981eye indicating participants fit a trend-line closer to the estimated regression line with the slope of based on the first principal axis than the estimated OLS regression line thus, providing support for "ensemble perception".
+These results are consistent to those found in @mosteller1981eye indicating participants fit a trend-line closer to the estimated regression line with a slope based on the first principal axis than the estimated OLS regression line thus, providing support for "ensemble perception".
 
@@
 
-+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
-+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
++ $y_{ijk,drawn}$ is the drawn $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value
++ $\hat y_{ijk,fit}$ is the fitted $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted $y$ values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
 + $\alpha_i$ is the intercept for the parameter choice $i$
 + $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
-+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $x_{ijk}$ is the $x$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
 + $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
@@
 By fitting smoothing splines, we can determine whether participants naturally fit a straight trend-line to the set of points or whether they deviate throughout the domain.
-In particular, in scatter-plots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and correcting for their fit toward the end of their trend-line.
+In particular, in scatter-plots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and corrected for their fit toward the end of their trend-line.
 In scatter-plots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PCA regression line.
@@
 
-The intent of research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
+The intent of this research was to adapt 'You Draw It' from the New York Times feature as a tool and method for testing graphics and introduce a method for statistically modeling the participant drawn lines.
 We provided support for the validity of the 'You Draw It' method by replicating the study found in @mosteller1981eye.
@@
 This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
-Further investigation is necessary to implement this method in non-linear settings and with real data.
+Further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication.
 This tool could also be used to evaluate human ability to extrapolate data from trends. 
-In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.
+We intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
 

### 2022-01-27 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
proof read of manuscript.

@@
 result. For instance, we must decide to highlight the relationship
-between two variables in a scatterplot by including a trend line, or
+between two variables in a scatter-plot by including a trend line, or
 adding color to highlight clustering \citep{vanderplas2017clusters}.
@@
 participants were asked to report the slope of the best-fitting
-regression line using a trackpad to adjust the tilt of a line on screen.
-Results indicated the slopes participants reported were always in excess
-of the ideal slopes, both in the positive and in the negative direction,
-and those biases increase with noise and with number of points. This
-supports the results found in \citet{mosteller1981eye} and suggest that
-participants might use Deming regression when fitting a line to a noisy
-scatterplot.
+regression line using a track-pad to adjust the tilt of a line on
+screen. Results indicated the slopes participants reported were always
+in excess of the ideal slopes, both in the positive and in the negative
+direction, and those biases increase with noise and with number of
+points. This supports the results found in \citet{mosteller1981eye} and
+suggest that participants might use Deming regression when fitting a
+line to a noisy scatter-plot.
 
@@
 ``ensemble perception,'' we hypothesize that visual regression tends to
-mimic principle component regression rather than a ordinary least
+mimic principle component regression rather than an ordinary least
 squares regression. In order to assess this hypothesis, we introduce a
@@
 
+In the study, participants are shown an interactive scatter-plot
+(\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
+in the trend in the yellow box region.'' The yellow box region moves
+along as the user draws their trend-line, providing a visual cue which
+indicates where the user still needs to complete a trend line. After the
+entire domain has been visually estimated or predicted, the yellow
+region disappears, indicating the participant has completed the task.
 Data Driven Documents (D3), a JavaScript-based graphing framework that
-facilitates user interaction, is used to create the `You Draw It' task
-plots. Integrating this into RShiny using the \texttt{r2d3} package,
-participants are asked to draw a trend-line using their computer mouse
-through a scatter-plot shown on their screen. In the study, participants
-are shown an interactive scatter-plot (\cref{fig:ydi-stimuli}) along
-with the prompt, ``Use your mouse to fill in the trend in the yellow box
-region.'' The yellow box region moves along as the participant draws
-their trend-line until the yellow region disappears, indicating the
-participant has filled in the entire domain. Details of the development
-of the `You Draw It' task plots will be addressed in future work.
+facilitates user interaction, is used to create the `You Draw It'
+visual. In order to allow for user interaction and data collection, we
+integrate the D3 visual into RShiny using the \texttt{r2d3} package.
+Details of the development of the `You Draw It' task plots will be
+addressed in future work.
 
@@
 reflect the four data sets (F, N, S, and V) used in
-\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). Parameter
-choices F, N, and S simulated data across a domain of 0 to 20. Parameter
-choice F produces a trend with a positive slope and a large variance
-while N has a negative slope and a large variance. In comparison, S
-shows a trend with a positive slope with a small variance and V yields a
-steep positive slope with a small variance over the domain of 4 to 16.
-\cref{fig:eyefitting-simplot} illustrates an example of simulated data
-for all four parameter choices intended to reflect the trends in
-\citet{mosteller1981eye}. Aesthetic design choices were made consistent
-across each of the interactive `You Draw It' task plots. The y-axis
-range extended 10\% beyond (above and below) the range of the simulated
-data points to allow for users to draw outside the simulated data set
-range.
+\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtain
+\(\beta_0\) from the point-slope equation of a line using the mean of
+the generated \(x\) values and the predefined \(y\) value at \(\bar x\),
+denoted \(y_{\bar x}\). Parameter choices F, N, and S simulated data
+across a domain of 0 to 20. Parameter choice F produces a trend with a
+positive slope and a large variance while N has a negative slope and a
+large variance. In comparison, S shows a trend with a positive slope
+with a small variance and V yields a steep positive slope with a small
+variance over the domain of 4 to 16. \cref{fig:eyefitting-simplot}
+illustrates an example of simulated data for all four parameter choices
+intended to reflect the trends in \citet{mosteller1981eye}. Aesthetic
+design choices were made consistent across each of the interactive `You
+Draw It' task plots. The y-axis range extended 10\% beyond (above and
+below) the range of the simulated data points to allow for users to draw
+outside the simulated data set range and avoid anchoring their lines to
+the corners of the plot.
 
@@
 Each scatter-plot was the graphical representation of a data set that
-was generated randomly, independently for each participant at the start
-of the experiment. Participants in the study are shown two `You Draw It'
-practice plots in order to train participants in the skills associated
-with executing the task. During the practice session, participants are
-provided with instruction prompts accompanied by a .gif and a practice
-plot. Instructions guide participants to start at the edge of the yellow
-box, to make sure the yellow region is moving along with their mouse as
-they draw, and that they can draw over their already drawn line.
-Practice plots are then followed by four `You Draw It' task plots
-associated with the current study. The order of the task plots was
+was generated randomly and independently for each participant at the
+start of the experiment. Participants in the study are shown two `You
+Draw It' practice plots in order to train participants in the skills
+associated with executing the task. During the practice session,
+participants are provided with instruction prompts accompanied by a .gif
+and a practice plot. Instructions guide participants to start at the
+edge of the yellow box, to make sure the yellow region is moving along
+with their mouse as they draw, and that they can draw over their already
+drawn line. Practice plots are then followed by four `You Draw It' task
+plots associated with the current study. The order of the task plots was
 randomly assigned for each individual in a completely randomized design.
@@
 We compare the participant drawn line to two regression lines determined
-by ordinary least squares regression and regression based on the
-principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
-illustrates the difference between an OLS regression line which
-minimizes the vertical distance of points from the line and a regression
-line based on the principal axis which minimizes the Euclidean distance
-of points (orthogonal) from the line.
+by ordinary least squares (OLS) regression and regression based on the
+principal axis (PCA). \cref{fig:ols-vs-pca-example} illustrates the
+difference between an OLS regression line which minimizes the vertical
+distance of points from the line and a regression line based on the
+principal axis which minimizes the Euclidean distance of points
+(orthogonal) from the line.
 
@@
 of the linear regression line fit through the simulated points could
-differ from the predetermined slope. Therefore, we fit an ordinary least
-squares (OLS) regression to each scatter-plot to obtain estimated
-parameters \(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted
-values, \(\hat y_{k,OLS}\), are then obtained every 0.25 increments
-across the domain from the OLS regression equation,
+differ from the predetermined slope. Therefore, we fit an OLS regression
+to each scatter-plot to obtain estimated parameters
+\(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted values,
+\(\hat y_{k,OLS}\), are then obtained every 0.25 increments across the
+domain from the OLS regression equation,
 \(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
@@
 most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
-determined by the ratio of the axis rotation in y and axis rotation in x
-of the first principal component with the y-intercept,
+determined by the ratio of the axis rotation in \(y\) and axis rotation
+in \(x\) of the first principal component with the y-intercept,
 \(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
@@
 
-\caption{Illustrates the data associated with and collected for one 'You Draw It' task plot.
+\caption{Illustrates the data associated with and collected for one `You Draw It' task plot.
 Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
@@
 \item
-  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
-  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
+  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
+  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  \(x\) value
 \item
-  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
+  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
   parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  x-value corresponding to either the OLS or PCA fit
+  \(x\) value corresponding to either the OLS or PCA fit
 \item
-  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
-  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
-  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
-  fit
+  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
+  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
+  and \(k^{th}\) increment of \(x\) value corresponding to either the
+  OLS or PCA fit
 \item
@@
 \item
-  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
+  \(x_{ijk}\) is the \(x\) value for the \(i^{th}\) parameter choice,
   \(j^{th}\) participant, and \(k^{th}\) increment
@@
 indicating participants fit a trend-line closer to the estimated
-regression line with the slope of based on the first principal axis than
-the estimated OLS regression line thus, providing support for ``ensemble
+regression line with a slope based on the first principal axis than the
+estimated OLS regression line thus, providing support for ``ensemble
 perception''.
@@
 we can see that participants began at approximately the correct starting
-point then deviated away from the fitted regression lines and correcting
+point then deviated away from the fitted regression lines and corrected
 for their fit toward the end of their trend-line. In scatter-plots with
@@
 
-The intent of research was to adapt `You Draw It' from the New York
+The intent of this research was to adapt `You Draw It' from the New York
 Times feature as a tool and method for testing graphics and introduce a
@@
 models. Further investigation is necessary to implement this method in
-non-linear settings and with real data. This tool could also be used to
-evaluate human ability to extrapolate data from trends. In the future,
-an R package designed for easy implementation of `You Draw It' task
-plots will help make this tool accessible to other researchers.
+non-linear settings and with real data in order to facilitate scientific
+communication. This tool could also be used to evaluate human ability to
+extrapolate data from trends. We intend to create an R package designed
+for easy implementation of `You Draw It' task plots in order to make
+this tool accessible to other researchers.
 

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
Comments

@@
 When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-For instance, we must decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
-These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
+For instance, we \svp{may} decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
+These design choices require that we understand the \svp{perceptual} and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
 
-Much of the research on the perception of visual features in charts has been conducted in psychophysics and test for accuracy and quantitative comparisons when understanding a plot. 
+Much of the research on the perception of visual features in charts has been conducted in psychophysics and tests for accuracy and quantitative comparisons when understanding a plot. 
 Cleveland and McGill conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons [@cleveland1984graphical]. 
 For example, it is more effective to display information on an $x$ or $y$ axis rather than using color in order to reduce the visual effort necessary to make numerical comparisons.
-Other early studies evaluated human perception of lines. 
-For instance, @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
+@cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
+\svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}
 The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
@@
 
-Graphical tests are useful for studying the perception of statistical graphs. 
+\svp{We need human testing of graphics in order to draw broad conclusions, develop guidelines for graphical design, and improve graphical communication.}
 Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
 All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
-Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception]. 
-These studies generally tested participants ability to detect a stimulus or a difference between two stimuli.
-In statistical graphics research, @vanderplas2015signs uses methods of psychophysics to estimate the effect of sign illusions by adjusting the sine illusion up and down.
+Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], \svp{testing} participants' abilities to detect a stimulus or a difference between two stimuli.
+\svp{Psychophysical methods have been used to test graphical perception, as in } @vanderplas2015signs, \svp{which used the method of adjustment} to estimate the magnitude of the impact of the sine illusion.
+\svp{However, there are more modern testing methods that have been developed since the heyday of psychophysics.}
 
-A major development in statistical graphics research is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
+\svp{One} major development in statistical graphics \svp{which led to more advanced testing methods} is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
 The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
@@
 
-One useful tool for testing statistical graphics is the concept of a lineup.
+\svp{Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.} 
 @buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
@@
 The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
+
+\svp{While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".}
 
@@
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations}.
 
-In @finney1951subjective, it was of interest to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
+@finney1951subjective \svp{used graphical testing for computational purposes: to} determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
 Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
@@
 
-@mosteller1981eye, sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
+@mosteller1981eye sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
 Participants were asked to fit lines by eye to four scatter-plots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
@@
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot. \svp{Cite deming regression XXX}
 
-In 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
+\svp{While not explicitly intended for perceptual testing,} in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
 Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
+The New York Times team utilizes Data Driven Documents (D3) \svp{XXX cite the D3 paper} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
+\svp{While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.}
 
-## Research objectives
-
-In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing. 
-The 'You Draw It' method is validated by replicating the study conducted by @mosteller1981eye. 
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, \svp{validating the 'You Draw It' method} by replicating the study conducted by @mosteller1981eye.
 Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
@@
 The experimental tasks took approximately 15 minutes to complete.
-While convenience sampling took place, this is primarily a perceptual task and previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
-Data collected in this study serves the purpose of a pilot study intended to lay the foundation for future data collection and provides promising results for understanding the perception of regression.
+While \svp{this study does utilize a convenience sample, as} this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
+\svp{These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.} 
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through an RShiny application found [here](https://shiny.srvanderplas.com/you-draw-it/).
+The experiment was conducted and distributed through a Shiny application \svp{XXX Cite Shiny} found [here](https://shiny.srvanderplas.com/you-draw-it/).
+\svp{We should probably make a rstudioconnect version that saves data to e.g.
+a google spreadsheet?
+Just in case my server goes down or something?}
 
@@
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
-In order to allow for user interaction and data collection, we integrate the D3 visual into RShiny using the `r2d3` package.
-Details of the development of the 'You Draw It' task plots will be addressed in future work.
+In order to allow for user interaction and data collection, we integrate the D3 visual into Shiny using the `r2d3` package.
+\svp{While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.} 
+<!-- Details of the development of the 'You Draw It' task plots will be addressed in future work. -->
 
@@
 
-This experiment was conducted as part of a larger study; for simplicity, we focus on the study design and methods related to the current study. 
+This experiment was conducted as part of a larger study \svp{of the perception of log and linear scales}; for simplicity, we focus on the study design and methods related to the current study. 
 Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
-Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task. 
+Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - \svp{in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in}.
 During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
@@
 Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
+\svp{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).}
 These results provide support that humans perform "ensemble perception" in a statistical graphic setting.

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
accepted Susan's edits; added citations; indicated comments for future review

@@
 When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-For instance, we \svp{may} decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
-These design choices require that we understand the \svp{perceptual} and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
+For instance, we may decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
+These design choices require that we understand the perceptual and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
 
@@
 
-\svp{We need human testing of graphics in order to draw broad conclusions, develop guidelines for graphical design, and improve graphical communication.}
+We need human testing of graphics in order to draw broad conclusions, develop guidelines for graphical design, and improve graphical communication.
 Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
 All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
-Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], \svp{testing} participants' abilities to detect a stimulus or a difference between two stimuli.
-\svp{Psychophysical methods have been used to test graphical perception, as in } @vanderplas2015signs, \svp{which used the method of adjustment} to estimate the magnitude of the impact of the sine illusion.
-\svp{However, there are more modern testing methods that have been developed since the heyday of psychophysics.}
+Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli.
+Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment to estimate the magnitude of the impact of the sine illusion.
+However, there are more modern testing methods that have been developed since the heyday of psychophysics.
 
-\svp{One} major development in statistical graphics \svp{which led to more advanced testing methods} is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
+One major development in statistical graphics which led to more advanced testing methods is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
 The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
@@
 
-\svp{Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.} 
+Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.
 @buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
@@
 
-\svp{While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".}
+While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".
 
@@
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations}.
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations.} \ear{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}.
 
-@finney1951subjective \svp{used graphical testing for computational purposes: to} determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
+@finney1951subjective used graphical testing for computational purposes: to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
 Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
@@
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot. \svp{Cite deming regression XXX}
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot [@deming1943statistical; @linnet1998performance; @martin2000general] \ear{xxx added citation}.
 
-\svp{While not explicitly intended for perceptual testing,} in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
+While not explicitly intended for perceptual testing, in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
 Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) \svp{XXX cite the D3 paper} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
+The New York Times team utilizes Data Driven Documents (D3) [@bostock2011d3] \ear{xxx added citation} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
-\svp{While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.}
+While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.
 
-In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, \svp{validating the 'You Draw It' method} by replicating the study conducted by @mosteller1981eye.
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, validating the 'You Draw It' method by replicating the study conducted by @mosteller1981eye.
 Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
@@
 The experimental tasks took approximately 15 minutes to complete.
-While \svp{this study does utilize a convenience sample, as} this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
-\svp{These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.} 
+While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
+These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application \svp{XXX Cite Shiny} found [here](https://shiny.srvanderplas.com/you-draw-it/).
+The experiment was conducted and distributed through a Shiny application [@shinyPkg] \ear{xxx added citation} found [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/).
 \svp{We should probably make a rstudioconnect version that saves data to e.g.
 a google spreadsheet?
-Just in case my server goes down or something?}
+Just in case my server goes down or something?} \ear{XXX currently linked to my shinyapps.io page but will get this updated to a new repository.}
 
@@
 In order to allow for user interaction and data collection, we integrate the D3 visual into Shiny using the `r2d3` package.
-\svp{While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.} 
-<!-- Details of the development of the 'You Draw It' task plots will be addressed in future work. -->
+While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.
 
@@
 
-This experiment was conducted as part of a larger study \svp{of the perception of log and linear scales}; for simplicity, we focus on the study design and methods related to the current study. 
+This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focus on the study design and methods related to the current study.
 Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
-Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - \svp{in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in}.
+Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in.
 During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
@@
 + **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://shiny.srvanderplas.com/you-draw-it/).
++ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/).
++ **\ear{RShiny Applet Code:}** \ear{The code used to create the RShiny Applet for data collection can be found} [\ear{here}](https://github.com/earobinson95/you-draw-it-validation-applet).

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
accepted Susan's edits; added citations; indicated comments for future review

@@
 consider the design choices most effective for conveying the intended
-result. For instance, we must decide to highlight the relationship
+result. For instance, we may decide to highlight the relationship
 between two variables in a scatter-plot by including a trend line, or
 adding color to highlight clustering \citep{vanderplas2017clusters}.
-These design choices require that we understand the perception and
+These design choices require that we understand the perceptual and
 visual biases that come into play when creating graphics, and as
@@
 Much of the research on the perception of visual features in charts has
-been conducted in psychophysics and test for accuracy and quantitative
+been conducted in psychophysics and tests for accuracy and quantitative
 comparisons when understanding a plot. Cleveland and McGill conducted a
@@
 \(y\) axis rather than using color in order to reduce the visual effort
-necessary to make numerical comparisons. Other early studies evaluated
-human perception of lines. For instance, \citet{cleveland1985graphical}
+necessary to make numerical comparisons. \citet{cleveland1985graphical}
 found that assessing the position of points along an axis is easier than
-determining the slope of a line. The results of these cognitive tasks
-provide some consistent guidance for chart design; however, other
-methods of visual testing can further evaluate design choices and help
-us understand cognitive biases related to the evaluation of statistical
-charts.
+determining the slope of a line.
+{\textcolor{RedOrange}{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}}
+The results of these cognitive tasks provide some consistent guidance
+for chart design; however, other methods of visual testing can further
+evaluate design choices and help us understand cognitive biases related
+to the evaluation of statistical charts.
 
@@
 
-Graphical tests are useful for studying the perception of statistical
-graphs. Studies might ask participants to identify differences in
+We need human testing of graphics in order to draw broad conclusions,
+develop guidelines for graphical design, and improve graphical
+communication. Studies might ask participants to identify differences in
 graphs, read information off of a chart accurately, use data to make
@@
 studied graphs from a psychological perspective
-\citep{spence1990visual, lewandowsky1989perception}. These studies
-generally tested participants ability to detect a stimulus or a
-difference between two stimuli. In statistical graphics research,
-\citet{vanderplas2015signs} uses methods of psychophysics to estimate
-the effect of sign illusions by adjusting the sine illusion up and down.
+\citep{spence1990visual, lewandowsky1989perception}, testing
+participants' abilities to detect a stimulus or a difference between two
+stimuli. Psychophysical methods have been used to test graphical
+perception, as in \citet{vanderplas2015signs}, which used the method of
+adjustment to estimate the magnitude of the impact of the sine illusion.
+However, there are more modern testing methods that have been developed
+since the heyday of psychophysics.
 
-A major development in statistical graphics research is Wilkinson's
-Grammar of Graphics \citep{wilkinson2013grammar}. The grammar of
-graphics serves as the fundamental framework for data visualization with
-the notion that graphics are built from the ground up by specifying
-exactly how to create a particular graph from a given data set. Visual
-representations are constructed through the use of ``tidy data'' which
-is characterized as a data set in which each variable is in its own
-column, each observation is in its own row, and each value is in its own
-cell \citep{wickham2016r}. Graphics are viewed as a mapping from
-variables in a data set (or statistics computed from the data) to visual
-attributes such as the axes, colors, shapes, or facets on the canvas in
-which the chart is displayed. Software, such as Hadley Wickham's ggplot2
+One major development in statistical graphics which led to more advanced
+testing methods is Wilkinson's Grammar of Graphics
+\citep{wilkinson2013grammar}. The grammar of graphics serves as the
+fundamental framework for data visualization with the notion that
+graphics are built from the ground up by specifying exactly how to
+create a particular graph from a given data set. Visual representations
+are constructed through the use of ``tidy data'' which is characterized
+as a data set in which each variable is in its own column, each
+observation is in its own row, and each value is in its own cell
+\citep{wickham2016r}. Graphics are viewed as a mapping from variables in
+a data set (or statistics computed from the data) to visual attributes
+such as the axes, colors, shapes, or facets on the canvas in which the
+chart is displayed. Software, such as Hadley Wickham's ggplot2
 \citep{wickham2011ggplot2}, aims to implement the framework of creating
@@
 
-One useful tool for testing statistical graphics is the concept of a
-lineup. \citet{buja2009statistical} introduced the lineup protocol to
-provide a framework for inferential testing. Through experimentation,
-methods such as the lineup protocol allow researchers to conduct studies
-geared at understanding human ability to conduct tasks related to the
-perception of statistical charts such as differentiation, prediction,
-estimation, and extrapolation
+Combining the grammar of graphics with another tool for statistical
+graphics testing, the statistical lineup, yields interesting results.
+\citet{buja2009statistical} introduced the lineup protocol to provide a
+framework for inferential testing. Through experimentation, methods such
+as the lineup protocol allow researchers to conduct studies geared at
+understanding human ability to conduct tasks related to the perception
+of statistical charts such as differentiation, prediction, estimation,
+and extrapolation
 \citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
@@
 develop new methods of testing statistical graphics.
+
+While these testing methods are excellent, there is one particular
+subset of statistical graphics testing methods which we intend to
+develop further in this paper: assessing graphics by fitting statistical
+models ``by eye''.
 
@@
 set of studies investigating human ability to detect trends in graphical
-representations from a psychophysical approach.
+representations
+{\textcolor{RedOrange}{using these types of manipulations.}}
+{\textcolor{blue}{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}}.
 
-In \citet{finney1951subjective}, it was of interest to determine the
-effect of stopping iterative maximum likelihood calculations after one
-iteration. Many techniques in statistical analysis are performed with
-the aid of iterative calculations such as Newton's method or Fisher's
-scoring. The author was interested in whether one iteration of
-calculations was sufficient in the estimation of parameters connected
-with dose-response relationships. One measure of interest is the
-relative potency between a test preparation of doses and standard
-preparation of does; relative potency is calculated as the ratio of two
-equally effective doses between the two preparation methods. In this
-study, twenty-one scientists were recruited via postal mail and asked to
-``rule two lines'' in order to judge by eye the positions for a pair of
-parallel probit regression lines in a biological assay. The author then
-computed one iterative calculation of the relative potency based on
-starting values as indicated by the pair of lines provided by each
-participant and compared these relative potency estimates to that which
-was estimated by the full probit technique (reaching convergence through
-multiple iterations). Results indicated that one cycle of iterations for
-calculating the relative potency was sufficient based on the starting
-values provided by eye from the participants.
+\citet{finney1951subjective} used graphical testing for computational
+purposes: to determine the effect of stopping iterative maximum
+likelihood calculations after one iteration. Many techniques in
+statistical analysis are performed with the aid of iterative
+calculations such as Newton's method or Fisher's scoring. The author was
+interested in whether one iteration of calculations was sufficient in
+the estimation of parameters connected with dose-response relationships.
+One measure of interest is the relative potency between a test
+preparation of doses and standard preparation of does; relative potency
+is calculated as the ratio of two equally effective doses between the
+two preparation methods. In this study, twenty-one scientists were
+recruited via postal mail and asked to ``rule two lines'' in order to
+judge by eye the positions for a pair of parallel probit regression
+lines in a biological assay. The author then computed one iterative
+calculation of the relative potency based on starting values as
+indicated by the pair of lines provided by each participant and compared
+these relative potency estimates to that which was estimated by the full
+probit technique (reaching convergence through multiple iterations).
+Results indicated that one cycle of iterations for calculating the
+relative potency was sufficient based on the starting values provided by
+eye from the participants.
 
-\citet{mosteller1981eye}, sought to understand the properties of least
+\citet{mosteller1981eye} sought to understand the properties of least
 squares and other computed lines by establishing one systematic method
@@
 suggest that participants might use Deming regression when fitting a
-line to a noisy scatter-plot.
+line to a noisy scatter-plot
+\citep{deming1943statistical, linnet1998performance, martin2000general}
+{\textcolor{blue}{xxx added citation}}.
 
-In 2015, the New York Times introduced an interactive feature, called
-You Draw It
+While not explicitly intended for perceptual testing, in 2015, the New
+York Times introduced an interactive feature, called You Draw It
 \citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
@@
 and compare how these assumptions relate to reality. The New York Times
-team utilizes Data Driven Documents (D3) that allows readers to predict
+team utilizes Data Driven Documents (D3) \citep{bostock2011d3}
+{\textcolor{blue}{xxx added citation}} that allows readers to predict
 these metrics through the use of drawing a line on their computer screen
@@
 line, the actual observed values are revealed and the reader may check
-their estimated knowledge against the actual reported data.
-
-\hypertarget{research-objectives}{%
-\subsection{Research objectives}\label{research-objectives}}
+their estimated knowledge against the actual reported data. While this
+interactive feature is designed to get readers to confront their own
+intuitions about data in the news, we feel that the interactivity of
+this method may be useful for another purpose.
 
 In this paper, we establish `You Draw It', adapted from the New York
-Times feature, as a tool for graphical testing. The `You Draw It' method
-is validated by replicating the study conducted by
+Times feature, as a tool for graphical testing, validating the `You Draw
+It' method by replicating the study conducted by
 \citet{mosteller1981eye}. Based on previous research surrounding
@@
 vision and signed an informed consent form. The experimental tasks took
-approximately 15 minutes to complete. While convenience sampling took
-place, this is primarily a perceptual task and previous results have
-found few differences between expert and non-expert participants in this
-context \citep{vanderplas2015spatial}. Data collected in this study
-serves the purpose of a pilot study intended to lay the foundation for
-future data collection and provides promising results for understanding
-the perception of regression. Participants completed the experiment on
-their own computers in an environment of their choosing. The experiment
-was conducted and distributed through an RShiny application found
-\href{https://shiny.srvanderplas.com/you-draw-it/}{here}.
+approximately 15 minutes to complete. While this study does utilize a
+convenience sample, as this is primarily a perceptual task, previous
+results have found few differences between expert and non-expert
+participants in this context \citep{vanderplas2015spatial}. These data
+were collected to validate this method of graphical testing, with the
+hopes of providing a new tool to assess graphical perception
+interactively. Participants completed the experiment on their own
+computers in an environment of their choosing. The experiment was
+conducted and distributed through a Shiny application \citep{shinyPkg}
+{\textcolor{blue}{xxx added citation}} found
+\href{https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/}{here}.
+{\textcolor{RedOrange}{We should probably make a rstudioconnect version that saves data to e.g. a google spreadsheet? Just in case my server goes down or something?}}
+{\textcolor{blue}{XXX currently linked to my shinyapps.io page but will get this updated to a new repository.}}
 
@@
 visual. In order to allow for user interaction and data collection, we
-integrate the D3 visual into RShiny using the \texttt{r2d3} package.
-Details of the development of the `You Draw It' task plots will be
-addressed in future work.
+integrate the D3 visual into Shiny using the \texttt{r2d3} package.
+While the interface is highly customized to this particular task, we
+hope to generalize the code and provide a Shiny widget in an R package
+soon.
 
@@
 
-This experiment was conducted as part of a larger study; for simplicity,
-we focus on the study design and methods related to the current study.
-Each scatter-plot was the graphical representation of a data set that
-was generated randomly and independently for each participant at the
-start of the experiment. Participants in the study are shown two `You
-Draw It' practice plots in order to train participants in the skills
-associated with executing the task. During the practice session,
-participants are provided with instruction prompts accompanied by a .gif
-and a practice plot. Instructions guide participants to start at the
-edge of the yellow box, to make sure the yellow region is moving along
-with their mouse as they draw, and that they can draw over their already
-drawn line. Practice plots are then followed by four `You Draw It' task
-plots associated with the current study. The order of the task plots was
-randomly assigned for each individual in a completely randomized design.
+This experiment was conducted as part of a larger study of the
+perception of log and linear scales; for simplicity, we focus on the
+study design and methods related to the current study. Each scatter-plot
+was the graphical representation of a data set that was generated
+randomly and independently for each participant at the start of the
+experiment. Participants in the study are shown two `You Draw It'
+practice plots in order to train participants in the skills associated
+with executing the task - in particular, the responsiveness of the
+applet requires that participants draw a line at a certain speed,
+ensuring that all of the evenly spaced points along the hand-drawn line
+are filled in. During the practice session, participants are provided
+with instruction prompts accompanied by a .gif and a practice plot.
+Instructions guide participants to start at the edge of the yellow box,
+to make sure the yellow region is moving along with their mouse as they
+draw, and that they can draw over their already drawn line. Practice
+plots are then followed by four `You Draw It' task plots associated with
+the current study. The order of the task plots was randomly assigned for
+each individual in a completely randomized design.
 
@@
 the distance from the their regression line over both the \(x\) and
-\(y\) axis simultaneously. These results provide support that humans
-perform ``ensemble perception'' in a statistical graphic setting. We
-allowed participants to draw trend lines that deviated from a straight
-line and gained an insight into the curvature the human eye perceives in
-a set of points.
+\(y\) axis simultaneously.
+{\textcolor{RedOrange}{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).}}
+These results provide support that humans perform ``ensemble
+perception'' in a statistical graphic setting. We allowed participants
+to draw trend lines that deviated from a straight line and gained an
+insight into the curvature the human eye perceives in a set of points.
 
@@
   \textbf{Study Applet:} The shiny app used to conduct the study can be
-  accessed \href{https://shiny.srvanderplas.com/you-draw-it/}{here}.
+  accessed
+  \href{https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/}{here}.
+\item
+  \textbf{{\textcolor{blue}{RShiny Applet Code:}}}
+  {\textcolor{blue}{The code used to create the RShiny Applet for data collection can be found}}
+  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{{\textcolor{blue}{here}}}.
 \end{itemize}

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
updated shinyapps.io link to new repository rsconnect

@@
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application [@shinyPkg] \ear{xxx added citation} found [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/).
+The experiment was conducted and distributed through a Shiny application [@shinyPkg] \ear{xxx added citation} found [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
 \svp{We should probably make a rstudioconnect version that saves data to e.g.
 a google spreadsheet?
-Just in case my server goes down or something?} \ear{XXX currently linked to my shinyapps.io page but will get this updated to a new repository.}
+Just in case my server goes down or something?} \ear{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.}
 
@@
 + **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/).
++ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
 + **\ear{RShiny Applet Code:}** \ear{The code used to create the RShiny Applet for data collection can be found} [\ear{here}](https://github.com/earobinson95/you-draw-it-validation-applet).

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
updated shinyapps.io link to new repository rsconnect

@@
 {\textcolor{blue}{xxx added citation}} found
-\href{https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/}{here}.
+\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
 {\textcolor{RedOrange}{We should probably make a rstudioconnect version that saves data to e.g. a google spreadsheet? Just in case my server goes down or something?}}
-{\textcolor{blue}{XXX currently linked to my shinyapps.io page but will get this updated to a new repository.}}
+{\textcolor{blue}{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.}}
 
@@
   accessed
-  \href{https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/}{here}.
+  \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
 \item

### 2022-02-02 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on edits

@@
 \svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}
+
++ \ear{Viewers often mentally exaggerate the magnitude of correlations in scatterplots when the data appear dense [@cleveland1982variable; @lauer1989density].}
++  \ear{Similarly, the aspect ratio of a graph can have an influence on the patterns that viewers identify.
+@cleveland1993visualizing, for example, argues that viewers can most easily detect cyclical patterns when an aspect ratio that makes the curve closest to 45◦ is used.}
++ \ear{Gestalt principles of perceptual organization play a crucial role in data extraction [@kosslyn2006graph]: for example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].}
+
 The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.

### 2022-02-02 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
working on edits

@@
 {\textcolor{RedOrange}{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}}
+
+\begin{itemize}
+\item
+  {\textcolor{blue}{Viewers often mentally exaggerate the magnitude of correlations in scatterplots when the data appear dense [@cleveland1982variable; @lauer1989density].}}
+\item
+  {\textcolor{blue}{Similarly, the aspect ratio of a graph can have an influence on the patterns that viewers identify.
+@cleveland1993visualizing, for example, argues that viewers can most easily detect cyclical patterns when an aspect ratio that makes the curve closest to 45◦ is used.}}
+\item
+  {\textcolor{blue}{Gestalt principles of perceptual organization play a crucial role in data extraction [@kosslyn2006graph]: for example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].}}
+\end{itemize}
+
 The results of these cognitive tasks provide some consistent guidance

### 2022-02-04 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
just a few more edits to go.

--- 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
@@
 @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
-\svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}
+\ear{The design choices we make can affect a viewers ability to judge relationships between $x$ and $y$ coordinates in a scatter plot.
+For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient} [@cleveland1982variables; @lauer1989density].
+@cleveland1993visualizing.
+\ear{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.}
+<!-- \svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.} -->
+
++ \ear{Gestalt principles of perceptual organization play a crucial role in data extraction [@kosslyn2006graph]: for example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].}
+
 The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.

### 2022-02-04 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
just a few more edits to go.

@@
 determining the slope of a line.
-{\textcolor{RedOrange}{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}}
+{\textcolor{blue}{The design choices we make can affect a viewers ability to judge relationships between $x$ and $y$ coordinates in a scatter plot.
+For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient}}
+\citep{cleveland1982variables, lauer1989density}.
+\citet{cleveland1993visualizing}.
+{\textcolor{blue}{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.}}
 
 \begin{itemize}
-\item
-  {\textcolor{blue}{Viewers often mentally exaggerate the magnitude of correlations in scatterplots when the data appear dense [@cleveland1982variable; @lauer1989density].}}
-\item
-  {\textcolor{blue}{Similarly, the aspect ratio of a graph can have an influence on the patterns that viewers identify.
-@cleveland1993visualizing, for example, argues that viewers can most easily detect cyclical patterns when an aspect ratio that makes the curve closest to 45◦ is used.}}
 \item

### 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
finalized first round edits.

--- 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
@@
 @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
-\svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.}
+\ear{Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
+For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient} [@cleveland1982variables; @lauer1989density].
+@cleveland1993visualizing.
+\ear{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.} @kosslyn2006graph \ear{examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
+For example, data points that are closer in space are more likely to be perceived as grouped} [@ciccione2020grouping]\ear{, and vertical and horizontal lines are easier to discriminate than oblique ones} [@appelle1972perception].
+<!-- \svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.} -->
 The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
@@
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations.} \ear{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}.
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \ear{using manual adjustment and manipulation methods.}
+<!-- \svp{using these types of manipulations.}  -->
+<!-- \ear{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}. -->
 
@@
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot [@deming1943statistical; @linnet1998performance; @martin2000general] \ear{xxx added citation}.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot [@deming1943statistical; @linnet1998performance; @martin2000general].
 
@@
 Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) [@bostock2011d3] \ear{xxx added citation} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
+The New York Times team utilizes Data Driven Documents (D3) [@bostock2011d3] that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
@@
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application [@shinyPkg] \ear{xxx added citation} found [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-\svp{We should probably make a rstudioconnect version that saves data to e.g.
+The experiment was conducted and distributed through a Shiny application [@shinyPkg] found [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+<!-- \svp{We should probably make a rstudioconnect version that saves data to e.g.
 a google spreadsheet?
-Just in case my server goes down or something?} \ear{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.}
+Just in case my server goes down or something?} \ear{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.} -->
 
@@
 Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
-\svp{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).}
+\ear{We utilized modern technology to replicate a study conducted 40 years ago, and strengthend the original results with current analysis methods which allow for more flexibility and sophistication.}
+<!-- \svp{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).} -->
 These results provide support that humans perform "ensemble perception" in a statistical graphic setting.
@@
 
+\newpage
 \begin{center}
@@
 + **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-+ **\ear{RShiny Applet Code:}** \ear{The code used to create the RShiny Applet for data collection can be found} [\ear{here}](https://github.com/earobinson95/you-draw-it-validation-applet).
++ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found [here](https://github.com/earobinson95/you-draw-it-validation-applet).

### 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
finalized first round edits.

@@
 determining the slope of a line.
-{\textcolor{blue}{The design choices we make can affect a viewers ability to judge relationships between $x$ and $y$ coordinates in a scatter plot.
+{\textcolor{blue}{Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
 For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient}}
@@
 {\textcolor{blue}{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.}}
-
-\begin{itemize}
-\item
-  {\textcolor{blue}{Gestalt principles of perceptual organization play a crucial role in data extraction [@kosslyn2006graph]: for example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].}}
-\end{itemize}
-
-The results of these cognitive tasks provide some consistent guidance
-for chart design; however, other methods of visual testing can further
-evaluate design choices and help us understand cognitive biases related
-to the evaluation of statistical charts.
+\citet{kosslyn2006graph}
+{\textcolor{blue}{examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
+For example, data points that are closer in space are more likely to be perceived as grouped}}
+\citep{ciccione2020grouping}{\textcolor{blue}{, and vertical and horizontal lines are easier to discriminate than oblique ones}}
+\citep{appelle1972perception}. The results of these cognitive tasks
+provide some consistent guidance for chart design; however, other
+methods of visual testing can further evaluate design choices and help
+us understand cognitive biases related to the evaluation of statistical
+charts.
 
@@
 representations
-{\textcolor{RedOrange}{using these types of manipulations.}}
-{\textcolor{blue}{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}}.
+{\textcolor{blue}{using manual adjustment and manipulation methods.}}
 
@@
 line to a noisy scatter-plot
-\citep{deming1943statistical, linnet1998performance, martin2000general}
-{\textcolor{blue}{xxx added citation}}.
+\citep{deming1943statistical, linnet1998performance, martin2000general}.
 
@@
 and compare how these assumptions relate to reality. The New York Times
-team utilizes Data Driven Documents (D3) \citep{bostock2011d3}
-{\textcolor{blue}{xxx added citation}} that allows readers to predict
-these metrics through the use of drawing a line on their computer screen
-with their computer mouse. After the reader has completed drawing the
-line, the actual observed values are revealed and the reader may check
-their estimated knowledge against the actual reported data. While this
-interactive feature is designed to get readers to confront their own
-intuitions about data in the news, we feel that the interactivity of
-this method may be useful for another purpose.
+team utilizes Data Driven Documents (D3) \citep{bostock2011d3} that
+allows readers to predict these metrics through the use of drawing a
+line on their computer screen with their computer mouse. After the
+reader has completed drawing the line, the actual observed values are
+revealed and the reader may check their estimated knowledge against the
+actual reported data. While this interactive feature is designed to get
+readers to confront their own intuitions about data in the news, we feel
+that the interactivity of this method may be useful for another purpose.
 
@@
 conducted and distributed through a Shiny application \citep{shinyPkg}
-{\textcolor{blue}{xxx added citation}} found
+found
 \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
-{\textcolor{RedOrange}{We should probably make a rstudioconnect version that saves data to e.g. a google spreadsheet? Just in case my server goes down or something?}}
-{\textcolor{blue}{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.}}
 
@@
 \(y\) axis simultaneously.
-{\textcolor{RedOrange}{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).}}
+{\textcolor{blue}{We utilized modern technology to replicate a study conducted 40 years ago, and strengthend the original results with current analysis methods which allow for more flexibility and sophistication.}}
 These results provide support that humans perform ``ensemble
@@
 \item
-  \textbf{{\textcolor{blue}{RShiny Applet Code:}}}
-  {\textcolor{blue}{The code used to create the RShiny Applet for data collection can be found}}
-  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{{\textcolor{blue}{here}}}.
+  \textbf{RShiny Applet Code:} The code used to create the RShiny Applet
+  for data collection can be found
+  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{here}.
 \end{itemize}

### 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
adjusted minor wording

@@
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \ear{using manual adjustment and manipulation methods.}
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \ear{using physical adjustment and manipulation methods.}
 <!-- \svp{using these types of manipulations.}  -->

### 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
adjusted minor wording

@@
 representations
-{\textcolor{blue}{using manual adjustment and manipulation methods.}}
+{\textcolor{blue}{using physical adjustment and manipulation methods.}}
 

### 2022-02-10 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
edits

--- 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
@@
 
-- name: Susan VanderPlas 2
+- name: Reka Howard 2
   affiliation: Department of Statistics, University of Nebraska - Lincoln
 
-- name: Reka Howard 3
+- name: Susan VanderPlas 3
   affiliation: Department of Statistics, University of Nebraska - Lincoln
@@
 @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
-\ear{Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
-For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient} [@cleveland1982variables; @lauer1989density].
+Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
+For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient [@cleveland1982variables; @lauer1989density].
 @cleveland1993visualizing.
-\ear{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.} @kosslyn2006graph \ear{examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
-For example, data points that are closer in space are more likely to be perceived as grouped} [@ciccione2020grouping]\ear{, and vertical and horizontal lines are easier to discriminate than oblique ones} [@appelle1972perception].
-<!-- \svp{XXX we probably could include a few more papers and discussions here - the Cleveland 1984 and 1985 papers are based on the same study and are basically duplicates.} -->
+provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.
+@kosslyn2006graph examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
+For example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].
 The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
@@
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \ear{using manual adjustment and manipulation methods.}
-<!-- \svp{using these types of manipulations.}  -->
-<!-- \ear{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}. -->
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations using physical adjustment and manipulation methods.
 
@@
 The experiment was conducted and distributed through a Shiny application [@shinyPkg] found [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-<!-- \svp{We should probably make a rstudioconnect version that saves data to e.g.
-a google spreadsheet?
-Just in case my server goes down or something?} \ear{XXX linked to updated shinyapps.io through new repository - does not currently write data anywhere.} -->
 
@@
 Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
-\ear{We utilized modern technology to replicate a study conducted 40 years ago, and strengthend the original results with current analysis methods which allow for more flexibility and sophistication.}
-<!-- \svp{XXX some sentance that says that our results are much stronger than the original results because our analysis methods allow for more sophistication (and we have a slight technological advantage on the original paper's authors as well, after 40 years).} -->
+We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
 These results provide support that humans perform "ensemble perception" in a statistical graphic setting.

### 2022-02-10 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
edits

--- 2022-02-07 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
@@
     Department of Statistics, University of Nebraska - Lincoln\\
-     and \\     Susan VanderPlas 2 \\
+     and \\     Reka Howard 2 \\
     Department of Statistics, University of Nebraska - Lincoln\\
-     and \\     Reka Howard 3 \\
+     and \\     Susan VanderPlas 3 \\
     Department of Statistics, University of Nebraska - Lincoln\\
@@
 found that assessing the position of points along an axis is easier than
-determining the slope of a line.
-{\textcolor{blue}{Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
-For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient}}
-\citep{cleveland1982variables, lauer1989density}.
-\citet{cleveland1993visualizing}.
-{\textcolor{blue}{provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.}}
-\citet{kosslyn2006graph}
-{\textcolor{blue}{examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
-For example, data points that are closer in space are more likely to be perceived as grouped}}
-\citep{ciccione2020grouping}{\textcolor{blue}{, and vertical and horizontal lines are easier to discriminate than oblique ones}}
+determining the slope of a line. Other studies focus on the viewers
+ability to perceive the strength of the relationship between \(x\) and
+\(y\) coordinates in a scatter-plot. For instance, when the data appear
+dense, viewers tend to overestimate the magnitude of the correlation
+coefficient \citep{cleveland1982variables, lauer1989density}.
+\citet{cleveland1993visualizing}. provides an argument for displaying
+cyclical patterns with an aspect ratio which sets the curve closet to
+45\(^{\circ}\). \citet{kosslyn2006graph} examines how Gestalt principles
+of perceptual organization are instrumental in extracting data from a
+chart. For example, data points that are closer in space are more likely
+to be perceived as grouped \citep{ciccione2020grouping}, and vertical
+and horizontal lines are easier to discriminate than oblique ones
 \citep{appelle1972perception}. The results of these cognitive tasks
@@
 set of studies investigating human ability to detect trends in graphical
-representations
-{\textcolor{blue}{using manual adjustment and manipulation methods.}}
+representations using physical adjustment and manipulation methods.
 
@@
 the distance from the their regression line over both the \(x\) and
-\(y\) axis simultaneously.
-{\textcolor{blue}{We utilized modern technology to replicate a study conducted 40 years ago, and strengthend the original results with current analysis methods which allow for more flexibility and sophistication.}}
-These results provide support that humans perform ``ensemble
-perception'' in a statistical graphic setting. We allowed participants
-to draw trend lines that deviated from a straight line and gained an
-insight into the curvature the human eye perceives in a set of points.
+\(y\) axis simultaneously. We utilized modern technology to replicate a
+study conducted 40 years ago, and strengthened the original results with
+current analysis methods which allow for more flexibility and
+sophistication. These results provide support that humans perform
+``ensemble perception'' in a statistical graphic setting. We allowed
+participants to draw trend lines that deviated from a straight line and
+gained an insight into the curvature the human eye perceives in a set of
+points.
 

### 2022-02-13 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
Edits from Reka

@@
 - Statistical Graphics
+- Visualization
 
@@
 Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It', where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
+In 2015, the New York Times introduced an interactive feature, called 'You Draw It', where readers were asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
 In this paper, we validate ‘You Draw It’ as a method for graphical testing, comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.
@@
 Much of the research on the perception of visual features in charts has been conducted in psychophysics and tests for accuracy and quantitative comparisons when understanding a plot. 
-Cleveland and McGill conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons [@cleveland1984graphical]. 
+@cleveland1984graphical conducted a series of cognitive tasks designed to establish a hierarchy of visual components for making comparisons. 
 For example, it is more effective to display information on an $x$ or $y$ axis rather than using color in order to reduce the visual effort necessary to make numerical comparisons.
@@
 For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient [@cleveland1982variables; @lauer1989density].
-@cleveland1993visualizing.
-provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve closet to 45$^{\circ}$.
+@cleveland1993visualizing provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve close to 45$^{\circ}$.
 @kosslyn2006graph examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
@@
 All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
-Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli.
+Early research studies considered graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli.
 Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment to estimate the magnitude of the impact of the sine illusion.
@@
 
-Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.
+Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields a method for evaluating graphical design choices.
 @buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
@@
 Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
-The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with dose-response relationships. 
-One measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods.
+The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with pharmaceutical dose-response relationships. 
+In pharmaceuticals, one measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods.
 In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay.
-The author then computed one iterative calculation of the relative potency based on starting values as indicated by the pair of lines provided by each participant and compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
-Results indicated that one cycle of iterations for calculating the relative potency was sufficient based on the starting values provided by eye from the participants.
+The author then computed one iterative calculation of the relative potency based on starting values as determined by the pair of lines drawn by each participant.
+The author then compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
+Results of the study indicated that one cycle of iterations for calculating the relative potency was sufficient based on the starting values provided by eye from the participants.
 
@@
 In @ciccione2021can, participants were asked to judge trends, estimate slopes, and conduct extrapolation.
-To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a track-pad to adjust the tilt of a line on screen.
+To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a track-pad to adjust the tilt of a line on the screen.
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot [@deming1943statistical; @linnet1998performance; @martin2000general].
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression [@deming1943statistical; @linnet1998performance; @martin2000general], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.
 
 While not explicitly intended for perceptual testing, in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
-Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
+Readers were asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
 The New York Times team utilizes Data Driven Documents (D3) [@bostock2011d3] that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse.
 After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
-While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.
+While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for the purpose of graphical testing and measuring the patterns humans see in data.
 
-In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, validating the 'You Draw It' method by replicating the study conducted by @mosteller1981eye.
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing.
+Our visual system is naturally built to look for structure and identify patterns.
+For instance, points going down from left to right indicates a negative correlation between the x and y variables.
+Our research is intended to implement the 'You Draw It' feature as a way to measure the patterns we see in data.
+We validate the 'You Draw It' method by replicating the less technological study conducted by @mosteller1981eye.
 Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
@@
 Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
-A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
+A total of 35 individuals completed 119 unique 'You Draw It' task plots; data collected from all completed 'You Draw It' task plots were included in the analysis.
 All participants had normal or corrected to normal vision and signed an informed consent form.
@@
 
-In the study, participants are shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
+In the study, participants were shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
 The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line.
-After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
+After the entire domain has been visually estimated or predicted, the yellow shaded region disappears, indicating the participant has completed the task.
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
-In order to allow for user interaction and data collection, we integrate the D3 visual into Shiny using the `r2d3` package.
+In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` package.
 While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.
@@
 ```{r ydi-stimuli, include = T, fig.cap = "'You Draw It' task plot as shown to particpants during the study.
-The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow shaded region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
 The last frame (right) illustrates the participants finished trend-line before submission.", out.width="100%"}
@@
 
-This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focus on the study design and methods related to the current study.
+This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focused on the study design and methods related to the current study.
 Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
-Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in.
-During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
-Instructions guide participants to start at the edge of the yellow box, to make sure the yellow region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
-Practice plots are then followed by four 'You Draw It' task plots associated with the current study.
+Participants in the study were shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in.
+During the practice session, participants were provided with instruction prompts accompanied by a .gif and a practice plot.
+Instructions guided participants to start at the edge of the yellow box, to make sure the yellow shaded region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
+Practice plots were then followed by four 'You Draw It' task plots associated with the current study.
 The order of the task plots was randomly assigned for each individual in a completely randomized design.
@@
 Therefore, we fit an OLS regression to each scatter-plot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
-Fitted values, $\hat y_{k,OLS}$, are then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
+Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
 The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
-The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
-Fitted values, $\hat y_{k,PCA}$, are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
+The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of the simulated points, $(\bar x_i, \bar y_i)$.
+Fitted values, $\hat y_{k,PCA}$, were then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
 
@@
 
-For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
+For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
 Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
@@
 Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS residuals and PCA residuals, constraining the fit to a linear trend. 
-Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
+Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
 The LMM equation for each fit (OLS and PCA) is given by:
@@
 Allowing for flexibility in the residual trend, \cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PCA regression line.
-Estimated residual trends are overlaid on the observed individual participant residuals. 
+Estimated residual trends were overlaid on the observed individual participant residuals. 
 The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that for scatter-plots with more noise (F and N), estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue).
@@
 Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principle component regression.
-Our results indicate that participants minimized the distance from the their regression line over both the $x$ and $y$ axis simultaneously. 
+Our results indicate that participants minimized the distance from their drawn regression line over both the $x$ and $y$ axis simultaneously. 
 We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
@@
 This tool could also be used to evaluate human ability to extrapolate data from trends. 
-We intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
+In the future, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
 

### 2022-02-13 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
Edits from Reka

@@
 New York Times introduced an interactive feature, called `You Draw It',
-where readers are asked to input their own assumptions about various
+where readers were asked to input their own assumptions about various
 metrics and compare how these assumptions relate to reality. In this
@@
 {\it Keywords:} Cognitive Bias, Graph Perception, Graphical
-Testing, Linear Regression, Statistical Graphics
+Testing, Linear Regression, Statistical Graphics, Visualization
 \vfill
@@
 been conducted in psychophysics and tests for accuracy and quantitative
-comparisons when understanding a plot. Cleveland and McGill conducted a
-series of cognitive tasks designed to establish a hierarchy of visual
-components for making comparisons \citep{cleveland1984graphical}. For
-example, it is more effective to display information on an \(x\) or
-\(y\) axis rather than using color in order to reduce the visual effort
-necessary to make numerical comparisons. \citet{cleveland1985graphical}
-found that assessing the position of points along an axis is easier than
+comparisons when understanding a plot. \citet{cleveland1984graphical}
+conducted a series of cognitive tasks designed to establish a hierarchy
+of visual components for making comparisons. For example, it is more
+effective to display information on an \(x\) or \(y\) axis rather than
+using color in order to reduce the visual effort necessary to make
+numerical comparisons. \citet{cleveland1985graphical} found that
+assessing the position of points along an axis is easier than
 determining the slope of a line. Other studies focus on the viewers
@@
 coefficient \citep{cleveland1982variables, lauer1989density}.
-\citet{cleveland1993visualizing}. provides an argument for displaying
-cyclical patterns with an aspect ratio which sets the curve closet to
+\citet{cleveland1993visualizing} provides an argument for displaying
+cyclical patterns with an aspect ratio which sets the curve close to
 45\(^{\circ}\). \citet{kosslyn2006graph} examines how Gestalt principles
@@
 of these types of tests require different levels of use and manipulation
-of the information being presented in the chart. Early researchers
-studied graphs from a psychological perspective
+of the information being presented in the chart. Early research studies
+considered graphs from a psychological perspective
 \citep{spence1990visual, lewandowsky1989perception}, testing
@@
 Combining the grammar of graphics with another tool for statistical
-graphics testing, the statistical lineup, yields interesting results.
-\citet{buja2009statistical} introduced the lineup protocol to provide a
-framework for inferential testing. Through experimentation, methods such
-as the lineup protocol allow researchers to conduct studies geared at
-understanding human ability to conduct tasks related to the perception
-of statistical charts such as differentiation, prediction, estimation,
-and extrapolation
+graphics testing, the statistical lineup, yields a method for evaluating
+graphical design choices. \citet{buja2009statistical} introduced the
+lineup protocol to provide a framework for inferential testing. Through
+experimentation, methods such as the lineup protocol allow researchers
+to conduct studies geared at understanding human ability to conduct
+tasks related to the perception of statistical charts such as
+differentiation, prediction, estimation, and extrapolation
 \citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
@@
 interested in whether one iteration of calculations was sufficient in
-the estimation of parameters connected with dose-response relationships.
-One measure of interest is the relative potency between a test
-preparation of doses and standard preparation of does; relative potency
-is calculated as the ratio of two equally effective doses between the
-two preparation methods. In this study, twenty-one scientists were
-recruited via postal mail and asked to ``rule two lines'' in order to
-judge by eye the positions for a pair of parallel probit regression
-lines in a biological assay. The author then computed one iterative
-calculation of the relative potency based on starting values as
-indicated by the pair of lines provided by each participant and compared
-these relative potency estimates to that which was estimated by the full
-probit technique (reaching convergence through multiple iterations).
-Results indicated that one cycle of iterations for calculating the
-relative potency was sufficient based on the starting values provided by
-eye from the participants.
+the estimation of parameters connected with pharmaceutical dose-response
+relationships. In pharmaceuticals, one measure of interest is the
+relative potency between a test preparation of doses and standard
+preparation of does; relative potency is calculated as the ratio of two
+equally effective doses between the two preparation methods. In this
+study, twenty-one scientists were recruited via postal mail and asked to
+``rule two lines'' in order to judge by eye the positions for a pair of
+parallel probit regression lines in a biological assay. The author then
+computed one iterative calculation of the relative potency based on
+starting values as determined by the pair of lines drawn by each
+participant. The author then compared these relative potency estimates
+to that which was estimated by the full probit technique (reaching
+convergence through multiple iterations). Results of the study indicated
+that one cycle of iterations for calculating the relative potency was
+sufficient based on the starting values provided by eye from the
+participants.
 
@@
 participants were asked to report the slope of the best-fitting
-regression line using a track-pad to adjust the tilt of a line on
+regression line using a track-pad to adjust the tilt of a line on the
 screen. Results indicated the slopes participants reported were always
@@
 points. This supports the results found in \citet{mosteller1981eye} and
-suggest that participants might use Deming regression when fitting a
-line to a noisy scatter-plot
-\citep{deming1943statistical, linnet1998performance, martin2000general}.
+suggest that participants might use Deming regression
+\citep{deming1943statistical, linnet1998performance, martin2000general},
+which minimizes the Euclidean distance of points from the line, when
+fitting a line to a noisy scatter-plot.
 
@@
 \citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
-Readers are asked to input their own assumptions about various metrics
+Readers were asked to input their own assumptions about various metrics
 and compare how these assumptions relate to reality. The New York Times
@@
 readers to confront their own intuitions about data in the news, we feel
-that the interactivity of this method may be useful for another purpose.
+that the interactivity of this method may be useful for the purpose of
+graphical testing and measuring the patterns humans see in data.
 
 In this paper, we establish `You Draw It', adapted from the New York
-Times feature, as a tool for graphical testing, validating the `You Draw
-It' method by replicating the study conducted by
-\citet{mosteller1981eye}. Based on previous research surrounding
-``ensemble perception,'' we hypothesize that visual regression tends to
-mimic principle component regression rather than an ordinary least
-squares regression. In order to assess this hypothesis, we introduce a
-method for statistically modeling the participant drawn lines using
-generalized additive mixed models.
+Times feature, as a tool for graphical testing. Our visual system is
+naturally built to look for structure and identify patterns. For
+instance, points going down from left to right indicates a negative
+correlation between the x and y variables. Our research is intended to
+implement the `You Draw It' feature as a way to measure the patterns we
+see in data. We validate the `You Draw It' method by replicating the
+less technological study conducted by \citet{mosteller1981eye}. Based on
+previous research surrounding ``ensemble perception,'' we hypothesize
+that visual regression tends to mimic principle component regression
+rather than an ordinary least squares regression. In order to assess
+this hypothesis, we introduce a method for statistically modeling the
+participant drawn lines using generalized additive mixed models.
 
@@
 email in May 2021. A total of 35 individuals completed 119 unique `You
-Draw It' task plots; all completed you draw it task plots were included
-in the analysis. All participants had normal or corrected to normal
-vision and signed an informed consent form. The experimental tasks took
-approximately 15 minutes to complete. While this study does utilize a
-convenience sample, as this is primarily a perceptual task, previous
-results have found few differences between expert and non-expert
-participants in this context \citep{vanderplas2015spatial}. These data
-were collected to validate this method of graphical testing, with the
-hopes of providing a new tool to assess graphical perception
-interactively. Participants completed the experiment on their own
-computers in an environment of their choosing. The experiment was
-conducted and distributed through a Shiny application \citep{shinyPkg}
-found
+Draw It' task plots; data collected from all completed `You Draw It'
+task plots were included in the analysis. All participants had normal or
+corrected to normal vision and signed an informed consent form. The
+experimental tasks took approximately 15 minutes to complete. While this
+study does utilize a convenience sample, as this is primarily a
+perceptual task, previous results have found few differences between
+expert and non-expert participants in this context
+\citep{vanderplas2015spatial}. These data were collected to validate
+this method of graphical testing, with the hopes of providing a new tool
+to assess graphical perception interactively. Participants completed the
+experiment on their own computers in an environment of their choosing.
+The experiment was conducted and distributed through a Shiny application
+\citep{shinyPkg} found
 \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
@@
 
-In the study, participants are shown an interactive scatter-plot
+In the study, participants were shown an interactive scatter-plot
 (\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
@@
 entire domain has been visually estimated or predicted, the yellow
-region disappears, indicating the participant has completed the task.
-Data Driven Documents (D3), a JavaScript-based graphing framework that
-facilitates user interaction, is used to create the `You Draw It'
+shaded region disappears, indicating the participant has completed the
+task. Data Driven Documents (D3), a JavaScript-based graphing framework
+that facilitates user interaction, is used to create the `You Draw It'
 visual. In order to allow for user interaction and data collection, we
-integrate the D3 visual into Shiny using the \texttt{r2d3} package.
+integrated the D3 visual into Shiny using the \texttt{r2d3} package.
 While the interface is highly customized to this particular task, we
@@
 \caption{'You Draw It' task plot as shown to particpants during the study.
-The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow shaded region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
 The last frame (right) illustrates the participants finished trend-line before submission.}\label{fig:ydi-stimuli}
@@
 This experiment was conducted as part of a larger study of the
-perception of log and linear scales; for simplicity, we focus on the
+perception of log and linear scales; for simplicity, we focused on the
 study design and methods related to the current study. Each scatter-plot
@@
 randomly and independently for each participant at the start of the
-experiment. Participants in the study are shown two `You Draw It'
+experiment. Participants in the study were shown two `You Draw It'
 practice plots in order to train participants in the skills associated
@@
 ensuring that all of the evenly spaced points along the hand-drawn line
-are filled in. During the practice session, participants are provided
+are filled in. During the practice session, participants were provided
 with instruction prompts accompanied by a .gif and a practice plot.
-Instructions guide participants to start at the edge of the yellow box,
-to make sure the yellow region is moving along with their mouse as they
-draw, and that they can draw over their already drawn line. Practice
-plots are then followed by four `You Draw It' task plots associated with
-the current study. The order of the task plots was randomly assigned for
-each individual in a completely randomized design.
+Instructions guided participants to start at the edge of the yellow box,
+to make sure the yellow shaded region is moving along with their mouse
+as they draw, and that they can draw over their already drawn line.
+Practice plots were then followed by four `You Draw It' task plots
+associated with the current study. The order of the task plots was
+randomly assigned for each individual in a completely randomized design.
 
@@
 \(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted values,
-\(\hat y_{k,OLS}\), are then obtained every 0.25 increments across the
+\(\hat y_{k,OLS}\), were then obtained every 0.25 increments across the
 domain from the OLS regression equation,
@@
 \(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
-using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
-Fitted values, \(\hat y_{k,PCA}\), are then obtained every 0.25
-increment across the domain from the PCA regression equation,
+using the mean of the simulated points, \((\bar x_i, \bar y_i)\). Fitted
+values, \(\hat y_{k,PCA}\), were then obtained every 0.25 increment
+across the domain from the PCA regression equation,
 \(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
@@
 
-For each participant, the final data set used for analysis contains
+For each participant, the final data set used for analysis contained
 \(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
 for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
-\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
-mixed model and a generalized additive mixed model, comparisons of
-vertical residuals in relation to the OLS fitted values
+\(x_{ijk}\) value for increment \(k = 1, ...,4 x_{max} + 1\). Using both
+a linear mixed model and a generalized additive mixed model, comparisons
+of vertical residuals in relation to the OLS fitted values
 (\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
@@
 residuals, constraining the fit to a linear trend. Parameter choice,
-\(x\), and the interaction between \(x\) and parameter choice were
+\(x\), and the interaction between \(x\) and the parameter choice were
 treated as fixed effects with a random participant effect accounting for
@@
 for both the OLS regression line and PCA regression line. Estimated
-residual trends are overlaid on the observed individual participant
+residual trends were overlaid on the observed individual participant
 residuals. The results of the GAMM align with those shown in
@@
 component regression. Our results indicate that participants minimized
-the distance from the their regression line over both the \(x\) and
+the distance from their drawn regression line over both the \(x\) and
 \(y\) axis simultaneously. We utilized modern technology to replicate a
@@
 communication. This tool could also be used to evaluate human ability to
-extrapolate data from trends. We intend to create an R package designed
-for easy implementation of `You Draw It' task plots in order to make
-this tool accessible to other researchers.
+extrapolate data from trends. In the future, we intend to create an R
+package designed for easy implementation of `You Draw It' task plots in
+order to make this tool accessible to other researchers.
 

### 2022-02-14 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
final read through before submission!

@@
 The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
-
 While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".
@@
 Our visual system is naturally built to look for structure and identify patterns.
-For instance, points going down from left to right indicates a negative correlation between the x and y variables.
+For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables.
 Our research is intended to implement the 'You Draw It' feature as a way to measure the patterns we see in data.
@@
 In the study, participants were shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
-The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line.
-After the entire domain has been visually estimated or predicted, the yellow shaded region disappears, indicating the participant has completed the task.
-Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
+The yellow box region moved along as the user drew their trend-line, providing a visual cue which indicates where the user still needed to complete a trend line.
+After the entire domain had been visually estimated or predicted, the yellow shaded region disappeared, indicating the participant had completed the task.
+Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual. 
 In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` package.
@@
 ```{r ydi-stimuli, include = T, fig.cap = "'You Draw It' task plot as shown to particpants during the study.
-The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow shaded region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The first frame (left) illustrates what particpants first saw with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant saw while completing the task; the yellow shaded region provided a visual cue for participants indicating where the participant still needed to complete a trend-line.
 The last frame (right) illustrates the participants finished trend-line before submission.", out.width="100%"}
@@
 Model equation parameters, $\beta_0$ and $\beta_1$, were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
-We obtain $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
+We obtained $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
 Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
-Parameter choice F produces a trend with a positive slope and a large variance while N has a negative slope and a large variance. 
-In comparison, S shows a trend with a positive slope with a small variance and V yields a steep positive slope with a small variance over the domain of 4 to 16. 
+Parameter choice F produced a trend with a positive slope and a large variance while N had a negative slope and a large variance. 
+In comparison, S showed a trend with a positive slope and a small variance while V yielded a steep positive slope with a small variance over the domain of 4 to 16. 
 \cref{fig:eyefitting-simplot} illustrates an example of simulated data for all four parameter choices intended to reflect the trends in @mosteller1981eye.
@@
 During the practice session, participants were provided with instruction prompts accompanied by a .gif and a practice plot.
-Instructions guided participants to start at the edge of the yellow box, to make sure the yellow shaded region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
+Instructions guided participants to start at the edge of the yellow box, to make sure the yellow shaded region was moving along with their mouse as they drew, and that they could draw over their already drawn line.
 Practice plots were then followed by four 'You Draw It' task plots associated with the current study.
@@
 
-We compare the participant drawn line to two regression lines determined by ordinary least squares (OLS) regression and regression based on the principal axis (PCA).
+We compared the participant drawn line to two regression lines determined by ordinary least squares (OLS) regression and regression based on the principal axis (PCA).
 \cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
@@
 The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
-The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of the simulated points, $(\bar x_i, \bar y_i)$.
+The estimated slope, $\hat\beta_{1,PCA}$, was determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of the simulated points, $(\bar x_i, \bar y_i)$.
 Fitted values, $\hat y_{k,PCA}$, were then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
@@
 
-Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS residuals and PCA residuals, constraining the fit to a linear trend. 
-Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
+Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PCA residuals, constraining the fit to a linear trend. 
+Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect included to account for variation due to participant.
 The LMM equation for each fit (OLS and PCA) is given by:
@@
 
-Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PCA residuals to allow for estimation of smoothing splines.
+Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PCA residuals to allow for estimation of smoothing splines.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
-A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
+A random participant effect was included to account for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
 The GAMM equation for each fit (OLS and PCA) residuals is given by:
@@
 Estimated residual trends were overlaid on the observed individual participant residuals. 
-The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that for scatter-plots with more noise (F and N), estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue).
+The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue) for scatter-plots with more noise (F and N).
 By fitting smoothing splines, we can determine whether participants naturally fit a straight trend-line to the set of points or whether they deviate throughout the domain.
@@
 Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principle component regression.
+We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
 Our results indicate that participants minimized the distance from their drawn regression line over both the $x$ and $y$ axis simultaneously. 
-We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
 These results provide support that humans perform "ensemble perception" in a statistical graphic setting.

### 2022-02-14 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
final read through before submission!

@@
 The advancement of graphing software provides the tools necessary to
-develop new methods of testing statistical graphics.
-
-While these testing methods are excellent, there is one particular
-subset of statistical graphics testing methods which we intend to
-develop further in this paper: assessing graphics by fitting statistical
-models ``by eye''.
+develop new methods of testing statistical graphics. While these testing
+methods are excellent, there is one particular subset of statistical
+graphics testing methods which we intend to develop further in this
+paper: assessing graphics by fitting statistical models ``by eye''.
 
@@
 instance, points going down from left to right indicates a negative
-correlation between the x and y variables. Our research is intended to
-implement the `You Draw It' feature as a way to measure the patterns we
-see in data. We validate the `You Draw It' method by replicating the
-less technological study conducted by \citet{mosteller1981eye}. Based on
-previous research surrounding ``ensemble perception,'' we hypothesize
-that visual regression tends to mimic principle component regression
-rather than an ordinary least squares regression. In order to assess
-this hypothesis, we introduce a method for statistically modeling the
-participant drawn lines using generalized additive mixed models.
+correlation between the \(x\) and \(y\) variables. Our research is
+intended to implement the `You Draw It' feature as a way to measure the
+patterns we see in data. We validate the `You Draw It' method by
+replicating the less technological study conducted by
+\citet{mosteller1981eye}. Based on previous research surrounding
+``ensemble perception,'' we hypothesize that visual regression tends to
+mimic principle component regression rather than an ordinary least
+squares regression. In order to assess this hypothesis, we introduce a
+method for statistically modeling the participant drawn lines using
+generalized additive mixed models.
 
@@
 (\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
-in the trend in the yellow box region.'' The yellow box region moves
-along as the user draws their trend-line, providing a visual cue which
-indicates where the user still needs to complete a trend line. After the
-entire domain has been visually estimated or predicted, the yellow
-shaded region disappears, indicating the participant has completed the
+in the trend in the yellow box region.'' The yellow box region moved
+along as the user drew their trend-line, providing a visual cue which
+indicates where the user still needed to complete a trend line. After
+the entire domain had been visually estimated or predicted, the yellow
+shaded region disappeared, indicating the participant had completed the
 task. Data Driven Documents (D3), a JavaScript-based graphing framework
-that facilitates user interaction, is used to create the `You Draw It'
+that facilitates user interaction, was used to create the `You Draw It'
 visual. In order to allow for user interaction and data collection, we
@@
 \caption{'You Draw It' task plot as shown to particpants during the study.
-The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow shaded region provides a visual cue for participants indicating where the participant still needs to complete a trend-line.
+The first frame (left) illustrates what particpants first saw with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant saw while completing the task; the yellow shaded region provided a visual cue for participants indicating where the participant still needed to complete a trend-line.
 The last frame (right) illustrates the participants finished trend-line before submission.}\label{fig:ydi-stimuli}
@@
 reflect the four data sets (F, N, S, and V) used in
-\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtain
+\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtained
 \(\beta_0\) from the point-slope equation of a line using the mean of
@@
 denoted \(y_{\bar x}\). Parameter choices F, N, and S simulated data
-across a domain of 0 to 20. Parameter choice F produces a trend with a
-positive slope and a large variance while N has a negative slope and a
-large variance. In comparison, S shows a trend with a positive slope
-with a small variance and V yields a steep positive slope with a small
+across a domain of 0 to 20. Parameter choice F produced a trend with a
+positive slope and a large variance while N had a negative slope and a
+large variance. In comparison, S showed a trend with a positive slope
+and a small variance while V yielded a steep positive slope with a small
 variance over the domain of 4 to 16. \cref{fig:eyefitting-simplot}
@@
 Instructions guided participants to start at the edge of the yellow box,
-to make sure the yellow shaded region is moving along with their mouse
-as they draw, and that they can draw over their already drawn line.
+to make sure the yellow shaded region was moving along with their mouse
+as they drew, and that they could draw over their already drawn line.
 Practice plots were then followed by four `You Draw It' task plots
@@
 
-We compare the participant drawn line to two regression lines determined
-by ordinary least squares (OLS) regression and regression based on the
-principal axis (PCA). \cref{fig:ols-vs-pca-example} illustrates the
-difference between an OLS regression line which minimizes the vertical
-distance of points from the line and a regression line based on the
-principal axis which minimizes the Euclidean distance of points
-(orthogonal) from the line.
+We compared the participant drawn line to two regression lines
+determined by ordinary least squares (OLS) regression and regression
+based on the principal axis (PCA). \cref{fig:ols-vs-pca-example}
+illustrates the difference between an OLS regression line which
+minimizes the vertical distance of points from the line and a regression
+line based on the principal axis which minimizes the Euclidean distance
+of points (orthogonal) from the line.
 
@@
 axes from the first principal component (direction which captures the
-most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
+most variance). The estimated slope, \(\hat\beta_{1,PCA}\), was
 determined by the ratio of the axis rotation in \(y\) and axis rotation
@@
 Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
-linear mixed model (LMM) is fit separately to the OLS residuals and PCA
+linear mixed model (LMM) was fit separately to the OLS residuals and PCA
 residuals, constraining the fit to a linear trend. Parameter choice,
 \(x\), and the interaction between \(x\) and the parameter choice were
-treated as fixed effects with a random participant effect accounting for
-variation due to participant. The LMM equation for each fit (OLS and
-PCA) is given by: \begin{equation}
+treated as fixed effects with a random participant effect included to
+account for variation due to participant. The LMM equation for each fit
+(OLS and PCA) is given by: \begin{equation}
 y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
@@
 Eliminating the linear trend constraint, the \texttt{bam} function in
-the mgcv package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to
+the mgcv package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} was used to
 fit a generalized additive mixed model (GAMM) separately to the OLS
@@
 estimated intercept and a separate smoothing spline for \(x\) was
-estimated for each parameter choice. A random participant effect
-accounting for variation due to participant and a random spline for each
-participant accounted for variation in spline for each participant. The
-GAMM equation for each fit (OLS and PCA) residuals is given by:
-\begin{equation}
+estimated for each parameter choice. A random participant effect was
+included to account for variation due to participant and a random spline
+for each participant accounted for variation in spline for each
+participant. The GAMM equation for each fit (OLS and PCA) residuals is
+given by: \begin{equation}
 y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
@@
 residuals. The results of the GAMM align with those shown in
-\cref{fig:eyefitting-lmer-residualplots} providing support that for
-scatter-plots with more noise (F and N), estimated trends of PCA
-residuals (orange) appear to align closer to the \(y=0\) horizontal
-(dashed) line than the OLS residuals (blue). By fitting smoothing
-splines, we can determine whether participants naturally fit a straight
+\cref{fig:eyefitting-lmer-residualplots} providing support that
+estimated trends of PCA residuals (orange) appear to align closer to the
+\(y=0\) horizontal (dashed) line than the OLS residuals (blue) for
+scatter-plots with more noise (F and N). By fitting smoothing splines,
+we can determine whether participants naturally fit a straight
 trend-line to the set of points or whether they deviate throughout the
@@
 support for a participants tendency to visually fit the first principle
-component regression. Our results indicate that participants minimized
-the distance from their drawn regression line over both the \(x\) and
-\(y\) axis simultaneously. We utilized modern technology to replicate a
-study conducted 40 years ago, and strengthened the original results with
+component regression. We utilized modern technology to replicate a study
+conducted 40 years ago, and strengthened the original results with
 current analysis methods which allow for more flexibility and
-sophistication. These results provide support that humans perform
+sophistication. Our results indicate that participants minimized the
+distance from their drawn regression line over both the \(x\) and \(y\)
+axis simultaneously. These results provide support that humans perform
 ``ensemble perception'' in a statistical graphic setting. We allowed

### 2022-07-19 manuscript-revisions/response-to-reviewers.qmd
working on revisions

---
title: "Response to reviewers of JCGS-22-045"
subtitle: "Eye Fitting Straight Lines in the Modern Era"
author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
date: "2022"
# output: pdf_document
format: 
  pdf:
    include-in-header:
      - revision-edits.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reviewer(s)' Comments to Author

All responses to reviewers' comments are written in \response{blue}.

### Reviewer: 1

Summary:
The authors present a study where human participants had to fit straight lines through points in a scatterplot. On average,
the participants' fitted lines closely matched a principal axis line and not the least squares regression line. Overall, the article is well written.

I have a series of minor comments, though - see below. Specific minor comments:

+ p.3, l.11 (& more): your spelling of "scatter-plot" with the hyphen is unconventional; simply use without hyphen (that will also match the 3 occurrences in your references)
  + \response{Changed all spellings of "scatter-plot" to "scatterplot" within the manuscript and figure captions.}
+ p.3, l.32: use past tense when referring to published past work, i.e., "focused" here
  + \response{Accepted change to past tense: "focus" to "focused".}
+ p.3, l.37: -> "Cleveland (1993) provided" (& more)
  + \response{Accepted changes to past tense: Cleveland (1993) "provides" to "provided"; Kosslin (2006) "examines" to "examined"; cognitive tasks "provide" to "provided".}
+ p.3, l.32: "viewers ability" -> "viewers' ability"
  + \response{Accepted grammatical change.}
+ p.3 or p.6, bottom: add some references from the late 1980s/early 1990s that are related to the visual assessment of structural changes in time series visualizations; this reference comes to mind first (although different sources cite it differently): Unwin, A.R., and Wills, G. (1988) “Eyeballing Time Series,” in Proceedings of the 1988 ASA Statistical Computing Section, 263-268. A google scholar search for "Eyeballing Time Series" results in a few related references.
  + \response{Added "Eyeballing Time Series" reference to reference list on p.4,l18.}
+ p.4, l.26 (& more): use consistent capitalization here: "Wilkinson’s Grammar of Graphics (Wilkinson 2013). The grammar of graphics"; I would always capitalize it as "Grammar of Graphics" as a proper name
  + \response{Capitalized all "grammar of graphics" to "Grammar of Graphics".}
+ p.6, l.42: use single quotes for `You Draw It` as in the abstract
  + \response{Added single quotes to 'You Draw It'.}
+ \todo{p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention
the supplementary materials in 1 or 2 sentences)}
+ \todo{p.7, l.34: you had 35 participants and got 119 plots; later (p.10, l.38) you state that each particpant had to evaluate 4 plots; provide this information here immediately; can you summarize why there are 21 plots missing, e.g., some plots were incomplete, some participants were dropping out after just 1 or 2 plots, etc.}
+ p.7, l.51 & p.17, l.27 & l.32: "here" is meaningless for printed materials; list the full URLs (but still make them clickable)
  + \response{All URLs have been changed from "here" to show URL text.}
+ p.8, l43: provide reference for the r2d3 R package
  + \response{Added reference for r2d3 R package.}
+ p.8, l.53: "R statistical software" -> "R software environment for statistical computing and graphics" (based on https://www.r-project.org/); also add the typical reference for R 
  + \response{Changed wording for use of R software and included typical reference for R.}
+ p.8, l.55: "i = 1, ...N" -> "i = 1, ..., N,"
  + \response{Added "," into indexing on p11, l34 and p7, l55}
+ \check{p.9, top: y_x-bar in Table 1 and beta_0 in (1) are not properly defined / used; are they the same?}
  + \response{We aimed to replicate the trends from Mosteller et al. (1981). This original manuscript gives vague descriptions of the coefficients used to describe their trends. The way they defined their paramters was by using the y-value at the mean of the x-values (y_x-bar) and the slope. When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter. In turn, the actual y-intercept (beta_0) will vary slightly from data set to data set, but each trend will go through the same (x-bar, y_x-bar) point. We changed "obtained" to "computed" to clarify that the intercept was computed using the point-slope form of a linear equation.}
+ p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?
  + \response{}
+ p.10, l.38-39: I suppose each participant got one plot of type S, F, V, N ? - say so
+ p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below
+ p.11, l.35 & l.48: be consistent, e.g., "increment" vs. "increments" and missing specification of k
in the 2nd part
+ p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations
+ p.13, l.26: keep the same order of F, S, V, N as before, i.e., S F V N from Table 1 to avoid confusing the reader
+ p.14, Fig 5: this uses even another order (F N S V); stay with the same order (S F V N from Table 1) in all consecutive models and figures to avoid confusing the reader (same for Fig. 6)
+ p.14, l.28: use one single Wood reference for the R package and the most important one that describes the method; leave out the three other Wood references here 
+ p.14, bottom - p.15, top: this model is very similar to the previous one; simply state y_i,j,k,drawn etc. are the same as for model (2) and only specify what is new/different here, e.g., the s_i
+ p.16, Discussion: Add some references from cognitive/human movement sciences that human arm movement is a
complex task. Drawing a straight line is not as easy as it sounds. Could this lead to the large residuals when x approaches 20 ?!? Here are some candidate articles (based on the abstracts). Do some further reading and find additional related references. 
  + https://doi.org/10.1007/BF00241501
  + https://doi.org/10.1145/2820619.2820633
+ p.17, Future work: 
  + I worked through the 12 plots at https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/
and had considerable problems to draw a straight line with a mouse, even when I intended to do so. Similarly, I fail in the classroom when I try to add a regression line to points sketched out on a blackboard. My best solution is to visually estimate the point of means (x-bar, y-bar), use a long ruler, and rotate it so that it goes through the point of means and most of the points in the scatterplot. 
  + This leads to a modified version of this study: Give the participants a horizontal line segment at the bottom of the graph and let them shift and rotate this line segment until they think they have found the best fit for the point cloud. This likely would remove some of the big errors we see in Figs 5 & 6 as x approaches 20.
  + A second option for future work that goes beyond Mosteller et al. (1981) is to create scatterplots with one (or multiple) extreme outliers. What would humans fit then? Still, the PCA line or rather a robust regression line that ignores the outliers (likely not the OLS line that may be totally off from a line that goes through most of the data). [Apparently, this will also depend on the instructions and the feedback that is provided in the training part of such a study.]
  + Neither of these has to be implemented in the current study, but listing both options as natural extensions for the straight line assessment seems to be worthwhile.
+ p.17-21, References:
  + capitalize journal names, e.g., Psychological bulletin & IEEE transactions on visualization and computer graphics
  + capitalize proper names, e.g., in obama’s presidency
  + for books, include the city/state or city/country, e.g., in Cleveland (1993)
  + Deming (1943) is incomplete (is this a book or a journal article?)
  + in Kosslyn & Kosslyn, spell out OUP (and indicate the city/state and not USA)
  + check capitalization of nouns; if I am not mistaken, JCGS uses upper case letters for all nouns [some of these apply to other references as well]

### Reviewer: 2
The authors provide empirical evidence that data from subjects performing a “You draw it” task replicate previous findings in graph perception, namely the finding that, when asked to fit a line on a scatterplot, participants draw a line that minimizes the Euclidean distance of the points to the fit instead of the vertical one (such as in OLS). The paper is written in a good English although the sections describing the statistical analyses are quite complex and a little
verbose.

My main concern about the manuscript is about its novelty. In fact, authors state that they confirm previous findings by Mosteller et al. (1981). However, the study by Ciccione and Dehaene (2021) that they cite in the manuscript, has already provided the exact same results: people minimize the orthogonal distance of points to the fit when asked to provide their best estimation of the regression line. The authors are surely aware of this (since they account for it on page 5, line 24); however, throughout the abstract, the rest of the introduction and the discussion, the authors present their results as novel and simply extending Mosteller study. In fact, what they called “a principal axis component” is mathematically equivalent to the “Deming regression” (as it is called in Ciccione & Dehaene), that minimizes the orthogonal
distance of the points.

In my opinion, the actual “novel” aspects of the presented research work are: 1) the fact that
an online tool initially meant for the general public can replicate empirical findings from the literature on graph perception; 2) the use of generalized additive mixed models. Concerning this second aspect, I am seriously wondering whether it was necessary at all to use such complex models to prove that participants minimized the orthogonal distance of the points to the fit. Simpler statistical analyses could have been (or could be) provided. However, since I am not an expert of these models, I recognize that it might be of interest for a specialized readership.

In conclusion, I strongly believe that the manuscript should concentrate on the actual novel aspects mentioned above (the use of an online tool and the use of generalized additive mixed models) and not on an aspect that the literature has already empirically tested and
statistically confirmed and formalized. If such a major revision is undertaken by the authors, the article might be worth publishing (especially within the scope of the journal).

Below I provide a series of other aspects that should also be considered.

+ Authors should avoid any form of plagiarism: for example, lines 43 to 46 (page 2) and line 55 (page 8) and line 22 (page 9) are identical excerpts from Ciccione & Dehaene.
+ Page 3, line 20. The method of adjustment is introduced but not explained, leaving the reader wondering what it refers to.
+ Page 3, line 50. The line-up protocol is introduced but, again, no clear explanation is provided.
+ Page 5, line 21. The concept of “ensemble perception” is not clearly related to what is described just before: how can the finding itself of human fitting a principal axis on scatterplot “support” (as the authors say) the work on ensemble perception? The link needs to be clarified or avoided.
+ Page 8, table 1. I think it’s a little bit confusing to see that the intercept parameter is called yx whereas in the equation in the text is called b0. The authors should choose one notation or
the other and stay consistent through the text.
+ Page 10, line 39. The reader is invited to refer to an R function for the description of the regression equation based on the principal component axis. Authors should provide an actual formalization of the equation.
+ Page 12, line 48 and page 15, line 45. Again (as in page 5, line 21), the authors’ work is directly presented as providing support for ensemble perception. Ensemble perception simply states that human perception tends to average, in parallel, the information coming from various items/locations in the visual field. Evidence for a principal axis component extraction does not provide a direct piece of evidence for ensemble perception. Authors should better clarify this link or avoid introducing the concept of ensemble perception.

### Editor's Comments to Author

Thank you for your submission and interesting work. Two reviewers have provided feedback and suggestions on your manuscript. There are a few points I would like to emphasize and request clarification:

1. In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis. In particular, how does your work differ from Ciccione and Dehaene (2021)?(Ciccione, L. and Dehaene, S., 2021. Can humans perform mental regression on a graph? Accuracy and bias in theperception of scatterplots. Cognitive Psychology, 128, p.101406.)
2. In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe
the method of estimation. In other words, what are the R functions doing to estimate the model?
3. Could you clarify if the participants were known to be experts in statistics or related fields?
4. Throughout the manuscript, references to “principle components” should be “principal components.”

### 2022-07-19 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
okay small edit to supplementary material. now for submission!
updating analysis to incorporate deming regression instead.

@@
 When creating a graphic, we must consider the design choices most effective for conveying the intended result.
-For instance, we may decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
+For instance, we may decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
 These design choices require that we understand the perceptual and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.
@@
 @cleveland1985graphical found that assessing the position of points along an axis is easier than determining the slope of a line.
-Other studies focus on the viewers ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatter-plot.
+Other studies focused on the viewers' ability to perceive the strength of the relationship between $x$ and $y$ coordinates in a scatterplot.
 For instance, when the data appear dense, viewers tend to overestimate the magnitude of the correlation coefficient [@cleveland1982variables; @lauer1989density].
-@cleveland1993visualizing provides an argument for displaying cyclical patterns with an aspect ratio which sets the curve close to 45$^{\circ}$.
-@kosslyn2006graph examines how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
+@cleveland1993visualizing provided an argument for displaying cyclical patterns with an aspect ratio which sets the curve close to 45$^{\circ}$.
+@kosslyn2006graph examined how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
 For example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].
-The results of these cognitive tasks provide some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
+The results of these cognitive tasks provided some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
 
@@
 One major development in statistical graphics which led to more advanced testing methods is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
-The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
+The Grammar of Graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
 Visual representations are constructed through the use of “tidy data” which is characterized as a data set in which each variable is in its own column, each observation is in its own row, and each value is in its own cell [@wickham2016r].
 Graphics are viewed as a mapping from variables in a data set (or statistics computed from the data) to visual attributes such as the axes, colors, shapes, or facets on the canvas in which the chart is displayed.
-Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the grammar of graphics recommends.
+Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the Grammar of Graphics recommends.
 
-Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields a method for evaluating graphical design choices.
+Combining the Grammar of Graphics with another tool for statistical graphics testing, the statistical lineup, yields a method for evaluating graphical design choices.
 @buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
@@
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. 
+Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@unwin1988eyeballing; @finney1951subjective; @mosteller1981eye]. 
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
@@
 @mosteller1981eye sought to understand the properties of least squares and other computed lines by establishing one systematic method of fitting lines by eye. 
-Participants were asked to fit lines by eye to four scatter-plots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
+Participants were asked to fit lines by eye to four scatterplots using an 8.5 x 11 inch transparency with a straight line etched completely across the middle. 
 A latin square design with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
@@
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression [@deming1943statistical; @linnet1998performance; @martin2000general], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression [@deming1943statistical; @linnet1998performance; @martin2000general], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatterplot.
 
-While not explicitly intended for perceptual testing, in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
+While not explicitly intended for perceptual testing, in 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
 Readers were asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
@@
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application [@shinyPkg] found [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
 
@@
 
-In the study, participants were shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
+In the study, participants were shown an interactive scatterplot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
 The yellow box region moved along as the user drew their trend-line, providing a visual cue which indicates where the user still needed to complete a trend line.
@@
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual. 
-In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` package.
+In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` package [@r2d3].
 While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.
@@
 
-All data processing was conducted in R statistical software. 
-A total of $N = 30$ points $(x_i, y_i), i = 1,...N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
+All data processing was conducted in R software environment for statistical computing and graphics [@Rsoftware]. 
+A total of $N = 30$ points $(x_i, y_i), i = 1,...,N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
 Data were simulated based on a linear model with additive errors: 
@@
 Model equation parameters, $\beta_0$ and $\beta_1$, were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
-We obtained $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
+We computed $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
 Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
@@
 
-```{r eyefitting-simplot, fig.height = 3, fig.width = 9, fig.cap = "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.", out.width="100%"}
+```{r eyefitting-simplot, fig.height = 3, fig.width = 9, fig.cap = "Example of simulated data points displayed in a scatterplot illustrating the trends associated with the four selected parameter choices.", out.width="100%"}
 
@@
 This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focused on the study design and methods related to the current study.
-Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
+Each scatterplot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
 Participants in the study were shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in.
@@
 Due to the randomness in the data generation process, the actual slope of the linear regression line fit through the simulated points could differ from the predetermined slope.
-Therefore, we fit an OLS regression to each scatter-plot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
+Therefore, we fit an OLS regression to each scatterplot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
 Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
@@
 
-For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
+For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...,N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
 Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
@@
 Estimated residual trends were overlaid on the observed individual participant residuals. 
-The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue) for scatter-plots with more noise (F and N).
+The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue) for scatterplots with more noise (F and N).
 By fitting smoothing splines, we can determine whether participants naturally fit a straight trend-line to the set of points or whether they deviate throughout the domain.
-In particular, in scatter-plots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and corrected for their fit toward the end of their trend-line.
-In scatter-plots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PCA regression line.
+In particular, in scatterplots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and corrected for their fit toward the end of their trend-line.
+In scatterplots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PCA regression line.
 As participants continued their trend-line, they crossed through the OLS regression line indicating they estimated the slope in the extreme direction.
@@
 
-\newpage
 \begin{center}
@@
 
-+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
-+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-+ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found [here](https://github.com/earobinson95/you-draw-it-validation-applet).
++ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
++ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found at [earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
++ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
++ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).

### 2022-07-19 analysis/you-draw-it-eyefitting-analysis.Rmd
updating analysis to incorporate deming regression instead.

@@
 ---
-title: "Eye Fitting Straight Lines in the Modern Era"
+title: "Eye Fitting Straight Lines in the Modern Era (JCGS-22-045)"
 subtitle: "Data Analysis"
-author: "Emily Robinson"
+author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
 output: html_document
@@
 )
-library(knitr)
 ```
@@
 
-+ Load libraries
-+ Functions for computing confidence intervals
++ Load required libraries
 
 ```{r setup2}
+# formatting of tables
+library(knitr)
+library(kableExtra)
+
+# read in data
+library(readr)
+
+# data management, plotting, etc.
 library(tidyverse)
-library(kableExtra)
-library(readr)
-library(here)
+
+# used to calculate demming regression equaiton
+library(mcr)
+
+# used to fit linear mixed model to residual trend
+library(lme4)
+library(emmeans)
+
+# used to fit generalized additive mixed model to residual trend
 library(mgcv)
-library(lme4)
-library(lmerTest)
-library(emmeans)
-source(here("analysis/gamm-predict-function.R"))
+```
+
++ Source code to load functions for computing confidence intervals
+
+```{r setup3}
+#' library(lme4)
+#' library(mgcv)
+#' lmer_model <- lmer(Reaction ~ Days + (Days || Subject), data = sleepstudy)
+#' ga_model <- gam(Reaction ~ Days + s(Subject, bs = "re") + s(Days, Subject, bs = "re"),
+#'   data = sleepstudy,
+#'   method = "REML"
+#' )
+#'
+#' head(
+#'   data.frame(
+#'     lmer = predict(lmer_model),
+#'     gam = predict_gamm(ga_model)
+#'   )
+#' )
+#'
+#' head(
+#'   cbind(
+#'     lmer = predict(lmer_model, re.form = NA),
+#'     gam1 = predict_gamm(ga_model, re_form = NA),
+#'     gam2 = predict_gamm(ga_model,
+#'       exclude = c("s(Subject)", "s(Days,Subject)")
+#'     )
+#'   )
+#' )
+#'
+#' head(predict_gamm(ga_model, se = TRUE))
+#' @export
+predict_gamm <- function(
+  model,
+  newdata,
+  re_form = NULL,
+  se = FALSE,
+  include = NULL,
+  exclude = NULL,
+  keep_prediction_data = FALSE,
+  ...) {
+
+  # Note because predict doesn't use NULL, can't use NULL for new_data arg or
+  # even a differently named arg, and I'm not going into the weeds of rlang to
+  # find a hack.
+
+  # basic checks
+  if (!inherits(model, "gam")) stop("Need a gam object.")
+
+  if (!rlang::is_null(include) && !rlang::is_character(include)) {
+    stop("include must be NULL or character.")
+  }
+
+  if (!rlang::is_null(exclude) && !rlang::is_character(exclude)) {
+    stop("exclude must be NULL or character.")
+  }
+
+  if (!rlang::is_null(re_form) &&
+      !rlang::is_na(re_form) &
+      !rlang::is_character(re_form)) {
+    stop("re_form must be NULL, NA, or character.")
+  }
+
+  if (any(include %in% exclude)) {
+    stop("You can't include and exclude the same thing.")
+  }
+
+  if (!rlang::is_logical(se)) {
+    stop("se must be TRUE or FALSE")
+  }
+
+  if (!rlang::is_logical(keep_prediction_data)) {
+    stop("keep_prediction_data must be TRUE or FALSE")
+  }
+
+  # standard prediction would simply call predict.gam
+  if (rlang::is_null(re_form) | rlang::is_character(re_form)) {
+    if (rlang::is_null(re_form)) {
+      preds <- predict(model,
+                       newdata,
+                       se = se,
+                       terms = include,
+                       exclude = exclude,
+                       ...
+      )
+    } else {
+      preds <- predict(model,
+                       newdata,
+                       se = se,
+                       terms = c(include, re_form),
+                       exclude = exclude,
+                       ...
+      )
+    }
+  } else if (rlang::is_na(re_form)) {
+
+    # FE only
+    re_terms <- sapply(model$smooth, function(x) inherits(x, "random.effect"))
+    re_terms <- sapply(model$smooth[re_terms], function(x) x$label)
+
+    preds <- predict(model,
+                     newdata,
+                     se = se,
+                     terms = include,
+                     exclude = c(re_terms, exclude),
+                     ...
+    )
+  }
+
+  if (se) {
+    preds <- data.frame(prediction = preds$fit, se = preds$se)
+  } else {
+    preds <- data.frame(prediction = preds)
+  }
+
+  if (keep_prediction_data) {
+    if (missing(newdata)) {
+      base <- model$model
+    } else {
+      base <- newdata
+    }
+    preds <- data.frame(base, preds)
+  }
+
+  preds
+}
+```
+
+## COMPUTE DEMING REGRESSION
+
+This section only needed to be run once to compute the fitted values based on the principal axis (Deming regression).
+The ordinary least squares fitted values were computed in the original data simulation step.
+The resulting data set is the `eyefitting-model-data.csv` used for analysis.
+
+```{r demming-regression}
+
+# read in feedback and simulated data
+eyefitting_feedback_data <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-feedback-data.csv")
+eyefitting_simulated_data <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-simulated-data.csv")
+
+# fit deming regression
+demingFit <- function(data){
+  fit <- mcreg(data$x, data$y, method.reg = "Deming")
+  demingIntercept <- fit@para[1]
+  demingSlope     <- fit@para[2]
+
+  return(tibble(demingIntercept = demingIntercept, demingSlope = demingSlope))
+}
+
+pca_preds <- eyefitting_simulated_data %>%
+  filter(dataset == "point_data", participant_id %in% unique(eyefitting_feedback_data$participant_id)) %>%
+  as_tibble() %>%
+  tidyr::nest(-c(participant_id, parm_id)) %>%
+  dplyr::mutate(deming_fit = purrr::map(data, demingFit)) %>%
+  unnest(deming_fit) %>%
+  select(participant_id, parm_id, demingIntercept, demingSlope) %>%
+  expand_grid(x = seq(0,20,0.25)) %>%
+  mutate(ypca = demingIntercept + demingSlope*x)
+
+eyefitting_model_data <- eyefitting_feedback_data %>%
+  left_join(pca_preds, by = c("participant_id", "x", "parm_id")) %>%
+  mutate(residual_ols_drawn = ydrawn - yols,
+         residual_pca_drawn = ydrawn - ypca) %>%
+  dplyr::select(nick_name, study_starttime, age, gender, academic_study, recruitment, participant_id, parm_id, plot_id, x, yols, ypca, ydrawn, residual_ols_drawn, residual_pca_drawn)
+
+# save data to csv - already run once and pushed to GitHub.
+# write.csv(eyefitting_model_data, file = "eyefitting-model-data.csv", row.names = F, na = "")
 ```
@@
 
-De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
+De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-model-data.csv).
 
 ```{r data}
-eyefitting_model_data  <- read_csv(here("data/youdrawit-eyefitting-model-data.csv"))
+eyefitting_model_data  <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-model-data.csv")
 
-factorCols <- c("participantID", "nick_name", "study_starttime", "age", "gender", "academic_study",
-                "recruitment", "plotID", "parm_id")
+factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
+                "recruitment", "participant_id", "parm_id", "plot_id")
 eyefitting_model_data[,factorCols] <- lapply(eyefitting_model_data[,factorCols], factor)
 ```
+
+There are a total of `length(unique(eyefitting_model_data$participant_id))` participants and `length(unique(eyefitting_model_data$participant_id))` 'You Draw It' plots completed.
 
@@
   scroll_box(width = "100%", height = "200px")
-
-eyefitting_model_data <- eyefitting_model_data %>%
-  dplyr::mutate(`Parameter Choice` = parm_id)
 ```
@@
 
-### OLS with loess smoother
+### EXAMPLE
 
-```{r yloess-ols-plot}
+```{r}
+# Example
+eyefitting_model_data %>%
+  filter(participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>%
+  ggplot(aes(x = x)) +
+  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
+  geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
+  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
+  geom_point(data = eyefitting_simulated_data %>%
+               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F"),
+             aes(x = x, y = y)) +
+  facet_wrap(~ parm_id) +
+  theme_bw() +
+  theme(aspect.ratio = 1) +
+  scale_y_continuous("y") +
+  scale_color_manual("", values = c("black", "steelblue", "orange2")) +
+  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
+```
+
+### OLS Spaghetti Plot
+
+```{r y-ols-plot}
+
 eyefitting_model_data %>%
   ggplot(aes(x = x)) +
-  geom_line(aes(y = yloess, group = plotID), alpha = 0.5, color = "steelblue") +
-  geom_line(alpha = 0.2, aes(y = yols, group = participantID)) +
-  # geom_line(alpha = 0.2, aes(y = ypca, group = participantID), linetype = "dashed") +
+  geom_line(aes(y = ydrawn, group = plot_id), alpha = 0.5, color = "steelblue") +
+  geom_line(alpha = 0.2, aes(y = yols, group = participant_id)) +
   facet_wrap(~ parm_id) +
@@
 
-### PCA with loess smoother
+### Principal Axis Spaghetti Plot
 
-```{r yloess-pca-plot}
+```{r y-pca-plot}
 eyefitting_model_data %>%
   ggplot(aes(x = x)) +
-  geom_line(aes(y = yloess, group = plotID), alpha = 0.5, color = "steelblue") +
-  # geom_line(alpha = 0.2, aes(y = yols, group = participantID)) +
-  geom_line(alpha = 0.2, aes(y = ypca, group = participantID)) +
+  geom_line(aes(y = ydrawn, group = plot_id), alpha = 0.5, color = "steelblue") +
+  geom_line(alpha = 0.2, aes(y = ypca, group = participant_id)) +
   facet_wrap(~ parm_id) +
@@
   scale_x_continuous(limits = c(0, 20)) +
-  ggtitle("PCA")
+  ggtitle("PA")
 ```
 
-### Residuals with loess smoother
+### Residuals Spaghetti Plot
 
@@
   ggplot(aes(x = x)) +
-  geom_line(aes(y = residualols.loess, group = plotID, color = "OLS"), alpha = 0.3) +
-  geom_line(aes(y = residualpca.loess, group = plotID, color = "PCA"), alpha = 0.3) +
+  geom_line(aes(y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.3) +
+  geom_line(aes(y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.3) +
   geom_hline(yintercept = 0, linetype = "dashed") +
@@
 ```{r, anova-lmer, echo = F}
-sv <- c("participantID", "parm_id", "x", "x:parm_id", "residual")
+sv <- c("participant_id", "parm_id", "x", "x:parm_id", "residual")
 df <- c("(50 - 1) = 49", "(4 - 1) = 3", "1", "(4 - 1) = 3", "143 - by subtraction")
@@
 # OLS
-eyefitting.ols.lmer <- lmer(residualols ~ -1 + parm_id + x:parm_id + (1|participantID),
+eyefitting.ols.lmer <- lmer(residual_ols_drawn ~ -1 + parm_id + x:parm_id + (1|participant_id),
                        data = eyefitting_model_data)
@@
 # PCA
-eyefitting.pca.lmer <- lmer(residualpca ~ -1 + parm_id + x:parm_id + (1|participantID),
+eyefitting.pca.lmer <- lmer(residual_pca_drawn ~ -1 + parm_id + x:parm_id + (1|participant_id),
                             data = eyefitting_model_data)
@@
 
-# write.csv(eyefitting.preds.lmer, "results/youdrawit-eyefitting-lmerpred-data.csv", row.names = F, na = "")
+# write.csv(eyefitting.preds.lmer, "youdrawit-eyefitting-lmerpred-data.csv", row.names = F, na = "")
 
@@
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+            aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PCA"), alpha = 0.2) +
+  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
   geom_line(aes(y = emmean.ols, color = "OLS")) +
-  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
+  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PA"), color = NA, alpha = 0.4) +
   geom_line(aes(y = emmean.pca, color = "PCA")) +
@@
   scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))
+  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PA"))
 
@@
 # OLS
-eyefitting.ols.gamm <- bam(residualols ~ parm_id + s(x, by = parm_id) + 
-                             s(participantID, bs = "re") +
-                             s(x,participantID, bs = "re"),
+eyefitting.ols.gamm <- bam(residual_ols_drawn ~ parm_id + s(x, by = parm_id) + 
+                             s(participant_id, bs = "re") +
+                             s(x,participant_id, bs = "re"),
             method = "REML",
@@
 # PCA
-eyefitting.pca.gamm <- bam(residualpca ~ -1 + parm_id + s(x) +
+eyefitting.pca.gamm <- bam(residual_pca_drawn ~ -1 + parm_id + s(x) +
                              s(x, by = parm_id) +
-                             s(participantID, bs = "re") +
-                             s(x,participantID, bs = "re"),
+                             s(participant_id, bs = "re") +
+                             s(x,participant_id, bs = "re"),
                            method = "REML",
@@
                                     x = seq(0,20, 0.5),
-                                    participantID = eyefitting_model_data$participantID[1])
+                                    participant_id = eyefitting_model_data$participant_id[1])
 
@@
 
-# write.csv(eyefitting.grid.gamm, "results/youdrawit-eyefitting-gammpred-data.csv", row.names = F, na = "")
+# write.csv(eyefitting.grid.gamm, "youdrawit-eyefitting-gammpred-data.csv", row.names = F, na = "")
 
@@
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
   geom_line(aes(y = ols.pred, color = "OLS")) +
-  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
-  geom_line(aes(y = pca.pred, color = "PCA")) +
+  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = pca.pred, color = "PA")) +
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +

### 2022-07-19 analysis/gamm-predict-function.R

@@ -1,122 +1,0 @@
-#' library(lme4)
-#' library(mgcv)
-#' lmer_model <- lmer(Reaction ~ Days + (Days || Subject), data = sleepstudy)
-#' ga_model <- gam(Reaction ~ Days + s(Subject, bs = "re") + s(Days, Subject, bs = "re"),
-#'   data = sleepstudy,
-#'   method = "REML"
-#' )
-#'
-#' head(
-#'   data.frame(
-#'     lmer = predict(lmer_model),
-#'     gam = predict_gamm(ga_model)
-#'   )
-#' )
-#'
-#' head(
-#'   cbind(
-#'     lmer = predict(lmer_model, re.form = NA),
-#'     gam1 = predict_gamm(ga_model, re_form = NA),
-#'     gam2 = predict_gamm(ga_model,
-#'       exclude = c("s(Subject)", "s(Days,Subject)")
-#'     )
-#'   )
-#' )
-#'
-#' head(predict_gamm(ga_model, se = TRUE))
-#' @export
-predict_gamm <- function(
-  model,
-  newdata,
-  re_form = NULL,
-  se = FALSE,
-  include = NULL,
-  exclude = NULL,
-  keep_prediction_data = FALSE,
-  ...) {
-  
-  # Note because predict doesn't use NULL, can't use NULL for new_data arg or
-  # even a differently named arg, and I'm not going into the weeds of rlang to
-  # find a hack.
-  
-  # basic checks
-  if (!inherits(model, "gam")) stop("Need a gam object.")
-  
-  if (!rlang::is_null(include) && !rlang::is_character(include)) {
-    stop("include must be NULL or character.")
-  }
-  
-  if (!rlang::is_null(exclude) && !rlang::is_character(exclude)) {
-    stop("exclude must be NULL or character.")
-  }
-  
-  if (!rlang::is_null(re_form) &&
-      !rlang::is_na(re_form) &
-      !rlang::is_character(re_form)) {
-    stop("re_form must be NULL, NA, or character.")
-  }
-  
-  if (any(include %in% exclude)) {
-    stop("You can't include and exclude the same thing.")
-  }
-  
-  if (!rlang::is_logical(se)) {
-    stop("se must be TRUE or FALSE")
-  }
-  
-  if (!rlang::is_logical(keep_prediction_data)) {
-    stop("keep_prediction_data must be TRUE or FALSE")
-  }
-  
-  # standard prediction would simply call predict.gam
[…]

### 2022-07-20 manuscript-revisions/response-to-reviewers.qmd
working on revisions

@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
 
@@
   + \response{Added single quotes to 'You Draw It'.}
-+ \todo{p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention
-the supplementary materials in 1 or 2 sentences)}
++ \todo{p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention the supplementary materials in 1 or 2 sentences)}
 + \todo{p.7, l.34: you had 35 participants and got 119 plots; later (p.10, l.38) you state that each particpant had to evaluate 4 plots; provide this information here immediately; can you summarize why there are 21 plots missing, e.g., some plots were incomplete, some participants were dropping out after just 1 or 2 plots, etc.}
@@
   + \response{Added "," into indexing on p11, l34 and p7, l55}
-+ \check{p.9, top: y_x-bar in Table 1 and beta_0 in (1) are not properly defined / used; are they the same?}
++ \dubcheck{p.9, top: $y_{\bar{x}}$ in Table 1 and $\beta_0$ in (1) are not properly defined and used; are they the same?}
   + \response{We aimed to replicate the trends from Mosteller et al.
@@
 This original manuscript gives vague descriptions of the coefficients used to describe their trends.
-The way they defined their paramters was by using the y-value at the mean of the x-values (y_x-bar) and the slope.
+The way they defined their parameters was by using the y-value at the mean of the x-values $(y_{\bar{x}})$ and the slope.
 When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
-In turn, the actual y-intercept (beta_0) will vary slightly from data set to data set, but each trend will go through the same (x-bar, y_x-bar) point.
-We changed "obtained" to "computed" to clarify that the intercept was computed using the point-slope form of a linear equation.}
-+ p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?
-  + \response{}
+In turn, the actual y-intercept $(beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
+We changed "obtained" to "computed" to clarify that the intercept was computed using the point slope form of a linear equation.}
++ \dubcheck{p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?}
+  + \response{We double checked the data simulation process based on the shiny applet code.
+There was an extra line of code in the manuscript file that incorrrectly filtered points out.
+There are were 30 points simulated for each parameter choice which is stated on p7 l53.
+A new "set" of data was simulated randomly for each participant with the same underlying parameters and a fixed number of points.
+We added the domain for each parameter choice to the table.
+The parameter choice names were selected to reflect the labels used in Mosteller et al.
+(1981).
+We clarified this on p8 l25 with "Model equation parameters, $beta_0$ and $beta_1$, and parameter choice letter names (S, F, V, N) and V), were selected to reflect the four data sets used and labeled in Mosteller et al..." See analysis in supplementary material for code on the data simulation procedure.}
 + p.10, l.38-39: I suppose each participant got one plot of type S, F, V, N ? - say so
-+ p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below
+  + \response{We clarified that each participant gote one of each of the plot types.}
++ \todo{p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below}
 + p.11, l.35 & l.48: be consistent, e.g., "increment" vs. "increments" and missing specification of k
 in the 2nd part
-+ p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations
+  + \response{Changed to "increment" for both to be consistent and added specification of k in l.48.}
++ \todo{p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations}
 + p.13, l.26: keep the same order of F, S, V, N as before, i.e., S F V N from Table 1 to avoid confusing the reader
@@
 + p.14, bottom - p.15, top: this model is very similar to the previous one; simply state y_i,j,k,drawn etc. are the same as for model (2) and only specify what is new/different here, e.g., the s_i
-+ p.16, Discussion: Add some references from cognitive/human movement sciences that human arm movement is a
-complex task.
++ \todo{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task.
 Drawing a straight line is not as easy as it sounds.
@@
   + https://doi.org/10.1145/2820619.2820633
-+ p.17, Future work: 
++ \todo{p.17, Future work:} 
   + I worked through the 12 plots at https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/
@@
   + Neither of these has to be implemented in the current study, but listing both options as natural extensions for the straight line assessment seems to be worthwhile.
-+ p.17-21, References:
++ \todo{p.17-21, References:}
   + capitalize journal names, e.g., Psychological bulletin & IEEE transactions on visualization and computer graphics

### 2022-07-20 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on revisions

@@
 
-Model equation parameters, $\beta_0$ and $\beta_1$, were selected to reflect the four data sets (F, N, S, and V) used in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
+Model equation parameters, $\beta_0$ and $\beta_1$, and parameter choice letter names (S, F, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
 We computed $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
-Parameter choices F, N, and S simulated data across a domain of 0 to 20. 
+Parameter choices S, F, and N simulated data across a domain of 0 to 20. 
 Parameter choice F produced a trend with a positive slope and a large variance while N had a negative slope and a large variance. 
@@
                      slope = c(0.66, -0.70, 0.66, 1.98),
-                     sigma = c(1.98, 2.5, 1.3, 1.5)
+                     sigma = c(1.98, 2.5, 1.3, 1.5),
+                     x_range = c("(0,20)", "(0,20)", "(0,20)", "(4,16)")
           ) %>%
@@
                          label = "eyefitting-parameters",
-                         col.names = c("Parameter Choice", "$y_{\\bar{x}}$", "$\\beta_1$", "$\\sigma$"),
+                         col.names = c("Parameter Choice", "$y_{\\bar{x}}$", "$\\beta_1$", "$\\sigma$", "Domain"),
                          caption = "Designated model equation parameters for simulated data.")
@@
 
-eyefitting_example_sim <- read.csv(here("results/youdrawit-eyefitting-simdata-example.csv"))
+eyefitting_example_sim <- read.csv(here("analysis/eyefitting-simulated-data.csv"))
 
 eyefitting_example_simplot <- eyefitting_example_sim %>%
-  filter(data == "point_data") %>%
-  filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
-  mutate(dataset = factor(dataset, levels = c("S", "F", "V", "N"))) %>%
-  dplyr::rename(`Parameter Choice` = dataset) %>%
+  filter(dataset == "point_data") %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x, y = y)) +
@@
 Instructions guided participants to start at the edge of the yellow box, to make sure the yellow shaded region was moving along with their mouse as they drew, and that they could draw over their already drawn line.
-Practice plots were then followed by four 'You Draw It' task plots associated with the current study.
+Practice plots were then followed by one of each of the four 'You Draw It' task plots associated with the current study (S, F, V, and N).
 The order of the task plots was randomly assigned for each individual in a completely randomized design.
@@
 Therefore, we fit an OLS regression to each scatterplot to obtain estimated parameters $\hat\beta_{0,OLS}$ and $\hat\beta_{1,OLS}$. 
-Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increments across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$., for $k = 1, ..., 4 x_{max} +1$.
+Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increment across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$, for $k = 1, ..., 4 x_{max} +1$.
 The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
 The estimated slope, $\hat\beta_{1,PCA}$, was determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of the simulated points, $(\bar x_i, \bar y_i)$.
-Fitted values, $\hat y_{k,PCA}$, were then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
+Fitted values, $\hat y_{k,PCA}$, were then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$, for $k = 1, ..., 4 x_{max} +1$.
 

### 2022-07-20 analysis/you-draw-it-eyefitting-analysis.Rmd
working on revisions

@@
 }
+```
+
+## DATA SIMULATION PROCEDURE
+
+```{r}
+
+eyefitting_parameter_details <- read_csv("eyefitting_parameter_details.csv")
+eyefitting_parameter_details %>% knitr::kable()
+
+linearDataGen <-
+  function(y_xbar,
+           slope,
+           sigma,
+           N = 30,
+           x_min = 0,
+           x_max = 20,
+           x_by  = 0.25){
+
+    # Set up x values
+    xVals <- seq(x_min, x_max, length.out = floor(N*1))
+    xVals <- sample(xVals, N, replace = FALSE)
+    xVals <- jitter(xVals)
+    xVals <- ifelse(xVals < x_min, x_min, xVals)
+    xVals <- ifelse(xVals > x_max, x_max, xVals)
+
+    # From slope intercept form
+    # y-y_xbar = m(x-xbar)
+    # y = m(x-xbar) + y_xbar = mx - mxbar + y_xbar
+    yintercept = y_xbar - slope*mean(xVals)
+
+    # Generate "good" errors
+    repeat{
+      errorVals <- rnorm(N, 0, sigma)
+      if(mean(errorVals[floor(N/3)]) < 2*sigma & mean(errorVals[floor(N/3)] > -2*sigma)){
+        break
+      }
+    }
+
+    # Simulate point data
+    point_data <- tibble(dataset = "point_data",
+                         x = xVals,
+                         y = yintercept + slope*x + errorVals) %>%
+      arrange(x)
+
+    # Obtain least squares regression coefficients
+    lm.fit <- lm(y ~ x, data = point_data)
+    yintercepthat <- coef(lm.fit)[1] %>% as.numeric()
+    slopehat <- coef(lm.fit)[2] %>% as.numeric()
+
+    # Simulate best fit line data
+    line_data <- tibble(dataset = "line_data",
+                        x = seq(x_min, x_max, x_by),
+                        y = yintercepthat + slopehat*x)
+
+    data <- list(point_data = point_data, line_data = line_data)
+
+    return(data)
+  }
+
+set.seed(68505)
+eyefitting_sim_data <- eyefitting_parameter_details %>%
+  mutate(data = purrr::pmap(list(y_xbar = y_xbar,
+                                 slope = slope,
+                                 sigma = sigma,
+                                 x_min = x_min,
+                                 x_max = x_max,
+                                 x_by = x_by), linearDataGen)) %>%
+  expand_grid(linear = "true",
+              draw_start = 5,
+              free_draw = TRUE) %>%
+  unnest(data) %>%
+  unnest(data)
+
+npoints_data <- eyefitting_sim_data %>%
+  filter(dataset == "point_data")
+table(npoints_data$parm_id)
+
+# write.csv(eyefitting_sim_data, "eyefitting-simulated-data.csv", row.names = F, na = "")
+
+eyefitting_example_simplot <- eyefitting_sim_data %>%
+  filter(dataset == "point_data") %>%
+  # filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  dplyr::rename(`Parameter Choice` = parm_id) %>%
+  ggplot(aes(x = x, y = y)) +
+  geom_point(size = 1) +
+  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+  legend.position = "none",
+  plot.title   = element_text(size = 12, hjust = 0),
+  axis.text    = element_text(size = 12),
+  axis.title   = element_text(size = 12),
+  legend.title = element_text(size = 12),
+  legend.text  = element_text(size = 12),
+  # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+  # strip.background = element_rect(size = 0.5),
+  legend.key.size = unit(1, "line")
+) +
+  scale_y_continuous(breaks = seq(-10, 20, 5))
+eyefitting_example_simplot
 ```
@@
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, ncol = 2) +
+  facet_wrap(~`Parameter Choice`, ncol = 1) +
   theme_bw(base_size = 14) +

### 2022-08-02 manuscript-revisions/response-to-reviewers.qmd
working on revisions

@@
 All responses to reviewers' comments are written in \response{blue}.
+
+\todo{Need to address yet.} \dubcheck{Double check / review.}
 
@@
 In turn, the actual y-intercept $(beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
-We changed "obtained" to "computed" to clarify that the intercept was computed using the point slope form of a linear equation.}
+We changed the equation notation to follow the point-slope equation rather than changing notation.}
 + \dubcheck{p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?}
@@
 + p.10, l.38-39: I suppose each participant got one plot of type S, F, V, N ? - say so
-  + \response{We clarified that each participant gote one of each of the plot types.}
-+ \todo{p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below}
+  + \response{We clarified that each participant got one of each of the plot types.}
++ p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below
+  + \response{Notation was changed to "PA" to indicate the principal axis.}
 + p.11, l.35 & l.48: be consistent, e.g., "increment" vs. "increments" and missing specification of k
@@
 + p.13, l.26: keep the same order of F, S, V, N as before, i.e., S F V N from Table 1 to avoid confusing the reader
+  + \response{The order of the parameter choices has been changed to be consistent (F, S, V, N) for all tables, text, and figures.}
 + p.14, Fig 5: this uses even another order (F N S V); stay with the same order (S F V N from Table 1) in all consecutive models and figures to avoid confusing the reader (same for Fig. 6)
+  + \response{The order of the parameter choices has been changed to be consistent (F, S, V, N) for all tables, text, and figures.}
 + p.14, l.28: use one single Wood reference for the R package and the most important one that describes the method; leave out the three other Wood references here
+  + \response{Selected one reference (book) which we used for the analysis.}
 + p.14, bottom - p.15, top: this model is very similar to the previous one; simply state y_i,j,k,drawn etc. are the same as for model (2) and only specify what is new/different here, e.g., the s_i
+  + \response{Clarified similarities between the linear mixed model equation and the generalized additive mixed model equation by stating "Defining $e_{ijk,fit}$ the same as in equation (2) above, the GAMM equation...".}
 + \todo{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task.
@@
 + Authors should avoid any form of plagiarism: for example, lines 43 to 46 (page 2) and line 55 (page 8) and line 22 (page 9) are identical excerpts from Ciccione & Dehaene.
+  + \response{Thank you for catching this.
+Amongst literature review and edits, we did not realize we had directly quoted.
+We have rephrased the information in each of the indicated locations as they are important points to include.}
 + Page 3, line 20. The method of adjustment is introduced but not explained, leaving the reader wondering what it refers to.
+  + \response{We describe the method of adjustment and added a reference.}
 + Page 3, line 50. The line-up protocol is introduced but, again, no clear explanation is provided.
+  + \response{We added a couple sentences to describe the lineup protocol in addition to the Buja (2009) reference.
+"A statistical lineup is a plot consisting of smaller panels where the viewer is asked to identify the target panel containing the real data from among a set of decoy null plots which display data under the assumption there is no relationship.
+If the viewer can identify the target panel randomly embedded within the set of null panels, this suggests that the real data is visually distinct from data generated under the null model."}
 + Page 5, line 21.
@@
 The link needs to be clarified or avoided.
+  + \response{We removed the sentence discussing "ensemble perception" in relation to the Mosteller (1981) paper.}
 + Page 8, table 1. I think it’s a little bit confusing to see that the intercept parameter is called yx whereas in the equation in the text is called b0. The authors should choose one notation or
 the other and stay consistent through the text.
+  + \response{We aimed to replicate the trends from Mosteller et al.
+(1981).
+This original manuscript gives vague descriptions of the coefficients used to describe their trends.
+The way they defined their parameters was by using the y-value at the mean of the x-values $(y_{\bar{x}})$ and the slope.
+When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
+In turn, the actual y-intercept $(beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
+We changed the equation notation to follow the point-slope equation rather than changing notation.}
 + Page 10, line 39.
@@
 Authors should provide an actual formalization of the equation.
+  + \response{We reran the analysis using the mcreg function in the mcr package which implements Deming regression (equivalent to the principal axis regression).
+We clarified and provided a reference for this in the "Fitted Regressions" Results section.}
 + Page 12, line 48 and page 15, line 45.
@@
 Authors should better clarify this link or avoid introducing the concept of ensemble perception.
+  + \response{We agree that ensemble perception is not necessarily the focus of this paper and have removed introducing the concept.}
 
@@
 4. Throughout the manuscript, references to “principle components” should be “principal components.”
+  + \response{Based on a reviewer's suggestion, we have changed terminology to principal axis.}

### 2022-08-02 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
working on revisions

@@
 @kosslyn2006graph examined how Gestalt principles of perceptual organization are instrumental in extracting data from a chart.
-For example, data points that are closer in space are more likely to be perceived as grouped [@ciccione2020grouping], and vertical and horizontal lines are easier to discriminate than oblique ones [@appelle1972perception].
+For example, @ciccione2020grouping conducted a study to support data points located closer together are more likely to be perceived as the same group and @appelle1972perception found that it is easier to discriminate vertical and horizontal lines than oblique lines.
 The results of these cognitive tasks provided some consistent guidance for chart design; however, other methods of visual testing can further evaluate design choices and help us understand cognitive biases related to the evaluation of statistical charts.
@@
 Early research studies considered graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli.
-Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment to estimate the magnitude of the impact of the sine illusion.
+Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment - a technique which requires participants to alter a changing stimulus to match a given constant stimuli [@gescheider_1997]- to estimate the magnitude of the impact of the sine illusion.
 However, there are more modern testing methods that have been developed since the heyday of psychophysics.
@@
 @buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
+A statistical lineup is a plot consisting of smaller panels where the viewer is asked to identify the target panel containing the real data from among a set of decoy null plots which display data under the assumption there is no relationship.
+If the viewer can identify the target panel randomly embedded within the set of null panels, this suggests that the real data is visually distinct from data generated under the null model.
 Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
@@
 A latin square design with packets of the set of points stapled together in four different sequences was used to determine if there is an effect of order of presentation.
-It was found that order of presentation had no effect and that participants tended to fit the slope of the principal axis (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
-These results support previous research on "ensemble perception" indicating the visual system can compute averages of various features in parallel across the items in a set [@chong2003representation; @chong2005statistical; @van2011rapid].
+It was found that order of presentation had no effect and that participants tended to fit the slope of the principal axis (PA) (error minimized orthogonally, both horizontal and vertical, to the regression line) over the slope of the least squares regression line (error minimized vertically to the regression line).
+<!-- These results support previous research on "ensemble perception" indicating the visual system can compute averages of various features in parallel across the items in a set [@chong2003representation; @chong2005statistical; @van2011rapid].
+-->
 
@@
 We validate the 'You Draw It' method by replicating the less technological study conducted by @mosteller1981eye.
-Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
+Based on previous research, we hypothesize that visual regression tends to mimic regression based on the principal axis rather than an ordinary least squares regression.
 In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.
@@
 A total of $N = 30$ points $(x_i, y_i), i = 1,...,N$ were generated for $x_i \in [x_{min}, x_{max}]$ where $x$ and $y$ have a linear relationship.
-Data were simulated based on a linear model with additive errors: 
+Data were simulated based on the point-slope form of a linear model with additive errors: 
 \begin{align}
-y_i & = \beta_0 + \beta_1 x_i + e_i \\
+y_i = \beta_1(x_i-\bar{x}) + y_{\bar{x}} + e_i \\
 \text{with } e_i & \sim N(0, \sigma^2). \nonumber
@@
 
-Model equation parameters, $\beta_0$ and $\beta_1$, and parameter choice letter names (S, F, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
-We computed $\beta_0$ from the point-slope equation of a line using the mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$.
+Model equation parameters, $\beta_1$, $y_{\bar{x}}$, and parameter choice letter names (F, S, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
+The mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$ were used in the point-slope equation of a line.
 Parameter choices S, F, and N simulated data across a domain of 0 to 20. 
@@
           ) %>%
-            mutate(Parm = factor(Parm, levels = c("S", "F", "V", "N"))) %>%
+            mutate(Parm = factor(Parm, levels = c("F", "S", "V", "N"))) %>%
             arrange(Parm) %>%
@@
 
-eyefitting_example_sim <- read.csv(here("analysis/eyefitting-simulated-data.csv"))
+eyefitting_sim_data <- read_csv("analysis/eyefitting-simulated-data.csv")
 
-eyefitting_example_simplot <- eyefitting_example_sim %>%
+eyefitting_example_simplot <- eyefitting_sim_data %>%
   filter(dataset == "point_data") %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  # filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
@@
   geom_point(size = 1) +
-  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  facet_wrap(~`Parameter Choice`, ncol = 4, labeller = label_both) +
   theme_bw(base_size = 14) +
@@
 This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focused on the study design and methods related to the current study.
-Each scatterplot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
+Each data set was generated randomly and independently for each participant at the start of the experiment and mapped to a scatterplot.
 Participants in the study were shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in.
@@
 
-We compared the participant drawn line to two regression lines determined by ordinary least squares (OLS) regression and regression based on the principal axis (PCA).
-\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.
+We compared the participant drawn line to two regression lines determined by ordinary least squares (OLS) regression and regression based on the principal axis (PA).
+\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the PA which minimizes the Euclidean distance of points (orthogonal) from the line.
 
@@
 Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increment across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$, for $k = 1, ..., 4 x_{max} +1$.
-The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
-The estimated slope, $\hat\beta_{1,PCA}$, was determined by the ratio of the axis rotation in $y$ and axis rotation in $x$ of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of the simulated points, $(\bar x_i, \bar y_i)$.
-Fitted values, $\hat y_{k,PCA}$, were then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$, for $k = 1, ..., 4 x_{max} +1$.
+The PA regression slope, $\hat\beta_{1,PA}$, and y-intercept, $\hat\beta_{0,PA}$, were determined using the `mcreg` function in the `mcr` package in R [@mcr_pkg] which implements Deming regression (equivalent to a regression based on the slope of the first principal axis).
+Fitted values, $\hat y_{k,PA}$ were then obtained every 0.25 increment across the domain from the PA regression equation, $\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k$, for $k = 1, ..., 4 x_{max} +1$.
 
@@
         ) +
-  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
+  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PA")) +
+  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PA"))
 ```
@@
 ```{r eyefitting-plots}
-eyefitting_model_data <- read.csv("data/youdrawit-eyefitting-model-data.csv") %>%
-  dplyr::rename(`Parameter Choice` = parm_id)
+eyefitting_simulated_data <- read_csv("analysis/eyefitting-simulated-data.csv")
+
+eyefitting_model_data <- read.csv("analysis/eyefitting-model-data.csv")
+
+factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
+                "recruitment", "participant_id", "parm_id", "plot_id")
+eyefitting_model_data[,factorCols] <- lapply(eyefitting_model_data[,factorCols], factor)
 ```
 
-For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...,N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
-Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
+For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PA}$ for parameter choice $i = 1,2,3,4$, j = $1,...,N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
+Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PA fitted values ($e_{ijk,PA} = y_{ijk,drawn} - \hat y_{ijk,PA}$) were made across the domain.
 \cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.
@@
 ```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one `You Draw It' task plot.
-Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).", out.width="50%"}
-trial.feedback <- read.csv("results/youdrawit-eyefitting-example-feedback.csv") %>%
-    mutate(`Parameter Choice` = "F")
-trial.sim <- read.csv("results/youdrawit-eyefitting-example-simulated.csv") %>%
-    mutate(`Parameter Choice` = "F")
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).", out.width="50%"}
+# trial.feedback <- read.csv("results/youdrawit-eyefitting-example-feedback.csv") %>%
+#     mutate(`Parameter Choice` = "F")
+# trial.sim <- read.csv("results/youdrawit-eyefitting-example-simulated.csv") %>%
+#     mutate(`Parameter Choice` = "F")
+#     
+# trial.feedback %>%
+#   ggplot(aes(x = x)) +
+#   geom_point(data = trial.sim, aes(y = y), alpha = 0.7) +
+#   geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
+#   geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
+#   geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
+#   facet_wrap(~`Parameter Choice`) +
+#   theme_bw(base_size = 14) +
+#   theme(aspect.ratio = 1,
+#         legend.position = "bottom",
+#         axis.text    = element_text(size = 12),
+#         axis.title   = element_text(size = 12),
+#         legend.title = element_text(size = 12),
+#         legend.text  = element_text(size = 12),
+#         # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+#         # strip.background = element_rect(size = 0.8),
+#         legend.key.size = unit(1, "line")
+#         ) +
+#   scale_x_continuous(limits = c(0,20)) +
+#   scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
+#   scale_color_manual("", values = c("black", "steelblue", "orange")) +
+#   scale_linetype_manual("", values = c("dashed", "solid", "solid"))
 
-trial.feedback %>%
+eyefitting_model_data %>%
+  filter(participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>%
+  rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_point(data = trial.sim, aes(y = y), alpha = 0.7) +
-  geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
-  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
+  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
+  geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
   geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
-  facet_wrap(~`Parameter Choice`) +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "bottom",
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
-        # strip.background = element_rect(size = 0.8),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_x_continuous(limits = c(0,20)) +
-  scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
-  scale_color_manual("", values = c("black", "steelblue", "orange")) +
+  geom_point(data = eyefitting_simulated_data %>%
+               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F")  %>% rename(`Parameter Choice` = parm_id),
+             aes(x = x, y = y)) +
+  facet_wrap(~ `Parameter Choice`, labeller = label_both) +
+  theme_bw() +
+  theme(aspect.ratio = 1) +
+  scale_y_continuous("y") +
+  scale_color_manual("", values = c("black", "steelblue", "orange2")) +
   scale_linetype_manual("", values = c("dashed", "solid", "solid"))
@@
 
-Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PCA residuals, constraining the fit to a linear trend. 
+Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, constraining the fit to a linear trend. 
 Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect included to account for variation due to participant.
-The LMM equation for each fit (OLS and PCA) is given by:
+The LMM equation for each fit (OLS and PA) is given by:
 \begin{equation}
@@
 + $y_{ijk,drawn}$ is the drawn $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value
-+ $\hat y_{ijk,fit}$ is the fitted $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted $y$ values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
++ $\hat y_{ijk,fit}$ is the fitted $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted $y$ values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PA fit
 + $\gamma_0$ is the overall intercept
@@
 
-Constraining the residual trend to a linear fit, \cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PCA regression line.
+Constraining the residual trend to a linear fit, \cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PA regression line.
 Estimated residual trend lines are overlaid on the observed individual participant residuals. 
-Results indicate the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
+Results indicate the estimated trends of PA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
 In particular, this trend is more prominent in parameter choices with large variances (F and N).
-These results are consistent to those found in @mosteller1981eye indicating participants fit a trend-line closer to the estimated regression line with a slope based on the first principal axis than the estimated OLS regression line thus, providing support for "ensemble perception".
+These results are consistent to those found in @mosteller1981eye indicating participants fit a trend-line closer to the estimated regression line with a slope based on the first principal axis than the estimated OLS regression line.
 
-```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
+```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
 Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
-eyefitting.preds.lmer <- read.csv("results/youdrawit-eyefitting-lmerpred-data.csv")
+eyefitting.preds.lmer <- read.csv("analysis/youdrawit-eyefitting-lmerpred-data.csv")
+
 # Plot Predictions
@@
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+            aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PCA"), alpha = 0.2) +
+  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
   geom_line(aes(y = emmean.ols, color = "OLS")) +
-  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
+  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PA"), color = NA, alpha = 0.4) +
   geom_line(aes(y = emmean.pca, color = "PCA")) +
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  facet_wrap(~`Parameter Choice`, ncol = 4, labeller = label_both) +
   theme_bw(base_size = 14) +
@@
 
-Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PCA residuals to allow for estimation of smoothing splines.
+Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
 A random participant effect was included to account for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
-The GAMM equation for each fit (OLS and PCA) residuals is given by:
+Defining e_{ijk,fit} the same as in equation (2) above, the GAMM equation for each fit (OLS and PA) residuals is given by:
 \begin{equation}
-y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
 \end{equation}
@@
 
-+ $y_{ijk,drawn}$ is the drawn $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value
-+ $\hat y_{ijk,fit}$ is the fitted $y$ value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
-+ $e_{ijk,fit}$ is the residual between the drawn and fitted $y$ values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$ value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the same as equation
 + $\alpha_i$ is the intercept for the parameter choice $i$
@@
 
-Allowing for flexibility in the residual trend, \cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PCA regression line.
+Allowing for flexibility in the residual trend, \cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS regression line and PA regression line.
 Estimated residual trends were overlaid on the observed individual participant residuals. 
-The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue) for scatterplots with more noise (F and N).
+The results of the GAMM align with those shown in \cref{fig:eyefitting-lmer-residualplots} providing support that estimated trends of PA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue) for scatterplots with more noise (F and N).
 By fitting smoothing splines, we can determine whether participants naturally fit a straight trend-line to the set of points or whether they deviate throughout the domain.
 In particular, in scatterplots with smaller variance (S and V), we can see that participants began at approximately the correct starting point then deviated away from the fitted regression lines and corrected for their fit toward the end of their trend-line.
-In scatterplots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PCA regression line.
+In scatterplots with larger variance (F and N), participants estimated their starting value in the extreme direction of the OLS regression line based on the increasing or decreasing trend but more accurately represented the starting value of the PA regression line.
 As participants continued their trend-line, they crossed through the OLS regression line indicating they estimated the slope in the extreme direction.
@@
 
-```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
+```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
 Estimated residual trends with 95\\% confidence bands are overlaid on the observed individual participant residuals."}
-eyefitting.grid.gamm <- read.csv("results/youdrawit-eyefitting-gammpred-data.csv")
+eyefitting.grid.gamm <- read_csv("analysis/youdrawit-eyefitting-gammpred-data.csv")
+# Plot Predictions
 eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
   geom_line(aes(y = ols.pred, color = "OLS")) +
-  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
-  geom_line(aes(y = pca.pred, color = "PCA")) +
+  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = pca.pred, color = "PA")) +
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, ncol = 4) +
+  facet_wrap(~`Parameter Choice`, ncol = 4, labeller = label_both) +
   theme_bw(base_size = 14) +
@@
 Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
-Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principle component regression as opposed to ordinary least squares regression.
-Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principle component regression.
+Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principal axis regression as opposed to ordinary least squares regression.
+Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principal axis regression.
 We utilized modern technology to replicate a study conducted 40 years ago, and strengthened the original results with current analysis methods which allow for more flexibility and sophistication.
 Our results indicate that participants minimized the distance from their drawn regression line over both the $x$ and $y$ axis simultaneously. 
-These results provide support that humans perform "ensemble perception" in a statistical graphic setting.
+<!-- These results provide support that humans perform "ensemble perception" in a statistical graphic setting. -->
 We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.

### 2022-08-02 analysis/you-draw-it-eyefitting-analysis.Rmd
working on revisions

@@
 
+<!-- ## GET DATA FROM DATABASE -->
+
+<!-- ```{r db} -->
+<!-- library(RSQLite) -->
+<!-- library(DBI) -->
+<!-- library(digest) -->
+<!-- library(lubridate) -->
+
+<!-- last_date <- as.numeric(as_datetime("2021-05-18 18:00:00 CDT")) -->
+
+<!-- # Connect to database -->
+<!-- filename <- "you_draw_it_data.db" -->
+<!-- sqlite.driver <- dbDriver("SQLite") -->
+<!-- db_con <- dbConnect(sqlite.driver, dbname = filename) -->
+<!-- # dbListTables(db_con) -->
+
+<!-- # Import eyefitting parameter details -->
+<!-- eyefitting_parameter_details <- dbReadTable(db_con,"eyefitting_parameter_details") %>% -->
+<!--   mutate(parm_id = as.factor(as.character(parm_id))) -->
+
+<!-- # Import user data -->
+<!-- users <- dbReadTable(db_con,"users") %>% -->
+<!--   filter(study_starttime > 1620152231, study_starttime < last_date) %>% -->
+<!--   select(-ip_address) %>% -->
+<!--   rowwise() %>% -->
+<!--   mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) -->
+
+<!-- # Import exponential prediction feedback data -->
+<!-- eyefitting_feedback_data <- dbReadTable(db_con,"feedback") %>% -->
+<!--   filter(study_starttime > 1620152231, study_starttime < last_date) %>% -->
+<!--   filter(parm_id %in% c("S", "F", "V", "N")) %>% -->
+<!--   rename(yols = y) %>% -->
+<!--   rowwise() %>% -->
+<!--   mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>% -->
+<!--   left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>% -->
+<!--   filter(recruitment != "I am the researcher") %>% -->
+<!--   rowwise() %>% -->
+<!--   mutate(plot_id = as.factor(digest(paste(nick_name, start_time, end_time)))) %>% -->
+<!--   select(study_starttime, participant_id, nick_name, plot_id, start_time, end_time, parm_id, x, yols, ydrawn, age, gender, academic_study, recruitment) -->
+
+<!-- factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study", -->
+<!--                 "recruitment", "participant_id", "parm_id", "plot_id") -->
+<!-- eyefitting_feedback_data[,factorCols] <- lapply(eyefitting_feedback_data[,factorCols], factor) -->
+<!-- # write.csv(eyefitting_feedback_data, file = "eyefitting-feedback-data.csv", row.names = F, na = "") -->
+
+<!-- # Import exponential prediction simulated data -->
+<!-- eyefitting_simulated_data <- dbReadTable(db_con,"simulated_data") %>% -->
+<!--   filter(study_starttime > 1620152231, study_starttime < last_date) %>% -->
+<!--   filter(parm_id %in% c("S", "F", "V", "N")) %>% -->
+<!--   rowwise() %>% -->
+<!--   mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>% -->
+<!--   left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>% -->
+<!--   filter(recruitment != "I am the researcher") -->
+
+<!-- factorCols <- c("nick_name", "dataset", "study_starttime", "age", "gender", "academic_study", -->
+<!--                 "recruitment", "participant_id", "parm_id") -->
+<!-- eyefitting_simulated_data[,factorCols] <- lapply(eyefitting_simulated_data[,factorCols], factor) -->
+<!-- # write.csv(eyefitting_simulated_data, file = "eyefitting-simulated-data.csv", row.names = F, na = "") -->
+
+<!-- # Disconnect from database -->
+<!-- dbDisconnect(db_con) -->
+<!-- length(unique(eyefitting_feedback_data$participant_id)) -->
+<!-- length(unique(eyefitting_feedback_data$plot_id)) -->
+<!-- ``` -->
+
 ## DATA SIMULATION PROCEDURE
@@
 
-eyefitting_parameter_details <- read_csv("eyefitting_parameter_details.csv")
+eyefitting_parameter_details <- read_csv("eyefitting-parameter-details.csv")
 eyefitting_parameter_details %>% knitr::kable()
@@
 
-# write.csv(eyefitting_sim_data, "eyefitting-simulated-data.csv", row.names = F, na = "")
+write.csv(eyefitting_sim_data, "eyefitting-simulated-data.csv", row.names = F, na = "")
 
@@
 
-There are a total of `length(unique(eyefitting_model_data$participant_id))` participants and `length(unique(eyefitting_model_data$participant_id))` 'You Draw It' plots completed.
+There are a total of `r length(unique(eyefitting_model_data$participant_id))` participants and `r length(unique(eyefitting_model_data$participant_id))` 'You Draw It' plots completed.
 
@@
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, ncol = 1) +
+  facet_wrap(~`Parameter Choice`, ncol = 4) +
   theme_bw(base_size = 14) +

### 2022-08-03 manuscript-revisions/response-to-reviewers.qmd
editing analysis

@@
   + \response{Added single quotes to 'You Draw It'.}
-+ \todo{p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention the supplementary materials in 1 or 2 sentences)}
-+ \todo{p.7, l.34: you had 35 participants and got 119 plots; later (p.10, l.38) you state that each particpant had to evaluate 4 plots; provide this information here immediately; can you summarize why there are 21 plots missing, e.g., some plots were incomplete, some participants were dropping out after just 1 or 2 plots, etc.}
++ p.7, top: be specific and indicate in which of the sections of your paper each of these points is being addressed, in other words, provide a specific overview what can be found in Sections 2 - 5 of your paper (and also mention the supplementary materials in 1 or 2 sentences)
+  + \response{Added information about specific parts contained in each of the paper sections.}
++ p.7, l.34: you had 35 participants and got 119 plots; later (p.10, l.38) you state that each participant had to evaluate 4 plots; provide this information here immediately; can you summarize why there are 21 plots missing, e.g., some plots were incomplete, some participants were dropping out after just 1 or 2 plots, etc.
+  + \response{After double checking the data, it was actually 131 plots - we are working with pilot data and thus had this number from another set.
+We clarified and made this number now reproducible by using r-code in the rmd template file rather than mannually entering it.
+We also discussed the missing plots by adding: "Data were collected as a part of a pilot study meant to test the applet; therefore, either voluntary participant dropout or disconnection from a server not designed to accommodate large magnitudes of participants resulted in missing plots in our data set for analysis."}
 + p.7, l.51 & p.17, l.27 & l.32: "here" is meaningless for printed materials; list the full URLs (but still make them clickable)
@@
 When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
-In turn, the actual y-intercept $(beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
+In turn, the actual y-intercept $(\beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
 We changed the equation notation to follow the point-slope equation rather than changing notation.}
-+ \dubcheck{p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?}
++ p.9, l.25-37: can you further describe these four models with some more details and extend the table, e.g., list the domains in the table as well (and not only in the text); also indicate how many points were used in each scatterplot (e.g., was this random or fixed?; plot V in Fig 2 seems to have much fewer points than the other three); do the 4 letters have a certain meaning?
   + \response{We double checked the data simulation process based on the shiny applet code.
@@
 (1981).
-We clarified this on p8 l25 with "Model equation parameters, $beta_0$ and $beta_1$, and parameter choice letter names (S, F, V, N) and V), were selected to reflect the four data sets used and labeled in Mosteller et al..." See analysis in supplementary material for code on the data simulation procedure.}
+We clarified this on p8 l25 with "Model equation parameters, $beta_1$, $y_{\bar{x}}$, and parameter choice letter names (F, S, V, N) and V), were selected to reflect the four data sets used and labeled in Mosteller et al..." See analysis in supplementary material for code on the data simulation procedure.}
 + p.10, l.38-39: I suppose each participant got one plot of type S, F, V, N ? - say so
-  + \response{We clarified that each participant got one of each of the plot types.}
+  + \response{We clarified that each participant got one of each of the plot types in Section 2.4.}
 + p.10-p.11: use PCA or PC and then use these consistently; if "principal axis" is the full term, I would even use PA as abbreviation as PCA is a widely used abbreviation for principal component analysis and could confuse readers here; if you switch to PA, also adjust the main text below
@@
   + \response{Changed to "increment" for both to be consistent and added specification of k in l.48.}
-+ \todo{p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations}
++ \dubcheck{p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations}
+  + \response{We clarified what we meant by "constraining the fit to a linear trend" by adding explantion about participants being allowed to freely draw: "The 'You Draw It' method does not restrict participants to draw a straight line as other methods would, such as using a ruler.
+Instead, participants are allowed to freely draw a line with potential curvature.
+Using the `lmer` function in the lme4 package, a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend."}
 + p.13, l.26: keep the same order of F, S, V, N as before, i.e., S F V N from Table 1 to avoid confusing the reader
@@
   + \response{Clarified similarities between the linear mixed model equation and the generalized additive mixed model equation by stating "Defining $e_{ijk,fit}$ the same as in equation (2) above, the GAMM equation...".}
-+ \todo{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task.
++ \dubcheck{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task.
 Drawing a straight line is not as easy as it sounds.
@@
   + https://doi.org/10.1145/2820619.2820633
-+ \todo{p.17, Future work:} 
+  + \response{These are some really great suggestions that lead to some interesting and fun reads related to this work.
+We reviewed the suggested readings and followed some references to other related articles.
+A few sentences have been added to the discussion section to address some of the cognitive/human movement sciences and influcences on our study.}
++ p.17, Future work:
   + I worked through the 12 plots at https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/
@@
   + Neither of these has to be implemented in the current study, but listing both options as natural extensions for the straight line assessment seems to be worthwhile.
-+ \todo{p.17-21, References:}
+  + \response{These would be some great extensions to validating 'You Draw It' as a method for graphical testing.
+We added these suggestions as well as suggested a study which compares the indirect interaction (mouse and pointer on the screen) and direct interaction (digital pen on a tablet) method of using 'You Draw It'.}
++ \dubcheck{p.17-21, References:}
   + capitalize journal names, e.g., Psychological bulletin & IEEE transactions on visualization and computer graphics
@@
   + check capitalization of nouns; if I am not mistaken, JCGS uses upper case letters for all nouns [some of these apply to other references as well]
+  + \response{Reviewed references and adjusted bibtex file and style.}
 
@@
 When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
-In turn, the actual y-intercept $(beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
+In turn, the actual y-intercept $(\beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
 We changed the equation notation to follow the point-slope equation rather than changing notation.}
@@
 1.
-In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis.
+\dubcheck{In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis.
 In particular, how does your work differ from Ciccione and Dehaene (2021)?(Ciccione, L.
@@
 Can humans perform mental regression on a graph?
-Accuracy and bias in theperception of scatterplots.
-Cognitive Psychology, 128, p.101406.)
-2. In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe
-the method of estimation. In other words, what are the R functions doing to estimate the model?
+Accuracy and bias in the perception of scatterplots.
+Cognitive Psychology, 128, p.101406.)}
+  + \response{We clarifed the differences between our work and Ciccione (2021) as our study use a more flexible method of testing graphics and introduces a way to analyze the data using GAMM's.
+While this study was designed to replicate results found in previous studies as a way to validate our method, our intent is to set the foundation for using 'You Draw It' for testing statistical graphics as the tool extends nicely to the non-linear setting.
+We added this clarification into the end of Section 1.2.}
+2.
+\todo{In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe the method of estimation.
+In other words, what are the R functions doing to estimate the model?}
+  + \response{We added the method of estimations based on the R documentation for the lmer and bam functions.}
 3. Could you clarify if the participants were known to be experts in statistics or related fields?
+  + \response{We added a sentence describing that the data is from a pilot study and thus participants from the recruitment methods likely have an interest in advancing research in data visualization: "As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely have an interest in fields related to statistics and want to help advance research in graphics."}
 4. Throughout the manuscript, references to “principle components” should be “principal components.”

### 2022-08-03 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
knitted manuscript revisions

@@
 knitr::opts_chunk$set(
-  echo = F, eval = T, messages = F, warnings = F,
+  echo = F, eval = T, message = F, warning = F,
   fig.width = 6, fig.height = 4,  fig.align = 'center',
@@
 library(here)
+```
+
+```{r eyefitting-data}
+eyefitting_simulated_data <- read_csv("analysis/eyefitting-simulated-data.csv")
+
+eyefitting_model_data <- read.csv("data/eyefitting-model-data.csv")
+
+factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
+                "recruitment", "participant_id", "parm_id", "plot_id")
+eyefitting_model_data[,factorCols] <- lapply(eyefitting_model_data[,factorCols], factor)
 ```
@@
 
-In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing.
+In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a new tool for graphical testing. 
 Our visual system is naturally built to look for structure and identify patterns. 
@@
 Our research is intended to implement the 'You Draw It' feature as a way to measure the patterns we see in data.
+The graphical testing method used in this study differs from prior methods found in @mosteller1981eye and @ciccione2021can by allowing participants to freely draw estimated trend lines - a method which extends nicely to a nonlinear setting.
 We validate the 'You Draw It' method by replicating the less technological study conducted by @mosteller1981eye.
+In [Section 2](#methods) we describe our participant sample, the graphical task to be completed, and the data generation process and study design.
+[Section 3](#results) describes the participant data collected and shares results from the analyses of the data using mixed models.
+Overall conclusions and discussion of results are presented in [Section 4](#conclusion-discussion) with extensions to the current work suggested in [Section 5](#future-work). 
+The RShiny applet to complete the study, participant data used for analysis, and code to replicate the analysis can be found in the [Supplementary Material](#supplementary-material).
 Based on previous research, we hypothesize that visual regression tends to mimic regression based on the principal axis rather than an ordinary least squares regression.
-In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.
+In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models (GAMM).
+While the focus of this paper is to share the results from the validation study which uses the new 'You Draw It' method to evaluate visually fitted linear trends to statistical regression results, the intent of this work is to set the foundation and demonstrate the strength of the combination of the 'You Draw It' method and GAMM analyses for testing statistical graphics and to extend the use of the method beyond the linear setting.
 
-# Methods
+# Methods {#methods}
 
@@
 Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
-A total of 35 individuals completed 119 unique 'You Draw It' task plots; data collected from all completed 'You Draw It' task plots were included in the analysis.
+A total of 35 individuals completed `r length(unique(eyefitting_model_data$plot_id))` unique 'You Draw It' task plots.
+Data were collected as a part of a pilot study meant to test the applet; therefore, either voluntary participant dropout or disconnection from a server not designed to accommodate large magnitudes of participants resulted in missing plots in our data set for analysis.
 All participants had normal or corrected to normal vision and signed an informed consent form.
 The experimental tasks took approximately 15 minutes to complete.
+As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely have an interest in fields related to statistics and want to help advance research in graphics.
 While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
@@
 
-eyefitting_sim_data <- read_csv("analysis/eyefitting-simulated-data.csv")
+eyefitting_sim_data <- read_csv("analysis/eyefitting-simulated-data-example.csv")
 
@@
 
-# Results
+# Results {#results}
 
@@
 slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
-                       type=c("PC", "OLS"))
+                       type=c("PA", "OLS"))
 
@@
 
-segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PC"),
+segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PA"),
                   data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="OLS"))
@@
 
-```{r eyefitting-plots}
-eyefitting_simulated_data <- read_csv("analysis/eyefitting-simulated-data.csv")
-
-eyefitting_model_data <- read.csv("analysis/eyefitting-model-data.csv")
-
-factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
-                "recruitment", "participant_id", "parm_id", "plot_id")
-eyefitting_model_data[,factorCols] <- lapply(eyefitting_model_data[,factorCols], factor)
-```
-
 For each participant, the final data set used for analysis contained $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PA}$ for parameter choice $i = 1,2,3,4$, j = $1,...,N_{participant}$, and $x_{ijk}$ value for increment $k = 1, ...,4 x_{max} + 1$.
@@
 Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).", out.width="50%"}
-# trial.feedback <- read.csv("results/youdrawit-eyefitting-example-feedback.csv") %>%
-#     mutate(`Parameter Choice` = "F")
-# trial.sim <- read.csv("results/youdrawit-eyefitting-example-simulated.csv") %>%
-#     mutate(`Parameter Choice` = "F")
-#     
-# trial.feedback %>%
-#   ggplot(aes(x = x)) +
-#   geom_point(data = trial.sim, aes(y = y), alpha = 0.7) +
-#   geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
-#   geom_line(aes(y = ypca, color = "PA", linetype = "PA")) +
-#   geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
-#   facet_wrap(~`Parameter Choice`) +
-#   theme_bw(base_size = 14) +
-#   theme(aspect.ratio = 1,
-#         legend.position = "bottom",
-#         axis.text    = element_text(size = 12),
-#         axis.title   = element_text(size = 12),
-#         legend.title = element_text(size = 12),
-#         legend.text  = element_text(size = 12),
-#         # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
-#         # strip.background = element_rect(size = 0.8),
-#         legend.key.size = unit(1, "line")
-#         ) +
-#   scale_x_continuous(limits = c(0,20)) +
-#   scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
-#   scale_color_manual("", values = c("black", "steelblue", "orange")) +
-#   scale_linetype_manual("", values = c("dashed", "solid", "solid"))
-
 eyefitting_model_data %>%
   filter(participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>%
-  rename(`Parameter Choice` = parm_id) %>%
+  dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
@@
   geom_point(data = eyefitting_simulated_data %>%
-               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F")  %>% rename(`Parameter Choice` = parm_id),
+               filter(dataset == "point_data", participant_id == "60f510d6896cddc508ea6d3de4b58234", parm_id == "F") %>% dplyr::rename(`Parameter Choice` = parm_id),
              aes(x = x, y = y)) +
@@
 ### Linear Trend Constraint
+<!-- https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf -->
 
-Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, constraining the fit to a linear trend. 
+The 'You Draw It' method does not restrict participants to draw a straight line as other methods would, such as using a ruler.
+Instead, participants are allowed to freely draw a line with potential curvature.
+Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend.
+Both fixed and random parameter estimates in the LMM were determined by optimizing the restricted maximum likelihood (REML) through penalized least squares.
 Parameter choice, $x$, and the interaction between $x$ and the parameter choice were treated as fixed effects with a random participant effect included to account for variation due to participant.
@@
 Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
+The `bam` function is used to fit GAMM's to very large data sets and use lower memory than the `gam` function.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
@@
 
-# Discussion and Conclusion
+# Discussion and Conclusion {#conclusion-discussion}
 
@@
 We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.
+Researchers in cognitive and human movement sciences have found that human arm movement is a complex task [@miall1995curvature; @rousset2015study].
+The 'You Draw It' method described in this paper uses *indirect interaction* in which the mouse position and resulting visual line on the screen are dissociated.
+Therefore, curvature found in participant drawn lines from a straight lines could potentially be explained by the lack of coordination which results from the eye-hand dissociation from indirect drawing and the distortion of visual perception affecting the curvature of movements.
+Additionally, there is a training effect related to the completion of the 'You Draw It' task - the movement of the line must be slow so that the visual representation on the screen can accurately capture each movement.
+@de1991misdirections conducted a study in which participants moved their hand slowly from an initial position in front of them to a visual target (movement task); they were then asked to repeat the task using different sizes of pointers (perceptual task).
+Their results indicated that deviations from the shortest pointers were comparable to those of the movement task, but that bias increased as the length of the pointer increased.
+While we suggested participants use a mouse to complete the study, we could not require the use; therefore, some participants may have used a track-pad and results may have been influenced by the pressure placed on their track-pad [@easton1978finger].
 
-# Future Work
+# Future Work {#future-work}
 
 This study provided a basis for the use of 'You Draw It' as a tool for testing statistical graphics and introduced a method for statistically modeling participant drawn lines using generalized additive mixed models.
-Further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication.
+Additional studies related to the validation and use of the tool would be useful for providing insight into explanations of biases introduced by the task such as the deviation from a straight line.
+For instance, a variation on the current study could compare manual adjustment methods such as shifting and rotating a horizontal line segment until the fit is suitable to the 'You Draw It' method on the same set of data.
+This might explain the large deviation from the participant drawn line as $x$ approaches 20.
+Another useful extension study would be to compare the 'You Draw It' method as conducted by direct interaction - using a digital pen on a tablet - to indirect interaction - using a computer mouse to relate to a pointer on the screen.
+Further extensions to this work might ask participants to draw a trend-line through scatterplots with one (or multiple) extreme outliers in order to evaluate the perceptual system's resistance to outliers.
+
+While the focus of this study was on drawing linear trend-lines, further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication - a strength of the combination of the flexible 'You Draw It' method and GAMM analysis method.
 This tool could also be used to evaluate human ability to extrapolate data from trends. 
@@
 
+{#supplementary-material}
+
++ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
++ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).
 + **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
 + **Data Analysis Code:** The code used to replicate the analysis in this paper can be found at [earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-+ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).

### 2022-08-03 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
knitted manuscript revisions

@@
 result. For instance, we may decide to highlight the relationship
-between two variables in a scatter-plot by including a trend line, or
+between two variables in a scatterplot by including a trend line, or
 adding color to highlight clustering \citep{vanderplas2017clusters}.
@@
 assessing the position of points along an axis is easier than
-determining the slope of a line. Other studies focus on the viewers
+determining the slope of a line. Other studies focused on the viewers'
 ability to perceive the strength of the relationship between \(x\) and
-\(y\) coordinates in a scatter-plot. For instance, when the data appear
+\(y\) coordinates in a scatterplot. For instance, when the data appear
 dense, viewers tend to overestimate the magnitude of the correlation
 coefficient \citep{cleveland1982variables, lauer1989density}.
-\citet{cleveland1993visualizing} provides an argument for displaying
+\citet{cleveland1993visualizing} provided an argument for displaying
 cyclical patterns with an aspect ratio which sets the curve close to
-45\(^{\circ}\). \citet{kosslyn2006graph} examines how Gestalt principles
+45\(^{\circ}\). \citet{kosslyn2006graph} examined how Gestalt principles
 of perceptual organization are instrumental in extracting data from a
-chart. For example, data points that are closer in space are more likely
-to be perceived as grouped \citep{ciccione2020grouping}, and vertical
-and horizontal lines are easier to discriminate than oblique ones
-\citep{appelle1972perception}. The results of these cognitive tasks
-provide some consistent guidance for chart design; however, other
-methods of visual testing can further evaluate design choices and help
-us understand cognitive biases related to the evaluation of statistical
-charts.
+chart. For example, \citet{ciccione2020grouping} conducted a study to
+support data points located closer together are more likely to be
+perceived as the same group and \citet{appelle1972perception} found that
+it is easier to discriminate vertical and horizontal lines than oblique
+lines. The results of these cognitive tasks provided some consistent
+guidance for chart design; however, other methods of visual testing can
+further evaluate design choices and help us understand cognitive biases
+related to the evaluation of statistical charts.
 
@@
 perception, as in \citet{vanderplas2015signs}, which used the method of
-adjustment to estimate the magnitude of the impact of the sine illusion.
-However, there are more modern testing methods that have been developed
-since the heyday of psychophysics.
+adjustment - a technique which requires participants to alter a changing
+stimulus to match a given constant stimuli \citep{gescheider_1997}- to
+estimate the magnitude of the impact of the sine illusion. However,
+there are more modern testing methods that have been developed since the
+heyday of psychophysics.
 
@@
 testing methods is Wilkinson's Grammar of Graphics
-\citep{wilkinson2013grammar}. The grammar of graphics serves as the
+\citep{wilkinson2013grammar}. The Grammar of Graphics serves as the
 fundamental framework for data visualization with the notion that
@@
 \citep{wickham2011ggplot2}, aims to implement the framework of creating
-charts and graphics as the grammar of graphics recommends.
+charts and graphics as the Grammar of Graphics recommends.
 
-Combining the grammar of graphics with another tool for statistical
+Combining the Grammar of Graphics with another tool for statistical
 graphics testing, the statistical lineup, yields a method for evaluating
 graphical design choices. \citet{buja2009statistical} introduced the
-lineup protocol to provide a framework for inferential testing. Through
-experimentation, methods such as the lineup protocol allow researchers
-to conduct studies geared at understanding human ability to conduct
-tasks related to the perception of statistical charts such as
-differentiation, prediction, estimation, and extrapolation
+lineup protocol to provide a framework for inferential testing. A
+statistical lineup is a plot consisting of smaller panels where the
+viewer is asked to identify the target panel containing the real data
+from among a set of decoy null plots which display data under the
+assumption there is no relationship. If the viewer can identify the
+target panel randomly embedded within the set of null panels, this
+suggests that the real data is visually distinct from data generated
+under the null model. Through experimentation, methods such as the
+lineup protocol allow researchers to conduct studies geared at
+understanding human ability to conduct tasks related to the perception
+of statistical charts such as differentiation, prediction, estimation,
+and extrapolation
 \citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
@@
 eye through a set of points
-\citep{finney1951subjective, mosteller1981eye}. Common methods of
-fitting trends by eye involved maneuvering a string, black thread, or
-ruler until the fit is suitable, then drawing the line through the set
-of points. Recently, \citet{ciccione2021can} conducted a comprehensive
-set of studies investigating human ability to detect trends in graphical
-representations using physical adjustment and manipulation methods.
+\citep{unwin1988eyeballing, finney1951subjective, mosteller1981eye}.
+Common methods of fitting trends by eye involved maneuvering a string,
+black thread, or ruler until the fit is suitable, then drawing the line
+through the set of points. Recently, \citet{ciccione2021can} conducted a
+comprehensive set of studies investigating human ability to detect
+trends in graphical representations using physical adjustment and
+manipulation methods.
 
@@
 of fitting lines by eye. Participants were asked to fit lines by eye to
-four scatter-plots using an 8.5 x 11 inch transparency with a straight
+four scatterplots using an 8.5 x 11 inch transparency with a straight
 line etched completely across the middle. A latin square design with
@@
 presentation. It was found that order of presentation had no effect and
-that participants tended to fit the slope of the principal axis (error
-minimized orthogonally, both horizontal and vertical, to the regression
-line) over the slope of the least squares regression line (error
-minimized vertically to the regression line). These results support
-previous research on ``ensemble perception'' indicating the visual
-system can compute averages of various features in parallel across the
-items in a set
-\citep{chong2003representation, chong2005statistical, van2011rapid}.
+that participants tended to fit the slope of the principal axis (PA)
+(error minimized orthogonally, both horizontal and vertical, to the
+regression line) over the slope of the least squares regression line
+(error minimized vertically to the regression line).
 
@@
 which minimizes the Euclidean distance of points from the line, when
-fitting a line to a noisy scatter-plot.
+fitting a line to a noisy scatterplot.
 
 While not explicitly intended for perceptual testing, in 2015, the New
-York Times introduced an interactive feature, called You Draw It
+York Times introduced an interactive feature, called `You Draw It'
 \citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
@@
 In this paper, we establish `You Draw It', adapted from the New York
-Times feature, as a tool for graphical testing. Our visual system is
+Times feature, as a new tool for graphical testing. Our visual system is
 naturally built to look for structure and identify patterns. For
@@
 intended to implement the `You Draw It' feature as a way to measure the
-patterns we see in data. We validate the `You Draw It' method by
-replicating the less technological study conducted by
-\citet{mosteller1981eye}. Based on previous research surrounding
-``ensemble perception,'' we hypothesize that visual regression tends to
-mimic principle component regression rather than an ordinary least
-squares regression. In order to assess this hypothesis, we introduce a
-method for statistically modeling the participant drawn lines using
-generalized additive mixed models.
+patterns we see in data. The graphical testing method used in this study
+differs from prior methods found in \citet{mosteller1981eye} and
+\citet{ciccione2021can} by allowing participants to freely draw
+estimated trend lines - a method which extends nicely to a nonlinear
+setting. We validate the `You Draw It' method by replicating the less
+technological study conducted by \citet{mosteller1981eye}. In
+\protect\hyperlink{methods}{Section 2} we describe our participant
+sample, the graphical task to be completed, and the data generation
+process and study design. \protect\hyperlink{results}{Section 3}
+describes the participant data collected and shares results from the
+analyses of the data using mixed models. Overall conclusions and
+discussion of results are presented in
+\protect\hyperlink{conclusion-discussion}{Section 4} with extensions to
+the current work suggested in \protect\hyperlink{future-work}{Section
+5}. The RShiny applet to complete the study, participant data used for
+analysis, and code to replicate the analysis can be found in the
+\protect\hyperlink{supplementary-material}{Supplementary Material}.
+Based on previous research, we hypothesize that visual regression tends
+to mimic regression based on the principal axis rather than an ordinary
+least squares regression. In order to assess this hypothesis, we
+introduce a method for statistically modeling the participant drawn
+lines using generalized additive mixed models (GAMM). While the focus of
+this paper is to share the results from the validation study which uses
+the new `You Draw It' method to evaluate visually fitted linear trends
+to statistical regression results, the intent of this work is to set the
+foundation and demonstrate the strength of the combination of the `You
+Draw It' method and GAMM analyses for testing statistical graphics and
+to extend the use of the method beyond the linear setting.
 
@@
 Participants were recruited through through Twitter, Reddit, and direct
-email in May 2021. A total of 35 individuals completed 119 unique `You
-Draw It' task plots; data collected from all completed `You Draw It'
-task plots were included in the analysis. All participants had normal or
-corrected to normal vision and signed an informed consent form. The
-experimental tasks took approximately 15 minutes to complete. While this
-study does utilize a convenience sample, as this is primarily a
-perceptual task, previous results have found few differences between
+email in May 2021. A total of 35 individuals completed 131 unique `You
+Draw It' task plots. Data were collected as a part of a pilot study
+meant to test the applet; therefore, either voluntary participant
+dropout or disconnection from a server not designed to accommodate large
+magnitudes of participants resulted in missing plots in our data set for
+analysis. All participants had normal or corrected to normal vision and
+signed an informed consent form. The experimental tasks took
+approximately 15 minutes to complete. As this is a pilot study,
+participants from Twitter and Reddit pages related to data visualization
+voluntarily completed the study and likely have an interest in fields
+related to statistics and want to help advance research in graphics.
+While this study does utilize a convenience sample, as this is primarily
+a perceptual task, previous results have found few differences between
 expert and non-expert participants in this context
@@
 The experiment was conducted and distributed through a Shiny application
-\citep{shinyPkg} found
-\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
+\citep{shinyPkg} found at
+\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.
 
@@
 
-In the study, participants were shown an interactive scatter-plot
+In the study, participants were shown an interactive scatterplot
 (\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
@@
 visual. In order to allow for user interaction and data collection, we
-integrated the D3 visual into Shiny using the \texttt{r2d3} package.
-While the interface is highly customized to this particular task, we
-hope to generalize the code and provide a Shiny widget in an R package
-soon.
+integrated the D3 visual into Shiny using the \texttt{r2d3} package
+\citep{r2d3}. While the interface is highly customized to this
+particular task, we hope to generalize the code and provide a Shiny
+widget in an R package soon.
 
@@
 
-All data processing was conducted in R statistical software. A total of
-\(N = 30\) points \((x_i, y_i), i = 1,...N\) were generated for
+All data processing was conducted in R software environment for
+statistical computing and graphics \citep{Rsoftware}. A total of
+\(N = 30\) points \((x_i, y_i), i = 1,...,N\) were generated for
 \(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
-relationship. Data were simulated based on a linear model with additive
-errors: \begin{align}
-y_i & = \beta_0 + \beta_1 x_i + e_i \\
+relationship. Data were simulated based on the point-slope form of a
+linear model with additive errors: \begin{align}
+y_i = \beta_1(x_i-\bar{x}) + y_{\bar{x}} + e_i \\
 \text{with } e_i & \sim N(0, \sigma^2). \nonumber
@@
 
-Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
-reflect the four data sets (F, N, S, and V) used in
-\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtained
-\(\beta_0\) from the point-slope equation of a line using the mean of
-the generated \(x\) values and the predefined \(y\) value at \(\bar x\),
-denoted \(y_{\bar x}\). Parameter choices F, N, and S simulated data
-across a domain of 0 to 20. Parameter choice F produced a trend with a
-positive slope and a large variance while N had a negative slope and a
-large variance. In comparison, S showed a trend with a positive slope
-and a small variance while V yielded a steep positive slope with a small
-variance over the domain of 4 to 16. \cref{fig:eyefitting-simplot}
-illustrates an example of simulated data for all four parameter choices
-intended to reflect the trends in \citet{mosteller1981eye}. Aesthetic
-design choices were made consistent across each of the interactive `You
-Draw It' task plots. The y-axis range extended 10\% beyond (above and
-below) the range of the simulated data points to allow for users to draw
-outside the simulated data set range and avoid anchoring their lines to
-the corners of the plot.
+Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\), and parameter
+choice letter names (F, S, V, N), were selected to reflect the four data
+sets used and labeled in \citet{mosteller1981eye}
+(\cref{tab:eyefitting-parameters}). The mean of the generated \(x\)
+values and the predefined \(y\) value at \(\bar x\), denoted
+\(y_{\bar x}\) were used in the point-slope equation of a line.
+Parameter choices S, F, and N simulated data across a domain of 0 to 20.
+Parameter choice F produced a trend with a positive slope and a large
+variance while N had a negative slope and a large variance. In
+comparison, S showed a trend with a positive slope and a small variance
+while V yielded a steep positive slope with a small variance over the
+domain of 4 to 16. \cref{fig:eyefitting-simplot} illustrates an example
+of simulated data for all four parameter choices intended to reflect the
+trends in \citet{mosteller1981eye}. Aesthetic design choices were made
+consistent across each of the interactive `You Draw It' task plots. The
+y-axis range extended 10\% beyond (above and below) the range of the
+simulated data points to allow for users to draw outside the simulated
+data set range and avoid anchoring their lines to the corners of the
+plot.
 
@@
 \centering
-\begin{tabular}[t]{cccc}
+\begin{tabular}[t]{ccccc}
 \toprule
-Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$\\
+Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$ & Domain\\
 \midrule
-S & 3.88 & 0.66 & 1.30\\
-F & 3.90 & 0.66 & 1.98\\
-V & 3.89 & 1.98 & 1.50\\
-N & 4.11 & -0.70 & 2.50\\
+F & 3.90 & 0.66 & 1.98 & (0,20)\\
+S & 3.88 & 0.66 & 1.30 & (0,20)\\
+V & 3.89 & 1.98 & 1.50 & (4,16)\\
+N & 4.11 & -0.70 & 2.50 & (0,20)\\
 \bottomrule
@@
 
-\caption{Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
+\caption{Example of simulated data points displayed in a scatterplot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
 \end{figure}
@@
 perception of log and linear scales; for simplicity, we focused on the
-study design and methods related to the current study. Each scatter-plot
-was the graphical representation of a data set that was generated
-randomly and independently for each participant at the start of the
-experiment. Participants in the study were shown two `You Draw It'
-practice plots in order to train participants in the skills associated
-with executing the task - in particular, the responsiveness of the
-applet requires that participants draw a line at a certain speed,
-ensuring that all of the evenly spaced points along the hand-drawn line
-are filled in. During the practice session, participants were provided
-with instruction prompts accompanied by a .gif and a practice plot.
-Instructions guided participants to start at the edge of the yellow box,
-to make sure the yellow shaded region was moving along with their mouse
-as they drew, and that they could draw over their already drawn line.
-Practice plots were then followed by four `You Draw It' task plots
-associated with the current study. The order of the task plots was
-randomly assigned for each individual in a completely randomized design.
+study design and methods related to the current study. Each data set was
+generated randomly and independently for each participant at the start
+of the experiment and mapped to a scatterplot. Participants in the study
+were shown two `You Draw It' practice plots in order to train
+participants in the skills associated with executing the task - in
+particular, the responsiveness of the applet requires that participants
+draw a line at a certain speed, ensuring that all of the evenly spaced
+points along the hand-drawn line are filled in. During the practice
+session, participants were provided with instruction prompts accompanied
+by a .gif and a practice plot. Instructions guided participants to start
+at the edge of the yellow box, to make sure the yellow shaded region was
+moving along with their mouse as they drew, and that they could draw
+over their already drawn line. Practice plots were then followed by one
+of each of the four `You Draw It' task plots associated with the current
+study (S, F, V, and N). The order of the task plots was randomly
+assigned for each individual in a completely randomized design.
 
@@
 determined by ordinary least squares (OLS) regression and regression
-based on the principal axis (PCA). \cref{fig:ols-vs-pca-example}
+based on the principal axis (PA). \cref{fig:ols-vs-pca-example}
 illustrates the difference between an OLS regression line which
 minimizes the vertical distance of points from the line and a regression
-line based on the principal axis which minimizes the Euclidean distance
-of points (orthogonal) from the line.
+line based on the PA which minimizes the Euclidean distance of points
+(orthogonal) from the line.
 
@@
 differ from the predetermined slope. Therefore, we fit an OLS regression
-to each scatter-plot to obtain estimated parameters
-\(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted values,
-\(\hat y_{k,OLS}\), were then obtained every 0.25 increments across the
-domain from the OLS regression equation,
-\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
-\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
-principal axis was determined by using the \texttt{princomp} function in
-the stats package in base R to obtain the rotation of the coordinate
-axes from the first principal component (direction which captures the
-most variance). The estimated slope, \(\hat\beta_{1,PCA}\), was
-determined by the ratio of the axis rotation in \(y\) and axis rotation
-in \(x\) of the first principal component with the y-intercept,
-\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
-using the mean of the simulated points, \((\bar x_i, \bar y_i)\). Fitted
-values, \(\hat y_{k,PCA}\), were then obtained every 0.25 increment
-across the domain from the PCA regression equation,
-\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
+to each scatterplot to obtain estimated parameters \(\hat\beta_{0,OLS}\)
+and \(\hat\beta_{1,OLS}\). Fitted values, \(\hat y_{k,OLS}\), were then
+obtained every 0.25 increment across the domain from the OLS regression
+equation,
+\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\), for
+\(k = 1, ..., 4 x_{max} +1\). The PA regression slope,
+\(\hat\beta_{1,PA}\), and y-intercept, \(\hat\beta_{0,PA}\), were
+determined using the \texttt{mcreg} function in the \texttt{mcr} package
+in R \citep{mcr_pkg} which implements Deming regression (equivalent to a
+regression based on the slope of the first principal axis). Fitted
+values, \(\hat y_{k,PA}\) were then obtained every 0.25 increment across
+the domain from the PA regression equation,
+\(\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k\), for
+\(k = 1, ..., 4 x_{max} +1\).
 
@@
 For each participant, the final data set used for analysis contained
-\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
-for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
+\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PA}\)
+for parameter choice \(i = 1,2,3,4\), j = \(1,...,N_{participant}\), and
 \(x_{ijk}\) value for increment \(k = 1, ...,4 x_{max} + 1\). Using both
@@
 of vertical residuals in relation to the OLS fitted values
-(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
-values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
+(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PA fitted
+values (\(e_{ijk,PA} = y_{ijk,drawn} - \hat y_{ijk,PA}\)) were made
 across the domain. \cref{fig:eyefitting-example-plot} displays an
@@
 \caption{Illustrates the data associated with and collected for one `You Draw It' task plot.
-Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
 \end{figure}
@@
 
+The `You Draw It' method does not restrict participants to draw a
+straight line as other methods would, such as using a ruler. Instead,
+participants are allowed to freely draw a line with potential curvature.
 Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
-linear mixed model (LMM) was fit separately to the OLS residuals and PCA
-residuals, constraining the fit to a linear trend. Parameter choice,
-\(x\), and the interaction between \(x\) and the parameter choice were
-treated as fixed effects with a random participant effect included to
-account for variation due to participant. The LMM equation for each fit
-(OLS and PCA) is given by: \begin{equation}
+linear mixed model (LMM) was fit separately to the OLS residuals and PA
+residuals, emulating the effect of constraining participants to draw a
+linear trend. Both fixed and random parameter estimates in the LMM were
+determined by optimizing the restricted maximum likelihood (REML)
+through penalized least squares. Parameter choice, \(x\), and the
+interaction between \(x\) and the parameter choice were treated as fixed
+effects with a random participant effect included to account for
+variation due to participant. The LMM equation for each fit (OLS and PA)
+is given by: \begin{equation}
 y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
@@
   parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  \(x\) value corresponding to either the OLS or PCA fit
+  \(x\) value corresponding to either the OLS or PA fit
 \item
@@
   and \(k^{th}\) increment of \(x\) value corresponding to either the
-  OLS or PCA fit
+  OLS or PA fit
 \item
@@
 of the residuals between the participant drawn points and fitted values
-for both the OLS regression line and PCA regression line. Estimated
+for both the OLS regression line and PA regression line. Estimated
 residual trend lines are overlaid on the observed individual participant
-residuals. Results indicate the estimated trends of PCA residuals
+residuals. Results indicate the estimated trends of PA residuals
 (orange) appear to align closer to the \(y=0\) horizontal (dashed) line
@@
 regression line with a slope based on the first principal axis than the
-estimated OLS regression line thus, providing support for ``ensemble
-perception''.
+estimated OLS regression line.
 
@@
 
-\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
+\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PA (orange) regression line constrained to a linear fit modeled by a linear mixed model.
 Estimated residual trends with 95\% confidence bands are overlaid on the observed individual participant residuals.}\label{fig:eyefitting-lmer-residualplots}
@@
 Eliminating the linear trend constraint, the \texttt{bam} function in
-the mgcv package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} was used to
-fit a generalized additive mixed model (GAMM) separately to the OLS
-residuals and PCA residuals to allow for estimation of smoothing
-splines. Parameter choice was treated as a fixed effect with no
-estimated intercept and a separate smoothing spline for \(x\) was
-estimated for each parameter choice. A random participant effect was
+the mgcv package \citep{mgcv_pkg} was used to fit a generalized additive
+mixed model (GAMM) separately to the OLS residuals and PA residuals to
+allow for estimation of smoothing splines. The \texttt{bam} function is
+used to fit GAMM's to very large data sets and use lower memory than the
+\texttt{gam} function. Parameter choice was treated as a fixed effect
+with no estimated intercept and a separate smoothing spline for \(x\)
+was estimated for each parameter choice. A random participant effect was
 included to account for variation due to participant and a random spline
 for each participant accounted for variation in spline for each
-participant. The GAMM equation for each fit (OLS and PCA) residuals is
-given by: \begin{equation}
-y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+participant. Defining e\_\{ijk,fit\} the same as in equation (2) above,
+the GAMM equation for each fit (OLS and PA) residuals is given by:
+\begin{equation}
+e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
 \end{equation} \noindent where
@@
 \item
-  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
-  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  \(x\) value
-\item
-  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
-  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
-  \(x\) value corresponding to either the OLS or PCA fit
-\item
-  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
-  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
-  and \(k^{th}\) increment of \(x\) value corresponding to either the
-  OLS or PCA fit
+  \(e_{ijk,fit}\) is the same as equation
 \item
@@
 of the residuals between the participant drawn points and fitted values
-for both the OLS regression line and PCA regression line. Estimated
+for both the OLS regression line and PA regression line. Estimated
 residual trends were overlaid on the observed individual participant
@@
 \cref{fig:eyefitting-lmer-residualplots} providing support that
-estimated trends of PCA residuals (orange) appear to align closer to the
+estimated trends of PA residuals (orange) appear to align closer to the
 \(y=0\) horizontal (dashed) line than the OLS residuals (blue) for
-scatter-plots with more noise (F and N). By fitting smoothing splines,
-we can determine whether participants naturally fit a straight
-trend-line to the set of points or whether they deviate throughout the
-domain. In particular, in scatter-plots with smaller variance (S and V),
-we can see that participants began at approximately the correct starting
-point then deviated away from the fitted regression lines and corrected
-for their fit toward the end of their trend-line. In scatter-plots with
-larger variance (F and N), participants estimated their starting value
-in the extreme direction of the OLS regression line based on the
-increasing or decreasing trend but more accurately represented the
-starting value of the PCA regression line. As participants continued
-their trend-line, they crossed through the OLS regression line
-indicating they estimated the slope in the extreme direction. These
-results provide further insight into the curvature humans perceive in a
-set of points.
+scatterplots with more noise (F and N). By fitting smoothing splines, we
+can determine whether participants naturally fit a straight trend-line
+to the set of points or whether they deviate throughout the domain. In
+particular, in scatterplots with smaller variance (S and V), we can see
+that participants began at approximately the correct starting point then
+deviated away from the fitted regression lines and corrected for their
+fit toward the end of their trend-line. In scatterplots with larger
+variance (F and N), participants estimated their starting value in the
+extreme direction of the OLS regression line based on the increasing or
+decreasing trend but more accurately represented the starting value of
+the PA regression line. As participants continued their trend-line, they
+crossed through the OLS regression line indicating they estimated the
+slope in the extreme direction. These results provide further insight
+into the curvature humans perceive in a set of points.
 
@@
 
-\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PCA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
+\caption{Estimated trend line of the residuals between the participant drawn points and fitted values for both the OLS (blue) regression line and PA (orange) regression line determined by smoothing splines fit by a generalized additive mixed model.
 Estimated residual trends with 95\% confidence bands are overlaid on the observed individual participant residuals.}\label{fig:eyefitting-gamm-residualplots}
@@
 
-\hypertarget{discussion-and-conclusion}{%
-\section{Discussion and Conclusion}\label{discussion-and-conclusion}}
+\hypertarget{conclusion-discussion}{%
+\section{Discussion and Conclusion}\label{conclusion-discussion}}
 
@@
 participants visually fit a regression line that mimics the first
-principle component regression as opposed to ordinary least squares
+principal axis regression as opposed to ordinary least squares
 regression. Data simulated with a larger variance provided strong
-support for a participants tendency to visually fit the first principle
-component regression. We utilized modern technology to replicate a study
+support for a participants tendency to visually fit the first principal
+axis regression. We utilized modern technology to replicate a study
 conducted 40 years ago, and strengthened the original results with
@@
 distance from their drawn regression line over both the \(x\) and \(y\)
-axis simultaneously. These results provide support that humans perform
-``ensemble perception'' in a statistical graphic setting. We allowed
-participants to draw trend lines that deviated from a straight line and
-gained an insight into the curvature the human eye perceives in a set of
-points.
+axis simultaneously. We allowed participants to draw trend lines that
+deviated from a straight line and gained an insight into the curvature
+the human eye perceives in a set of points. Researchers in cognitive and
+human movement sciences have found that human arm movement is a complex
+task \citep{miall1995curvature, rousset2015study}. The `You Draw It'
+method described in this paper uses \emph{indirect interaction} in which
+the mouse position and resulting visual line on the screen are
+dissociated. Therefore, curvature found in participant drawn lines from
+a straight lines could potentially be explained by the lack of
+coordination which results from the eye-hand dissociation from indirect
+drawing and the distortion of visual perception affecting the curvature
+of movements. Additionally, there is a training effect related to the
+completion of the `You Draw It' task - the movement of the line must be
+slow so that the visual representation on the screen can accurately
+capture each movement. \citet{de1991misdirections} conducted a study in
+which participants moved their hand slowly from an initial position in
+front of them to a visual target (movement task); they were then asked
+to repeat the task using different sizes of pointers (perceptual task).
+Their results indicated that deviations from the shortest pointers were
+comparable to those of the movement task, but that bias increased as the
+length of the pointer increased. While we suggested participants use a
+mouse to complete the study, we could not require the use; therefore,
+some participants may have used a track-pad and results may have been
+influenced by the pressure placed on their track-pad
+\citep{easton1978finger}.
 
@@
 modeling participant drawn lines using generalized additive mixed
-models. Further investigation is necessary to implement this method in
-non-linear settings and with real data in order to facilitate scientific
-communication. This tool could also be used to evaluate human ability to
-extrapolate data from trends. In the future, we intend to create an R
-package designed for easy implementation of `You Draw It' task plots in
-order to make this tool accessible to other researchers.
+models. Additional studies related to the validation and use of the tool
+would be useful for providing insight into explanations of biases
+introduced by the task such as the deviation from a straight line. For
+instance, a variation on the current study could compare manual
+adjustment methods such as shifting and rotating a horizontal line
+segment until the fit is suitable to the `You Draw It' method on the
+same set of data. This might explain the large deviation from the
+participant drawn line as \(x\) approaches 20. Another useful extension
+study would be to compare the `You Draw It' method as conducted by
+direct interaction - using a digital pen on a tablet - to indirect
+interaction - using a computer mouse to relate to a pointer on the
+screen. Further extensions to this work might ask participants to draw a
+trend-line through scatterplots with one (or multiple) extreme outliers
+in order to evaluate the perceptual system's resistance to outliers.
 
-\newpage
+While the focus of this study was on drawing linear trend-lines, further
+investigation is necessary to implement this method in non-linear
+settings and with real data in order to facilitate scientific
+communication - a strength of the combination of the flexible `You Draw
+It' method and GAMM analysis method. This tool could also be used to
+evaluate human ability to extrapolate data from trends. In the future,
+we intend to create an R package designed for easy implementation of
+`You Draw It' task plots in order to make this tool accessible to other
+researchers.
+
 \begin{center}
@@
 
+\{\#supplementary-material\}
+
 \begin{itemize}
@@
 \item
+  \textbf{Study Applet:} The shiny app used to conduct the study can be
+  accessed at
+  \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.
+\item
+  \textbf{RShiny Applet Code:} The code used to create the RShiny Applet
+  for data collection can be found at
+  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{github.com/earobinson95/you-draw-it-validation-applet}.
+\item
   \textbf{Participant Data:} De-identified participant data collected in
   the study and used for analyses are available to be downloaded from
-  GitHub
-  \href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}{here}.
+  GitHub at
+  \href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}{github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}.
 \item
   \textbf{Data Analysis Code:} The code used to replicate the analysis
-  in this paper can be found
-  \href{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}{here}.
-\item
-  \textbf{Study Applet:} The shiny app used to conduct the study can be
-  accessed
-  \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{here}.
-\item
-  \textbf{RShiny Applet Code:} The code used to create the RShiny Applet
-  for data collection can be found
-  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{here}.
+  in this paper can be found at
+  \href{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}{earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}.
 \end{itemize}

### 2022-08-04 manuscript-revisions/response-to-reviewers.qmd
made changes based on Susan's review.

@@
   + \response{After double checking the data, it was actually 131 plots - we are working with pilot data and thus had this number from another set.
-We clarified and made this number now reproducible by using r-code in the rmd template file rather than mannually entering it.
+We clarified and made this number now reproducible by using r-code in the rmd template file rather than manually entering it.
 We also discussed the missing plots by adding: "Data were collected as a part of a pilot study meant to test the applet; therefore, either voluntary participant dropout or disconnection from a server not designed to accommodate large magnitudes of participants resulted in missing plots in our data set for analysis."}
@@
 The way they defined their parameters was by using the y-value at the mean of the x-values $(y_{\bar{x}})$ and the slope.
-When simulated our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
+When simulating our data, we first generate and jitter x-values across the domain, therefore, x-bar may differ for each unique data set due to the random jitter.
 In turn, the actual y-intercept $(\beta_0)$ will vary slightly from data set to data set, but each trend will go through the same $(\bar{x}, y_{\bar{x}})$ point.
@@
   + \response{Changed to "increment" for both to be consistent and added specification of k in l.48.}
-+ \dubcheck{p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations}
++ p.12, l.50: it is not clear to me what you mean by "constraining the fit to a linear trend", even after reading the following part; provide some additional explanations
   + \response{We clarified what we meant by "constraining the fit to a linear trend" by adding explantion about participants being allowed to freely draw: "The 'You Draw It' method does not restrict participants to draw a straight line as other methods would, such as using a ruler.
@@
   + \response{Clarified similarities between the linear mixed model equation and the generalized additive mixed model equation by stating "Defining $e_{ijk,fit}$ the same as in equation (2) above, the GAMM equation...".}
-+ \dubcheck{p.16, Discussion:} Add some references from cognitive/human movement sciences that human arm movement is a complex task.
++ p.16, Discussion: Add some references from cognitive/human movement sciences that human arm movement is a complex task.
 Drawing a straight line is not as easy as it sounds.
@@
 We added these suggestions as well as suggested a study which compares the indirect interaction (mouse and pointer on the screen) and direct interaction (digital pen on a tablet) method of using 'You Draw It'.}
-+ \dubcheck{p.17-21, References:}
++ p.17-21, References:
   + capitalize journal names, e.g., Psychological bulletin & IEEE transactions on visualization and computer graphics
@@
 1.
-\dubcheck{In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis.
+In your manuscript, it would be helpful if you could clarify and emphasize the novelty of your experiment, data collection, and analysis.
 In particular, how does your work differ from Ciccione and Dehaene (2021)?(Ciccione, L.
@@
 Accuracy and bias in the perception of scatterplots.
-Cognitive Psychology, 128, p.101406.)}
+Cognitive Psychology, 128, p.101406.)
   + \response{We clarifed the differences between our work and Ciccione (2021) as our study use a more flexible method of testing graphics and introduces a way to analyze the data using GAMM's.

### 2022-08-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
reran maunscript knit

@@
 
-Model equation parameters, $\beta_1$, $y_{\bar{x}}$, and parameter choice letter names (F, S, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
+Model equation parameters, $\beta_1$, $y_{\bar{x}}$, and parameter choice letter names (S, F, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
 The mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$ were used in the point-slope equation of a line.
@@
           ) %>%
-            mutate(Parm = factor(Parm, levels = c("F", "S", "V", "N"))) %>%
+            mutate(Parm = factor(Parm, levels = c("S", "F", "V", "N"))) %>%
             arrange(Parm) %>%
@@
   # filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
-  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
@@
 + $\gamma_0$ is the overall intercept
-+ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
++ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (S, F, V, N) on the intercept
 + $\gamma_1$ is the overall slope for $x$
@@
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
             aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
-            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PCA"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
   geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
@@
   geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PA"), color = NA, alpha = 0.4) +
-  geom_line(aes(y = emmean.pca, color = "PCA")) +
+  geom_line(aes(y = emmean.pca, color = "PA")) +
   geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
@@
   scale_y_continuous("Residual") +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PA")) +
   scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PA"))
@@
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
   geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
@@
   scale_y_continuous("Residual") +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PA")) +
+  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PA")) 
 eyefitting.gamm.plot

### 2022-08-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
reran maunscript knit

@@
 Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\), and parameter
-choice letter names (F, S, V, N), were selected to reflect the four data
+choice letter names (S, F, V, N), were selected to reflect the four data
 sets used and labeled in \citet{mosteller1981eye}
@@
 \midrule
+S & 3.88 & 0.66 & 1.30 & (0,20)\\
 F & 3.90 & 0.66 & 1.98 & (0,20)\\
-S & 3.88 & 0.66 & 1.30 & (0,20)\\
 V & 3.89 & 1.98 & 1.50 & (4,16)\\
@@
 \item
-  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
+  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (S, F,
   V, N) on the intercept

### 2022-08-06 analysis/you-draw-it-eyefitting-analysis.Rmd
reran maunscript knit

--- 2022-07-20 analysis/you-draw-it-eyefitting-analysis.Rmd
@@
 
-**Rmd file containing code can be accessed [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/analysis).**
+**Rmd file containing code can be accessed [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/blob/main/analysis/you-draw-it-eyefitting-analysis.Rmd).**
 
@@
 
+## GET DATA FROM DATABASE
+
+```{r db}
+library(RSQLite)
+library(DBI)
+library(digest)
+library(lubridate)
+
+last_date <- as.numeric(as_datetime("2021-05-18 18:00:00 CDT"))
+
+# Connect to database
+filename <- "you_draw_it_data.db"
+sqlite.driver <- dbDriver("SQLite")
+db_con <- dbConnect(sqlite.driver, dbname = filename)
+# dbListTables(db_con)
+
+# Import eyefitting parameter details
+eyefitting_parameter_details <- dbReadTable(db_con,"eyefitting_parameter_details") %>%
+  mutate(parm_id = as.factor(as.character(parm_id)))
+
+# Import user data
+users <- dbReadTable(db_con,"users") %>%
+  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
+  select(-ip_address) %>%
+  rowwise() %>%
+  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime))))
+
+# Import exponential prediction feedback data
+eyefitting_feedback_data <- dbReadTable(db_con,"feedback") %>%
+  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
+  filter(parm_id %in% c("S", "F", "V", "N")) %>%
+  dplyr::rename(yols = y) %>%
+  rowwise() %>%
+  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>%
+  left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>%
+  filter(recruitment != "I am the researcher") %>%
+  rowwise() %>%
+  mutate(plot_id = as.factor(digest(paste(nick_name, start_time, end_time)))) %>%
+  select(study_starttime, participant_id, nick_name, plot_id, start_time, end_time, parm_id, x, yols, ydrawn, age, gender, academic_study, recruitment)
+
+factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
+                "recruitment", "participant_id", "parm_id", "plot_id")
+eyefitting_feedback_data[,factorCols] <- lapply(eyefitting_feedback_data[,factorCols], factor)
+# write.csv(eyefitting_feedback_data, file = "eyefitting-feedback-data.csv", row.names = F, na = "")
+
+# Import exponential prediction simulated data
+eyefitting_simulated_data <- dbReadTable(db_con,"simulated_data") %>%
+  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
+  filter(parm_id %in% c("S", "F", "V", "N")) %>%
+  rowwise() %>%
+  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>%
+  left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>%
+  filter(recruitment != "I am the researcher")
+
+factorCols <- c("nick_name", "dataset", "study_starttime", "age", "gender", "academic_study",
+                "recruitment", "participant_id", "parm_id")
+eyefitting_simulated_data[,factorCols] <- lapply(eyefitting_simulated_data[,factorCols], factor)
+# write.csv(eyefitting_simulated_data, file = "eyefitting-simulated-data.csv", row.names = F, na = "")
+
+# Disconnect from database
+dbDisconnect(db_con)
+length(unique(eyefitting_feedback_data$participant_id))
+length(unique(eyefitting_feedback_data$plot_id))
+```
+
 ## DATA SIMULATION PROCEDURE
 
-```{r}
+```{r, fig.width = 12, fig.height = 4}
 
-eyefitting_parameter_details <- read_csv("eyefitting_parameter_details.csv")
+eyefitting_parameter_details <- read_csv("eyefitting-parameter-details.csv")
 eyefitting_parameter_details %>% knitr::kable()
@@
 
-# write.csv(eyefitting_sim_data, "eyefitting-simulated-data.csv", row.names = F, na = "")
+# write.csv(eyefitting_sim_data, "eyefitting-simulated-data-example.csv", row.names = F, na = "")
 
@@
   # filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
@@
 # save data to csv - already run once and pushed to GitHub.
-# write.csv(eyefitting_model_data, file = "eyefitting-model-data.csv", row.names = F, na = "")
+# write.csv(eyefitting_model_data, file = "../data/eyefitting-model-data.csv", row.names = F, na = "")
 ```
@@
 
-De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-model-data.csv).
+De-identified participant data collected in the study and used for analyses are available to be viewed and downloaded from GitHub at [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data).
 
 ```{r data}
-eyefitting_model_data  <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-model-data.csv")
+eyefitting_model_data  <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/eyefitting-model-data.csv")
 
@@
 
-```{r}
+```{r, fig.height = 4, fig.width = 4}
 # Example
@@
 
-```{r y-ols-plot}
+```{r y-ols-plot, fig.width = 12, fig.height = 4}
 
@@
   geom_line(alpha = 0.2, aes(y = yols, group = participant_id)) +
-  facet_wrap(~ parm_id) +
+  facet_wrap(~ parm_id, ncol = 4) +
   theme_bw() +
@@
 
-```{r y-pca-plot}
+```{r y-pca-plot, fig.width = 12, fig.height = 4}
 eyefitting_model_data %>%
@@
   geom_line(alpha = 0.2, aes(y = ypca, group = participant_id)) +
-  facet_wrap(~ parm_id) +
+  facet_wrap(~ parm_id, ncol = 4) +
   theme_bw() +
@@
 
-```{r residual-plot}
+```{r residual-plot, fig.width = 12, fig.height = 4}
 eyefitting_model_data %>%
@@
   geom_hline(yintercept = 0, linetype = "dashed") +
-  facet_wrap(~parm_id) +
+  facet_wrap(~parm_id, ncol = 4) +
   theme_bw() +
@@
 
-+ parm_ID (S, F, V, N)
++ parm_ID (F, S, V, N)
 + x (0, 20)
@@
 
-```{r lmer, fig.height = 9}
+```{r lmer, fig.width = 12, fig.height = 4}
 # OLS
@@
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
             aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
             aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PCA"), alpha = 0.2) +
@@
 
-```{r gamm, fig.height = 9}
+```{r gamm, fig.width = 12, fig.height = 4}
 # OLS
@@
   filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
+  mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>%
   dplyr::rename(`Parameter Choice` = parm_id) %>%
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("F", "S", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
   geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +

### 2022-08-06 analysis/you-draw-it-eyefitting-analysis.Rmd
idk

@@
 }
-```
-
-## GET DATA FROM DATABASE
-
-```{r db}
-library(RSQLite)
-library(DBI)
-library(digest)
-library(lubridate)
-
-last_date <- as.numeric(as_datetime("2021-05-18 18:00:00 CDT"))
-
-# Connect to database
-filename <- "you_draw_it_data.db"
-sqlite.driver <- dbDriver("SQLite")
-db_con <- dbConnect(sqlite.driver, dbname = filename)
-# dbListTables(db_con)
-
-# Import eyefitting parameter details
-eyefitting_parameter_details <- dbReadTable(db_con,"eyefitting_parameter_details") %>%
-  mutate(parm_id = as.factor(as.character(parm_id)))
-
-# Import user data
-users <- dbReadTable(db_con,"users") %>%
-  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
-  select(-ip_address) %>%
-  rowwise() %>%
-  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime))))
-
-# Import exponential prediction feedback data
-eyefitting_feedback_data <- dbReadTable(db_con,"feedback") %>%
-  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
-  filter(parm_id %in% c("S", "F", "V", "N")) %>%
-  dplyr::rename(yols = y) %>%
-  rowwise() %>%
-  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>%
-  left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>%
-  filter(recruitment != "I am the researcher") %>%
-  rowwise() %>%
-  mutate(plot_id = as.factor(digest(paste(nick_name, start_time, end_time)))) %>%
-  select(study_starttime, participant_id, nick_name, plot_id, start_time, end_time, parm_id, x, yols, ydrawn, age, gender, academic_study, recruitment)
-
-factorCols <- c("nick_name", "study_starttime", "age", "gender", "academic_study",
-                "recruitment", "participant_id", "parm_id", "plot_id")
-eyefitting_feedback_data[,factorCols] <- lapply(eyefitting_feedback_data[,factorCols], factor)
-# write.csv(eyefitting_feedback_data, file = "eyefitting-feedback-data.csv", row.names = F, na = "")
-
-# Import exponential prediction simulated data
-eyefitting_simulated_data <- dbReadTable(db_con,"simulated_data") %>%
-  filter(study_starttime > 1620152231, study_starttime < last_date) %>%
-  filter(parm_id %in% c("S", "F", "V", "N")) %>%
-  rowwise() %>%
-  mutate(participant_id = as.factor(digest(paste(nick_name, study_starttime)))) %>%
-  left_join(users, by = c("participant_id", "nick_name", "study_starttime")) %>%
-  filter(recruitment != "I am the researcher")
-
-factorCols <- c("nick_name", "dataset", "study_starttime", "age", "gender", "academic_study",
-                "recruitment", "participant_id", "parm_id")
-eyefitting_simulated_data[,factorCols] <- lapply(eyefitting_simulated_data[,factorCols], factor)
-# write.csv(eyefitting_simulated_data, file = "eyefitting-simulated-data.csv", row.names = F, na = "")
-
-# Disconnect from database
-dbDisconnect(db_con)
-length(unique(eyefitting_feedback_data$participant_id))
-length(unique(eyefitting_feedback_data$plot_id))
 ```

### 2022-08-10 manuscript-revisions/response-to-reviewers.qmd
finalizing revisions - need to get the manuscript-revisions/revision-diffs.R file to create a diff.pdf

@@
 
-\todo{Need to address yet.} \dubcheck{Double check / review.}
+<!-- \todo{Need to address yet.} \dubcheck{Double check / review.} -->
 
@@
   + \response{Added "," into indexing on p11, l34 and p7, l55}
-+ \dubcheck{p.9, top: $y_{\bar{x}}$ in Table 1 and $\beta_0$ in (1) are not properly defined and used; are they the same?}
++ p.9, top: $y_{\bar{x}}$ in Table 1 and $\beta_0$ in (1) are not properly defined and used; are they the same?
   + \response{We aimed to replicate the trends from Mosteller et al.
@@
 2.
-\todo{In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe the method of estimation.
-In other words, what are the R functions doing to estimate the model?}
+In Section 3, it is helpful to note the R functions used as comments or footnotes, but you should also clearly describe the method of estimation.
+In other words, what are the R functions doing to estimate the model?
   + \response{We added the method of estimations based on the R documentation for the lmer and bam functions.}

### 2022-08-10 manuscript-revisions/revision-diffs.R

@@ -1,0 +1,12 @@
+library(latexdiffr)
+
+latexdiff(
+  "original-submission-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  "Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  output = "diff",
+  open = TRUE,
+  clean = TRUE,
+  quiet = TRUE,
+  output_format = NULL,
+  ld_opts = "--replace-context2cmd=\"none\""
+)

### 2022-10-06 manuscript-revisions/response-to-reviewers-2.qmd
initial round of final edits

---
title: "Second response to reviewers of JCGS-22-045"
subtitle: "Eye Fitting Straight Lines in the Modern Era"
author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
date: "2022"
# output: pdf_document
format: 
  pdf:
    include-in-header:
      - revision-edits.tex
---

## Reviewer(s)' Comments to Author

All responses to reviewers' comments are written in \response{blue}.

<!-- \todo{Need to address yet.} \dubcheck{Double check / review.} -->

### Reviewer: 1

I sincerely appreciate the revision made by the authors. They addressed my main concerns and they better clarified the added value of their work in the context of previous and very similar studies. Given the scope of the journal, I think that the article is now suitable for publication. I just have two final suggestions for the authors:

1) I invite them to carefully re-read the entire manuscript, since a few typos can still be found.

+ \response{We have carefully reviewed the manuscript for typos.}

2) My previous concern about the lack of clear recognition of previous similar work on the topic (which has been very well addressed by the authors, concerning the introduction, the methods and the results sections) seems somehow still applicable to the two last sections. For example, the possible role of training in the task is only discussed in terms of "movements". What about the training itself in performing the task? (i.e., becoming better and better at performing mental regressions? For example, Cui et al., 2018 precisely studied the question of training in correlation estimations in scatterplots). In the "future work" section, some of the suggestions made by the authors have indeed already been tested: concerning the resistance to outliers, quite a few studies already investigated human ability to perform regressions over scatterplots with outliers (as an example, but other studies can be found: Bobko & Karren, 1979 and Correll & Heer, 2017). Lastly, authors suggest to conduct studies on human ability to extrapolate data from trends; again, studies exist on the topic (as an example, Ciccione et al., 2022, who asked participants precisely to extrapolate new datapoints from exponential and linear trends plotted on log and linear scales). I am aware that the studies I mention come from the literature in cognitive psychology and/or psychophysics, rather than from the literature in data visualization, statistics or interactive visualization. However, I think that the authors, given the research work they conducted and presented, are ideally placed to fill the gap between the fields of cognitive psychology and data visualization. Therefore, they could consider adding those and/or other references to their last sections in order to better acknowledge the work conducted in fields other than data visualization.

+ \response{We included and acknowledged the work suggested from cognitive psychology. Thank you for bringing close attention to this field of work!}

Good job and good luck with your future work !

### Reviewer: 2

The authors present a study where human participants had to fit straight lines through points in a scatterplot. On average, the participants' fitted lines closely matched a principal axis line and not the least squares regression line.

The authors addressed my previous comments very well. Particular thanks for the latexdiff version that made it easy to review the revision!

Only a few very minor corrections remain to be done.

+ p.7, l.23-24: use past tense here: "participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely HAD an interest in fields related to statistics and WANTED to help advance research in graphics."
  + \response{Changed from present to past tense: have -> had and want -> wanted.}

+ p.7, l.37: use the full URL (not every pdf viewer / web browser correctly translates an incomplete URL): https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/ [same for the URLs on p.17]
  + \response{Double checked and provided full URLs. We also provided the publishers with a .zip file of supplementary material.}

+ p.8, l.46: after eq (1), continue with "..., where b_1 represents ... and y_x-bar ...", i.e., verbally explain earlier in the text what these mean [your explanation comes too late later on]
  + \response{This was a very helpful suggestion, we clarified under equation 1 with "where $\beta_1$ represents the slope of the trend and $y_{\bar{x}}$ represents the $y$ value at the mean of the generated $x$ values." and later stated "The point $(x, y_{\bar{x})}$ was used in the point-slope equation of a line."}

+ p.8, l.27 & p. 11, l.54 & p.14, l.5: speak of "R" package as many authors these days make joint use of R and Python (and others) for their work (so always clarify which "package" from which software environment is used) [check elsewhere]}
  + \response{We clarified the use of R packages after each use of a package from R statistical software.}

+ p.14, l.11: "REML is used" is not clear - is this another R function from the mgcv package or the method in general? - be more specific}
  + \response{We clarified "...; the `bam` function implements restricted maximum likelihood (REML) to estimate parameters and smoothing splines."}

+ p.14, l.19: place e_{ijk,fit} into $...$
  + \response{Added math typeset around e\_{ijk,fit} for $e_{ijk,fit}$.}

+ p.14, l.30: "eijk,fit is the same as equation" likely should be "eijk,fit is the same as IN equation **(2)**"
  + \response{Added 'in' and '(2)' to defining $e_{ijk,fit}$ after equation (3).}

+ p.19, l.54: remove the extra author from "Kosslyn, S. M. & Kosslyn, S. M. (2006)" - this is just Kosslyn
  + \response{Removed extra author in references.}

+ p.20, l.45: for Rousset et al., add "ACM, New York, NY" as publisher for these proceedings - see the citation at https://dl.acm.org/doi/10.1145/2820619.2820633
  + \response{Exported bibtex citation from the reccommended link.}

\response{Thank you for your valuable feedback, suggestions, and review of our manuscript.}

### Associate Editor's Comments to Author

(There are no comments.)

### 2022-10-06 supplementary-materials/you-draw-it-eyefitting-analysis.Rmd
supplementary material

--- 2022-08-06 analysis/you-draw-it-eyefitting-analysis.Rmd
@@
 ---
-title: "Eye Fitting Straight Lines in the Modern Era (JCGS-22-045)"
+title: "Eye Fitting Straight Lines in the Modern Era"
 subtitle: "Data Analysis"
@@
 <br>
-
-**Rmd file containing code can be accessed [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/blob/main/analysis/you-draw-it-eyefitting-analysis.Rmd).**
 
@@
 
-eyefitting_parameter_details <- read_csv("eyefitting-parameter-details.csv")
+# eyefitting_parameter_details <- read_csv("eyefitting-parameter-details.csv")
+eyefitting_parameter_details <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-feedback-data.csv")
 eyefitting_parameter_details %>% knitr::kable()
@@
 ```{r data}
-eyefitting_model_data  <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/eyefitting-model-data.csv")
+# eyefitting_model_data  <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/eyefitting-model-data.csv")
+eyefitting_model_data  <- read_csv("eyefitting-model-data.csv")
 

### 2022-10-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
finalizing revisions - need to get the manuscript-revisions/revision-diffs.R file to create a diff.pdf
Supplementary material reference fix
updated supplementary materials

@@
 The experimental tasks took approximately 15 minutes to complete.
-As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely have an interest in fields related to statistics and want to help advance research in graphics.
+As this is a pilot study, participants from Twitter and Reddit pages related to data visualization voluntarily completed the study and likely had an interest in fields related to statistics and wanted to help advance research in graphics.
 While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
@@
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
 
@@
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual. 
-In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` package [@r2d3].
+In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` R package [@r2d3].
 While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.
@@
 
+where $\beta_1$ represents the slope of the trend and $y_{\bar{x}}$ represents the $y$ value at the mean of the generated $x$ values.
 Model equation parameters, $\beta_1$, $y_{\bar{x}}$, and parameter choice letter names (S, F, V, N), were selected to reflect the four data sets used and labeled in @mosteller1981eye \pcref{tab:eyefitting-parameters}.
-The mean of the generated $x$ values and the predefined $y$ value at $\bar x$, denoted $y_{\bar x}$ were used in the point-slope equation of a line.
+The point $(x, y_{\bar{x})}$ was used in the point-slope equation of a line.
 Parameter choices S, F, and N simulated data across a domain of 0 to 20. 
@@
 Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increment across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$, for $k = 1, ..., 4 x_{max} +1$.
-The PA regression slope, $\hat\beta_{1,PA}$, and y-intercept, $\hat\beta_{0,PA}$, were determined using the `mcreg` function in the `mcr` package in R [@mcr_pkg] which implements Deming regression (equivalent to a regression based on the slope of the first principal axis).
+The PA regression slope, $\hat\beta_{1,PA}$, and y-intercept, $\hat\beta_{0,PA}$, were determined using the `mcreg` function in the mcr package in R [@mcr_pkg] which implements Deming regression (equivalent to a regression based on the slope of the first principal axis).
 Fitted values, $\hat y_{k,PA}$ were then obtained every 0.25 increment across the domain from the PA regression equation, $\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k$, for $k = 1, ..., 4 x_{max} +1$.
@@
 ```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Illustrates the data associated with and collected for one `You Draw It' task plot.
-Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).", out.width="50%"}
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (long dashed orange).", out.width="50%"}
 eyefitting_model_data %>%
@@
   scale_color_manual("", values = c("black", "steelblue", "orange2")) +
-  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
+  scale_linetype_manual("", values = c("dashed", "solid", "longdash"))
 ```
@@
 Instead, participants are allowed to freely draw a line with potential curvature.
-Using the `lmer` function in the lme4 package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend.
+Using the `lmer` function in the lme4 R package [@lme4], a linear mixed model (LMM) was fit separately to the OLS residuals and PA residuals, emulating the effect of constraining participants to draw a linear trend.
 Both fixed and random parameter estimates in the LMM were determined by optimizing the restricted maximum likelihood (REML) through penalized least squares.
@@
 
-Eliminating the linear trend constraint, the `bam` function in the mgcv package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
-The `bam` function is used to fit GAMM's to very large data sets and use lower memory than the `gam` function.
+Eliminating the linear trend constraint, the `bam` function in the mgcv R package [@mgcv_pkg] was used to fit a generalized additive mixed model (GAMM) separately to the OLS residuals and PA residuals to allow for estimation of smoothing splines.
+The `bam` function is used to fit GAMM's to very large data sets and use lower memory than the `gam` function in the mgcv R package; the `bam` function implements restricted maximum likelihood (REML) to estimate parameters and smoothing splines.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice.
 A random participant effect was included to account for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
-Defining e_{ijk,fit} the same as in equation (2) above, the GAMM equation for each fit (OLS and PA) residuals is given by:
+Defining $e_{ijk,fit}$ the same as in equation (2) above, the GAMM equation for each fit (OLS and PA) residuals is given by:
 \begin{equation}
@@
 
-+ $e_{ijk,fit}$ is the same as equation
++ $e_{ijk,fit}$ is the same as in equation (2)
 + $\alpha_i$ is the intercept for the parameter choice $i$
@@
   facet_wrap(~`Parameter Choice`, ncol = 4, labeller = label_both) +
+  # facet_wrap(`Parameter Choice`, ncol = 4, labeller = label_both) +
   theme_bw(base_size = 14) +
@@
 While we suggested participants use a mouse to complete the study, we could not require the use; therefore, some participants may have used a track-pad and results may have been influenced by the pressure placed on their track-pad [@easton1978finger].
+In cognitive psychology, @cui_2018 studied the improvement in estimating correlations in scatter plots through perceptual learning interventions.  
+Their study provided support for an improved proficiency in correlation estimation after training.
+We used a randomized complete design experimental structure, while @mosteller1981eye implemented a latin square design to test for an order effect due to practice.
 
@@
 Another useful extension study would be to compare the 'You Draw It' method as conducted by direct interaction - using a digital pen on a tablet - to indirect interaction - using a computer mouse to relate to a pointer on the screen.
-Further extensions to this work might ask participants to draw a trend-line through scatterplots with one (or multiple) extreme outliers in order to evaluate the perceptual system's resistance to outliers.
-
+Previous studies investigated human ability to perform regressions over scatterplots with outliers and found participant estimated trend lines to be closer to a robust trend line than the statistically fitted trend line where outliers were included [@bobko_1979; @correll_2017].
+This work could provide an extension and further evaluate the perceptual system's resistance to outliers by allowing participants to deviate from a straight line, potentially capturing the outlier(s).
+<!-- Further extensions to this work might ask participants to draw a trend-line through scatterplots with one (or multiple) extreme outliers in order to further evaluate the perceptual system's resistance to outliers [@bobko_1979; @correll_2017].
+-->
 While the focus of this study was on drawing linear trend-lines, further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication - a strength of the combination of the flexible 'You Draw It' method and GAMM analysis method.
-This tool could also be used to evaluate human ability to extrapolate data from trends. 
+@ciccione2021can evaluated human ability to extrapolate data from trends by asking participants to provide a single point estimate; this tool could be used to evaluate a continuous extrapolation of points from a trend line for various linear and non-linear data structures.
 In the future, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
 
-\begin{center}
-{\large\bf SUPPLEMENTARY MATERIAL}
-\end{center}
+# Supplementary Material {#supplementary-material}
 
-{#supplementary-material}
-
-+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-+ **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/you-draw-it-validation-applet](https://github.com/earobinson95/you-draw-it-validation-applet).
-+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
-+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found at [earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
+<!-- + **Study Applet:** The shiny app used to conduct the study can be accessed at  [https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+-->
+<!-- + **RShiny Applet Code:** The code used to create the RShiny Applet for data collection can be found at [https://github.com/earobinson95/you-draw-it-validation-applet(https://github.com/earobinson95/you-draw-it-validation-applet).
+-->
++ **Participant Data:** De-identified participant data collected in the study and used for analyses are available in `eyefitting-model-data.csv` or may be downloaded from GitHub at  [https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
++ **Data Analysis Code:** The code used to replicate the analysis in this paper can be in `you-draw-it-eyefitting-analysis.Rmd` or found at [https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).

### 2022-10-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
finalizing revisions - need to get the manuscript-revisions/revision-diffs.R file to create a diff.pdf
updated supplementary materials

@@
 participants from Twitter and Reddit pages related to data visualization
-voluntarily completed the study and likely have an interest in fields
-related to statistics and want to help advance research in graphics.
+voluntarily completed the study and likely had an interest in fields
+related to statistics and wanted to help advance research in graphics.
 While this study does utilize a convenience sample, as this is primarily
@@
 \citep{shinyPkg} found at
-\href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.
+\url{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}.
 
@@
 visual. In order to allow for user interaction and data collection, we
-integrated the D3 visual into Shiny using the \texttt{r2d3} package
+integrated the D3 visual into Shiny using the \texttt{r2d3} R package
 \citep{r2d3}. While the interface is highly customized to this
@@
 
+where \(\beta_1\) represents the slope of the trend and \(y_{\bar{x}}\)
+represents the \(y\) value at the mean of the generated \(x\) values.
 Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\), and parameter
@@
 sets used and labeled in \citet{mosteller1981eye}
-(\cref{tab:eyefitting-parameters}). The mean of the generated \(x\)
-values and the predefined \(y\) value at \(\bar x\), denoted
-\(y_{\bar x}\) were used in the point-slope equation of a line.
-Parameter choices S, F, and N simulated data across a domain of 0 to 20.
-Parameter choice F produced a trend with a positive slope and a large
-variance while N had a negative slope and a large variance. In
-comparison, S showed a trend with a positive slope and a small variance
-while V yielded a steep positive slope with a small variance over the
-domain of 4 to 16. \cref{fig:eyefitting-simplot} illustrates an example
-of simulated data for all four parameter choices intended to reflect the
-trends in \citet{mosteller1981eye}. Aesthetic design choices were made
-consistent across each of the interactive `You Draw It' task plots. The
-y-axis range extended 10\% beyond (above and below) the range of the
-simulated data points to allow for users to draw outside the simulated
-data set range and avoid anchoring their lines to the corners of the
-plot.
+(\cref{tab:eyefitting-parameters}). The point \((x, y_{\bar{x})}\) was
+used in the point-slope equation of a line. Parameter choices S, F, and
+N simulated data across a domain of 0 to 20. Parameter choice F produced
+a trend with a positive slope and a large variance while N had a
+negative slope and a large variance. In comparison, S showed a trend
+with a positive slope and a small variance while V yielded a steep
+positive slope with a small variance over the domain of 4 to 16.
+\cref{fig:eyefitting-simplot} illustrates an example of simulated data
+for all four parameter choices intended to reflect the trends in
+\citet{mosteller1981eye}. Aesthetic design choices were made consistent
+across each of the interactive `You Draw It' task plots. The y-axis
+range extended 10\% beyond (above and below) the range of the simulated
+data points to allow for users to draw outside the simulated data set
+range and avoid anchoring their lines to the corners of the plot.
 
@@
 \(\hat\beta_{1,PA}\), and y-intercept, \(\hat\beta_{0,PA}\), were
-determined using the \texttt{mcreg} function in the \texttt{mcr} package
-in R \citep{mcr_pkg} which implements Deming regression (equivalent to a
+determined using the \texttt{mcreg} function in the mcr package in R
+\citep{mcr_pkg} which implements Deming regression (equivalent to a
 regression based on the slope of the first principal axis). Fitted
@@
 \caption{Illustrates the data associated with and collected for one `You Draw It' task plot.
-Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
+Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PA regression line based on the principal axis (long dashed orange).}\label{fig:eyefitting-example-plot}
 \end{figure}
@@
 participants are allowed to freely draw a line with potential curvature.
-Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
+Using the \texttt{lmer} function in the lme4 R package \citep{lme4}, a
 linear mixed model (LMM) was fit separately to the OLS residuals and PA
@@
 Eliminating the linear trend constraint, the \texttt{bam} function in
-the mgcv package \citep{mgcv_pkg} was used to fit a generalized additive
-mixed model (GAMM) separately to the OLS residuals and PA residuals to
-allow for estimation of smoothing splines. The \texttt{bam} function is
-used to fit GAMM's to very large data sets and use lower memory than the
-\texttt{gam} function. Parameter choice was treated as a fixed effect
-with no estimated intercept and a separate smoothing spline for \(x\)
-was estimated for each parameter choice. A random participant effect was
-included to account for variation due to participant and a random spline
-for each participant accounted for variation in spline for each
-participant. Defining e\_\{ijk,fit\} the same as in equation (2) above,
-the GAMM equation for each fit (OLS and PA) residuals is given by:
-\begin{equation}
+the mgcv R package \citep{mgcv_pkg} was used to fit a generalized
+additive mixed model (GAMM) separately to the OLS residuals and PA
+residuals to allow for estimation of smoothing splines. The \texttt{bam}
+function is used to fit GAMM's to very large data sets and use lower
+memory than the \texttt{gam} function in the mgcv R package; the
+\texttt{bam} function implements restricted maximum likelihood (REML) to
+estimate parameters and smoothing splines. Parameter choice was treated
+as a fixed effect with no estimated intercept and a separate smoothing
+spline for \(x\) was estimated for each parameter choice. A random
+participant effect was included to account for variation due to
+participant and a random spline for each participant accounted for
+variation in spline for each participant. Defining \(e_{ijk,fit}\) the
+same as in equation (2) above, the GAMM equation for each fit (OLS and
+PA) residuals is given by: \begin{equation}
 e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
@@
 \item
-  \(e_{ijk,fit}\) is the same as equation
+  \(e_{ijk,fit}\) is the same as in equation (2)
 \item
@@
 influenced by the pressure placed on their track-pad
-\citep{easton1978finger}.
+\citep{easton1978finger}. In cognitive psychology, \citet{cui_2018}
+studied the improvement in estimating correlations in scatter plots
+through perceptual learning interventions.\\
+Their study provided support for an improved proficiency in correlation
+estimation after training. We used a randomized complete design
+experimental structure, while \citet{mosteller1981eye} implemented a
+latin square design to test for an order effect due to practice.
 
@@
 interaction - using a computer mouse to relate to a pointer on the
-screen. Further extensions to this work might ask participants to draw a
-trend-line through scatterplots with one (or multiple) extreme outliers
-in order to evaluate the perceptual system's resistance to outliers.
+screen. Previous studies investigated human ability to perform
+regressions over scatterplots with outliers and found participant
+estimated trend lines to be closer to a robust trend line than the
+statistically fitted trend line where outliers were included
+\citep{bobko_1979, correll_2017}. This work could provide an extension
+and further evaluate the perceptual system's resistance to outliers by
+allowing participants to deviate from a straight line, potentially
+capturing the outlier(s). While the focus of this study was on drawing
+linear trend-lines, further investigation is necessary to implement this
+method in non-linear settings and with real data in order to facilitate
+scientific communication - a strength of the combination of the flexible
+`You Draw It' method and GAMM analysis method. \citet{ciccione2021can}
+evaluated human ability to extrapolate data from trends by asking
+participants to provide a single point estimate; this tool could be used
+to evaluate a continuous extrapolation of points from a trend line for
+various linear and non-linear data structures. In the future, we intend
+to create an R package designed for easy implementation of `You Draw It'
+task plots in order to make this tool accessible to other researchers.
 
-While the focus of this study was on drawing linear trend-lines, further
-investigation is necessary to implement this method in non-linear
-settings and with real data in order to facilitate scientific
-communication - a strength of the combination of the flexible `You Draw
-It' method and GAMM analysis method. This tool could also be used to
-evaluate human ability to extrapolate data from trends. In the future,
-we intend to create an R package designed for easy implementation of
-`You Draw It' task plots in order to make this tool accessible to other
-researchers.
-
-\begin{center}
-{\large\bf SUPPLEMENTARY MATERIAL}
-\end{center}
-
-\{\#supplementary-material\}
+\hypertarget{supplementary-material}{%
+\section{Supplementary Material}\label{supplementary-material}}
 
@@
 \item
-  \textbf{Study Applet:} The shiny app used to conduct the study can be
-  accessed at
-  \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.
-\item
-  \textbf{RShiny Applet Code:} The code used to create the RShiny Applet
-  for data collection can be found at
-  \href{https://github.com/earobinson95/you-draw-it-validation-applet}{github.com/earobinson95/you-draw-it-validation-applet}.
-\item
   \textbf{Participant Data:} De-identified participant data collected in
-  the study and used for analyses are available to be downloaded from
-  GitHub at
-  \href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}{github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}.
+  the study and used for analyses are available in
+  \texttt{eyefitting-model-data.csv} or may be downloaded from GitHub at
+  \url{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}.
 \item
   \textbf{Data Analysis Code:} The code used to replicate the analysis
-  in this paper can be found at
-  \href{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}{earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}.
+  in this paper can be in \texttt{you-draw-it-eyefitting-analysis.Rmd}
+  or found at
+  \url{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}.
 \end{itemize}

### 2022-10-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
updated supplementary-material to follow the email guidelines: "You should provide an archive (.zip or .tar) that contains exactly the files listed under supplementary materials. Each item listed should include (1) a title, (2) a brief description, and (3) in parentheses, the file name and file type separated by a comma."

@@
 -->
-+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available in `eyefitting-model-data.csv` or may be downloaded from GitHub at  [https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
-+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be in `you-draw-it-eyefitting-analysis.Rmd` or found at [https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
++ **Participant Data:** De-identified participant data collected in the study and used for analyses (eyefitting-model-data.csv).
+<!-- `eyefitting-model-data.csv` or may be downloaded from GitHub at  [https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data).
+-->
++ **Data Analysis Code:** The code used to replicate the analysis in this paper (you-draw-it-eyefitting-analysis.Rmd).
+<!-- or found at [https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/analysis/you-draw-it-eyefitting-analysis.Rmd](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/supplementary-materials/you-draw-it-eyefitting-analysis.Rmd).
+-->
++ **README:** File containing detailed descriptions of the supplementary material (README.html).

### 2022-10-06 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
updated supplementary-material to follow the email guidelines: "You should provide an archive (.zip or .tar) that contains exactly the files listed under supplementary materials. Each item listed should include (1) a title, (2) a brief description, and (3) in parentheses, the file name and file type separated by a comma."

@@
   \textbf{Participant Data:} De-identified participant data collected in
-  the study and used for analyses are available in
-  \texttt{eyefitting-model-data.csv} or may be downloaded from GitHub at
-  \url{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data}.
+  the study and used for analyses (eyefitting-model-data.csv).
 \item
   \textbf{Data Analysis Code:} The code used to replicate the analysis
-  in this paper can be in \texttt{you-draw-it-eyefitting-analysis.Rmd}
-  or found at
-  \url{https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html}.
+  in this paper (you-draw-it-eyefitting-analysis.Rmd).
+\item
+  \textbf{README:} File containing detailed descriptions of the
+  supplementary material (README.html).
 \end{itemize}

### 2022-10-06 manuscript-revisions/final-submission-cover-letter.qmd
adjusted naming of supplementary-material

---
title: "Final JCGS Cover Letter"
subtitle: "Eye Fitting Straight Lines in the Modern Era"
author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
date: "2022"
format: pdf
---

## manuscript.zip

+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd`
+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex`
+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
+ `bibliography.bib`
+ `agsm.bst`
+ `images/ydi-stimuli.png`
+ `figure-latex/eyefitting-example-plot-1.pdf`
+ `figure-latex/eyefitting-gamm-residualplots-1.pdf`
+ `figure-latex/eyefitting-lmer-residualplots-1.pdf`
+ `figure-latex/eyefitting-simplot-1.pdf`
+ `figure-latex/ols-vs-pca-example-1.pdf`

## supplementary-material.zip

+ `README.html` 
+ `eyefitting-model-data.csv`
+ `you-draw-it-eyefitting-analysis.Rmd`

### 2022-10-06 supplementary-material/you-draw-it-eyefitting-analysis.Rmd
adjusted naming of supplementary-material

--- 2022-10-06 supplementary-materials/you-draw-it-eyefitting-analysis.Rmd
@@
 # eyefitting_parameter_details <- read_csv("eyefitting-parameter-details.csv")
-eyefitting_parameter_details <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-feedback-data.csv")
+eyefitting_parameter_details <- read_csv("https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/analysis/eyefitting-parameter-details.csv")
 eyefitting_parameter_details %>% knitr::kable()

### 2022-10-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
made final revisions.

@@
 In this paper, we validate ‘You Draw It’ as a method for graphical testing, comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.
-Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line.
+Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the principal axis over the slope of the least-squares regression line.
 This trend was most prominent when shown data simulated with larger variances.
@@
 Early research studies considered graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli.
-Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment - a technique which requires participants to alter a changing stimulus to match a given constant stimuli [@gescheider_1997]- to estimate the magnitude of the impact of the sine illusion.
+Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment - a technique which requires participants to alter a changing stimulus to match a given constant stimuli [@gescheider_1997] - to estimate the magnitude of the impact of the sine illusion.
 However, there are more modern testing methods that have been developed since the heyday of psychophysics.
@@
 Graphics are viewed as a mapping from variables in a data set (or statistics computed from the data) to visual attributes such as the axes, colors, shapes, or facets on the canvas in which the chart is displayed.
-Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the Grammar of Graphics recommends.
+Software, such as Hadley Wickham’s `ggplot2` package in R [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the Grammar of Graphics recommends.
 
@@
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points.
-This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression [@deming1943statistical; @linnet1998performance; @martin2000general], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatterplot.
+This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression [@deming1943statistical; @linnet1998performance; @martin2000general], which is equivalent to the principal axis and minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatterplot.
 
@@
 
-Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
+Participants were recruited through Twitter, Reddit, and direct email in May 2021.
 A total of 35 individuals completed `r length(unique(eyefitting_model_data$plot_id))` unique 'You Draw It' task plots.
@@
 Participants completed the experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application [@shinyPkg] found at [https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+The experiment was conducted and distributed through a R Shiny application [@shinyPkg] found at [https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
 
@@
 Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual. 
-In order to allow for user interaction and data collection, we integrated the D3 visual into Shiny using the `r2d3` R package [@r2d3].
+In order to allow for user interaction and data collection, we integrated the D3 visual into R Shiny using the `r2d3` R package [@r2d3].
 While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.
@@
 \end{align} 
-
 where $\beta_1$ represents the slope of the trend and $y_{\bar{x}}$ represents the $y$ value at the mean of the generated $x$ values.
@@
 Aesthetic design choices were made consistent across each of the interactive 'You Draw It' task plots. 
-The y-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range and avoid anchoring their lines to the corners of the plot.
+The $y$-axis range extended 10\% beyond (above and below) the range of the simulated data points to allow for users to draw outside the simulated data set range and avoid anchoring their lines to the corners of the plot.
 
@@
 Fitted values, $\hat y_{k,OLS}$, were then obtained every 0.25 increment across the domain from the OLS regression equation, $\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k$, for $k = 1, ..., 4 x_{max} +1$.
-The PA regression slope, $\hat\beta_{1,PA}$, and y-intercept, $\hat\beta_{0,PA}$, were determined using the `mcreg` function in the mcr package in R [@mcr_pkg] which implements Deming regression (equivalent to a regression based on the slope of the first principal axis).
+The PA regression slope, $\hat\beta_{1,PA}$, and $y$-intercept, $\hat\beta_{0,PA}$, were determined using the `mcreg` function in the mcr package in R [@mcr_pkg] which implements Deming regression (equivalent to a regression based on the slope of the principal axis).
 Fitted values, $\hat y_{k,PA}$ were then obtained every 0.25 increment across the domain from the PA regression equation, $\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k$, for $k = 1, ..., 4 x_{max} +1$.
@@
 Using generalized additive mixed models, we assessed the deviation of the participant drawn lines from the statistically fitted regression lines.
-Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the first principal axis regression as opposed to ordinary least squares regression.
+Our results found that when shown points following a linear trend, participants visually fit a regression line that mimics the principal axis regression line as opposed to ordinary least squares regression line.
 Data simulated with a larger variance provided strong support for a participants tendency to visually fit the first principal axis regression.
@@
 We allowed participants to draw trend lines that deviated from a straight line and gained an insight into the curvature the human eye perceives in a set of points.
+
 Researchers in cognitive and human movement sciences have found that human arm movement is a complex task [@miall1995curvature; @rousset2015study].
@@
 In cognitive psychology, @cui_2018 studied the improvement in estimating correlations in scatter plots through perceptual learning interventions.
-Their study provided support for an improved proficiency in correlation estimation after training.
+Results form their study provided support for an improved proficiency in correlation estimation after training.
 We used a randomized complete design experimental structure, while @mosteller1981eye implemented a latin square design to test for an order effect due to practice.
@@
 Previous studies investigated human ability to perform regressions over scatterplots with outliers and found participant estimated trend lines to be closer to a robust trend line than the statistically fitted trend line where outliers were included [@bobko_1979; @correll_2017].
-This work could provide an extension and further evaluate the perceptual system's resistance to outliers by allowing participants to deviate from a straight line, potentially capturing the outlier(s).
+The work in this paper could provide an extension and further evaluate the perceptual system's resistance to outliers by allowing participants to deviate from a straight line, potentially capturing the outlier(s).
 <!-- Further extensions to this work might ask participants to draw a trend-line through scatterplots with one (or multiple) extreme outliers in order to further evaluate the perceptual system's resistance to outliers [@bobko_1979; @correll_2017].
 -->
-While the focus of this study was on drawing linear trend-lines, further investigation is necessary to implement this method in non-linear settings and with real data in order to facilitate scientific communication - a strength of the combination of the flexible 'You Draw It' method and GAMM analysis method.
-@ciccione2021can evaluated human ability to extrapolate data from trends by asking participants to provide a single point estimate; this tool could be used to evaluate a continuous extrapolation of points from a trend line for various linear and non-linear data structures.
+While the focus of this study was on drawing linear trend lines, further investigation is necessary to implement this method in nonlinear settings and with real data in order to facilitate scientific communication - a strength of the combination of the flexible 'You Draw It' method and GAMM analysis method.
+@ciccione2021can evaluated human ability to extrapolate data from trends by asking participants to provide a single point estimate; this tool could be used to evaluate a continuous extrapolation of points from a trend line for various linear and nonlinear data structures.
 In the future, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
@@
 -->
-+ **README:** File containing detailed descriptions of the supplementary material (README.html).
++ **readme:** File containing detailed descriptions of the supplementary material (README.html).

### 2022-10-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
made final revisions.

@@
 in the previous study; when shown points following a linear trend,
-participants tended to fit the slope of the first principal component
-over the slope of the least-squares regression line. This trend was most
+participants tended to fit the slope of the principal axis over the
+slope of the least-squares regression line. This trend was most
 prominent when shown data simulated with larger variances. This study
@@
 adjustment - a technique which requires participants to alter a changing
-stimulus to match a given constant stimuli \citep{gescheider_1997}- to
+stimulus to match a given constant stimuli \citep{gescheider_1997} - to
 estimate the magnitude of the impact of the sine illusion. However,
@@
 such as the axes, colors, shapes, or facets on the canvas in which the
-chart is displayed. Software, such as Hadley Wickham's ggplot2
-\citep{wickham2011ggplot2}, aims to implement the framework of creating
-charts and graphics as the Grammar of Graphics recommends.
+chart is displayed. Software, such as Hadley Wickham's \texttt{ggplot2}
+package in R \citep{wickham2011ggplot2}, aims to implement the framework
+of creating charts and graphics as the Grammar of Graphics recommends.
 
@@
 \citep{deming1943statistical, linnet1998performance, martin2000general},
-which minimizes the Euclidean distance of points from the line, when
-fitting a line to a noisy scatterplot.
+which is equivalent to the principal axis and minimizes the Euclidean
+distance of points from the line, when fitting a line to a noisy
+scatterplot.
 
@@
 
-Participants were recruited through through Twitter, Reddit, and direct
-email in May 2021. A total of 35 individuals completed 131 unique `You
-Draw It' task plots. Data were collected as a part of a pilot study
-meant to test the applet; therefore, either voluntary participant
-dropout or disconnection from a server not designed to accommodate large
-magnitudes of participants resulted in missing plots in our data set for
-analysis. All participants had normal or corrected to normal vision and
-signed an informed consent form. The experimental tasks took
-approximately 15 minutes to complete. As this is a pilot study,
-participants from Twitter and Reddit pages related to data visualization
-voluntarily completed the study and likely had an interest in fields
-related to statistics and wanted to help advance research in graphics.
-While this study does utilize a convenience sample, as this is primarily
-a perceptual task, previous results have found few differences between
-expert and non-expert participants in this context
-\citep{vanderplas2015spatial}. These data were collected to validate
-this method of graphical testing, with the hopes of providing a new tool
-to assess graphical perception interactively. Participants completed the
-experiment on their own computers in an environment of their choosing.
-The experiment was conducted and distributed through a Shiny application
-\citep{shinyPkg} found at
+Participants were recruited through Twitter, Reddit, and direct email in
+May 2021. A total of 35 individuals completed 131 unique `You Draw It'
+task plots. Data were collected as a part of a pilot study meant to test
+the applet; therefore, either voluntary participant dropout or
+disconnection from a server not designed to accommodate large magnitudes
+of participants resulted in missing plots in our data set for analysis.
+All participants had normal or corrected to normal vision and signed an
+informed consent form. The experimental tasks took approximately 15
+minutes to complete. As this is a pilot study, participants from Twitter
+and Reddit pages related to data visualization voluntarily completed the
+study and likely had an interest in fields related to statistics and
+wanted to help advance research in graphics. While this study does
+utilize a convenience sample, as this is primarily a perceptual task,
+previous results have found few differences between expert and
+non-expert participants in this context \citep{vanderplas2015spatial}.
+These data were collected to validate this method of graphical testing,
+with the hopes of providing a new tool to assess graphical perception
+interactively. Participants completed the experiment on their own
+computers in an environment of their choosing. The experiment was
+conducted and distributed through a R Shiny application \citep{shinyPkg}
+found at
 \url{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}.
@@
 visual. In order to allow for user interaction and data collection, we
-integrated the D3 visual into Shiny using the \texttt{r2d3} R package
+integrated the D3 visual into R Shiny using the \texttt{r2d3} R package
 \citep{r2d3}. While the interface is highly customized to this
@@
 \text{with } e_i & \sim N(0, \sigma^2). \nonumber
-\end{align}
-
-where \(\beta_1\) represents the slope of the trend and \(y_{\bar{x}}\)
-represents the \(y\) value at the mean of the generated \(x\) values.
-Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\), and parameter
-choice letter names (S, F, V, N), were selected to reflect the four data
-sets used and labeled in \citet{mosteller1981eye}
+\end{align} where \(\beta_1\) represents the slope of the trend and
+\(y_{\bar{x}}\) represents the \(y\) value at the mean of the generated
+\(x\) values. Model equation parameters, \(\beta_1\), \(y_{\bar{x}}\),
+and parameter choice letter names (S, F, V, N), were selected to reflect
+the four data sets used and labeled in \citet{mosteller1981eye}
 (\cref{tab:eyefitting-parameters}). The point \((x, y_{\bar{x})}\) was
@@
 \citet{mosteller1981eye}. Aesthetic design choices were made consistent
-across each of the interactive `You Draw It' task plots. The y-axis
+across each of the interactive `You Draw It' task plots. The \(y\)-axis
 range extended 10\% beyond (above and below) the range of the simulated
@@
 \(k = 1, ..., 4 x_{max} +1\). The PA regression slope,
-\(\hat\beta_{1,PA}\), and y-intercept, \(\hat\beta_{0,PA}\), were
+\(\hat\beta_{1,PA}\), and \(y\)-intercept, \(\hat\beta_{0,PA}\), were
 determined using the \texttt{mcreg} function in the mcr package in R
 \citep{mcr_pkg} which implements Deming regression (equivalent to a
-regression based on the slope of the first principal axis). Fitted
-values, \(\hat y_{k,PA}\) were then obtained every 0.25 increment across
-the domain from the PA regression equation,
+regression based on the slope of the principal axis). Fitted values,
+\(\hat y_{k,PA}\) were then obtained every 0.25 increment across the
+domain from the PA regression equation,
 \(\hat y_{k,PA} = \hat\beta_{0,PA} + \hat\beta_{1,PA} x_k\), for
@@
 Our results found that when shown points following a linear trend,
-participants visually fit a regression line that mimics the first
-principal axis regression as opposed to ordinary least squares
-regression. Data simulated with a larger variance provided strong
-support for a participants tendency to visually fit the first principal
-axis regression. We utilized modern technology to replicate a study
-conducted 40 years ago, and strengthened the original results with
-current analysis methods which allow for more flexibility and
-sophistication. Our results indicate that participants minimized the
-distance from their drawn regression line over both the \(x\) and \(y\)
-axis simultaneously. We allowed participants to draw trend lines that
-deviated from a straight line and gained an insight into the curvature
-the human eye perceives in a set of points. Researchers in cognitive and
-human movement sciences have found that human arm movement is a complex
-task \citep{miall1995curvature, rousset2015study}. The `You Draw It'
-method described in this paper uses \emph{indirect interaction} in which
-the mouse position and resulting visual line on the screen are
-dissociated. Therefore, curvature found in participant drawn lines from
-a straight lines could potentially be explained by the lack of
-coordination which results from the eye-hand dissociation from indirect
-drawing and the distortion of visual perception affecting the curvature
-of movements. Additionally, there is a training effect related to the
-completion of the `You Draw It' task - the movement of the line must be
-slow so that the visual representation on the screen can accurately
-capture each movement. \citet{de1991misdirections} conducted a study in
-which participants moved their hand slowly from an initial position in
-front of them to a visual target (movement task); they were then asked
-to repeat the task using different sizes of pointers (perceptual task).
+participants visually fit a regression line that mimics the principal
+axis regression line as opposed to ordinary least squares regression
+line. Data simulated with a larger variance provided strong support for
+a participants tendency to visually fit the first principal axis
+regression. We utilized modern technology to replicate a study conducted
+40 years ago, and strengthened the original results with current
+analysis methods which allow for more flexibility and sophistication.
+Our results indicate that participants minimized the distance from their
+drawn regression line over both the \(x\) and \(y\) axis simultaneously.
+We allowed participants to draw trend lines that deviated from a
+straight line and gained an insight into the curvature the human eye
+perceives in a set of points.
+
+Researchers in cognitive and human movement sciences have found that
+human arm movement is a complex task
+\citep{miall1995curvature, rousset2015study}. The `You Draw It' method
+described in this paper uses \emph{indirect interaction} in which the
+mouse position and resulting visual line on the screen are dissociated.
+Therefore, curvature found in participant drawn lines from a straight
+lines could potentially be explained by the lack of coordination which
+results from the eye-hand dissociation from indirect drawing and the
+distortion of visual perception affecting the curvature of movements.
+Additionally, there is a training effect related to the completion of
+the `You Draw It' task - the movement of the line must be slow so that
+the visual representation on the screen can accurately capture each
+movement. \citet{de1991misdirections} conducted a study in which
+participants moved their hand slowly from an initial position in front
+of them to a visual target (movement task); they were then asked to
+repeat the task using different sizes of pointers (perceptual task).
 Their results indicated that deviations from the shortest pointers were
@@
 studied the improvement in estimating correlations in scatter plots
-through perceptual learning interventions.\\
-Their study provided support for an improved proficiency in correlation
-estimation after training. We used a randomized complete design
-experimental structure, while \citet{mosteller1981eye} implemented a
-latin square design to test for an order effect due to practice.
+through perceptual learning interventions. Results form their study
+provided support for an improved proficiency in correlation estimation
+after training. We used a randomized complete design experimental
+structure, while \citet{mosteller1981eye} implemented a latin square
+design to test for an order effect due to practice.
 
@@
 statistically fitted trend line where outliers were included
-\citep{bobko_1979, correll_2017}. This work could provide an extension
-and further evaluate the perceptual system's resistance to outliers by
-allowing participants to deviate from a straight line, potentially
-capturing the outlier(s). While the focus of this study was on drawing
-linear trend-lines, further investigation is necessary to implement this
-method in non-linear settings and with real data in order to facilitate
-scientific communication - a strength of the combination of the flexible
-`You Draw It' method and GAMM analysis method. \citet{ciccione2021can}
-evaluated human ability to extrapolate data from trends by asking
-participants to provide a single point estimate; this tool could be used
-to evaluate a continuous extrapolation of points from a trend line for
-various linear and non-linear data structures. In the future, we intend
-to create an R package designed for easy implementation of `You Draw It'
-task plots in order to make this tool accessible to other researchers.
+\citep{bobko_1979, correll_2017}. The work in this paper could provide
+an extension and further evaluate the perceptual system's resistance to
+outliers by allowing participants to deviate from a straight line,
+potentially capturing the outlier(s). While the focus of this study was
+on drawing linear trend lines, further investigation is necessary to
+implement this method in nonlinear settings and with real data in order
+to facilitate scientific communication - a strength of the combination
+of the flexible `You Draw It' method and GAMM analysis method.
+\citet{ciccione2021can} evaluated human ability to extrapolate data from
+trends by asking participants to provide a single point estimate; this
+tool could be used to evaluate a continuous extrapolation of points from
+a trend line for various linear and nonlinear data structures. In the
+future, we intend to create an R package designed for easy
+implementation of `You Draw It' task plots in order to make this tool
+accessible to other researchers.
 
@@
 \item
-  \textbf{README:} File containing detailed descriptions of the
+  \textbf{readme:} File containing detailed descriptions of the
   supplementary material (README.html).

### 2022-10-11 manuscript-revisions/final-submission-cover-letter.qmd
made final revisions.

@@
 
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
-+ `bibliography.bib`
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf` Final pdf version of manuscript.
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd` Final rmarkdown document used to write the manuscript with the proper formatting.
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex` Final latex file outputted from the rmarkdown document.
++ `bibliography.bib` File containing references in bibtex format.
 + `agsm.bst`
-+ `images/ydi-stimuli.png`
-+ `figure-latex/eyefitting-example-plot-1.pdf`
-+ `figure-latex/eyefitting-gamm-residualplots-1.pdf`
-+ `figure-latex/eyefitting-lmer-residualplots-1.pdf`
-+ `figure-latex/eyefitting-simplot-1.pdf`
-+ `figure-latex/ols-vs-pca-example-1.pdf`
++ `images/ydi-stimuli.png` Image for Figure 1 in the manuscript.
++ `figure-latex/eyefitting-simplot-1.pdf` Image for Figure 2 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
++ `figure-latex/ols-vs-pca-example-1.pdf` Image for Figure 3 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
++ `figure-latex/eyefitting-example-plot-1.pdf` Image for Figure 4 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
++ `figure-latex/eyefitting-lmer-residualplots-1.pdf` Image for Figure 5 in the manuscript; originally created using R ggplot2 code within the rmarkdown file. This image is to be printed in color.
++ `figure-latex/eyefitting-gamm-residualplots-1.pdf` Image for Figure 6 in the manuscript; originally created using R ggplot2 code within the rmarkdown file. This image is to be printed in color.
 
@@
 
-+ `README.html` 
-+ `eyefitting-model-data.csv`
-+ `you-draw-it-eyefitting-analysis.Rmd`
++ `README.html` File containing detailed descriptions of the supplementary material.
++ `eyefitting-model-data.csv` De-identified participant data collected in the study and used for analyses.
++ `you-draw-it-eyefitting-analysis.Rmd` The .Rmd code file used to replicate the analysis in this paper.

### 2022-10-13 latexdiff.R
trial-latexdiff

@@ -1,0 +1,11 @@
+library(latexdiffr)
+
+latexdiff(path1 = "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex", 
+          path2 = "Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex", 
+          output = "diff", 
+          open = TRUE, 
+          output_format = "tex")
+
+# latexdiff("intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd", 
+#           "Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd")
+

### 2022-10-13 latexdiff.R
latexdiffs

@@ -1,11 +1,0 @@
-library(latexdiffr)
-
-latexdiff(path1 = "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex", 
-          path2 = "Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex", 
-          output = "diff", 
-          open = TRUE, 
-          output_format = "tex")
-
-# latexdiff("intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd", 
-#           "Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd")
-

### 2022-10-13 manuscript-revisions/revision-diffs.R

@@ -1,9 +1,24 @@
 library(latexdiffr)
 
+# first revision comparison -----------------------
+
 latexdiff(
   "original-submission-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  output = "diff-1",
+  open = TRUE,
+  clean = TRUE,
+  quiet = TRUE,
+  output_format = NULL,
+  ld_opts = "--replace-context2cmd=\"none\""
+)
+
+# second revision comparison -------------------------
+
+latexdiff(
+  "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
   "Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
-  output = "diff",
+  output = "manuscript-revisions/diff-2",
   open = TRUE,
   clean = TRUE,
   quiet = TRUE,

### 2022-10-14 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
final revisions submtted :) yay

@@
 - name: Emily A. Robinson 1
-  # thanks: The authors gratefully acknowledge ...
+  thanks: The authors report there are no competing interests to declare.
   affiliation: Department of Statistics, University of Nebraska - Lincoln
@@
   geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
-            aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
+            aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.1) +
   geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), 
-            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+            aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.1) +
   geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
@@
   ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.2) +
-  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.2) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_ols_drawn, group = plot_id, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data %>% mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>% dplyr::rename(`Parameter Choice` = parm_id), aes(x = x, y = residual_pca_drawn, group = plot_id, color = "PA"), alpha = 0.1) +
   geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
@@
 -->
-+ **readme:** File containing detailed descriptions of the supplementary material (README.html).
++ **README:** File containing detailed descriptions of the supplementary material (README.html).

### 2022-10-14 manuscript-revisions/final-submission-cover-letter.qmd
final revisions submtted :) yay

--- 2022-10-06 manuscript-revisions/final-submission-cover-letter.qmd
@@
 
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex`
-+ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf`
-+ `bibliography.bib`
-+ `agsm.bst`
-+ `images/ydi-stimuli.png`
-+ `figure-latex/eyefitting-example-plot-1.pdf`
-+ `figure-latex/eyefitting-gamm-residualplots-1.pdf`
-+ `figure-latex/eyefitting-lmer-residualplots-1.pdf`
-+ `figure-latex/eyefitting-simplot-1.pdf`
-+ `figure-latex/ols-vs-pca-example-1.pdf`
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.pdf` Final pdf version of manuscript.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex` Final latex file outputted from the rmarkdown template.
+
++ `bibliography.bib` File containing references in bibtex format.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era.bbl` File containing the references.
+
++ `images/ydi-stimuli.png` Image for Figure 1 in the manuscript.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/`
+
+    `eyefitting-simplot-1.pdf` Image for Figure 2 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/`
+
+    `ols-vs-pca-example-1.pdf` Image for Figure 3 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/`
+
+    `eyefitting-example-plot-1.pdf` Image for Figure 4 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/`
+
+    `eyefitting-lmer-residualplots-1.pdf` Image for Figure 5 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
+This image is to be printed in color; we have emailed co-editors Galin Jones and Faming Liang about the appeal process for color costs.
+
++ `Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/`
+
+    `eyefitting-gamm-residualplots-1.pdf` Image for Figure 6 in the manuscript; originally created using R ggplot2 code within the rmarkdown file.
+This image is to be printed in color; we have emailed co-editors Galin Jones and Faming Liang about the appeal process for color costs.
 
@@
 
-+ `README.html` 
-+ `eyefitting-model-data.csv`
-+ `you-draw-it-eyefitting-analysis.Rmd`
++ `README.html` File containing detailed descriptions of the supplementary material.
+
++ `eyefitting-model-data.csv` De-identified participant data collected in the study and used for analyses.
+
++ `you-draw-it-eyefitting-analysis.Rmd` The .Rmd code file used to replicate the analysis in this paper.

### 2022-10-14 manuscript/Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
final revisions submtted :) yay

--- 2022-10-11 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
@@
   \author{
-        Emily A. Robinson 1 \\
+        Emily A. Robinson 1 \thanks{The authors report there are no
+competing interests to declare.} \\
     Department of Statistics, University of Nebraska - Lincoln\\
@@
 \item
-  \textbf{readme:} File containing detailed descriptions of the
+  \textbf{README:} File containing detailed descriptions of the
   supplementary material (README.html).

### 2022-10-14 manuscript-revisions/revision-diffs.R

@@ -25,3 +25,10 @@ latexdiff(
   output_format = NULL,
   ld_opts = "--replace-context2cmd=\"none\""
 )
+
+# in terminal
+
+# latexdiff intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex > diff-2.tex
+
+
+
